### 2021-03-16 inst/analysis/dna-per-cell.Rmd
finish cells_per_dna analysis

---
title: "Cell-DNA Standard Curves"
subtitle: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::pdf_document2:
    number_sections: true
    toc: true
    toc_depth: 1
    df_print: tibble
    dev: cairo_pdf
    latex_engine: xelatex
geometry: margin = 0.5in
fontsize: 11pt
mainfont: Calibri
params:
  print_plots: true
---

```{r setup, include = FALSE}
# load libraries
suppressPackageStartupMessages({
  devtools::load_all()
  library(tidyverse)
  library(wmo)
  library(targets)
})

# set global chunk options
knitr::opts_chunk$set(
  echo = FALSE, 
  message = FALSE, 
  warning = FALSE, 
  fig.align = "center", 
  out.width = "49%"
)

options(knitr.table.format = function() {
  if (knitr::is_latex_output()) 
    "latex" else "html"
})

theme_set(theme_wmo(base_family = "Calibri"))
```

```{r}
withr::with_dir(here::here(), {
  std <- tar_read(dna_per_cell_std)
  sample <- tar_read(dna_per_cell_clean)
  cells_per_dna <- tar_read(cells_per_dna)
})
```

\newpage

# Overview

Cell counts were estimated from total cellular DNA measured by PicoGreen fluorescence (Quant-iT PicoGreen dsDNA Assay Kit, P11496, Thermo). Here, we determine the relationship between total DNA and cell count by quantifying DNA extracted from cells seeded at different densities in basal growth medium. For the lung fibroblasts, cells were extracted with 100 μL or 200 μL of TE buffer with 0.2% Triton X-100. Earlier experiments used the smaller volume of extraction buffer and this is adjusted for during the extracellular flux calculations.

# Analysis

## Standard Curves

```{r}
std %>% 
  unnest(c(summary)) %>% 
  ggplot() +
  aes(y = reorder(date, desc(date)), x = r.squared) +
  geom_point() +
  labs(
    y = "Date", 
    x = expression(paste("R"^2))
  )
```

## Samples

```{r}
sample %>% 
  ggplot() +
  aes(x = cells, y = conc) +
  facet_wrap(~ cell_type * volume, labeller = as_labeller(toupper)) +
  geom_point(alpha = 0.3) +
  stat_summary(
    geom = "pointrange", 
    fun.data = "mean_se"
  ) +
  geom_smooth(
    method = "lm", 
    formula = y ~ 0 + x, 
    se = FALSE
  ) +
  labs(
    x = "Cell count", 
    y = "DNA (ng)"
  )
```

These plots suggest that, for the 100 μL extraction buffer, the maximum cell count is ~ 250,000. We will exclude the 300,000 value prior to calculate the slopes of the curves. For the PASMC samples, the 400,000 cell sample is quite high compared to the other samples. Will also remove these values prior to calculation of the slopes.

```{r}
usethis::use_data(cells_per_dna, overwrite = TRUE)

cells_per_dna  %>% 
  knitr::kable(
    digits = 4, 
    col.names = c("Cell Type", "Volume", "Cell / DNA"), 
    align = "c", 
    booktabs = TRUE
  ) %>% 
  kableExtra::kable_styling(position = "center") 
```

### 2021-03-16 R/cells_per_dna.R

@@ -1,0 +1,13 @@
+#' Cell count per DNA
+#'
+#' A data set containing the slopes to interpolate cell number from DNA
+#' measurements.
+#'
+#' \describe{
+#'   \item{cell_type}{
+#'   `lf`    = lung fibroblasts \cr
+#'   `pasmc` = pulmonary artery smooth muscle cells}
+#'   \item{volume}{volume of lysis buffer used to extract DNA}
+#'   \item{slope}{cells / ng DNA}
+#'   }
+"cells_per_dna"

### 2021-03-16 R/functions.R

@@ -1,5 +1,21 @@
 # functions.R
 
+path_to_data <- function(nm) {
+  system.file(
+    stringr::str_c("extdata/", nm),
+    package = "Copeland.2021.hypoxia.flux"
+  )
+}
+
+
+path_to_reports <- function(nm) {
+  system.file(
+    stringr::str_c("analysis/", nm),
+    package = "Copeland.2021.hypoxia.flux"
+  )
+}
+
+
 read_multi_excel <- function(excel_file) {
   sheets <- readxl::excel_sheets(excel_file)
   purrr::map(sheets, ~readxl::read_excel(excel_file, sheet = .x)) %>%
@@ -27,14 +43,111 @@ clean_technical_replicates <- function(tbl) {
     dplyr::ungroup()
 }
 
-data_file <-
-    system.file(
-      "extdata/dna-per-cell-number.xlsx",
-      package = "Copeland.2021.hypoxia.flux"
+make_std_curves <- function(df) {
+  df %>%
+    dplyr::filter(!is.na(.data$conc)) %>%
+    dplyr::select(where(~all(!is.na(.)))) %>%
+    dplyr::group_by(dplyr::across(-c(.data$conc, .data$value))) %>%
+    tidyr::nest() %>%
+    dplyr::mutate(
+      title = stringr::str_c(!!!rlang::syms(dplyr::groups(.)), sep = "_")
+    ) %>%
+    dplyr::ungroup() %>%
+    dplyr::mutate(
+      model = furrr::future_map(
+        .data$data,
+        ~lm(value ~ conc, data = .x, na.action = modelr::na.warn)
+      ),
+      summary = furrr::future_map(.data$model, broom::glance),
+      plots = furrr::future_map2(.data$data, .data$title, make_std_plots)
+    ) %>%
+    dplyr::group_by(
+      dplyr::across(
+        -c(.data$data, .data$title, .data$model, .data$summary, .data$plots)
+      )
     )
+}
 
-read_multi_excel(data_file) %>%
-  purrr::map(clean_technical_replicates) %>%
-  bind_rows(.id = "id") %>%
-  separate(id, c("cell_type", "volume"), sep = "_", convert = TRUE) %>%
-  mutate(cells = 1000 * cells)
+
+make_std_plots <- function(df, title = NULL) {
+  ggplot2::ggplot(df) +
+    ggplot2::aes(x = .data$conc, y = .data$value) +
[…]

### 2021-03-16 _targets.R

@@ -1,16 +1,35 @@
 # _targets.R
 
 #setup
+devtools::load_all()
 library(targets)
+library(tarchetypes)
 source("R/functions.R")
 
+conflicted::conflict_prefer("filter", "dplyr")
+
+options(tidyverse.quiet = TRUE)
+
+ppp <- TRUE
+
+future::plan(future::multisession(workers = future::availableCores() - 1))
+
 # target-specific options
-tar_option_set(packages = c(
-  "dplyr"
-))
+tar_option_set(
+  packages = c("tidyverse")
+)
 
 # list of target objects
 list(
-  tar_target(data, data.frame(x = sample.int(100), y = sample.int(100))),
-  tar_target(summary, summ(data)) # Call your custom functions as needed.
+  tar_target(dna_per_cell_file, path_to_data("dna-per-cell-number.xlsx")),
+  tar_target(dna_per_cell_raw, clean_dna_per_cell(dna_per_cell_file)),
+  tar_target(dna_per_cell_std, make_std_curves(dna_per_cell_raw)),
+  tar_target(dna_per_cell_clean, interp_data(dna_per_cell_raw, dna_per_cell_std)),
+  tar_target(cells_per_dna, calculate_cells_per_dna(dna_per_cell_clean)),
+  tar_render(
+    dna_per_cell_report,
+    path = path_to_reports("dna-per-cell.Rmd"),
+    params = list(print_plots = ppp),
+    output_dir = system.file("analysis/pdfs", package = "Copeland.2021.hypoxia.flux")
+  )
 )

### 2021-03-17 inst/analysis/dna-per-cell.Rmd
Remove ignored files

@@
     toc_depth: 1
-    df_print: tibble
+    df_print: kable
     dev: cairo_pdf
@@
 mainfont: Calibri
-params:
-  print_plots: true
 ---

### 2021-03-17 R/functions.R

@@ -22,6 +22,7 @@ read_multi_excel <- function(excel_file) {
     rlang::set_names(sheets)
 }
 
+
 clean_technical_replicates <- function(tbl) {
   tidyr::pivot_longer(
     data = tbl,
@@ -42,6 +43,7 @@ clean_technical_replicates <- function(tbl) {
     dplyr::summarise(value = mean(.data$value, na.rm = TRUE)) %>%
     dplyr::ungroup()
 }
+
 
 make_std_curves <- function(df) {
   df %>%
@@ -71,7 +73,10 @@ make_std_curves <- function(df) {
 
 make_std_plots <- function(df, title = NULL) {
   ggplot2::ggplot(df) +
-    ggplot2::aes(x = .data$conc, y = .data$value) +
+    ggplot2::aes(
+      x = .data$conc,
+      y = .data$value
+    ) +
     ggplot2::geom_smooth(
       method = "lm",
       formula = y ~ x,
@@ -129,25 +134,25 @@ clean_dna_per_cell <- function(filename) {
   filename %>%
     read_multi_excel() %>%
     purrr::map(clean_technical_replicates) %>%
-    bind_rows(.id = "id") %>%
-    separate(id, c("cell_type", "volume"), sep = "_", convert = TRUE) %>%
-    mutate(cells = 1000 * cells)
+    dplyr::bind_rows(.id = "id") %>%
+    tidyr::separate(id, c("cell_type", "volume"), sep = "_", convert = TRUE) %>%
+    dplyr::mutate(cells = 1000 * cells)
 }
 
 
 calculate_cells_per_dna <- function(tbl) {
   tbl %>%
-    filter(!(cell_type == "lf" & volume == "100" & cells == 300000)) %>%
-    filter(!(cell_type == "pasmc" & volume == "200" & cells == 400000)) %>%
-    group_by(cell_type, volume) %>%
-    nest() %>%
-    mutate(
+    dplyr::filter(!(cell_type == "lf" & volume == "100" & cells == 300000)) %>%
+    dplyr::filter(!(cell_type == "pasmc" & volume == "200" & cells == 400000)) %>%
+    dplyr::group_by(cell_type, volume) %>%
+    tidyr::nest() %>%
+    dplyr::mutate(
       model = map(data, ~lm(conc ~ 0 + cells, data = .x, na.action = modelr::na.warn)),
       glance = map(model, broom::tidy)
     ) %>%
[…]

### 2021-03-17 _targets.R

@@ -8,15 +8,17 @@ source("R/functions.R")
 
 conflicted::conflict_prefer("filter", "dplyr")
 
-options(tidyverse.quiet = TRUE)
-
-ppp <- TRUE
+options(
+  tidyverse.quiet = TRUE,
+  usethis.quiet = TRUE
+)
 
 future::plan(future::multisession(workers = future::availableCores() - 1))
 
 # target-specific options
 tar_option_set(
-  packages = c("tidyverse")
+  packages = c("tidyverse"),
+  format = "qs"
 )
 
 # list of target objects
@@ -29,7 +31,6 @@ list(
   tar_render(
     dna_per_cell_report,
     path = path_to_reports("dna-per-cell.Rmd"),
-    params = list(print_plots = ppp),
     output_dir = system.file("analysis/pdfs", package = "Copeland.2021.hypoxia.flux")
   )
 )

### 2021-03-20 inst/analysis/dna-per-cell.Rmd
complete extracellular fluxes analysis

@@
 
+\newpage
+
 ```{r}
@@
 cells_per_dna  %>% 
-  knitr::kable(
-    digits = 4, 
+  my_kable(
+    digits = 0, 
     col.names = c("Cell Type", "Volume", "Cell / DNA"), 
     align = "c", 
-    booktabs = TRUE
+    caption = "Cells per DNA (ng)"
   ) %>% 

### 2021-03-20 inst/analysis/extracellular-fluxes.Rmd
complete extracellular fluxes analysis

---
title: "Extracellular Fluxes"
subtitle: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::pdf_document2:
    number_sections: true
    toc: true
    toc_depth: 1
    df_print: kable
    dev: cairo_pdf
    latex_engine: xelatex
geometry: margin = 0.5in
fontsize: 11pt
mainfont: Calibri
---

```{r setup, include = FALSE}
# load libraries
suppressPackageStartupMessages({
  devtools::load_all()
  library(tidyverse)
  library(wmo)
  library(targets)
})

# resolve conflicts
conflicted::conflict_prefer("filter", "dplyr")

# set global chunk options
knitr::opts_chunk$set(
  echo = FALSE, 
  message = FALSE, 
  warning = TRUE, 
  fig.align = "center", 
  out.width = "49%"
)

options(knitr.table.format = function() {
  if (knitr::is_latex_output()) 
    "latex" else "html"
})

theme_set(theme_wmo(base_family = "Calibri"))
```

```{r}
withr::with_dir(here::here(), {
  data_raw <- tar_read(conc_raw)
  std <- tar_read(conc_std)
  std_fld <- tar_read(conc_std_clean_fld)
  std_clean <- tar_read(conc_std_clean)
  conc <- tar_read(conc_with_missing)
  evap <- tar_read(evap_raw)
  flux_measurements <- tar_read(flux_measurements)
  growth_curves <- tar_read(growth_curves)
  growth_rates <- tar_read(growth_rates)
  degradation_rates <- tar_read(degradation_rates)
  k <- tar_read(k)
  fluxes <- tar_read(fluxes)
})
```

\newpage

# Overview

Extracellular fluxes are determined from serial measurements of cell count and metabolite mass in conditioned medium accounting for evaporation and metabolite degradation/accumulation. Here, we calculate the extracellular fluxes included in the manuscript. Metabolic flux models assume metabolic steady state associated with exponential cell growth. Pilot experiments yielded the following conditions to monitor this time period:

*Lung fibroblasts.* Lung fibroblasts (LFs) were seeded at 25,000 cells per well of six-well plates on Day -1. Treatment was initiated on this day by placing plates in hypoxia (0.2% or 0.5% ambient oxygen) or treating with the prolyl hydroxylase inhibitor molidustat (10 μM, BAY, BAY85-3934, Cayman) with 0.1% DMSO as the vehicle control. The medium was changed on Day 0 and samples were collected this day and every 24 h for 72 h. The cell culture medium was MCDB131 (genDepot) supplemented with glucose (8 mM) and glutamine (1 mM), similar to medium used for isotope labeling experiments.

*Pulmonary artery smooth muscle cells.* Pulmonary artery smooth muscle cells (PASMCs) were seeded at 25,000 cells per well of six-well plates on Day -1. Treatment was initiated on this day by placing plates in hypoxia (0.5% ambient oxygen). The medium was changed on Day 0 and samples were collected at this time and every 12 h for 48 h. The cell culture medium was MCDB131 (genDepot) supplemented with glucose (5.551 mM) and glutamine (10 mM), similar to medium used for isotope labeling experiments.

*Hypoxia plus BAY.* When hypoxia was combined with BAY treatment to measure glucose and lactate fluxes, cells were cultured in complete growth medium (*e.g.*, FGM-2 for LFs).

```{r}
data_raw %>% 
  select(cell_type, experiment, date) %>% 
  distinct() %>% 
  group_by(cell_type, experiment) %>% 
  summarise(count = n()) %>% 
  my_kable(
    col.names = c("Cell type", "Experiment", "Count"), 
    caption = "Biological replicates"
  )
```

# Data Acquisition

Raw data are indexed to metadata by sample number, which corresponds to a unique combination of measurement type, treatment, time, and well. The sample type, `cells` or `empty`, refers to whether the medium was obtained from wells containing cells for flux measurements (`cells`) or from empty wells for degradation rate measurements (`empty`). The raw data for each experiment are contained in multi-sheet Excel files.

## Cell count

Cell counts were estimated from total cellular DNA measured by PicoGreen fluorescence (Quant-iT PicoGreen dsDNA Assay Kit, P11496, Thermo). The relationship between total DNA and cell count was determined by measuring DNA from cells seeded at different densities in basal growth medium. In the table below, the DNA standards were adjusted to incorporate the conversion from [DNA] to cell count. Lysis buffer volume changed between the two experimental batches, which affected the slope of the [DNA] *v*. cell count standard curve. Here, `conc` refers to cell count.

## Lactate

Lactate concentration was determined using an enzymatic assay (\textsc{l}-Lactate Assay Kit, 700510, Cayman). For batch `a`, samples were deproteinized with MPA per the manufacturer's protocol (2.1-fold dilution) and then diluted 5-fold prior to analysis (10.5-fold dilution total). Control experiments indicated that, for medium samples, deproteinization is not necessary (*i.e.*, signal intensities for unconditioned and conditioned medium were the same regardless of the deproteinization step). For subsequent batches, samples were diluted 10-fold in PBS prior to analysis. As for cell count, the lactate standard concentrations were adjusted to account for sample dilution.

## Glucose

Glucose concentration was determined using an enzymatic assay (Glucose Colorimetric Assay Kit, 10009582, Cayman). All samples were diluted 10-fold in water (batch `a`) or PBS (batches `b` and `c`) prior to analysis. Preliminary studies suggested no significant degradation of glucose. Glucose standard concentrations were adjusted to reflect unit conversion from mg/dL to μM and sample dilution prior to analysis.

## Pyruvate

Pyruvate concentration was measured multiple ways. For hypoxia batch `a` samples, it was determined by HPLC analysis of deproteinized OPD-derivatized samples. This analysis utilized ketovalerate (`KV`) as an internal standard (10 μL, 100 μM final concentration), which was added at the same time as the amino acid internal standards. For `bay` batch `a`, pyruvate concentration was determined by LC-MS analysis with `d8-valine` as an internal standard. For all batch `b` and `c` samples, pyruvate concentration was determined by enzymatic assay (Pyruvate Assay Kit, 700470, Cayman). For the enzymatic assay, samples were diluted 20-fold in PBS.

## Amino acids

Amino acid concentrations were determined by HPLC analysis of FMOC- and OPA-derivatized samples, producing chemicals detectable by fluorescent or UV/visible spectrophotometery, respectively. Prior to HPLC, internal standards (norvaline and sarcosine, 10 μL, 100 μM final concentration) were added to the samples (180 μL for batch `a` samples where pyruvate was measured by chromatography or 190 μL for all other samples) and they were deproteinized using ice-cold acetone (400 μL). Insoluble material was removed by centrifugation and the supernatant was evaporated to < 50% of the initial volume (< 100 μL) prior to HPLC.

\newpage

# Data Analysis

All measured values are interpolated to concentration in μM for metabolites or cell count for DNA measurements.

## Standard curves

A review of the r^2^ values for the standard curves will identify problematic fits.

```{r, results = "hide", fig.show = "hold", fig.align = "left"}
plot_r_squared <- function(df, title) {
  ggplot(df) +
    aes(
      y = reorder(metabolite, desc(metabolite)), 
      x = r.squared, 
      color = interaction(date, batch, run)
    ) +
    geom_point(
      aes(shape = detector), 
      size = 3, 
      alpha = 0.6
    ) +
    labs(
      title = title, 
      x = expression(paste("r"^2)), 
      y = "Metabolite"
    ) +
    guides(color = guide_legend(title = NULL, nrow = 10))
}

std %>% 
  select(metabolite:run, summary) %>% 
  unnest(c(summary)) %>% 
  group_by(cell_type, experiment) %>% 
  nest() %>% 
  mutate(
    title = str_c(cell_type, experiment, sep = " "), 
    plots = map2(data, title, plot_r_squared)
  ) %>% 
  pull(plots)
```

These data suggest worse fits from fluorescent detection.

```{r, results = "hide", fig.show = "hold", fig.align = "left", out.width = "24.5%"}
std %>% 
  filter(detector == "fld", date == "2018-02-12", experiment == "02") %>% 
  pull(plots)
```

The fluorescence signal is plateauing between 250 and 1000 μM. The highest standard concentration will be removed from the curves.

In addition, there were a few standard curves that had poor fits:

```{r, fig.show = "hold", fig.align = "left", results = "hide"}
std_fld %>% 
  select(metabolite:run, summary) %>% 
  unnest(c(summary)) %>% 
  group_by(cell_type, experiment) %>% 
  nest() %>% 
  mutate(
    title = str_c(cell_type, experiment, sep = " "), 
    plots = map2(data, title, plot_r_squared)
  ) %>% 
  pull(plots)
```

After removing the high fluorescence standard, the remaining poorest fits were:

```{r}
poor_fits <- 
  std_fld %>% 
  ungroup() %>% 
  select(metabolite:run, summary, plots) %>% 
  unnest(c(summary)) %>% 
  slice_min(order_by = r.squared, n = 7)

poor_fits %>% 
  select(metabolite:r.squared) %>% 
  my_kable(caption = "Poor standard curve fits")
```

```{r, fig.show = "hold", fig.align = "left", results = "hide", out.width = "32.9%"}
poor_fits %>% 
  pull(plots)
```

When an apparent outlier was identified, it was removed from the standard curve.

```{r, fig.show = "hold", fig.align = "left", results = "hide", out.width = "32.9%"}
std_clean %>% 
  semi_join(poor_fits, by = c("metabolite", "experiment", "date", "batch", "run", "detector")) %>% 
  pull(plots)
```

## Missing data

The sample data frame was expanded to identify missing samples.

- For the 0.5% oxygen experiments, only three `empty` time 0 samples were collected and analyzed. The measurements from these samples will be used as time 0 to calculate degradation and accumulation rates for both treatments, 21% and 0.5%.

- For the LF 0.5% oxygen experiment performed 2017-11-06, the 48 h 21% `empty` sample was lost.

- Specific `empty` samples were not collected for the BAY treatment condition, only for the DMSO condition. Thus, the DMSO degradation rates will be duplicated and used for BAY treatment in the flux calculations.

- For the LF BAY experiment performed 2018-11-02, only one of three time 0 `empty` samples were collected.

# Analysis

## Pyruvate assays

Pyruvate was ultimately measured by three different methods, HPLC, LC-MS, and enzymatic assay.

```{r}
pyruvate <- 
  conc %>% 
  filter(metabolite == "pyruvate") %>%
  mutate(detector = factor(
    detector, 
    levels = c("enzyme", "hplc", "lcms"), 
    labels = c("Enzyme", "HPLC", "LC-MS")
  ))

pyruvate %>% 
  filter(conc < 1500) %>% 
  filter(time == 0) %>% 
  ggplot() + 
  aes(x = detector, y = conc, color = detector) +
  geom_jitter(show.legend = FALSE) +
  labs(
    title = "Pyruvate concentrations at time 0", 
    x = "Detection method", 
    y = "Concentration (μM)"
  )
```

Compared to the enzymatic assay, the HPLC measurements tended to underestimate the pyruvate concentration while the LC-MS measurements tended to overestimate the pyruvate concentration. I suspect the pyruvate determination using the enzymatic assay is more accurate and we will preferentially use these measurements to determine the fluxes. Unfortunately, for the 0.2% oxygen experiments, the only pyruvate measurements available were obtained using the HPLC assay, and so this assay will be used to estimate fluxes from those experiments.

## Amino acid measurements

Amino acid metabolites were measured using two detectors, fluorescence (`fld`) and multi-wavelength (`mwd`). We need to decide which detector is optimal for which amino acids based on the metabolite and its concentration in medium. Cystine is derivatized poorly by FMOC, and so it not detectable by fluorescence.

The fluorescence response is linear up to a concentration of ~ 250 μM on the basis of the standard curves. First, we will assess the metabolite concentrations across all of the samples.

```{r, fig.show = "hold"}
amino_acids <-
  conc %>% 
  filter(detector %in% c("mwd", "fld"))

plot_aa <- function(df) {
  df %>% 
    filter(!is.na(conc)) %>% 
    ggplot() + 
    aes(y = conc, x = metabolite, color = detector) + 
    geom_hline(yintercept = 300, linetype = 2, color = "gray") +
    geom_jitter(alpha = 0.1) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    labs(
      x = "Amino acid", 
      y = "Concentration (μM)"
    ) + 
    scale_color_discrete(name = "Detector", labels = c("FLD", "MWD")) +
    guides(color = guide_legend(override.aes = list(alpha = 1)))
}

amino_acids %>% 
  filter(cell_type == "lf") %>% 
  plot_aa() + ggtitle("LF amino acids")

amino_acids %>% 
  filter(cell_type == "pasmc") %>% 
  plot_aa() + ggtitle("PASMC amino acids")
```

Fluorescence detection appears to be associated with smaller variance when amino acid concentration is low. At higher concentrations, fluorescence detection is likely underestimating the concentration based on the standard curve analysis performed above. Based on these data, we will use concentrations determined by fluorescence detection for metabolites with an average concentration < 300 μM.

```{r}
amino_acids %>% 
  group_by(metabolite) %>% 
  summarise(conc = mean(conc, na.rm = TRUE)) %>% 
  mutate(detector = case_when(
    metabolite == "cystine" ~ "mwd", 
    conc > 300 ~ "mwd", 
    conc <= 300 ~ "fld"
  )) %>% 
  filter(detector == "mwd") %>% 
  select(metabolite) %>% 
  my_kable(caption = "MWD detection", col.names = "Metabolite")
```

## Evaporation

Preliminary experiments suggested that medium evaporation rates would have a significant impact on the interpretation of metabolite concentration measurements and that the evaporation rates differed between the normoxia and hypoxia incubators. To determine evaporation rate, a six-well plate was filled with medium and weighed during each sample collection. For 0.2% oxygen experiments, no evaporation controls were measured and so 0.5% oxygen rates were used instead. For BAY experiments, only one evaporation plate containing medium supplemented with 0.1% DMSO was used and this evaporation rate was applied to both conditions. Predicted well volumes were calculated from experimental measurements of plate weight assuming a starting volume of 2 mL per well.

```{r}
evap %>% 
  ggplot() +
  aes(
    x = time, 
    y = volume, 
    color = interaction(experiment, oxygen, treatment, sep = " | ")
  ) +
  geom_point(
    alpha = 0.3, 
    size = 2
  ) +
  geom_smooth(
    method = "lm", 
    formula = y ~ x, 
    se = FALSE, 
    show.legend = FALSE
  ) +
  stat_summary(
    geom = "pointrange", 
    fun.data = "mean_se", 
    alpha = 1, 
    fatten = 1, 
    show.legend = FALSE
  ) +
  guides(color = guide_legend(
    title = "Treatment", 
    override.aes = list(alpha = 1)
  )) +
  labs(
    tite = "Predicted well volumes", 
    x = "Time (h)", 
    y = "Volume (ml)"
  )
```

Specific evaporation rates for BAY-treated cells were not obtained, so the DMSO condition was duplicated. Evaporation controls were not obtained for BAY batch `a` or 0.2% oxygen experiments and so the average evaporation rates from BAY batch `b` and 0.5% oxygen experiments were duplicated and substituted, respectively. The predicted well volumes are attached to the interpolated sample concentrations or cell counts and made available as `flux_measurements`.

```{r}
usethis::use_data(flux_measurements, overwrite = TRUE)
```

# Flux Calculations

Determination of fluxes is based on the solution of the following differential equations:

$$ 
\begin{aligned}
\frac{dX}{dt} &= \mu X \\[6 pt]
\frac{dM}{dt} &= -kM+vX 
\end{aligned}
$$

where $X$ is the cell number, $\mu$ is cell growth rate, $M$ is moles of substrate, $k$ is the metabolite degradation rate, and $v$ is the extracellular flux rate.

These equations are solved as follows:

$$
\begin{aligned}
X &= X_0e^{\mu t} \\[6 pt]
Me^{k t} &= \frac{v X_0}{\mu + k} (e^{(\mu + k)t}-1) + M_0
\end{aligned}
$$

The data can be fit to a linear equation and the slope is used to calculate $v$.

## Growth rates

```{r, fig.show = "hold", fig.align = "left", results = "hide", out.width = "32.9%"}
growth_curves %>% 
  select(cell_type:data) %>% 
  unnest(c(data)) %>% 
  group_by(cell_type, experiment, batch, date, oxygen, treatment, time) %>%
  summarize(conc = mean(conc, na.rm = TRUE)) %>% 
  filter(!is.na(time)) %>% 
  group_by(cell_type, experiment) %>% 
  nest() %>% 
  mutate(plot = map2(
    data, 
    cell_type, 
    ~ggplot(.x) + 
      aes(
        x = time, 
        y = conc/1000, 
        color = interaction(oxygen, treatment, sep = " | ")
      ) +
      stat_summary(
        geom = "pointrange", 
        fun.data = "mean_se"
      ) +
      stat_summary(
        geom = "line", 
        fun.data = "mean_se", 
        show.legend = FALSE
      ) +
      labs(
        title = toupper(.y), 
        x = "Time (h)", 
        y = expression(paste("Cell count (\U00D7", "10"^3, ")")), 
        color = "condition"
      )
  )) %>% 
  pull(plot)
```

Robust linear models were fit to the logarithm of cell count *v*. time to determine growth rate (μ) and initial cell count (X0). The calculated growth rates are available as `growth_rates`.

```{r}
usethis::use_data(growth_rates, overwrite = TRUE)
```

```{r, fig.show = "hold", out.width = "49%"}
plot_growth_info <- function(df, y){
  ggplot(df) +
    aes(
      x = interaction(oxygen, treatment), 
      y = .data[[y]], 
      color = interaction(oxygen, treatment)
    ) +
    facet_wrap(
      ~cell_type, 
      ncol = 1
    ) +
    geom_point(
      aes(shape = batch), 
      size = 4, 
      alpha = 0.5, 
      show.legend = FALSE
    ) +
    stat_summary(
      fun.data = "mean_se", 
      geom = "crossbar", 
      width = 0.3, 
      show.legend = FALSE
    ) +
    labs(
      x = "Treatment", 
      y = y
    )
}

plot_growth_info(growth_rates, "mu") + ylab(expression(paste("Growth rate (h"^-1, ")")))
plot_growth_info(growth_rates, "X0") + ylab("Cell count at time 0")
```

## Degradation rates

Metabolite concentrations from unconditioned medium (`type == empty`) were used to calculate metabolite degradation or accumulation rates. Glutamine, for example, is known to degrade over time. When available, 96 h samples were included in these analyses to improve fits. Metabolite concentration was multiplied by predicted well volume to determine metabolite mass in nanomoles.

```{r, out.width = "100%"}
ggplot(degradation_rates) +
  aes(
    x = reorder(abbreviation, k), 
    y = k, 
    color = interaction(oxygen, treatment, sep = " | ")
  ) +
  geom_hline(yintercept = 0) +
  geom_point(
    aes(shape = batch), 
    alpha = 0.3
  ) +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  labs(
    title = "Degradation and accumulation rates", 
    x = "Metabolite", 
    y = expression(paste("Rate constant (h"^"-1", ")")), 
    color = "condition"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

There appears to be some treatment-dependent effects on degradation rates. Will group rates by treatment, remove outliers, and determine if the mean rates are different from 0. If so, these rates will be included in the metabolic flux calculations.

```{r}
usethis::use_data(k, overwrite = TRUE)
```

## Fluxes

```{r}
usethis::use_data(fluxes, overwrite = TRUE)
```

```{r, out.width = "75%"}
plot_fluxes <- function(df, title) {
  ggplot(df) + 
    aes(
      x = reorder(abbreviation, flux),
      y = flux,
      color = interaction(oxygen, treatment, sep = " | ")
    ) +
    geom_hline(yintercept = 0) +
    geom_point(
      aes(group = interaction(oxygen, treatment)),
      position = position_dodge(width = 0.5),
      size = 2,
      alpha = 0.3
    ) +
    stat_summary(
      fun.data = "mean_se",
      geom = "crossbar",
      width = 0.3,
      position = position_dodge(width = 0.5),
      show.legend = FALSE
    ) +
    labs(
      title = title,
      x = "Metabolite",
      y = "Flux (fmol / cell / h)",
      color = "Condition"
    ) +
    guides(colour = guide_legend(override.aes = list(alpha = 1))) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

fluxes %>%
  filter(metabolite %in% c("glucose", "lactate")) %>%
  plot_fluxes("Metabolite flux") + facet_wrap(~cell_type, nrow = 1)
```

```{r, results = "hide", out.width = "100%"}
fluxes %>% 
  filter(!(metabolite %in% c("glucose", "lactate"))) %>% 
  group_by(cell_type, experiment) %>% 
  nest() %>% 
  mutate(plots = map2(data, toupper(cell_type), plot_fluxes)) %>% 
  pull(plots)
```

\newpage

```{r}
fluxes %>%
  ungroup() %>%
  group_by(metabolite, cell_type, oxygen, treatment) %>%
  summarise(
    v = mean(flux, na.rm = TRUE),
    sd = sd(flux, na.rm = TRUE),
    n = n(),
    se = sd / sqrt(n()),
    cv = sd / abs(v)
  ) %>%
  arrange(cell_type, metabolite, oxygen, treatment) %>%
  my_kable(
    caption = "Flux rate summary",
    col.names = c(
      "Metabolite", 
      "Cell type", 
      "Oxygen", 
      "Treatment", 
      "Flux", 
      "SD", 
      "n", 
      "SEM", 
      "CV"
    ),
    align = "lcccccccc",
    digits = 2,
    longtable = TRUE,
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("repeat_header"),
    font_size = 9
  )
```

### 2021-03-20 R/flux_measurements.R

@@ -1,0 +1,40 @@
+#' Measurements for extracellular flux calculations
+#'
+#' A data set containing the interpolated metabolite concentrations, cell counts,
+#' and evaporation volumes for extracellular flux determinations.
+#'
+#' \describe{
+#'   \item{cell_type}{
+#'   `lf` = lung fibroblasts \cr
+#'   `pasmc` = pulmonary artery smooth muscle cells}
+#'   \item{experiment}{
+#'   `02` = 0.2% oxygen for hypoxia \cr
+#'   `05` = 0.5% oxygen for hypoxia \cr
+#'   `bay` = molidustat treatment \cr
+#'   `05-bay` = 0.5% oxygen plus molidustat}
+#'   \item{batch}{a group of biological replicates analyzed similarly}
+#'   \item{date}{start date of an experiment}
+#'   \item{metabolite}{name of measurement}
+#'   \item{abbreviation}{abbreviated metabolite name}
+#'   \item{detector}{
+#'   `fld` = HPLC fluorescence detector \cr
+#'   `mwd` = HPLC multi-wavelength detector \cr
+#'   `enzyme` = enzymatic assay \cr
+#'   `hplc` = OPD-derivatized HPLC detection \cr
+#'   `lcms` = liquid chromatography-mass spectrometry assay \cr
+#'   `picogreen` = fluorescence dye labeling of DNA}
+#'   \item{type}{
+#'   `cells` = conditioned medium \cr
+#'   `empty` = unconditioned medium}
+#'   \item{oxygen}{ambient oxygen level}
+#'   \item{treatment}{
+#'   `None` = no treatment \cr
+#'   `DMSO` = 0.1% DMSO \cr
+#'   `BAY` = 10 μM molidustat}
+#'   \item{time}{hours}
+#'   \item{well}{denotes technical replicates in each experiment}
+#'   \item{conc}{measured in number for cells and μM for metabolites}
+#'   \item{volume}{measured in mL, extrapolated from evaporation controls}
+#'   \item{nmol}{metabolite mass}
+#'   }
+"flux_measurements"

### 2021-03-20 R/fluxes.R

@@ -1,0 +1,26 @@
+#' Calculated fluxes
+#'
+#' A data set containing metabolite fluxes.
+#'
+#'  \describe{
+#'   \item{metabolite}{name of measurement}
+#'   \item{abbreviation}{abbreviated metabolite name}
+#'   \item{cell_type}{
+#'   `lf` = lung fibroblasts \cr
+#'   `pasmc` = pulmonary artery smooth muscle cells}
+#'   \item{experiment}{
+#'   `02` = 0.2% oxygen for hypoxia \cr
+#'   `05` = 0.5% oxygen for hypoxia \cr
+#'   `bay` = molidustat treatment \cr
+#'   `05-bay` = 0.5% oxygen plus molidustat}
+#'   \item{batch}{a group of biological replicates analyzed similarly}
+#'   \item{date}{start date of an experiment}
+#'   \item{oxygen}{ambient oxygen level}
+#'   \item{treatment}{
+#'   `None` = no treatment \cr
+#'   `DMSO` = 0.1% DMSO \cr
+#'   `BAY` = 10 μM molidustat}
+#'   \item{flux}{fmol / cell / h, positive fluxes indicate secretion, negative fluxes
+#'   indicate uptake}
+#'   }
+"fluxes"

### 2021-03-20 R/functions.R

@@ -1,12 +1,44 @@
 # functions.R
 
+# %nin% -------------------------------------------------------------------
+
+#' Negate %in%
+#'
+#' Binary operator that returns a logical vector returning `TRUE` if there is
+#' not a match for its left operand.
+#'
+#' @param x The values to be matched
+#' @param y The values to be matched against
+#'
+"%nin%" <- function(x, table) {
+  match(x, table, nomatch = 0L) == 0L
+}
+
+# my_kable ----------------------------------------------------------------
+
+my_kable <- function(data, ...) {
+  kableExtra::kable(data, booktabs = TRUE, linesep = "", ...) %>%
+    kableExtra::kable_styling(
+      latex_options = c("hold_position"),
+      font_size = 9
+    )
+}
+
+# path_to_data ------------------------------------------------------------
+
 path_to_data <- function(nm) {
-  system.file(
-    stringr::str_c("extdata/", nm),
-    package = "Copeland.2021.hypoxia.flux"
+  list.files(
+    system.file(
+      "extdata/",
+      package = "Copeland.2021.hypoxia.flux"
+    ),
+    pattern = nm,
+    full.names = TRUE,
+    recursive = TRUE
   )
 }
 
+# path_to_reports ---------------------------------------------------------
 
 path_to_reports <- function(nm) {
   system.file(
@@ -15,6 +47,28 @@ path_to_reports <- function(nm) {
   )
 }
 
+# path_to_plots -----------------------------------------------------------
+
+path_to_plots <- function(nm) {
+  path <-
+    stringr::str_c(
+      system.file(
+        stringr::str_c("analysis/"),
+        package = "Copeland.2021.hypoxia.flux"
+      ),
+      "/figures/",
+      nm
+    )
+
+  if (dir.exists(path)) unlink(path, recursive = TRUE)
+
+  if (!dir.exists(path)) dir.create(path = path, recursive = TRUE)
+
+  path
+
+}
+
+# read_multi_excel --------------------------------------------------------
 
 read_multi_excel <- function(excel_file) {
   sheets <- readxl::excel_sheets(excel_file)
@@ -22,6 +76,8 @@ read_multi_excel <- function(excel_file) {
     rlang::set_names(sheets)
 }
 
+
[…]

### 2021-03-20 R/growth_rates.R

@@ -1,0 +1,25 @@
+#' Growth rate parameters
+#'
+#' A data set containing the growth rate and X0 values from linear fitting of
+#' cell count data.
+#'
+#' \describe{
+#'   \item{cell_type}{
+#'   `lf` = lung fibroblasts \cr
+#'   `pasmc` = pulmonary artery smooth muscle cells}
+#'   \item{experiment}{
+#'   `02` = 0.2% oxygen for hypoxia \cr
+#'   `05` = 0.5% oxygen for hypoxia \cr
+#'   `bay` = molidustat treatment \cr
+#'   `05-bay` = 0.5% oxygen plus molidustat}
+#'   \item{batch}{a group of biological replicates analyzed similarly}
+#'   \item{date}{start date of an experiment}
+#'   \item{oxygen}{ambient oxygen level}
+#'   \item{treatment}{
+#'   `None` = no treatment \cr
+#'   `DMSO` = 0.1% DMSO \cr
+#'   `BAY` = 10 μM molidustat}
+#'   \item{X0}{cell count at time 0}
+#'   \item{mu}{growth rate per hour}
+#'   }
+"growth_rates"

### 2021-03-20 R/k.R

@@ -1,0 +1,15 @@
+#' Spontaneous degradation and accumulation rates
+#'
+#' A data set containing the rates of significant metabolite accumulation or
+#' degradation in cell-free medium.
+#'
+#' \describe{
+#'   \item{metabolite}{name}
+#'   \item{oxygen}{ambient oxygen level}
+#'   \item{treatment}{
+#'   `None` = no treatment \cr
+#'   `DMSO` = 0.1% DMSO \cr
+#'   `BAY` = 10 μM molidustat}
+#'   \item{k}{accumulation or decay rate per hour, positive values indicate decay}
+#'   }
+"k"

### 2021-03-20 _targets.R

@@ -23,14 +23,154 @@ tar_option_set(
 
 # list of target objects
 list(
-  tar_target(dna_per_cell_file, path_to_data("dna-per-cell-number.xlsx")),
-  tar_target(dna_per_cell_raw, clean_dna_per_cell(dna_per_cell_file)),
-  tar_target(dna_per_cell_std, make_std_curves(dna_per_cell_raw)),
-  tar_target(dna_per_cell_clean, interp_data(dna_per_cell_raw, dna_per_cell_std)),
-  tar_target(cells_per_dna, calculate_cells_per_dna(dna_per_cell_clean)),
+
+  # dna per cell analysis
+  tar_target(
+    dna_per_cell_file,
+    path_to_data("dna-per-cell-number.xlsx"),
+    format = "file"
+  ),
+  tar_target(
+    dna_per_cell_raw,
+    clean_dna_per_cell(dna_per_cell_file)
+  ),
+  tar_target(
+    dna_per_cell_std,
+    make_std_curves(dna_per_cell_raw)
+  ),
+  tar_target(
+    dna_per_cell_clean,
+    interp_data(dna_per_cell_raw, dna_per_cell_std)
+  ),
+  tar_target(
+    cells_per_dna,
+    calculate_cells_per_dna(dna_per_cell_clean)
+  ),
   tar_render(
     dna_per_cell_report,
     path = path_to_reports("dna-per-cell.Rmd"),
+    output_dir = system.file("analysis/pdfs", package = "Copeland.2021.hypoxia.flux")
+  ),
+
+  # extracellular fluxes
+  tar_target(
+    fluxes_meta_files,
+    path_to_data("(lf|pasmc)_.*_meta\\.csv"),
+    format = "file"
+  ),
+  tar_target(
+    fluxes_meta,
+    clean_flux_meta(fluxes_meta_files)
+  ),
+  tar_target(
+    fluxes_data_files,
+    path_to_data("(lf|pasmc)_.*_[a-z]_\\d{4}-\\d{2}-\\d{2}\\.xlsx"),
+    format = "file"
+  ),
+  tar_target(
+    fluxes_data,
+    assemble_flux_data(fluxes_data_files)
+  ),
+  tar_target(
+    conc_raw,
+    clean_fluxes(fluxes_data)
+  ),
+  tar_target(
+    conc_std,
+    make_std_curves(conc_raw)
+  ),
+  tar_target(
+    conc_std_plots,
[…]

### 2021-03-20 inst/analysis/qbias-correction-factors.Rmd
complete qbias analysis

---
title: "Q Bias Correction Factors"
subtitle: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::pdf_document2:
    number_sections: true
    toc: true
    toc_depth: 1
    df_print: kable
    dev: cairo_pdf
    latex_engine: xelatex
geometry: margin = 0.5in
fontsize: 11pt
mainfont: Calibri
---

```{r setup, include = FALSE}
# load libraries
suppressPackageStartupMessages({
  devtools::load_all()
  library(tidyverse)
  library(wmo)
  library(targets)
  library(mzrtools)
})

# resolve conflicts
conflicted::conflict_prefer("filter", "dplyr")

# set global chunk options
knitr::opts_chunk$set(
  echo = FALSE, 
  message = FALSE, 
  warning = TRUE, 
  fig.align = "center", 
  out.width = "49%"
)

options(knitr.table.format = function() {
  if (knitr::is_latex_output()) 
    "latex" else "html"
})

theme_set(theme_wmo(base_family = "Calibri"))
```

```{r}
withr::with_dir(here::here(), {
  qbias_ratios <- tar_read(qbias_ratios)
  correction_factors <- tar_read(correction_factors)
})
```

\newpage

# Overview

A selected ion monitoring method was utilized to increase sensitivity for isotope detection. Others have noted quadrupole bias on the Q Exactive using this detection method (*i.e.*, the sensitivity for each isotope differs relative to its position within the m/*z* window). To account for this, naturally labeled standards were analyzed using methods where the m/*z* window is shifted -1.003355 units, the ^13^C mass defect.

# Analysis

The M1/M0 ratio was calculated for each frame shift. The resulting data were fit with a third-degree polynomial function.

```{r}
make_plots <- function(data, batch, metabolite) {
  ggplot(data) +
    aes(x = window, y = ratio) +
    geom_smooth(
      method = MASS::rlm, 
      formula = y ~ poly(x, 3),
      color = "blue", 
      se = FALSE
    ) +
    geom_point(
      size = 2, 
      alpha = 0.5, 
      color = "black"
    ) +
    geom_point(
      size = 5, 
      stat = "summary", 
      fun.data = "mean_se", 
      color = "blue"
    ) +
    labs(
      title = str_c(metabolite, batch, sep = " "), 
      x = "Window", 
      y = "Ratio"
    ) 
}
```

```{r, results = "hide", out.width = "24.9%", fig.show = "hold"}
qbias_ratios %>% 
  nest() %>% 
  mutate(plots = pmap(list(data, batch, metabolite), make_plots)) %>% 
  pull(plots)
```

```{r}
qbias_ratios %>% 
  select(-ratio) %>% 
  pivot_longer(c(M0, M1), names_to = "isotope", values_to = "value") %>% 
  mutate(
    M = str_extract(isotope, "\\d") %>% as.numeric, 
    M = M + window, 
    iso = str_c("M", M)
  ) %>% 
  left_join(correction_factors, by = c("metabolite", "batch", "iso" = "M")) %>% 
  mutate(value = value * cf) %>% 
  select(-c(cf, iso, M)) %>% 
  pivot_wider(names_from = isotope, values_from = value) %>% 
  mutate(ratio = M1 / M0) %>% 
  group_by(metabolite, batch) %>% 
  nest() %>% 
  mutate(plots = pmap(list(data, batch, metabolite), make_plots)) %>% 
  pull(plots)
```

```{r}
usethis::use_data(correction_factors, overwrite = TRUE)
```

### 2021-03-20 R/correction_factors.R

@@ -1,0 +1,14 @@
+#' Quadrupole bias correction factors
+#'
+#' A data set containing experimentally determined correction factors to adjust
+#' peak areas for isotopes detected in selected ion monitoring mode. Isotope peak
+#' areas are multiplied by the correction factor. The corrected peak area is
+#' subsequently used to determine the mass isotope distribution.
+#'
+#'  \describe{
+#'    \item{batch}{correction factors associated with specific experiments}
+#'    \item{metabolite}{name of measurement}
+#'    \item{M}{isotope}
+#'    \item{cf}{correction factor}
+#'   }
+"correction_factors"

### 2021-03-20 R/functions.R

@@ -967,3 +967,116 @@ calculate_fluxes <- function(flux_curves) {
       .data$date
     )
 }
+
+
+# calculate_ratios --------------------------------------------------------
+
+calculate_ratios <- function(file_name) {
+  readr::read_csv(file_name) %>%
+    dplyr::filter(!is.na(area)) %>%
+    tidyr::separate(
+      .data$filename,
+      into = c(NA, "window", "replicate"),
+      sep = c(1, 2),
+      convert = TRUE
+    ) %>%
+    tidyr::separate(
+      .data$ion,
+      into = c("metabolite", "isotope"),
+      sep = " ",
+      convert = TRUE
+    ) %>%
+    dplyr::mutate(carbons = case_when(
+      .data$metabolite %in% c("citrate") ~ 6,
+      .data$metabolite %in% c("2HG", "2OG", "glutamate", "glutamine") ~ 5,
+      .data$metabolite %in% c("aspartate", "malate", "succinate") ~ 4,
+      .data$metabolite %in% c("lactate", "pyruvate", "alanine", "serine") ~ 3
+    )) %>%
+    dplyr::filter(.data$window <= .data$carbons) %>%
+    tidyr::pivot_wider(names_from = .data$isotope, values_from = .data$area) %>%
+    dplyr::mutate(ratio = .data$M1 / .data$M0) %>%
+    dplyr::group_by(.data$metabolite)
+}
+
+
+# import_qbias ------------------------------------------------------------
+
+import_qbias <- function(file_list) {
+  file_list %>%
+    rlang::set_names(
+      stringr::str_extract(basename(.), pattern = "(?<=_)\\w(?=\\.csv)")
+    ) %>%
+    purrr::map_dfr(calculate_ratios, .id = "batch") %>%
+    dplyr::group_by(batch, metabolite) %>%
+    dplyr::arrange(metabolite)
+}
+
+
+# calculate_predicted_ratios ----------------------------------------------
+
+calculate_predicted_ratios <- function() {
+  isotope_library <- tibble::tribble(
+    ~ name, ~ formula, ~ polarity,
+    "2HG", "C5H8O5", "negative",
+    "2OG", "C5H6O5", "negative",
+    "alanine", "C3H7NO2", "negative",
+    "aspartate", "C4H7NO4", "negative",
[…]

### 2021-03-20 _targets.R

@@ -24,7 +24,8 @@ tar_option_set(
 # list of target objects
 list(
 
-  # dna per cell analysis
+  # dna per cell ------------------------------------------------------------
+
   tar_target(
     dna_per_cell_file,
     path_to_data("dna-per-cell-number.xlsx"),
@@ -52,7 +53,8 @@ list(
     output_dir = system.file("analysis/pdfs", package = "Copeland.2021.hypoxia.flux")
   ),
 
-  # extracellular fluxes
+  # extracellular fluxes ----------------------------------------------------
+
   tar_target(
     fluxes_meta_files,
     path_to_data("(lf|pasmc)_.*_meta\\.csv"),
@@ -171,6 +173,31 @@ list(
   tar_render(
     extracellular_fluxes_report,
     path = path_to_reports("extracellular-fluxes.Rmd"),
+    output_dir = system.file("analysis/pdfs", package = "Copeland.2021.hypoxia.flux")
+  ),
+
+  # q bias correction -------------------------------------------------------
+
+  tar_target(
+    qbias_files,
+    path_to_data("q-bias-correction"),
+    format = "file"
+  ),
+  tar_target(
+    qbias_ratios,
+    import_qbias(qbias_files)
+  ),
+  tar_target(
+    pred_ratios,
+    calculate_predicted_ratios()
+  ),
+  tar_target(
+    correction_factors,
+    calculate_correction_factors(qbias_ratios, pred_ratios)
+  ),
+  tar_render(
+    qbias_correction_factor_report,
+    path = path_to_reports("qbias-correction-factors.Rmd"),
     output_dir = system.file("analysis/pdfs", package = "Copeland.2021.hypoxia.flux")
   )
 )

### 2021-03-22 inst/analysis/mass-isotope-distributions.Rmd
complete mid analysis

---
title: "Mass Isotopomer Distributions"
subtitle: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::pdf_document2:
    number_sections: true
    toc: true
    toc_depth: 1
    df_print: kable
    dev: cairo_pdf
    latex_engine: xelatex
geometry: margin = 0.5in
fontsize: 11pt
mainfont: Calibri
---

```{r setup, include = FALSE}
# load libraries
suppressPackageStartupMessages({
  devtools::load_all()
  library(tidyverse)
  library(wmo)
  library(targets)
})

# resolve conflicts
conflicted::conflict_prefer("filter", "dplyr")

# set global chunk options
knitr::opts_chunk$set(
	echo = FALSE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE,
	out.width = "49%"
)

options(knitr.table.format = function() {
  if (knitr::is_latex_output()) 
    "latex" else "html"
})

theme_set(theme_wmo(base_family = "Calibri"))
```

```{r}
withr::with_dir(here::here(), {
  mids <- tar_read(mids)
})
```

\newpage

# Overview

Mass isotopomer distributions (MIDs) were determined for key intracellular metabolites. Cells were seeded on Day -1 in "light" labeling medium. On Day 0, medium was changed to "heavy" labeling medium containing an equimolar concentration of the stable isotope substrate. These experiments utilized the following substrates: [1,2-^13^C~2~] glucose, [U-^13^C~6~] glucose, [U-^13^C~5~] glutamine, and [U-^13^C~3~] lactate. Intracellular metabolites for every time point were extracted in 80% MeOH pre-cooled to -80 °C and analyzed by LC-MS. Each sample was run using two different LC-MS methods. First, selected ion monitoring (SIM) was used to enhance the sensitivity for key metabolites used in modeling. Second, a polarity-switching, full scan (FS) method was used. The full scan method provided labeling insights for metabolites that were not previously targeted.

# Data Processing

Peaks were interpolated using either Thermo Xcalibur or TraceFinder software.

## Quadrupole bias correction

A selected ion monitoring method was utilized to increase sensitivity for isotope detection. Others have noted quadrupole bias on the Q Exactive using this detection method (*i.e.*, the sensitivity for each isotope differs relative to its position within the m/*z* window). To account for this, naturally labeled standards were analyzed using methods where the m/*z* window is shifted -1.003355 units, the ^13^C mass defect. The M1/M0 ratio was calculated for each frame shift and corrected to the theoretical ratio. These correction factors are applied to the raw MID data.

## Natural abundance correction matrices

Isotope correction matrices are generated from the metabolite's molecular formula using a custom R package (`wmoldham/mzrtools`).

## Correct for natural isotope abundance

MIDs are calculated from the peak areas corrected for quadrupole bias for each metabolite within each run. The correction matrices are are then used to adjust experiment MIDs for the natural abundance of ^13^C.

# Analysis

```{r}
mids_filtered <-
  mids %>% 
  group_by(across(-c(date, mid))) %>% 
  remove_nested_outliers(mid, remove = TRUE)
```

```{r}
mids_filtered %>% 
  pivot_wider(names_from = method, values_from = mid) %>% 
  filter(!is.na(sim)) %>%  
  ggplot() + 
  aes(
    x = sim, 
    y = fs, 
    color = isotope
  ) +
  facet_wrap(~metabolite) +
  geom_abline(slope = 1) +
  geom_point() + 
  scale_color_manual(values = viridis::viridis(7)) +
  labs(
    x = "Selected ion monitoring", 
    y = "Full scan", 
    title = "Method comparison"
  )
```

There is reasonably good agreement between full scan and SIM methods for several of the metabolites. This observation suggests that those metabolites detected in full scan mode only are reasonable approximations. Certainly, SIM mode is preferred for metabolites detected with it, particularly as the MID correction seems to be better.

```{r remove-excess-fs}
fs <-
  mids_filtered %>% 
  filter(method == "fs") %>% 
  pull(metabolite) %>% 
  unique()

sim <- 
  mids_filtered %>% 
  filter(method == "sim") %>% 
  pull(metabolite) %>% 
  unique()

fs <- setdiff(fs, sim)

method_filter <-
  tibble(
    metabolite = c(fs, sim), 
    method = c(rep("fs", length(fs)), rep("sim", length(sim)))
  )

mids_method <-
  mids_filtered %>% 
  semi_join(method_filter, by = c("metabolite", "method"))
```

```{r tracer-label-key}
tracer_labels <- c(
  expression(paste("[1,2-"^13, "C"[2], "] glucose")),
  expression(paste("[U-"^13, "C"[6], "] glucose")), 
  expression(paste("[U-"^13, "C"[5], "] glutamine")), 
  expression(paste("[U-"^13, "C"[3], "] lactate"))
)
tracer_levels <- c("glc2", "glc6", "q5", "lac3")
```

```{r, results = "hide", fig.show = "hold", out.width = "49%"}
mids_method %>% 
  arrange(metabolite) %>% 
  filter((cell_type == "lf" & time == 72) | (cell_type == "pasmc" & time == 48)) %>% 
  group_by(method, cell_type, metabolite) %>% 
  mutate(tracer = factor(tracer, levels = tracer_levels, labels = tracer_labels)) %>% 
  nest() %>% 
  mutate(plots = pmap(
    list(data, metabolite, method, cell_type),   
    ~ ggplot(data = ..1) + 
      aes(
        x = isotope, 
        y = mid, 
        color = interaction(oxygen, treatment, sep = " | ")
      ) +
      facet_wrap(
        ~tracer, 
        nrow = 2, 
        ncol = 2, 
        scales = "free_x", 
        labeller = label_parsed
      ) +
      stat_summary(
        geom = "pointrange", 
        position = position_dodge(width = 0.3), 
        fun.data = "mean_se"
      ) +
      labs(
        x = "Isotope", 
        y = "Mole fraction", 
        title = stringr::str_c(toupper(..3), " ", ..2, " ", toupper(..4)), 
        color = NULL
      )
  )) %>% 
    pull(plots)
```

### 2021-03-22 R/functions.R

@@ -89,13 +89,7 @@ clean_technical_replicates <- function(tbl) {
     dplyr::group_by(
       dplyr::across(-c(.data$replicate, .data$value))
     ) %>%
-    dplyr::mutate(
-      value = replace(
-        .data$value,
-        which(abs(.data$value - median(.data$value)) / mad(.data$value) > 2),
-        NA
-      )
-    ) %>%
+    dplyr::mutate(value = replace_outliers(value)) %>%
     dplyr::summarise(value = mean(.data$value, na.rm = TRUE)) %>%
     dplyr::ungroup()
 }
@@ -1015,22 +1009,6 @@ import_qbias <- function(file_list) {
 # calculate_predicted_ratios ----------------------------------------------
 
 calculate_predicted_ratios <- function() {
-  isotope_library <- tibble::tribble(
-    ~ name, ~ formula, ~ polarity,
-    "2HG", "C5H8O5", "negative",
-    "2OG", "C5H6O5", "negative",
-    "alanine", "C3H7NO2", "negative",
-    "aspartate", "C4H7NO4", "negative",
-    "citrate", "C6H8O7", "negative",
-    "glutamate", "C5H9NO4", "negative",
-    "glutamine", "C5H10N2O3", "negative",
-    "lactate", "C3H6O3", "negative",
-    "malate", "C4H6O5", "negative",
-    "pyruvate", "C3H4O3", "negative",
-    "serine", "C3H7NO3", "negative",
-    "succinate", "C4H6O4", "negative"
-  )
-
   predicted_ratios <-
     isotope_library %>%
     dplyr::mutate(table = purrr::map2(
@@ -1039,7 +1017,7 @@ calculate_predicted_ratios <- function() {
       ~mzrtools::mz_iso_quant(molecule = .x, polarity = .y)[["prob_matrix"]])
     ) %>%
     dplyr::mutate(pred_ratio = purrr::map_dbl(table, ~ .x[[2, 1]]/.x[[1, 1]])) %>%
-    dplyr::select(.data$name, .data$pred_ratio)
+    dplyr::select(.data$metabolite, .data$pred_ratio)
 }
 
 
@@ -1055,7 +1033,7 @@ calculate_correction_factors <- function(qbias_ratios, pred_ratios) {
     tidyr::unnest(c(data, predict)) %>%
     dplyr::select(.data$batch, .data$metabolite, .data$window, .data$carbons, .data$predict) %>%
[…]

### 2021-03-22 R/mids.R

@@ -1,0 +1,29 @@
+#' Corrected mass isotopomer distributions
+#'
+#' A data set containing corrected mass isotope distributions. Technical
+#' replicates have been filtered from the data.
+#'
+#' \describe{
+#'   \item{method}{
+#'   `fs` = full scan \cr
+#'   `sim` = selected ion monitoring}
+#'   \item{cell_type}{
+#'   `lf` = lung fibroblasts \cr
+#'   `pasmc` = pulmonary artery smooth muscle cells}
+#'   \item{date}{start date of an experiment}
+#'   \item{tracer}{
+#'   `glc2` = \[1,2-13C2\] glucose \cr
+#'   `glc6` = \[U-13C6\] glucose \cr
+#'   `q5` = \[U-13C5\] glutamine \cr
+#'   `lac3` = \[U-13C3\] lactate}
+#'   \item{oxygen}{ambient oxygen level}
+#'   \item{treatment}{
+#'   `none` = no treatment \cr
+#'   `DMSO` = 0.1% DMSO \cr
+#'   `BAY` = 10 μM molidustat}
+#'   \item{time}{hours}
+#'   \item{metabolite}{name of measurement}
+#'   \item{isotope}{mass shift of detected isotope}
+#'   \item{mid}{mole fraction of the indicated isotope}
+#'   }
+"mids"

### 2021-03-22 _targets.R

@@ -199,5 +199,52 @@ list(
     qbias_correction_factor_report,
     path = path_to_reports("qbias-correction-factors.Rmd"),
     output_dir = system.file("analysis/pdfs", package = "Copeland.2021.hypoxia.flux")
+  ),
+
+  # mass isotopomer distributions -------------------------------------------
+
+  tar_target(
+    mid_files,
+    path_to_data("(a|b)_(fs|sim)_(lf|pasmc)_.*\\.csv"),
+    format = "file"
+  ),
+  tar_target(
+    mid_clean,
+    clean_mids(mid_files)
+  ),
+  tar_target(
+    mid_correct,
+    correct_mid(mid_clean)
+  ),
+  tar_target(
+    mids,
+    remove_mid_outliers(mid_correct)
+  ),
+  tar_target(
+    mid_curves,
+    plot_mid_curves(mids)
+  ),
+  tar_target(
+    mid_curve_plots,
+    print_plots(mid_curves$plots, mid_curves$title, "mids"),
+    format = "file"
+  ),
+  tar_render(
+    mid_report,
+    path = path_to_reports("mass-isotope-distributions.Rmd"),
+    output_dir = system.file("analysis/pdfs", package = "Copeland.2021.hypoxia.flux")
+  ),
+
+  # biomass -----------------------------------------------------------------
+
+  tar_target(
+    biomass_file,
+    path_to_data("cell-mass.csv"),
+    format = "file"
+  ),
+  tar_target(
+    biomass_clean,
+    clean_biomass(biomass_file)
   )
+
 )

### 2021-03-22 data-raw/internal-data.R

@@ -1,0 +1,120 @@
+# isotope correction ------------------------------------------------------
+
+isotope_library <-
+  tibble::tribble(
+    ~ metabolite, ~ formula, ~ polarity,
+    "2HG", "C5H8O5", "negative",
+    "2OG", "C5H6O5", "negative",
+    "alanine", "C3H7NO2", "negative",
+    "aspartate", "C4H7NO4", "negative",
+    "citrate", "C6H8O7", "negative",
+    "glutamate", "C5H9NO4", "negative",
+    "glutamine", "C5H10N2O3", "negative",
+    "lactate", "C3H6O3", "negative",
+    "malate", "C4H6O5", "negative",
+    "pyruvate", "C3H4O3", "negative",
+    "serine", "C3H7NO3", "negative",
+    "succinate", "C4H6O4", "negative",
+    "3PG", "C3H7O7P", "negative",
+    "aconitate", "C6H6O6", "negative",
+    "FBP", "C6H14O12P2", "negative",
+    "G3P", "C3H9O6P", "negative",
+    "palmitate", "C16H32O2", "negative",
+    "PEP", "C3H5O6P", "negative",
+    "sedoheptulose", "C7H14O7", "negative",
+    "DHAP", "C3H7O6P", "negative",
+    "GAP", "C3H7O6P", "negative",
+    "G1P", "C6H13O9P", "negative",
+    "G6P", "C6H13O9P", "negative",
+    "R5P", "C5H11O8P", "negative"
+  )
+
+
+# biomass calculation -----------------------------------------------------
+
+cell_composition <-
+  tibble::tribble(
+    ~ metabolite, ~ μmol_per_gDW,
+    "ALA", 0.6,
+    "ARG", 0.377,
+    "ASP", 0.359,
+    "ASN", 0.288,
+    "CYS", 0.145,
+    "GLN", 0.322,
+    "GLU", 0.386,
+    "GLY", 0.538,
+    "HIS", 0.143,
+    "ILE", 0.324,
+    "LEU", 0.564,
+    "LYS", 0.57,
+    "MET", 0.138,
+    "PHE", 0.219,
+    "PRO", 0.313,
+    "SER", 0.43,
+    "THR", 0.386,
+    "TRP", 0.044,
+    "TYR", 0.182,
+    "VAL", 0.416,
+    "glycogen", 0.279,
+    "dAMP", 0.0148,
+    "dGMP", 0.0099,
[…]

### 2021-03-22 R/functions.R
finish cell biomass calculations

@@ -41,10 +41,16 @@ path_to_data <- function(nm) {
 # path_to_reports ---------------------------------------------------------
 
 path_to_reports <- function(nm) {
-  system.file(
-    stringr::str_c("analysis/", nm),
-    package = "Copeland.2021.hypoxia.flux"
-  )
+  path <-
+    stringr::str_c(
+      system.file(
+        stringr::str_c("analysis"),
+        package = "Copeland.2021.hypoxia.flux"
+      ),
+      "/",
+      nm
+    )
+  path
 }
 
 # path_to_plots -----------------------------------------------------------
@@ -1288,4 +1294,95 @@ clean_biomass <- function(biomass_file) {
       diff = post-pre,
       cell_mass = diff / cell_number * 1E12
     )
+}
+
+
+# calculate_biomass -------------------------------------------------------
+
+calculate_biomass <- function(biomass_clean) {
+  biomass_clean %>%
+    dplyr::group_by(.data$cell_type, .data$date) %>%
+    dplyr::mutate(cell_mass = replace_outliers(.data$cell_mass)) %>%
+    dplyr::summarise(cell_mass = mean(.data$cell_mass, na.rm = TRUE)) %>%
+    dplyr::mutate(cell_mass = replace_outliers(.data$cell_mass)) %>%
+    dplyr::summarise(biomass = mean(.data$cell_mass, na.rm = TRUE))
+}
+
+
+#  calculate_biomass_equations --------------------------------------------
+
+calculate_biomass_equations <- function(biomass) {
+  biomass_coefs <- function(biomass, metabolites) {
+    μmol_per_mass %>%
+      dplyr::mutate(coefficient = .data$μmol_per_gDW * biomass) %>%
+      dplyr::filter(.data$metabolite %in% metabolites) %>%
+      dplyr::select(.data$metabolite, .data$coefficient) %>%
+      dplyr::mutate(
+        metabolite = dplyr::case_when(
+          .data$metabolite %in% c("AcCoA") ~ stringr::str_c(metabolite, ".c"),
+          TRUE ~ metabolite
+        ))
+  }
+
+  biomass_eq <- function(biomass_coefs) {
+    biomass_coefs %>%
+      dplyr::mutate(
+        combined = stringr::str_c(
+          as.character(format(coefficient, digits = 0)),
+          metabolite,
[…]

### 2021-03-22 _targets.R

@@ -245,6 +245,24 @@ list(
   tar_target(
     biomass_clean,
     clean_biomass(biomass_file)
+  ),
+  tar_target(
+    biomass,
+    calculate_biomass(biomass_clean)
+  ),
+  tar_target(
+    biomass_equations,
+    calculate_biomass_equations(biomass)
+  ),
+  tar_target(
+    biomass_equations_out,
+    save_biomass_equations(biomass_equations),
+    format = "file"
+  ),
+  tar_render(
+    biomass_report,
+    path = path_to_reports("biomass.Rmd"),
+    output_dir = system.file("analysis/pdfs", package = "Copeland.2021.hypoxia.flux")
   )
 
 )

### 2021-03-22 data-raw/internal-data.R

@@ -102,7 +102,7 @@ metabolite_composition <-
   dplyr::left_join(cell_composition, metabolite_composition, by = "class") %>%
   dplyr::mutate(
     component = dplyr::if_else(is.na(.data$component), .data$metabolite, .data$component),
-    stoichiometry = dplyr::replace_na(stoichiometry, 1),
+    stoichiometry = tidyr::replace_na(stoichiometry, 1),
     μmol_per_g = .data$μmol_per_gDW * .data$stoichiometry
   ) %>%
   dplyr::group_by(.data$component) %>%

### 2021-03-24 R/functions.R
prepare data for model generation

@@ -1367,22 +1367,211 @@ calculate_biomass_equations <- function(biomass) {
 }
 
 
-# save_biomass_equations --------------------------------------------------
 
-save_biomass_equations <- function(biomass_equations) {
+# write_matlab_input ------------------------------------------------------
+
+write_matlab_input <- function(x, column, suffix) {
   path <- path_to_reports("modeling/matlab-input")
+  column <- ensym(column)
 
   purrr::walk2(
-    biomass_equations$coefs,
-    biomass_equations$cell_type,
+    x[[column]],
+    x[["cell_type"]],
     ~readr::write_csv(
       .x,
-      file = file.path(path, stringr::str_c(.y, "_biomass.csv"))
+      file = file.path(path, stringr::str_c(.y, suffix))
     )
   )
 
   purrr::map_chr(
-    biomass_equations$cell_type,
-    ~ file.path(path, stringr::str_c(.x, "_biomass.csv"))
+    x[["cell_type"]],
+    ~ file.path(path, stringr::str_c(.x, suffix))
   )
+
+}
+
+
+# format_reactions --------------------------------------------------------
+
+format_reactions <- function(reactions_file) {
+  model_reactions <-
+    readr::read_csv(reactions_file) %>%
+    dplyr::filter(!is.na(.data$name)) %>%
+    dplyr::mutate(
+      pathway = factor(pathway, levels = c(
+        "transport",
+        "glycolysis",
+        "pentose phosphate pathway",
+        "anaplerosis",
+        "tricarboxylic acid cycle",
+        "amino acid metabolism",
+        "biomass",
+        "mixing",
+        "dilution"
+      ))
+    )
+
+  usethis::use_data(model_reactions, overwrite = TRUE)
+}
+
+
+# format_fluxes -----------------------------------------------------------
+
+format_fluxes <- function(growth_rates, fluxes) {
+  growth <-
+    growth_rates %>%
+    dplyr::mutate(metabolite = "biomass") %>%
+    dplyr::select(-.data$X0) %>%
+    dplyr::rename(flux = .data$mu)
+
+  fluxes_final <-
+    fluxes %>%
[…]

### 2021-03-24 R/model_reactions.R

@@ -1,0 +1,11 @@
+#' Model reactions
+#'
+#' A data table of model reactions formatted for INCA analysis.
+#'
+#'  \describe{
+#'   \item{index}{reaction identifier}
+#'   \item{name}{reaction name}
+#'   \item{equation}{equation for INCA input}
+#'   \item{pathway}{metabolic pathway}
+#'   }
+"model_reactions"

### 2021-03-24 _packages.R

@@ -1,12 +1,13 @@
 # Generated by targets::tar_renv(): do not edit by hand
 library(bs4Dash)
 library(clustermq)
-library(dplyr)
 library(future)
 library(gt)
 library(pingr)
+library(qs)
 library(rstudioapi)
 library(shiny)
 library(shinycssloaders)
 library(shinyWidgets)
+library(tidyverse)
 library(visNetwork)

### 2021-03-24 _targets.R

@@ -256,13 +256,43 @@ list(
   ),
   tar_target(
     biomass_equations_out,
-    save_biomass_equations(biomass_equations),
+    write_matlab_input(biomass_equations, coefs, "_biomass.csv"),
     format = "file"
   ),
   tar_render(
     biomass_report,
     path = path_to_reports("biomass.Rmd"),
     output_dir = system.file("analysis/pdfs", package = "Copeland.2021.hypoxia.flux")
+  ),
+
+  # matlab input ------------------------------------------------------------
+
+  tar_target(
+    reactions_file,
+    path_to_reports("modeling/matlab-input/reactions.csv"),
+    format = "file"
+  ),
+  tar_target(
+    model_reactions,
+    format_reactions(reactions_file)
+  ),
+  tar_target(
+    model_fluxes,
+    format_fluxes(growth_rates, fluxes)
+  ),
+  tar_target(
+    model_fluxes_out,
+    write_matlab_input(model_fluxes, data, "_fluxes.csv"),
+    format = "file"
+  ),
+  tar_target(
+    model_mids,
+    format_mids(mids)
+  ),
+  tar_target(
+    model_mids_out,
+    write_matlab_input(model_mids, data, "_mids.csv"),
+    format = "file"
   )
 
 )

### 2021-03-30 inst/manuscript/manuscript.Rmd
start working on manuscript

---
title: "Hypoxia uncouples HIF gene transcription and metabolic flux in proliferating primary cells"

author:
- Courtney A. Copeland: 
    institute: 1
- Benjamin A. Olenchock: 
    institute: 1, 2
- David R. Ziehr:
    institute: 1, 4
- Sarah McGarrity: 
    institute: 1, 5
- Jamey D. Young: 
    institute: 3
- Joseph Loscalzo: 
    institute: 1
- William M. Oldham: 
    institute: 1
    email: woldham@bwh.harvard.edu
    correspondence: true
institute:
- 1: Department of Medicine, Brigham and Women's Hospital and Harvard Medical School, Boston, MA
- 2: Regeneron Pharmaceuticals, Tarrytown, NY
- 3: Departments of Chemical & Biomolecular Engineering and Molecular Physiology & Biophysics, Vanderbilt University, Nashville, TN
- 4: Department of Medicine, Massachusetts General Hospital and Harvard Medical School, Boston, MA
- 5: Center for Systems Biology, School of Health Sciences, University of Iceland, Reykjavik, Iceland

bibliography_main: library.bib
bibliography_software: packages.bib
csl: cell-metabolism.csl
link-citations: yes

output: 
  bookdown::pdf_document2:
    template: template.tex
    keep_tex: false
    df_print: kable
    toc: false
    number_sections: false
    latex_engine: xelatex
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
      - '--lua-filter=pagebreak.lua'
      - '--lua-filter=multiple-bibliographies.lua'
compact-title: true
geometry: margin = 0.75in
fontsize: 11pt
mainfont: Calibri
graphics: yes
header-includes:
- \usepackage{lineno}
- \linenumbers
- \setlength\linenumbersep{1cm}
- \usepackage{setspace}
- \doublespacing
- \makeatletter\renewcommand*{\fps@figure}{htbp}\makeatother
- \setlength{\parskip}{0.75\baselineskip plus 2pt minus 2pt}
- \usepackage{float}
---

```{r setup, include = FALSE}
# library(Copeland.2021.hypoxia.flux)
devtools::load_all()

knitr::opts_chunk$set(
  echo = FALSE, 
  warning = FALSE, 
  message = FALSE, 
  out.width = "100%", 
  fig.align = "center", 
  fig.pos = "H", 
  out.extra = ""
)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

knitr::opts_knit$set(kable.force.latex = FALSE)

options(tinytex.verbose = TRUE)

# kable function
my_kable <- function(data, ...) {
  kableExtra::kable(data, booktabs = TRUE, linesep = "", ...) %>% 
    kableExtra::kable_styling(latex_options = c("HOLD_position"), font_size = 11)
}
```

# Summary 
(150 words)

# Keywords

hypoxia, metabolic flux analysis, hypoxia-inducible factor, prolyl hydroxylase, metabolism

\newpage

# Introduction

Metazoan cells depend on aerobic respiration to meet cellular energy demands. With an inadequate oxygen supply, or hypoxia, cells must reduce energy consumption and shift energy production away from oxidative phosphorylation. Cells accomplish this goal through stabilization of the hypoxia-inducible transcription factor 1α (HIF-1α), which activates the transcription of glucose transporters, glycolytic enzymes, lactate dehydrogenase, and pyruvate dehydrogenase kinase, while decreasing the expression of proteins in the tricarboxylic acid (TCA) cycle and electron transport chain [@RN1980, @RN1198]. Overall, these changes in gene transcription should increase glycolytic capacity and divert glucose-derived pyruvate from oxidative phosphorylation toward lactate fermentation to maintain energy production and minimize the formation of reactive oxygen species [@RN2525].

While this "glycolytic shift" of primary carbon metabolism is well-described, the effects of hypoxia on other metabolic pathways are an area of active investigation [@RN953, @RN1603, @RN1198]. Since hypoxia is a prominent feature of cancer biology as tumor growth outstrips blood supply, most detailed metabolic studies of cell metabolism in hypoxia have used tumor cell models [@RN2395, @RN1447, @RN1438, @RN634, @RN1200, @RN975]. For example, stable isotope tracing and metabolic flux analyses identified a critical role for reductive carboxylation of glutamine-derived 2-oxoglutarate for lipid biosynthesis in tumor growth [@RN1447; @RN628; @RN1966; @RN2395], and metabolomic studies identified aspartate as a limiting metabolite for cancer cell proliferation under hypoxia [@RN634]. By contrast, comparatively little is known about metabolic adaptations of primary cells to hypoxia and the importance of the metabolic pathways described above remain to be elucidated. This information would provide important context for understanding the extent to which cancer cell metabolism responds differently to hypoxic stress. Given the metabolic adaptations required for rapid proliferation in cancer cells, we hypothesized that hypoxia would elicit different metabolic responses in primary cells than has been observed previously in studies of cancer cell metabolism.

To test this hypothesis, we developed models of bioenergetic carbon flux in human primary cells cultured under 21% or 0.5% oxygen conditions. We found that hypoxia fails to increase glycolysis in primary cells despite robust up-regulation of the HIF-1α transcriptional program. In normoxia, HIF-1α activation by the prolyl hydroxylase inhibitor molidustat (BAY-85-3934, "BAY") [@RN580] did increase glycolysis and lactate efflux; however, hypoxia abrogated this response. These findings suggest the existence of hypoxia-dependent metabolic regulatory mechanisms that override the effects of HIF-1α-dependent up-regulation of glycolytic gene expression.

# Results

The goal of this study was to identify the metabolic changes associated with hypoxia in proliferating primary cells. Cells were seeded and placed into hypoxia for 24 h prior to sample collection to provide adequate time for activation of the hypoxia-dependent transcriptional program. From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (Figure \ref{fig:f1}A), thought to be an indicator of metabolic steady state. Lung fibroblasts cultured in 0.5% oxygen grew more slowly (Figure \ref{fig:f1}B), but slower growth was not associated with decreased cell viability (Figure \ref{fig:f1}C). These cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1, (GLUT1), and lactate dehydrogenase A (LDHA) (Figure \ref{fig:f1}D-H). These changes persisted for the duration of the experimental time course.

## Extracellular flux analysis reveals little effect of hypoxia on glycolysis

Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (Figure \ref{fig:f1}I-J and Supplementary Figure 1). Flux calculations incorporated the changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@RN1523] (Supplementary Figure 1). Surprisingly, while glucose uptake was modestly increased in hypoxia, lactate efflux was decreased (Figure \ref{fig:f1}I) despite activation of the HIF-1 transcriptional program. Similar findings were observed in pulmonary artery smooth muscle cells (Supplementary Figure 2) and when the ambient oxygen level was decreased further to 0.2% (Supplementary Figure 3). In addition to glucose and lactate, extracellular fluxes of pyruvate and amino acids were generally decreased in hypoxia, including a marked decrease in glutamine uptake. Notably, hypoxia was previously shown to increase glutamine uptake in studies of cancer cell metabolism [@RN1447; @RN2395; @RN628].

Given that hypoxia did not increase glucose and lactate fluxes as expected from the associated changes in glycolytic gene expression, we next assessed the capacity of HIF-1α to augment glycolysis in lung fibroblasts. Cells were treated with BAY to stabilize HIF-1α under 21% oxygen conditions (Figure \ref{fig:f2}). Similar to hypoxia, BAY decreased cell growth rate (Figure \ref{fig:f2}A-B) and activated the HIF-1 transcriptional program (Figure \ref{fig:f2}C-G). Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1 target gene transcription and protein expression. In normoxia, this transcriptional program was associated with increased glucose uptake and lactate efflux (Figure \ref{fig:f2}H). Relatively modest effects on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (Figure \ref{fig:f2}J).

# Discussion

# Acknowledgements

# Author Contributions

# Declaration of Interests

# Figure Legends

# Tables

# STAR Methods

### 2021-04-02 inst/analysis/flux-differences.Rmd
work on mid figures for manuscript

--- 2021-03-22 inst/analysis/mass-isotope-distributions.Rmd
@@
 ---
-title: "Mass Isotopomer Distributions"
+title: "Model Analysis"
 subtitle: "`r format(Sys.time(), '%B %d, %Y')`"
@@
 withr::with_dir(here::here(), {
-  mids <- tar_read(mids)
+  model_fluxes <- tar_read(model_fluxes)
 })
@@
 
-Mass isotopomer distributions (MIDs) were determined for key intracellular metabolites.
-Cells were seeded on Day -1 in "light" labeling medium.
-On Day 0, medium was changed to "heavy" labeling medium containing an equimolar concentration of the stable isotope substrate.
-These experiments utilized the following substrates: [1,2-^13^C~2~] glucose, [U-^13^C~6~] glucose, [U-^13^C~5~] glutamine, and [U-^13^C~3~] lactate.
-Intracellular metabolites for every time point were extracted in 80% MeOH pre-cooled to -80 °C and analyzed by LC-MS.
-Each sample was run using two different LC-MS methods.
-First, selected ion monitoring (SIM) was used to enhance the sensitivity for key metabolites used in modeling.
-Second, a polarity-switching, full scan (FS) method was used.
-The full scan method provided labeling insights for metabolites that were not previously targeted.
+In this document, we will analyze the output from the metabolic flux maps.
 
-# Data Processing
-
-Peaks were interpolated using either Thermo Xcalibur or TraceFinder software.
-
-## Quadrupole bias correction
-
-A selected ion monitoring method was utilized to increase sensitivity for isotope detection.
-Others have noted quadrupole bias on the Q Exactive using this detection method (*i.e.*, the sensitivity for each isotope differs relative to its position within the m/*z* window).
-To account for this, naturally labeled standards were analyzed using methods where the m/*z* window is shifted -1.003355 units, the ^13^C mass defect.
-The M1/M0 ratio was calculated for each frame shift and corrected to the theoretical ratio.
-These correction factors are applied to the raw MID data.
-
-## Natural abundance correction matrices
-
-Isotope correction matrices are generated from the metabolite's molecular formula using a custom R package (`wmoldham/mzrtools`).
-
-## Correct for natural isotope abundance
-
-MIDs are calculated from the peak areas corrected for quadrupole bias for each metabolite within each run.
-The correction matrices are are then used to adjust experiment MIDs for the natural abundance of ^13^C.
+```{r}
+usethis::use_data(model_fluxes, overwrite = TRUE)
+```
 
@@
 
-```{r}
-mids_filtered <-
-  mids %>% 
-  group_by(across(-c(date, mid))) %>% 
-  remove_nested_outliers(mid, remove = TRUE)
-```
+## Significant Differences
 
-```{r}
-mids_filtered %>% 
-  pivot_wider(names_from = method, values_from = mid) %>% 
-  filter(!is.na(sim)) %>%  
-  ggplot() + 
-  aes(
-    x = sim, 
-    y = fs, 
-    color = isotope
-  ) +
-  facet_wrap(~metabolite) +
-  geom_abline(slope = 1) +
-  geom_point() + 
-  scale_color_manual(values = viridis::viridis(7)) +
-  labs(
-    x = "Selected ion monitoring", 
-    y = "Full scan", 
-    title = "Method comparison"
-  )
-```
-
-There is reasonably good agreement between full scan and SIM methods for several of the metabolites.
-This observation suggests that those metabolites detected in full scan mode only are reasonable approximations.
-Certainly, SIM mode is preferred for metabolites detected with it, particularly as the MID correction seems to be better.
-
-```{r remove-excess-fs}
-fs <-
-  mids_filtered %>% 
-  filter(method == "fs") %>% 
-  pull(metabolite) %>% 
-  unique()
-
-sim <- 
-  mids_filtered %>% 
-  filter(method == "sim") %>% 
-  pull(metabolite) %>% 
-  unique()
-
-fs <- setdiff(fs, sim)
-
-method_filter <-
-  tibble(
-    metabolite = c(fs, sim), 
-    method = c(rep("fs", length(fs)), rep("sim", length(sim)))
-  )
-
-mids_method <-
-  mids_filtered %>% 
-  semi_join(method_filter, by = c("metabolite", "method"))
-```
-
-```{r tracer-label-key}
-tracer_labels <- c(
-  expression(paste("[1,2-"^13, "C"[2], "] glucose")),
-  expression(paste("[U-"^13, "C"[6], "] glucose")), 
-  expression(paste("[U-"^13, "C"[5], "] glutamine")), 
-  expression(paste("[U-"^13, "C"[3], "] lactate"))
-)
-tracer_levels <- c("glc2", "glc6", "q5", "lac3")
-```
-
-```{r, results = "hide", fig.show = "hold", out.width = "49%"}
-mids_method %>% 
-  arrange(metabolite) %>% 
-  filter((cell_type == "lf" & time == 72) | (cell_type == "pasmc" & time == 48)) %>% 
-  group_by(method, cell_type, metabolite) %>% 
-  mutate(tracer = factor(tracer, levels = tracer_levels, labels = tracer_labels)) %>% 
-  nest() %>% 
-  mutate(plots = pmap(
-    list(data, metabolite, method, cell_type),   
-    ~ ggplot(data = ..1) + 
-      aes(
-        x = isotope, 
-        y = mid, 
-        color = interaction(oxygen, treatment, sep = " | ")
-      ) +
-      facet_wrap(
-        ~tracer, 
-        nrow = 2, 
-        ncol = 2, 
-        scales = "free_x", 
-        labeller = label_parsed
-      ) +
-      stat_summary(
-        geom = "pointrange", 
-        position = position_dodge(width = 0.3), 
-        fun.data = "mean_se"
-      ) +
-      labs(
-        x = "Isotope", 
-        y = "Mole fraction", 
-        title = stringr::str_c(toupper(..3), " ", ..2, " ", toupper(..4)), 
-        color = NULL
-      )
-  )) %>% 
-    pull(plots)
-```
+Identify reactions that are significantly different based on whether the confidence intervals overlap. This is more conservative than p < 0.05.

### 2021-04-02 inst/manuscript/manuscript.Rmd
work on mid figures for manuscript

@@
 In normoxia, HIF-1α activation by the prolyl hydroxylase inhibitor molidustat (BAY-85-3934, "BAY") [@RN580] did increase glycolysis and lactate efflux; however, hypoxia abrogated this response.
-These findings suggest the existence of hypoxia-dependent metabolic regulatory mechanisms that override the effects of HIF-1α-dependent up-regulation of glycolytic gene expression.
+These findings suggest the existence of hypoxia-dependent regulatory mechanisms that override the metabolic consequences of HIF-1α-dependent up-regulation of glycolytic gene expression in human primary cells.
 
@@
 
-## Extracellular flux analysis reveals little effect of hypoxia on glycolysis
+## Extracellular flux analysis reveals little impact of hypoxia on glycolysis
 
-Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (Figure \ref{fig:f1}I-J and Supplementary Figure 1).
+Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (Figure \ref{fig:f1}I-J, Supplementary Figures 1-3).
 Flux calculations incorporated the changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@RN1523] (Supplementary Figure 1).
 Surprisingly, while glucose uptake was modestly increased in hypoxia, lactate efflux was decreased (Figure \ref{fig:f1}I) despite activation of the HIF-1 transcriptional program.
-Similar findings were observed in pulmonary artery smooth muscle cells (Supplementary Figure 2) and when the ambient oxygen level was decreased further to 0.2% (Supplementary Figure 3).
-In addition to glucose and lactate, extracellular fluxes of pyruvate and amino acids were generally decreased in hypoxia, including a marked decrease in glutamine uptake.
+Similar findings were observed when the ambient oxygen level was decreased further to 0.2% (Supplementary Figure 2) and in pulmonary artery smooth muscle cells (PASMCs; Supplementary Figure 3).
+In addition to glucose and lactate, the absolute extracellular fluxes of pyruvate and amino acids were generally decreased in hypoxia, including a marked decrease in glutamine uptake in lung fibroblasts.
 Notably, hypoxia was previously shown to increase glutamine uptake in studies of cancer cell metabolism [@RN1447; @RN2395; @RN628].
+Glutamine uptake did increase in PASMCs, as did the uptake of branch chain amino acids and arginine, suggesting differential responses of these mesenchymal cells to hypoxia.
 
-Given that hypoxia did not increase glucose and lactate fluxes as expected from the associated changes in glycolytic gene expression, we next assessed the capacity of HIF-1α to augment glycolysis in lung fibroblasts.
+Given that hypoxia did not increase glucose and lactate fluxes as expected from the associated increases in glycolytic gene expression, we next assessed the capacity of HIF-1α to augment glycolysis in lung fibroblasts.
 Cells were treated with BAY to stabilize HIF-1α under 21% oxygen conditions (Figure \ref{fig:f2}).
@@
 Relatively modest effects on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (Figure \ref{fig:f2}J).
+
+## Stable isotope tracing
+
+To characterize further the extent of hypoxia-induced metabolic reprogramming, we next treated lung fibroblasts with stable carbon isotopes of glucose and glutamine to trace label incorporation into key carbon utilization pathways (Figure \ref{fig:f3} and Supplementary Figure 4).
+Overall, hypoxia-treated cells had decreased label incorporation into downstream metabolites (*i.e.*, the unlabeled, or M0, fraction was greater).
+This finding is consistent with the extracellular flux measurements suggesting slower substrate utilization by hypoxic cells.
+BAY treatment recapitulated the labeling pattern observed with hypoxia, suggesting similar intracellular metabolite flux between these two conditions.
+Beyond this observation, the labeling patterns in hypoxia- and BAY-treated cells were similar to their respective controls, arguing against marked metabolic reprogramming in response to prolyl hydroxylase inhibition by either hypoxia or BAY.
+Compared to previous studies of metabolic flux in cancer cells [@RN2395; @RN1447], no hypoxia-mediated increase in M5-labeled citrate from [U-^13^C~5~] glutamine was observed, indicating no increase in reductive carboxylation for lipid synthesis.
+The overall fraction of M5-citrate in these cells was low (< 6%).
 

### 2021-04-02 R/functions.R

@@ -1414,6 +1414,8 @@ format_reactions <- function(reactions_file) {
     )
 
   usethis::use_data(model_reactions, overwrite = TRUE)
+
+  model_reactions
 }
 
 # format_fluxes -----------------------------------------------------------
@@ -1516,28 +1518,18 @@ format_mids <- function(mids) {
         (.data$method == "fs" & .data$metabolite %in% c("FBP", "3PG"))
     )
 
-  mids_pasmc_05_new <-
+  mids_pasmc_05 <-
     mids_filtered %>%
     dplyr::filter(
       .data$cell_type == "pasmc" &
-        .data$oxygen %in% c("21%", "0.5%") &
-        .data$date %in% c("2019-12-06", "2019-12-11", "2019-12-16")
+        .data$oxygen %in% c("21%", "0.5%")
     ) %>%
     dplyr::filter(
       .data$method == "sim" |
         (.data$method == "fs" & .data$metabolite %in% c("FBP", "3PG"))
     )
 
-  mids_pasmc_05_old <-
-    mids_filtered %>%
-    dplyr::filter(
-      .data$cell_type == "pasmc" &
-        .data$oxygen %in% c("21%", "0.5%") &
-        .data$date %in% c("2018-05-25", "2018-05-29", "2018-06-02", "2018-06-08")
-    ) %>%
-    dplyr::filter(!(.data$method == "fs" & .data$metabolite == "2OG"))
-
-  dplyr::bind_rows(mids_lf_new, mids_pasmc_05_new, mids_pasmc_05_old) %>%
+  dplyr::bind_rows(mids_lf_new, mids_pasmc_05) %>%
     dplyr::group_by(
       .data$method,
       .data$cell_type,
@@ -2392,3 +2384,190 @@ arrange_s1 <- function(a, b, c, d) {
       theme = theme(plot.margin = margin(-3, -3, -3, -3))
     )
 }
+
+# plot_mids ---------------------------------------------------------------
+
+plot_mids <- function(df) {
+
+  tracer_labels <-
+    c(expression(paste("[1,2-"^13, "C"[2], "] glucose")),
+      expression(paste("[U-"^13, "C"[6], "] glucose")),
+      expression(paste("[U-"^13, "C"[5], "] glutamine")),
+      expression(paste("[U-"^13, "C"[3], "] lactate")))
+
+  tracer_levels <-
+    c("glc2", "glc6", "q5", "lac3")
+
+  df %>%
+    dplyr::mutate(
[…]

### 2021-04-02 _targets.R

@@ -625,5 +625,56 @@ list(
     s3_figure,
     write_figures(s3, "s3.pdf", width = 5, height = 5),
     format = "file"
+  ),
+
+  # M3 ----------------------------------------------------------------------
+
+  tar_target(
+    m3,
+    plot_manuscript_mids(model_mids)
+  ),
+  tar_target(
+    m3_figure,
+    write_figures(m3, "m3.pdf", width = 7, height = 5),
+    format = "file"
+  ),
+
+  # S4 ----------------------------------------------------------------------
+
+  tar_target(
+    s4,
+    plot_lf_mids(model_mids)
+  ),
+  tar_target(
+    s4_figure,
+    write_figures(s4, "s4.pdf", width = 7, height = 5)
+  ),
+
+  # S5 ----------------------------------------------------------------------
+
+  tar_target(
+    s5,
+    plot_pasmc_mids(model_mids)
+  ),
+  tar_target(
+    s5_figure,
+    write_figures(s5, "s5.pdf", width = 5, height = 9.5)
+  ),
+
+  # model fluxes ------------------------------------------------------------
+
+  tar_target(
+    map_flux_files,
+    path_to_data("(lf|pasmc)_(21|05|dmso|bay)\\.csv"),
+    format = "file"
+  ),
+  tar_target(
+    map_fluxes,
+    clean_model_fluxes(map_flux_files, model_reactions)
   )
+  # tar_target(
+  #   flux_differences,
+  #   calculate_flux_differences(map_fluxes)
+  # )
+
 )

### 2021-04-07 inst/analysis/flux-differences.Rmd
run check()

@@
   model_fluxes <- tar_read(model_fluxes)
+  flux_differences <- tar_read(flux_differences)
 })
@@
 usethis::use_data(model_fluxes, overwrite = TRUE)
+usethis::use_data(flux_differences, overwrite = TRUE)
 ```
@@
 Identify reactions that are significantly different based on whether the confidence intervals overlap. This is more conservative than p < 0.05.
+
+```{r}
+calculate_ratio <- function(norm) {
+  flux_differences %>% 
+    filter(normalization == norm) %>% 
+    # filter(!is.na(ratio)) %>% 
+    select(-c(normalization, equation, ctl, contains(c("flux", "lb", "ub")))) %>% 
+    pivot_wider(names_from = exp, values_from = ratio)
+}
+
+print_flux_table <- function(df, caption) {
+  df %>% 
+    mutate(across(c(`0.5%`, BAY), ~as.character(signif(., 3)))) %>% 
+    kableExtra::kable(
+      caption = caption, 
+      col.names = c("Cell type", "Pathway", "Index", "ID", "Type", "0.5%", "BAY"), 
+      align = "clccccc", 
+      escape = TRUE, 
+      digits = 2, 
+      booktabs = TRUE, 
+      longtable = TRUE, 
+      linesep = ""
+    ) %>% 
+    kableExtra::kable_styling(font_size = 9)
+}
+```
+
+```{r}
+calculate_ratio("none") %>% print_flux_table("Significant Flux Differences")
+```
+
+Do any of these differences have a negative ratio?
+
+```{r}
+calculate_ratio("none") %>% 
+  filter(`0.5%` < 0 | BAY < 0) %>% 
+  print_flux_table("Reversed fluxes")
+```
+
+For both of these models, some of the PPP fluxes reverse.
+
+## Normalized differences
+
+It doesn't look like there is any dramatic reprogramming of the intracellular metabolic pathways secondary to hypoxia or BAY treatment.
+In other words, most of the differences are related to absolute rates rather than differential usage.
+Another way to look at this might be to normalize the fluxes to glucose uptake or biomass fluxes.
+This would show all fluxes in terms of total glucose use or growth rate, for example.
+
+```{r}
+calculate_ratio("growth") %>% print_flux_table("Significant differences in growth-normalized fluxes")
+```
+
+```{r}
+calculate_ratio("glucose") %>% print_flux_table("Significant differences in glucose-normalized fluxes")
+```

### 2021-04-13 R/0_figures.R
rename files as load_all wasn't detecting

@@ -10,7 +10,7 @@ theme_plots <- function() {
     ggplot2::theme(
       plot.margin = ggplot2::margin(5, 5, 5, 5),
       plot.tag = ggplot2::element_text(face = "bold"),
-      axis.title.y = ggplot2::element_text(margin = ggplot2::margin(r = 3))
+      axis.title.y.left = ggplot2::element_text(margin = ggplot2::margin(r = 3))
     )
 }
 
@@ -58,12 +58,14 @@ write_figures <- function(plot, filename) {
   ggplot2::ggsave(
     filename = filename,
     plot = plot,
+    device = cairo_pdf,
     path = path,
     width = overall_width,
     height = overall_height,
-    units = "in",
-    device = cairo_pdf
+    units = "in"
   )
+
+  if (file.exists("Rplots.pdf")) unlink("Rplots.pdf")
 
   stringr::str_c(path, "/", filename)
 
@@ -160,12 +162,13 @@ arrange_s6 <- function(a, b, c, d) {
 
 # arrange_m5 --------------------------------------------------------------
 
-arrange_m5 <- function(a, b, c, d, e) {
+arrange_m5 <- function(a, b, c, d, e, f) {
   layout <- "
   abc
-  de#
+  def
+  ##f
   "
-  a + b + c + d + e +
+  a + b + c + d + e + f +
     theme_patchwork(
       design = layout,
       widths = unit(1.25, "in"),

### 2021-04-13 R/0_rnaseq.R

@@ -4,23 +4,25 @@
 
 count_rnaseq <- function() {
 
-  count_data <- lf.hypoxia.molidustat.rnaseq::count_data
+  dds <- rnaseq.lf.hypoxia.molidustat::lf_hyp_bay_se
 
-  pheno_data <-
-    lf.hypoxia.molidustat.rnaseq::pheno_data %>%
+  SummarizedExperiment::colData(dds) <-
+    SummarizedExperiment::colData(dds) %>%
+    tibble::as_tibble() %>%
     dplyr::mutate(
+      experiment = factor(experiment),
       oxygen = factor(oxygen, levels = c("21%", "0.5%"), labels = c("N", "H")),
       treatment = factor(treatment, labels = c("DMSO", "BAY")),
       group = stringr::str_c(oxygen, treatment, sep = "."),
       group = factor(group, levels = c("N.DMSO", "H.DMSO", "N.BAY", "H.BAY"))
-    )
+    ) %>%
+    as("DataFrame")
 
   design <- ~ experiment + group
 
   dds <-
-    DESeq2::DESeqDataSetFromMatrix(
-      countData = count_data,
-      colData = pheno_data,
+    DESeq2::DESeqDataSet(
+      dds,
       design = design
     )
 
@@ -144,7 +146,11 @@ identify_deg <- function(dds, expr) {
     tidy = TRUE,
     parallel = TRUE
   ) %>%
-    dplyr::rename(symbol = row) %>%
+    dplyr::left_join(
+      tibble::as_tibble(SummarizedExperiment::rowData(dds), rownames = "row"),
+      by = "row"
+    ) %>%
+    dplyr::rename(symbol = hgnc_symbol) %>%
     dplyr::arrange(padj)
 
 }
@@ -182,20 +188,18 @@ plot_rnaseq_venn <- function(dds) {
 
 # plot_rnaseq_volcano -----------------------------------------------------
 
-plot_rnaseq_volcano <- function(dds) {
+plot_rnaseq_volcano <- function(results) {
 
   df <-
-    identify_deg(dds, expr((h.dmso - n.dmso) - (n.bay - n.dmso))) %>%
-    dplyr::arrange(desc(abs(stat))) %>%
-    dplyr::mutate(fc = 2 ^ log2FoldChange)
+    tibble::as_tibble(results)
 
   ggplot2::ggplot(df) +
     ggplot2::aes(
-      x = fc,
+      x = 2 ^ log2FoldChange,
       y = padj
     ) +
[…]

### 2021-04-13 inst/analysis/rnaseq.Rmd
finish gsea analysis

---
title: "RNA-seq Analysis"
subtitle: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::pdf_document2:
    number_sections: true
    toc: true
    toc_depth: 1
    df_print: kable
    dev: cairo_pdf
    latex_engine: xelatex
geometry: margin = 0.5in
fontsize: 11pt
mainfont: Calibri
---

```{r setup, include = FALSE}
# load libraries
suppressPackageStartupMessages({
  devtools::load_all()
  library(tidyverse)
  library(wmo)
  library(targets)
})

# resolve conflicts
conflicted::conflict_prefer("filter", "dplyr")

# set global chunk options
knitr::opts_chunk$set(
  echo = FALSE, 
  message = FALSE, 
  warning = FALSE, 
  fig.align = "center", 
  out.width = "49%"
)

options(knitr.table.format = function() {
  if (knitr::is_latex_output()) 
    "latex" else "html"
})

theme_set(theme_wmo(base_family = "Calibri"))
```

```{r}
withr::with_dir(here::here(), {
  rnaseq_pca <- tar_read(rnaseq_pca)
  rnaseq_different_differences <- tar_read(rnaseq_different_differences)
})
```

\newpage

# Overview

We observed that proliferating primary cells exposed to hypoxia do not increase glucose uptake and lactate efflux despite up-regulation of glucose transporters and glycolytic genes. When these cells are treated with the prolyl hydroxylase inhibitor molidustat in normoxia, the expected increases in glycolytic flux are observed. Interestingly, when molidustat-treated cells are cultured in hypoxia, hypoxia blocks molidustat-mediated increases in glycolysis. In an effort to identify the mechanism mediating this effect, we performed RNA-seq on lung fibroblasts treated for three days with 0.5% oxygen or molidustat (10 μM) with 21% and DMSO (0.1%) controls. RNA was collected and purified and sent for RNA-seq analysis.

# Analysis

We will use the `DESeq2` package to identify differentially expresssed genes. We will use the model `~ experiment + oxygen * treatment`.

## PCA

```{r}
rnaseq_pca
```

## Differential Expression

In this analysis, will identify those genes where the effects of hypoxia and molidustat are different, which will hopefully point to mechanisms by which hypoxia may regulate cell metabolism differently than PHD2 inhibition.

```{r}
head(rnaseq_different_differences, 20)
```

### 2021-04-13 _targets.R

@@ -6,7 +6,7 @@ devtools::load_all()
 library(targets)
 library(tarchetypes)
 
-src <- list.files("R", pattern = "^_.*\\.R", full.names = TRUE)
+src <- list.files("R", pattern = "^0_.*\\.R", full.names = TRUE)
 invisible(lapply(src, source))
 
 conflicted::conflict_prefer("filter", "dplyr")
@@ -21,7 +21,7 @@ future::plan(future::multisession(workers = future::availableCores() - 1))
 # target-specific options
 tar_option_set(
   packages = c("tidyverse", "patchwork"),
-  imports = c("lf.hypoxia.molidustat.rnaseq"),
+  imports = c("rnaseq.lf.hypoxia.molidustat"),
   format = "qs"
 )
 
@@ -422,12 +422,24 @@ list(
     plot_rnaseq_pca(pca_data)
   ),
   tar_target(
+    rnaseq_different_differences,
+    identify_deg(dds, expr((h.dmso - n.dmso) - (n.bay - n.dmso)))
+  ),
+  tar_target(
     rnaseq_volcano,
-    plot_rnaseq_volcano(dds)
+    plot_rnaseq_volcano(rnaseq_different_differences)
   ),
   tar_target(
     rnaseq_venn,
     plot_rnaseq_venn(dds)
+  ),
+  tar_target(
+    rnaseq_gsea,
+    run_gsea(rnaseq_different_differences)
+  ),
+  tar_target(
+    rnaseq_gsea_plot,
+    plot_gsea(rnaseq_gsea, "HALLMARK")
   ),
 
   # M1 ----------------------------------------------------------------------
@@ -836,7 +848,7 @@ list(
   # ),
   tar_target(
     m5,
-    arrange_m5(m5a, m5b, m5c, rnaseq_pca, rnaseq_volcano)
+    arrange_m5(m5a, m5b, m5c, rnaseq_pca, rnaseq_volcano, rnaseq_gsea_plot)
   ),
   tar_target(
     m5_figure,

### 2021-04-14 inst/analysis/rnaseq.Rmd
work on TF binding analysis

@@
   rnaseq_different_differences <- tar_read(rnaseq_different_differences)
+  rnaseq_gsea_plot <- tar_read(rnaseq_gsea_plot)
 })
@@
 ```{r}
-head(rnaseq_different_differences, 20)
+head(rnaseq_different_differences, 20) %>% 
+  my_kable()
 ```
+
+## Gene Set Enrichment Analysis
+
+```{r}
+rnaseq_gsea_plot
+```
+
+## Transcription Factor Analysis
+
+```{r}
+
+```
+
+```{r}
+
+```
+
+```{r}
+tf_rnk$TFranking_plot
+```

### 2021-04-14 R/0_figures.R

@@ -162,13 +162,14 @@ arrange_s6 <- function(a, b, c, d) {
 
 # arrange_m5 --------------------------------------------------------------
 
-arrange_m5 <- function(a, b, c, d, e, f) {
+arrange_m5 <- function(a, b, c, d, e, f, g, h) {
   layout <- "
   abc
-  def
-  ##f
+  deg
+  ffg
+  hh#
   "
-  a + b + c + d + e + f +
+  a + b + c + d + e + f + g + h +
     theme_patchwork(
       design = layout,
       widths = unit(1.25, "in"),

### 2021-04-14 R/0_rnaseq.R

@@ -138,6 +138,10 @@ identify_deg <- function(dds, expr) {
   con <- eval(expr)
   # con <- (h.dmso - n.dmso) - (n.bay - n.dmso)
 
+  annots <-
+    tibble::as_tibble(SummarizedExperiment::rowData(dds), rownames = "row") %>%
+    dplyr::select(row, hgnc_symbol, description)
+
   DESeq2::results(
     dds,
     contrast = con,
@@ -146,11 +150,9 @@ identify_deg <- function(dds, expr) {
     tidy = TRUE,
     parallel = TRUE
   ) %>%
-    dplyr::left_join(
-      tibble::as_tibble(SummarizedExperiment::rowData(dds), rownames = "row"),
-      by = "row"
-    ) %>%
+    dplyr::left_join(annots, by = "row") %>%
     dplyr::rename(symbol = hgnc_symbol) %>%
+    dplyr::relocate(symbol, description, .after = "row") %>%
     dplyr::arrange(padj)
 
 }
@@ -193,28 +195,63 @@ plot_rnaseq_volcano <- function(results) {
   df <-
     tibble::as_tibble(results)
 
+  left <-
+    df %>%
+    dplyr::filter(log2FoldChange < 0) %>%
+    dplyr::slice_min(stat, n = 10)
+
+  right <-
+    df %>%
+    dplyr::filter(log2FoldChange > 0) %>%
+    dplyr::slice_max(stat, n = 10)
+
   ggplot2::ggplot(df) +
     ggplot2::aes(
-      x = 2 ^ log2FoldChange,
+      # x = 2 ^ log2FoldChange,
+      x = log2FoldChange,
       y = padj
+    ) +
+    # ggrepel::geom_text_repel(
+    #   data = head(df, 40),
+    #   ggplot2::aes(label = symbol),
+    #   size = 4/ggplot2::.pt,
+    #   max.overlaps = 20,
+    #   segment.size = 0.1,
+    # ) +
+    ggrepel::geom_text_repel(
+      data = left,
+      ggplot2::aes(label = symbol),
+      size = 4/ggplot2::.pt,
+      max.overlaps = 20,
+      segment.size = 0.1,
+      nudge_x = -4,
+      direction = "y"
+    ) +
+    ggrepel::geom_text_repel(
+      data = right,
+      ggplot2::aes(label = symbol),
+      size = 4/ggplot2::.pt,
+      max.overlaps = 20,
+      segment.size = 0.1,
+      nudge_x = 4.5,
[…]

### 2021-04-14 _targets.R

@@ -438,8 +438,20 @@ list(
     run_gsea(rnaseq_different_differences)
   ),
   tar_target(
+    rnaseq_goi,
+    plot_rnaseq_goi(dds, c("EPAS1", "P4HA2", "RBM3"))
+  ),
+  tar_target(
     rnaseq_gsea_plot,
     plot_gsea(rnaseq_gsea, "HALLMARK")
+  ),
+  tar_target(
+    rnaseq_tfea,
+    run_tfea(rnaseq_different_differences)
+  ),
+  tar_target(
+    rnaseq_tfea_plot,
+    plot_tfea(rnaseq_tfea)
   ),
 
   # M1 ----------------------------------------------------------------------
@@ -848,7 +860,7 @@ list(
   # ),
   tar_target(
     m5,
-    arrange_m5(m5a, m5b, m5c, rnaseq_pca, rnaseq_volcano, rnaseq_gsea_plot)
+    arrange_m5(m5a, m5b, m5c, rnaseq_pca, rnaseq_volcano, rnaseq_goi, rnaseq_gsea_plot, rnaseq_tfea_plot)
   ),
   tar_target(
     m5_figure,

### 2021-04-15 inst/analysis/rnaseq.Rmd
format rnaseq analysis report

--- 2021-04-13 inst/analysis/rnaseq.Rmd
@@
 withr::with_dir(here::here(), {
-  rnaseq_pca <- tar_read(rnaseq_pca)
-  rnaseq_different_differences <- tar_read(rnaseq_different_differences)
+  pca <- tar_read(rnaseq_pca)
+  deg <- tar_read(rnaseq_different_differences)
+  gsea_p <- tar_read(rnaseq_gsea_plot)
+  tfea_p <- tar_read(rnaseq_tfea_plot)
 })
@@
 ```{r}
-rnaseq_pca
+pca
 ```
@@
 ```{r}
-head(rnaseq_different_differences, 20)
+deg %>% 
+  dplyr::select(row, symbol, description, lg2fc = log2FoldChange, padj) %>% 
+  head(20) %>% 
+  my_kable()
 ```
+
+## Gene Set Enrichment Analysis
+
+```{r}
+gsea_p
+```
+
+## Transcription Factor Analysis
+
+```{r}
+tfea_p
+```

### 2021-04-23 inst/analysis/metabolomics.Rmd
add metabolomics data and process targeted analysis

--- 2021-04-13 inst/analysis/rnaseq.Rmd
@@
 ---
-title: "RNA-seq Analysis"
+title: "Metabolomics"
 subtitle: "`r format(Sys.time(), '%B %d, %Y')`"
@@
 withr::with_dir(here::here(), {
-  rnaseq_pca <- tar_read(rnaseq_pca)
-  rnaseq_different_differences <- tar_read(rnaseq_different_differences)
+  raw <- tar_read(metab_targeted_raw)
+  clean <- tar_read(metab_targeted_clean)
 })
@@
 Interestingly, when molidustat-treated cells are cultured in hypoxia, hypoxia blocks molidustat-mediated increases in glycolysis.
-In an effort to identify the mechanism mediating this effect, we performed RNA-seq on lung fibroblasts treated for three days with 0.5% oxygen or molidustat (10 μM) with 21% and DMSO (0.1%) controls.
-RNA was collected and purified and sent for RNA-seq analysis.
+In an effort to identify the mechanism mediating this effect, we performed metabolomics on lung fibroblasts treated for three days with 0.5% oxygen or molidustat (10 μM) with 21% and DMSO (0.1%) controls.
 
-# Analysis
-
-We will use the `DESeq2` package to identify differentially expresssed genes. We will use the model `~ experiment + oxygen * treatment`.
+# Data Processing
 
@@
 ```{r}
-rnaseq_pca
+pcaMethods::pca(t(SummarizedExperiment::assay(raw)), scale = "uv", center = TRUE) %>% 
+  pcaMethods::scores() %>% 
+  merge(SummarizedExperiment::colData(raw), by = 0) %>% 
+  filter(type %nin% c("qc", "blank")) %>% 
+  ggplot() +
+  aes(
+    x = PC1, 
+    y = PC2, 
+    shape = interaction(type, treatment, oxygen, sep = "|"),
+    color = replicate
+  ) + 
+  geom_point()
 ```
+## Cleaning
 
-## Differential Expression
-
-In this analysis, will identify those genes where the effects of hypoxia and molidustat are different, which will hopefully point to mechanisms by which hypoxia may regulate cell metabolism differently than PHD2 inhibition.
+1. drift correction
+1. remove low quality features (QC RSD > 0.2, dispersion ratio > 0.4)
+1. impute missing values with random forest
+1. probabilistic quotient normalization
 
 ```{r}
-head(rnaseq_different_differences, 20)
+pcaMethods::pca(t(SummarizedExperiment::assay(clean)), scale = "uv", center = TRUE) %>% 
+  pcaMethods::scores() %>% 
+  merge(SummarizedExperiment::colData(clean), by = 0) %>% 
+  ggplot() +
+  aes(
+    x = PC1, 
+    y = PC2, 
+    color = interaction(oxygen, treatment, sep = "|")
+  ) + 
+  geom_point() +
+  labs(
+    color = "Group"
+  )
 ```

### 2021-04-23 _targets.R

@@ -475,6 +475,31 @@ list(
     output_dir = system.file("analysis/pdfs", package = "Copeland.2021.hypoxia.flux")
   ),
 
+  # metabolomics ------------------------------------------------------------
+
+  tar_target(
+    metab_targeted_files,
+    path_to_data("lf_05-bay_metabolomics-targeted.xlsx"),
+    format = "file"
+  ),
+  tar_target(
+    metab_targeted_raw,
+    format_metab_targeted(metab_targeted_files)
+  ),
+  tar_target(
+    metab_targeted_clean,
+    remove_missing_metab(metab_targeted_raw) %>%
+      correct_drift() %>%
+      quality_control() %>%
+      impute_missing() %>%
+      normalize()
+  ),
+  tar_render(
+    metabolomics_report,
+    path = path_to_reports("metabolomics.Rmd"),
+    output_dir = system.file("analysis/pdfs", package = "Copeland.2021.hypoxia.flux")
+  ),
+
   # M1 ----------------------------------------------------------------------
 
   tar_target(

### 2021-04-23 inst/src/metabolomics.R

@@ -1,0 +1,231 @@
+# metabolomics.R
+
+# format_metab_targeted ---------------------------------------------------
+
+format_metab_targeted <- function(metab_targeted_files) {
+  readxl::read_excel(
+    metab_targeted_files,
+    sheet = 2
+  ) %>%
+    dplyr::select(
+      id = `Raw File Name`,
+      sample = `Sample ID`,
+      metabolite = `Compound Name`,
+      mz = `Detected Mass`,
+      rt = RT,
+      area = `Peak Area`
+    ) %>%
+    dplyr::left_join(wmo::hmdb_mappings, by = "metabolite") %>%
+    dplyr::filter(!is.na(hmdb)) %>%
+    dplyr::mutate(
+      id = stringr::str_c("S", sprintf("%02d", id)),
+      type = dplyr::case_when(
+        sample == "water" ~ "blank",
+        stringr::str_detect(sample, "mix") ~ "qc",
+        TRUE ~ "sample"
+      ),
+      oxygen = stringr::str_extract(sample, "hyp|norm"),
+      oxygen = factor(oxygen, levels = c("norm", "hyp"), labels = c("N", "H")),
+      treatment = stringr::str_extract(sample, "bay|dmso"),
+      treatment = factor(treatment, levels = c("dmso", "bay"), labels = c("DMSO", "BAY")),
+      replicate = stringr::str_extract(sample, "(?<=\\.)\\w{1}$"),
+      mz = replace(mz, mz == "N/F", NA_real_),
+      mz = as.numeric(mz),
+      dplyr::across(c(rt, area), ~replace(., . == 0, NA_real_))
+    ) %>%
+    dplyr::select(-sample) %>%
+    new_tbl_se(
+      a_data = "area",
+      f_names = "hmdb",
+      f_data = c("metabolite", "mz", "rt"),
+      s_names = "id",
+      s_data = c("type", "oxygen", "treatment", "replicate")
+    ) %>%
+    tbl_to_se()
+
+}
+
+# new_tbl_se --------------------------------------------------------------
+
+new_tbl_se <- function(
+  tbl,
+  a_data,
+  f_names,
+  f_data = NULL,
+  s_names,
+  s_data = NULL
+) {
+  structure(
+    tbl,
+    class = c("tbl_se", class(tbl)),
+    a_data = a_data,
+    f_names = f_names,
+    s_names = s_names,
+    f_data = f_data,
[…]

### 2021-04-28 inst/analysis/metabolomics.Rmd
differential metabolite analysis

@@
   clean <- tar_read(metab_targeted_clean)
+  metab_pca <- tar_read(metab_targeted_pca)
+  metab_limma <- tar_read(metab_targeted_limma)
+  # metab_tt <- tar_read(metab_different_differences)
+  volcano <- tar_read(metab_volcano)
 })
@@
   merge(SummarizedExperiment::colData(raw), by = 0) %>% 
-  filter(type %nin% c("qc", "blank")) %>% 
+  # filter(type %nin% c("qc", "blank")) %>% 
   ggplot() +
@@
     y = PC2, 
-    shape = interaction(type, treatment, oxygen, sep = "|"),
-    color = replicate
+    shape = type,
+    color = interaction(oxygen, treatment, sep = "|")
   ) + 
-  geom_point()
+  geom_point() +
+  labs(
+    color = "Group", 
+    shape = "Type"
+  )
 ```
@@
 1. probabilistic quotient normalization
+1. batch correction
 
 ```{r}
-pcaMethods::pca(t(SummarizedExperiment::assay(clean)), scale = "uv", center = TRUE) %>% 
-  pcaMethods::scores() %>% 
-  merge(SummarizedExperiment::colData(clean), by = 0) %>% 
-  ggplot() +
-  aes(
-    x = PC1, 
-    y = PC2, 
-    color = interaction(oxygen, treatment, sep = "|")
-  ) + 
-  geom_point() +
-  labs(
-    color = "Group"
-  )
+metab_pca
 ```
+
+## Differentially regulated metabolites
+
+```{r}
+volcano
+```
+
+```{r}
+plot_mois(clean, "malate")
+```

### 2021-04-28 _targets.R

@@ -492,7 +492,27 @@ list(
       correct_drift() %>%
       quality_control() %>%
       impute_missing() %>%
-      normalize()
+      pqn()
+  ),
+  tar_target(
+    metab_targeted_pca,
+    plot_metab_pca(metab_targeted_clean)
+  ),
+  tar_target(
+    metab_targeted_limma,
+    fit_metab_limma(metab_targeted_clean)
+  ),
+  tar_target(
+    metab_different_differences,
+    metab_top_table(metab_targeted_clean, metab_targeted_limma, "deltas")
+  ),
+  tar_target(
+    metab_volcano,
+    plot_metab_volcano(metab_different_differences)
+  ),
+  tar_target(
+    metab_moi,
+    plot_mois(metab_targeted_clean, c("glyceraldehyde 3-phosphate", "2-hydroxyglutarate", "aconitate", "taurine", "hydroxyproline", "malate"))
   ),
   tar_render(
     metabolomics_report,

### 2021-04-28 inst/src/metabolomics.R

@@ -17,6 +17,7 @@ format_metab_targeted <- function(metab_targeted_files) {
     ) %>%
     dplyr::left_join(wmo::hmdb_mappings, by = "metabolite") %>%
     dplyr::filter(!is.na(hmdb)) %>%
+    dplyr::filter(!stringr::str_detect(hmdb, "ISTD")) %>%
     dplyr::mutate(
       id = stringr::str_c("S", sprintf("%02d", id)),
       type = dplyr::case_when(
@@ -28,6 +29,8 @@ format_metab_targeted <- function(metab_targeted_files) {
       oxygen = factor(oxygen, levels = c("norm", "hyp"), labels = c("N", "H")),
       treatment = stringr::str_extract(sample, "bay|dmso"),
       treatment = factor(treatment, levels = c("dmso", "bay"), labels = c("DMSO", "BAY")),
+      group = stringr::str_c(oxygen, treatment, sep = "."),
+      group = factor(group, levels = c("N.DMSO", "N.BAY", "H.DMSO", "H.BAY")),
       replicate = stringr::str_extract(sample, "(?<=\\.)\\w{1}$"),
       mz = replace(mz, mz == "N/F", NA_real_),
       mz = as.numeric(mz),
@@ -39,7 +42,7 @@ format_metab_targeted <- function(metab_targeted_files) {
       f_names = "hmdb",
       f_data = c("metabolite", "mz", "rt"),
       s_names = "id",
-      s_data = c("type", "oxygen", "treatment", "replicate")
+      s_data = c("type", "oxygen", "treatment", "replicate", "group")
     ) %>%
     tbl_to_se()
 
@@ -222,7 +225,7 @@ impute_missing <- function(qc) {
 
 # normalize ---------------------------------------------------------------
 
-normalize <- function(imputed) {
+pqn <- function(imputed) {
   mat <- SummarizedExperiment::assay(imputed)
   quotients <- mat / SummarizedExperiment::rowData(imputed)$reference
   quotient_medians <- apply(quotients, 2, median)
@@ -230,4 +233,282 @@ normalize <- function(imputed) {
   imputed
 }
 
+# plot_metab_pca ----------------------------------------------------------
 
+plot_metab_pca <- function(clean) {
+
+  input <-
+    SummarizedExperiment::assay(clean) %>%
+    log() %>%
+    t() %>%
[…]

### 2021-04-30 inst/manuscript/manuscript.Rmd
work on manuscript methods

@@
 output: 
-  bookdown::pdf_document2:
-    template: template.tex
-    keep_tex: false
+  # bookdown::pdf_document2:
+  #   # template: template.tex
+  #   keep_tex: false
+  #   df_print: kable
+  #   toc: false
+  #   number_sections: false
+  #   latex_engine: xelatex
+  #   pandoc_args:
+  #     - '--lua-filter=scholarly-metadata.lua'
+  #     - '--lua-filter=author-info-blocks.lua'
+  #     - '--lua-filter=pagebreak.lua'
+  #     - '--lua-filter=multiple-bibliographies.lua'
+
+  bookdown::word_document2:
+    reference_docx: template.docx
     df_print: kable
-    toc: false
     number_sections: false
-    latex_engine: xelatex
     pandoc_args:
@@
       - '--lua-filter=multiple-bibliographies.lua'
+
 compact-title: true
@@
 devtools::load_all()
+library(magrittr)
 
@@
 }
+
+withr::with_dir(here::here(), {
+  biomass <- targets::tar_read(biomass)
+  resources <- targets::tar_read(resources_table)
+})
+
+biomass <- tibble::deframe(biomass)
 ```
@@
 
+\newpage
+
 # STAR Methods
+
+## Resource Availability
+
+### Lead Contact
+
+Further information and requests for resources and reagents should be directed to and will be fulfilled by the lead contact, William Oldham (woldham@bwh.harvard.edu).
+
+### Materials Availability
+
+This study did not generate new unique reagents.
+
+### Data and code availability
+
+The original data and analysis code are available as a reproducible research compendium formatted as an R package available at OldhamLab/Copeland.2021.hypoxia.flux (https://github.com/oldhamlab/Copeland.2021.hypoxia.flux).
+RNA-seq data has been deposited in the NIH Short Read Archive (PRJNA721596).
+
+## Experimental Model Details
+
+### Lung fibrolbasts
+
+Primary normal human lung fibroblasts (LFs) were purchased from Lonza (CC-2512) and cultured in FGM-2 (Lonza CC-3132) in a standard tissue culture incubator in 5% CO~2~ at 37 °C.
+Cells from two donors were used in these studies: #33652 (56, male) and #29132 (19, female).
+Cell authentication was performed by the vendor.
+
+### Pulmonary artery smooth muscle cells
+
+Primary human pulmonary artery smooth muscle cells were purchased from Lonza (CC-2581) and cultured in SmGM-2 (Lonza CC-3182) in a standard tissue culture incubator in 5% CO~2~ at 37 °C.
+Cells from multiple donors were used in these studies: #30020 (64, male), #27662 (35, male), #26698 (51, male), and #19828 (51, male).
+Cell authentication was performed by the vendor.
+
+## Method Details
+
+### Metabolic flux protocol
+
+For extracellular flux measurements, cells were seeded in either standard growth medium or MCDB131 medium without glucose, glutamine, or phenol red (genDEPOT) supplemented with 2% dialyzed fetal bovine serum (Mediatech) and naturally labeled glucose and glutamine ("light" labeling medium).
+For LFs, glucose was supplemented at 8 mM and glutamine was supplemented at 1 mM.
+For PASMCs, glucose was supplemented at 5.55 mM and glutamine was supplemented at 10 mM.
+These concentrations match the concentrations of these substrates determined in standard growth medium.
+Preliminary experiments were performed to identify the optimal cell seeding density, exponential growth phase, and labeling duration consistent with metabolic and isotopic steady state.
+On Day -1, 25,000 cells were seeded in a 35 mm dish in "light" labeling medium.
+Hypoxia-treated cells were transferred to a tissue culture glovebox set to 0.5% oxygen and 5% CO~2~ (Coy Lab Products).
+Medium was supplemented with DMSO 0.1% or BAY (10 μM) for DMSO and BAY treatment conditions.
+On Day 0, cells were washed with PBS and the medium was changed to either "light" medium for flux measurements or "heavy" medium for tracer experiments.
+For LFs, samples were collected on Day 0 and every 24 h for 72 h.
+For PASMCs, samples were collected on Day 0 and every 12 h for 48 h.
+Medium and cell lysates were collected at each time point for intra- and extracellular metabolite measurements and total DNA quantification.
+Dishes without cells were weighed daily to correct for evaporative medium losses and to empirically determine degradation and accumulation rates of metabolites.
+Medium samples and cell lysates for DNA measurement were stored at -80 °C until analysis.
+Each individual experiment included triplicate wells for each treatment and time point, and each experiment was repeated 4-8 times.
+
+### Cell count
+
+Direct cell counts of trypsizined cell suspensions in PBS were obtained following staining with propidium iodide and acridine orange using a LUNA-FL fluorescence cell counter (Logos Biosystems).
+Indirect cell counts for flux measurements were interpolated from total DNA quantified using the Quant-iT PicoGreen dsDNA Assay Kit (Thermo).
+Cells were washed once with two volumes of PBS, lysed with Tris-EDTA buffer containing 2% Triton X-100, and collected by scraping.
+Total DNA in 10 μL of lysate was determined by adding 100 μL of 1X PicoGreen dye in Tris-EDTA buffer and interpolating the fluorescence intensity with a standard curve generated using the λ DNA standard.
+Cell counts were interpolated from a standard curve of DNA obtained from known cell numbers seeded in basal medium (Supplementary Figure 1A).
+No difference in total cellular DNA was identified between normoxia and hypoxia cultures (Supplementary Figure 1B).
+
+### Immunoblots
+
+Cells were washed with one volume of PBS and collected by scraping in PBS.
+Cell suspensions were centrifuged at 5,000 ×*g* for 5 min at 4 °C.
+Pellets were lysed in buffer containing Tris 10 mM, pH 7.4, NaCl 150 mM, EDTA 1 mM, EGTA 1 mM, Triton X-100 1% v/v, NP-40 0.5% v/v, and Halt Protease Inhibitor Cocktail (Thermo).
+Protein concentrations were determined by BCA Protein Assay (Thermo).
+Lysates were normalized for protein concentration and subjected to SDS-PAGE separation on stain-free tris-glycine gels (Bio-Rad), cross-linked and imaged with the Chemidoc system (Bio-Rad), transferred to PVDF membranes with the Trans-Blot Turbo transfer system (Bio-Rad), imaged, blocked in 5% blocking buffer (Bio-Rad), blotted in primary and secondary antibodies, and developed using WesternBright ECL (Advansta).
+Band signal intensity was normalized to total protein per lane as determined from the stain-free gel or membrane images.
+
+### RT-qPCR
+
+Total RNA was isolated from cells with the RNeasy Mini Kit (Qiagen).
+cDNA was synthesized from 0.25-1.00 ng RNA with the High Capacity cDNA Reverse Transcription Kit (Applied Biosystems).
+RT-qPCR analysis was performed with an Applied Biosystems 7500 Fast Real Time PCR System with TaqMan Universal PCR Master Mix and pre-designed TaqMan gene expression assays (Life Technologies).
+Relative expression levels were calculated using the comparative cycle threshold method referenced to *ACTB*.
+
+###  Glucose assay
+
+Medium samples were diluted 10-fold in PBS.
+Glucose concentration was determined using the Glucose Colorimetric Assay Kit (Cayman) according to the manufacturer's protocol.
+Standards were prepared in PBS.
+
+### Lactate assay
+
+Medium samples were diluted 10-fold in PBS.
+Glucose concentration was determined using the ʟ-Lactate Assay Kit (Cayman).
+Medium samples did not require deproteinization, otherwise the samples were analyzed according to the manufacturer's protocol.
+Standards were prepared in PBS.
+
+### Pyruvate assay
+
+Pyruvate was measured using either an enzymatic assay (most samples) or an HPLC-based assay (medium from 0.2% oxygen experiments).
+For the enzymatic assay, medium samples were diluted 20-fold in PBS.
+Pyruvate concentration was determined using the Pyruvate Assay Kit (Cayman).
+Medium samples did not require deproteinization, otherwise the samples were analyzed according to the manufacturer's protocol.
+Standards were prepared in PBS.
+For the HPLC assay, 2-oxovaleric acid was added to medium samples as an internal standard.
+Samples were subsequently deproteinized with 2 volumes of ice-cold acetone.
+Supernatants were evaporated to < 50% of the starting volume at 43 °C in a SpeedVac concentrator (Thermo Savant) and reconstituted to the starting volume with HPLC-grade water prior to derivatization.
+Samples were derivatized 1:1 by volume with *o*-phenylenediamine (25 mM in 2 M HCl) for 30 min at 80 °C.
+Derivatized pyruvate was separated with a Poroshell HPH C-18 column (2.1 × 100 mm, 2.7 μm) on an Infinity II high-performance liquid chromatography system with fluorescence detection of OPD-derivatized α-keto acids as described previously [@RN730].
+
+### Amino acid assay
+
+Medium amino acid concentrations were determined following the addition of norvaline and sarcosine internal standards and deproteinization with 2 volumes of ice-cold acetone.
+Supernatants were evaporated to < 50% of the starting volume at 43 °C in a SpeedVac concentrator (Thermo Savant) and reconstituted to the starting volume with HPLC-grade water prior to analysis.
+Amino acids in deproteinized medium were derivatized with *o*-phthalaldehyde (OPA) and 9-fluorenylmethylchloroformate (FMOC) immediately prior to separation with a Poroshell HPH-C18 column (4.6 × 100 mm, 2.7 μm) on an Infinity II high-performance liquid chromatography system with ultraviolet and fluorescence detection of OPA- and FMOC-derivatized amino acids, respectively, according to the manufacturer's protocol (Agilent) [@RN1294].
+
+### Flux calculations
+
+The growth rate ($\mu$) and flux (*v*) for each measured metabolite were defined as follows [@RN1523]:
+
+\begin{align}
+\frac{dX}{dt} &= \mu X (\#eq:1) \\[6 pt] 
+\frac{dM}{dt} &= -kM+vX (\#eq:2) 
+\end{align}
+
+where *X* is the cell density, *k* is the first-order degradation or accumulation rate, and *M* is the mass of the metabolite. These equations are solved as follows:
+
+\begin{align}
+X &= X_0e^{\mu t} (\#eq:3)\\[6 pt]
+Me^{k t} &= \frac{v X_0}{\mu + k} (e^{(\mu + k)t}-1) + M_0 (\#eq:4) 
+\end{align}
+
+Growth rate ($\mu$) and cell count at time 0 ($X_0$) were determined by robust linear modeling of the logarithm of cell count as a function of time ($t$).
+Metabolite mass was calculated from the measured metabolite concentrations and predicted well volume accounting for evaporative losses (Supplementary Figure 1C).
+First-order degradation and accumulation rates were obtained from robust linear modeling of metabolite mass *v*.
+time in unconditioned culture medium.
+Rates that significantly differed from 0 using Student's *t*-test were incorporated into the flux calculations.
+Final fluxes were obtained by robust linear modeling of $Me^{k t}$ *versus* $(e^{(\mu + k)t}-1)$ to determine the slope from which $v$ was calculated using equation \@ref(eq:4).
+
+### Metabolomics
+
+#### Metabolite extraction
+
+Intracellular metabolites were obtained after washing cells with 2 volumes of ice-cold PBS and floating on liquid nitrogen.
+Plates were stored at -80 °C until extraction.
+Metabolites were extracted with 1 mL 80% MeOH pre-cooled to -80 °C containing 10 nmol [D~8~]-\textsc{dl}-valine as an internal standard (Cambridge Isotope Labs).
+Insoluble material was removed by centrifugation at 21,000 ×*g* for 15 min at 4 °C.
+The supernatant was evaporated to dryness at 42 °C using a SpeedVac concentrator (Thermo Savant).
+Samples were resuspended in 35 μL LC-MS-grade water prior to analysis.
+
+#### Acquisition parameters
+
+LC-MS analysis was performed on a Vanquish ultra-high-performance liquid chromatography system coupled to a Q Exactive orbitrap mass spectrometer by a HESI-II electrospray ionization probe (Thermo).
+External mass calibration was performed weekly.
+Metabolite samples (2.5 μL) were separated using a ZIC-pHILIC stationary phase (2.1 × 150 mm, 5 μm) (Merck).
+The autosampler temperature was 4 °C and the column compartment was maintained at 25 °C.
+Mobile phase A was 20 mM ammonium carbonate and 0.1% ammonium hydroxide.
+Mobile phase B was acetonitrile.
+The flow rate was 0.1 mL/min.
+Solvent was introduced to the mass spectrometer *via* electrospray ionization with the following source parameters: sheath gas 40, auxiliary gas 15, sweep gas 1, spray voltage +3.0 kV for positive mode and -3.1 kV for negative mode, capillary temperature 275 °C, S-lens RF level 40, and probe temperature 350 °C.
+Data were acquired and peaks integrated using TraceFinder 4.1 (Thermo).
+
+#### Stable isotope quantification
+
+All metabolites except fructose 2,6-bisphosphate (FBP) and 3-phosphoglycerate (3PG) were measured using the following mobile phase gradient: 0 min, 80% B; 5 min, 80% B; 30 min, 20% B; 31 min, 80% B; 42 min, 80% B.
+The mass spectrometer was operated in selected ion monitoring mode with an m/*z* window width of 9.0 centered 1.003355-times half the number of carbon atoms in the target metabolite.
+The resolution was set at 70,000 and AGC target was 1 × 10^5^ ions.
+Peak areas were corrected for quadrupole bias as previously described [@RN1063].
+Mass isotope distributions for FBP and 3PG were calculated from full scan chromatograms as described below.
+Raw mass isotopomer distributions were corrected for natural isotope abundance using a custom R package (mzrtools, https://github.com/oldhamlab/mzrtools) employing the method of Fernandez, *et al.* [@RN557].
+
+#### Metabolomic profiling
+
+For metabolomic profiling and quantification of isotopic enrichment for FBP and 3PG, the following mobile phase gradient was used: 0 min, 80% B; 20 min, 20% B; 20.5 min, 80% B; 28 min, 80% B; 42 min, 80% B.
+The mass spectrometer was operated in polarity switching full scan mode from 70-1000 m/*z*.
+Resolution was set to 70,000 and the AGC target was 1x10^6^ ions.
+Peak identifications were based on an in-house library of authentic metabolite standards previously analyzed utilizing this method.
+For metabolomics studies, pooled quality control (QC) samples were injected at the beginning, end, and between every four samples of the run.
+Raw peak areas for each metabolite were corrected for instrument drift using a cubic spline model of QC peak areas.
+Low quality features were removed on the basis of a relative standard deviation greater than 0.2 in the QC samples and a dispersion ratio greater than 0.4 [@RN3077].
+Missing values were imputed using random forrest.
+Samples peak areas were normalized using probabilistic quotient normalization [@RN3094].
+Differentially regulated metabolites were identified using limma [@limma].
+
+### Biomass determination
+
+The dry weight of LFs was determined to be `r round(biomass[["lf"]])` pg/cell.
+The dry weight of PASMCs was determined to be `r round(biomass[["pasmc"]])` pg/cell.
+These values were estimated by washing 3 × 10^6^ cells twice in PBS and thrice in ice-cold acetone prior to drying overnight in a SpeedVac.
+The composition of the dry cell mass was estimated from the literature [@RN1741; @RN2005], and stoichiometric coefficients were determined as described [@RN1522; @RN2517].
+
+### Metabolic flux analysis
+
+Metabolic flux analysis was performed using the elementary metabolite unit-based software package INCA [@RN2501].
+Inputs to the model include the chemical reactions and atom transitions of central carbon metabolism, extracellular fluxes, the identity and composition of ^13^C-labeled tracers, and the MIDs of labeled intracellular metabolites.
+The metabolic network was adapted from previously published networks [@RN1522; @RN2266] and comprises `r nrow(dplyr::filter(model_reactions, !(pathway %in% c("mixing", "dilution"))))` reactions representing glycolysis, the pentose phosphate pathway, the tricarboxylic acid cycle, anaplerotic pathways, serine metabolism, and biomass synthesis.
+The network includes seven extracellular substrates (aspartate, cystine, glucose, glutamine, glycine, pyruvate, serine) and five metabolic products (alanine, biomass, glutamate, lactate, lipid).
+Models were fit using three ^13^C-labeled tracers, [1,2-^13^C~2~] glucose, [U-^13^C~6~] glucose, and [U-^13^C~5~] glutamine.
+The MIDs of twelve metabolites (2-oxoglutarate, 3-phosphoglycerate, alanine, aspartate, citrate, fructose bisphosphate, glutamate, glutamine, lactate, malate, pyruvate, serine) were used to constrain intracellular fluxes.
+The following assumptions were made:
+
+1. Metabolism was at steady state.
+1. Labeled CO~2~ produced during decarboxylation reactions left the system and did not re-incorporate during carboxylation reactions.
+1. Protein turnover occurred at a negligible rate compared to glucose and glutamine consumption.
+1. Acetyl-CoA, aspartate, fumarate, malate, oxaloacetate, pyruvate existed in cytosolic and mitochondrial pools. Aspartate and malate were allowed to exchange freely between the compartments.
+1. The per cell biomass requirements of proliferating lung fibroblasts were similar to published estimated in other cells [@RN1741; @RN2005]. 
+1.
+Dilution of alanine, aspartate, glutamate, glutamine, lactate, and pyruvate was allowed to occur through reversible exchange with unlabeled substrates in the medium as these metabolites were present in unconditioned medium or serum.
+1. Succinate and fumarate are symmetric molecules that have interchangeable orientations when metabolized by TCA cycle enzymes.
+
+Flux estimation was repeated a minimum of 50 times from random initial values.
+Results were subjected to a χ^2^ statistical test to assess goodness-of-fit.
+Accurate 95% confidence intervals were computed for estimated parameters by evaluating the sensitivity of the sum-of-square residuals to parameter variations [@RN60; @RN1522].
+
+### NAD(H) assay
+
+Cellular NAD^+^ and NADH were measured using an enzymatic fluorimetric cycling assay based on the reduction of NAD^+^ to NADH by alcohol dehydrogenase (ADH) and subsequent electron transfer to generate the fluorescent molecule resorufin [@RN1603].
+Briefly, cells were washed twice with one volume PBS.
+Pyridine nucleotides were extracted on ice with buffer containing 50% by volume PBS and 50% lysis buffer (100 mM sodium carbonate, 20 mM sodium bicarbonate, 10 mM nicotinamide, 0.05% by volume Triton-X-100, 1% by mass dodecyltrimethylammonium bromide) and collected by scraping.
+Extracts were divided equally and 0.5 volume of 0.4 N HCl was added to one sample.
+Both extracts were heated at 65 °C for 15 min to degrade selectively either the oxidized (buffer) or reduced (HCl) nucleotides.
+The reaction was cooled on ice and quenched by adding 0.5 M Tris-OH to the acid-treated samples or 0.2 N HCl plus 0.25 M Tris-OH to the buffer samples.
+Samples were then diluted in reaction buffer (50 mM EDTA and 10 mM Tris, pH 7.06).
+Cell debris was pelleted by centrifugation, and 50 μL was incubated for 2 h with 100 μL reaction buffer containing 0.6 M EtOH, 0.5 mM phenazine methosulfate, 0.05 mM resazurin, and 0.1 mg/mL ADH.
+Fluorescence intensities were measured with a Spectramax Gemini XPS (Moelcular Devices) with excitation 540 nm, emission 588 nm, and 550 nm excitation cut-off filter.
+Sample intensities were compared to a standard curve generated from known concentrations of NADH.
+The ratio of fluorescence in buffer-extracted to acid-extracted samples corresponds to the NADH/NAD^+^ ratio.
+Absolute NADH and NAD^+^ were normalized to cell count from cells estimated from total DNA quantification as described above.
+
+### RNA-seq
+
+RNA was collected from LFs treated for three days ± hypoxia ± BAY as described above.
+Four biological replicates were analyzed.
+Library construction and sequencing was performed by BGI Genomics using 100 bp paired end analysis and a read depth of 50M reads per sample.
+Sequences were deposited in the NIH SRA (PRJNA721596).
+Sequences were mapped to the human GRCh38 primary assembly and counts summarized using Rsubread [@Rsubread].
+This data is available from the Oldham Lab GitHub repository (https://github.com/oldhamlab/rnaseq.lf.hypoxia.molidustat).
+Differentially expressed transcripts were identified using DESeq2 [@DESeq2].
+Gene set enrichment and transcription factor enrichment was performed using the fgsea and TFEA.ChIP R packages, respectively [@fgsea; @TFEA.ChIP].
+
+## Quantification and Statistical Analysis
+
+The raw data and annotated analysis code necessary to reproduce this manuscript are contained in an R package research compendium available from the Oldham Lab GitHub repository (https://github.com/oldhamlab/Copeland.2021.hypoxia.flux).
+Data analysis, statistical comparisons, and visualization were performed in R [@R-base] using the packages listed below.
+Experiments included technical and biological replicates as noted above.
+The number of biological replicates (n) is indicated in the figure legends.
+Summary data show the mean ± SEM.
+Outliers were identified using twice the median absolute deviation.
+Two group comparisons (*e.g.*, 21% *v.* 0.5% oxygen) were performed using Student's *t*-test on log-transformed values with pairing among biological replicates.
+Multifactor comparisons (*e.g.*, BAY and oxygen) were performed using linear mixed-effects models with Tukey's *post hoc* testing.
+Metabolomics and RNA-seq data were analyzed as described above.
+Probability values less than 0.05 were considered significant.
+
+\newpage
+
+## Key Resources Table
+
+<br>
+
+```{r, results = "asis"}
+flextable::flextable_to_rmd(resources)
+```
+
+\newpage
+
+# Supplemental Items Titles
+
+\newpage
+
+# References
+
+::: {#refs_main}
+:::
+
+# Software
+
+::: {#refs_software}
+:::

### 2021-04-30 _targets.R

@@ -963,6 +963,13 @@ list(
   tar_target(
     m6_figure,
     write_figures(m6, "m6.pdf")
+  ),
+
+  # resources table ---------------------------------------------------------
+
+  tar_target(
+    resources_table,
+    create_resources()
   )
 
 )

### 2021-04-30 inst/src/figures.R

@@ -197,3 +197,69 @@ arrange_m6 <- function(a, b, c, d, e, f) {
       heights = unit(c(1.25), "in")
     )
 }
+
+# create_resources --------------------------------------------------------
+
+create_resources <- function() {
+  tibble::tribble(
+    ~category, ~`REAGENT or RESOURCE`, ~SOURCE, ~IDENTIFIER,
+    "Antibodies", "HIF-1α", "BD Biosciences", "610958",
+    "Antibodies", "c-MYC", "Cell Signaling Technologies", "D84C12",
+    "Antibodies", "LDHA", "Cell Signaling Technologies", "2012",
+    "Antibodies", "HRP-α-Rabbit IgG", "Cell Signaling Technologies", "7074",
+    "Antibodies", "HRP-α-Mouse IgG", "Cell Signaling Technologies", "7076",
+    "Chemicals, peptides, and recombinant proteins", "[1,2-^13^C~1~] glucose", "Cambridge Isotope Labs", "CLM-504-PK",
+    "Chemicals, peptides, and recombinant proteins", "[U-^13^C~6~] glucose", "Cambridge Isotope Labs", "CLM-1396-PK",
+    "Chemicals, peptides, and recombinant proteins", "[U-^13^C~5~] glutamine", "Cambridge Isotope Labs", "CLM-1822-H-PK",
+    "Chemicals, peptides, and recombinant proteins", "[U-^13^C~3~] lactate", "Sigma", "485926",
+    "Chemicals, peptides, and recombinant proteins", "Molidustat (BAY-85-3934)", "Cayman", "15297",
+    "Critical commercial assays", "Glucose colorimetric assay kit", "Cayman", "10009582",
+    "Critical commercial assays", "ʟ-Lactate assay kit", "Cayman", "700510",
+    "Critical commercial assays", "Pyruvate assay kit", "Cayman", "700470",
+    "Depositied data", "Raw and analyzed data", "This paper", "https://github.com/oldhamlab/Copeland.2021.hypoxia.flux",
+    "Depositied data", "RNA-seq reads", "This paper", "SRA: PRJNA721596",
+    "Depositied data", "Summarized RNA-seq data", "This paper", "https://github.com/oldhamlab/rnaseq.lf.hypoxia.molidustat",
+    "Experimental models: Cell lines", "Normal human lung fibroblasts", "Lonza", "CC-2512",
[…]

### 2021-05-03 inst/manuscript/manuscript.Rmd
incorporating figures and tables

@@
   message = FALSE, 
-  out.width = "100%", 
+  out.width = "50%", 
   fig.align = "center", 
@@
 
+\newpage
+
 # Summary 
@@
 Since hypoxia is a prominent feature of cancer biology as tumor growth outstrips blood supply, most detailed metabolic studies of cell metabolism in hypoxia have used tumor cell models [@RN2395, @RN1447, @RN1438, @RN634, @RN1200, @RN975].
-For example, stable isotope tracing and metabolic flux analyses identified a critical role for reductive carboxylation of glutamine-derived 2-oxoglutarate for lipid biosynthesis in tumor growth [@RN1447; @RN628; @RN1966; @RN2395], and metabolomic studies identified aspartate as a limiting metabolite for cancer cell proliferation under hypoxia [@RN634].
-By contrast, comparatively little is known about metabolic adaptations of primary cells to hypoxia and the importance of the metabolic pathways described above remain to be elucidated.
+For example, stable isotope tracing and metabolic flux analyses identified a critical role for reductive carboxylation of glutamine-derived α-ketoglutarate for lipid biosynthesis in tumor growth [@RN1447; @RN628; @RN1966; @RN2395], and metabolomic studies identified aspartate as a limiting metabolite for cancer cell proliferation under hypoxia [@RN634].
+By contrast, comparatively little is known about metabolic adaptations of primary cells to hypoxia and the importance of the metabolic pathways described above remain to be elucidated in these cells.
 This information would provide important context for understanding the extent to which cancer cell metabolism responds differently to hypoxic stress.
@@
 In normoxia, HIF-1α activation by the prolyl hydroxylase inhibitor molidustat (BAY-85-3934, "BAY") [@RN580] did increase glycolysis and lactate efflux; however, hypoxia abrogated this response.
-These findings suggest the existence of hypoxia-dependent regulatory mechanisms that override the metabolic consequences of HIF-1α-dependent up-regulation of glycolytic gene expression in human primary cells.
+These findings suggest the existence of important hypoxia-dependent regulatory mechanisms that override the metabolic consequences of HIF-1α-dependent up-regulation of glycolytic gene expression in human primary cells.
 
@@
 From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (Figure \ref{fig:f1}A), thought to be an indicator of metabolic steady state.
-Lung fibroblasts cultured in 0.5% oxygen grew more slowly (Figure \ref{fig:f1}B), but slower growth was not associated with decreased cell viability (Figure \ref{fig:f1}C).
+Lung fibroblasts (LFs) cultured in 0.5% oxygen grew more slowly (Figure \ref{fig:f1}B), but slower growth was not associated with decreased cell viability (Figure \ref{fig:f1}C).
 These cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1, (GLUT1), and lactate dehydrogenase A (LDHA) (Figure \ref{fig:f1}D-H).
@@
 Similar findings were observed when the ambient oxygen level was decreased further to 0.2% (Supplementary Figure 2) and in pulmonary artery smooth muscle cells (PASMCs; Supplementary Figure 3).
-In addition to glucose and lactate, the absolute extracellular fluxes of pyruvate and amino acids were generally decreased in hypoxia, including a marked decrease in glutamine uptake in lung fibroblasts.
+In addition to lactate, the absolute extracellular fluxes of pyruvate and amino acids were generally decreased in hypoxia, including a marked decrease in glutamine uptake in LFs.
 Notably, hypoxia was previously shown to increase glutamine uptake in studies of cancer cell metabolism [@RN1447; @RN2395; @RN628].
-Glutamine uptake did increase in PASMCs, as did the uptake of branch chain amino acids and arginine, suggesting differential responses of these mesenchymal cells to hypoxia.
+Glutamine uptake did increase in PASMCs, as did the uptake of branch chain amino acids and arginine (Supplementary Figure 3), suggesting differential responses of these mesenchymal cells to hypoxia.
 
-Given that hypoxia did not increase glucose and lactate fluxes as expected from the associated increases in glycolytic gene expression, we next assessed the capacity of HIF-1α to augment glycolysis in lung fibroblasts.
+Given that hypoxia did not increase glucose and lactate fluxes as expected from the associated increases in glycolytic gene expression, we next assessed the capacity of HIF-1α to augment glycolysis in LFs.
 Cells were treated with BAY to stabilize HIF-1α under 21% oxygen conditions (Figure \ref{fig:f2}).
@@
 Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1 target gene transcription and protein expression.
-In normoxia, this transcriptional program was associated with increased glucose uptake and lactate efflux (Figure \ref{fig:f2}H).
-Relatively modest effects on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (Figure \ref{fig:f2}J).
+In normoxia, this transcriptional program was associated with substantially increased glucose uptake and lactate efflux (Figure \ref{fig:f2}H).
+Comparatively modest effects of BAY on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (Figure \ref{fig:f2}J).
 
-## Stable isotope tracing
+## Stable isotope incorporation suggests preservation of the intracellular metabolic program in hypoxia
 
-To characterize further the extent of hypoxia-induced metabolic reprogramming, we next treated lung fibroblasts with stable carbon isotopes of glucose and glutamine to trace label incorporation into key carbon utilization pathways (Figure \ref{fig:f3} and Supplementary Figure 4).
-Overall, hypoxia-treated cells had decreased label incorporation into downstream metabolites (*i.e.*, the unlabeled, or M0, fraction was greater).
+To validate our findings from extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate of label incorporation and trace its incorporation into key carbon utilization pathways (Figure \ref{fig:f3} and Supplementary Figures 4-5).
+Consistent with our extracellular flux measurements, we found that the rate of label incorporation from [U-^13^C~6~] glucose into pyruvate was similar across 21%, 0.5%, and DMSO-treated conditions, but markedly increased with BAY treatment (\ref{fig:f3}A-B).
+
+Overall, hypoxia-treated cells generally had decreased label incorporation into downstream metabolites (*i.e.*, the unlabeled, or M0, fraction was greater).
 This finding is consistent with the extracellular flux measurements suggesting slower substrate utilization by hypoxic cells.
-BAY treatment recapitulated the labeling pattern observed with hypoxia, suggesting similar intracellular metabolite flux between these two conditions.
+BAY treatment recapitulated the labeling patterns observed with hypoxia, suggesting similar effects on intracellular metabolite distribution between these two conditions.
 Beyond this observation, the labeling patterns in hypoxia- and BAY-treated cells were similar to their respective controls, arguing against marked metabolic reprogramming in response to prolyl hydroxylase inhibition by either hypoxia or BAY.
-Compared to previous studies of metabolic flux in cancer cells [@RN2395; @RN1447], no hypoxia-mediated increase in M5-labeled citrate from [U-^13^C~5~] glutamine was observed, indicating no increase in reductive carboxylation for lipid synthesis.
+These findings were similar in PASMCs (Supplementary Figure 5) where stable isotope incorporation was attenuated in hypoxia and the overall labeling pattern was quite similar in both conditions.
+Compared to previous studies of metabolic flux in cancer cells [@RN2395; @RN1447], no marked increase in M5-labeled citrate from [U-^13^C~5~] glutamine was observed with hypoxia treatment, indicating no increase in reductive carboxylation for lipid synthesis.
 The overall fraction of M5-citrate in these cells was low (< 6%).
+
+## Metabolic flux in hypoxia is closely coupled to cell growth rate
+
+To clarify changes in intracellular metabolite fluxes, we next generated metabolic flux models incorporating the extracellular flux measurements and stable isotope tracing data described above.
+Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (Figure \ref{fig:f3A}, Supplementary Figure 6).
+Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by INCA [@RN1522; @RN2501; @RN964] (Figure \ref{fig:f4}, Supplementary Figure 6, and Supplementary Tables 1-3).
+Overall, proliferating lung fibroblasts demonstrated high rates of glucose uptake and glycolysis.
+Approximately 15% of cytoplasmic pyruvate enters the TCA cycle with the balance converted to lactate.
+In hypoxia, significant reductions in glycolysis, the TCA cycle, and amino acid metabolism were observed (Figure \ref{fig:f4}A).
+To facilitate the identification of differential carbon utilization between treatment groups, metabolite fluxes were normalized to cell growth rate.
+In this analysis, a modest increase in glycolytic flux is observed (Supplementary Figure 4D).
+This observation suggests that the effects of the HIF-1 transcriptional program are evident only after adjusting for differences in cell growth rate.
+
+Metabolite fluxes in DMSO-treated cells were similar to 21% oxygen controls.
+Compared to hypoxia, BAY treatment was associated with an increase in glycolysis and lactate fermentation in 21% oxygen, while similar decreases in serine and glutamine incorporation were observed (Figure \ref{fig:f4}D).
+
+```{r exchange-fluxes}
+exch_df <- dplyr::filter(model_fluxes, type == "exch" & id == "MCT")
+nm <- c("norm", "hyp", "dmso", "bay")
+mct_flux <- dplyr::pull(exch_df, flux) %>% prettyNum(big.mark = ",") %>% rlang::set_names(nm)
+mct_lb <- dplyr::pull(exch_df, lb) %>% prettyNum(big.mark = ",") %>% rlang::set_names(nm)
+mct_ub <- dplyr::pull(exch_df, ub) %>% prettyNum(big.mark = ",") %>% rlang::set_names(nm)
+```
 
@@
 
+\newpage
+
 # Figure Legends
 
+(ref:f1) **Extracellular fluxes of lung fibroblasts in hypoxia.** (**A**) Growth curves of lung fibroblasts (LFs) cultured in 21% and 0.5% oxygen (n = 8).
+(**B**) Growth rates were determined by linear fitting of log-transformed growth curves.
+(**C**) Cell viability, assessed by acridine orange plus propidium iodide staining, did not differ between 21% and 0.5% oxygen culture conditions (n = 3 technical replicates).
+(**D**) Representative immunoblot of LF protein lysates collected at the indicated times.
+(**E**, **F**) Relative change in HIF-1α (E) and LDHA (F) protein levels compared to 21% oxygen time 0 (n = 4).
+(**G**, **H**) Relative changes in GLUT1 (G) and LDHA (H) mRNA levels compared to 21% oxygen time 0 (n = 4).
+(**I**, **J**) Extracellular fluxes of the indicated metabolites (n = 8).
+Data are mean ± SEM.
+Comparisons were performed using Student's paired *t*-test and p < 0.05 indicated by *.
+Flux probability values were not corrected for multiple comparisons.
+Abbreviations as noted in the text.
+
+(ref:f2) **Prolyl hydroxylase inhibition of lung fibroblasts in normoxia.** (**A**) Growth curves of lung fibroblasts (LFs) cultured in 21% oxygen and treated with molidustat (BAY, 10 μM) or vehicle (DMSO, 0.1%) (n = 8).
+(**B**) Growth rates were determined by linear fitting of log-transformed growth curves.
+(**C**) Representative immunoblot of LF protein lysates collected at the indicated times.
+(**D**, **E**) Relative change in HIF-1α (D) and LDHA (E) protein levels compared to DMSO time 0.
+(**F**, **G**) Relative changes in GLUT1 (F) and LDHA (G) mRNA levels compared to DMSO time 0.
+(**H**, **I**) Extracellular fluxes of the indicated metabolites (n = 8).
+Data are mean ± SEM.
+Comparisons were performed using Student's paired *t*-test and p < 0.05 indicated by *.
+Flux probability values were not corrected for multiple comparisons.
+Abbreviations as noted in the text.
+
+(ref:f3)**Stable isotope tracing of lung fibroblasts.** (**A**) Fraction of pyruvate labeling following treatment of lung fibroblasts (LFs) with [U-^13^C~6~] glucose.
+(**B**) Rate values determined from asymptotic regression fit of the data from (A).
+(**C**) Isotopic labeling of key intracellular metabolites after 72 h of treatment with the indicated tracers.
+Data are the mean ± SEM of 4 biological replicates.
+FBP, fructose bisphosphate; 3PG, 3-phosphoglycerate; AKG, α-ketoglutarate.
+
+(ref:f4) **Metabolic flux maps of lung fibroblasts.** (**A**) Ratio of metabolic fluxes in 0.5% oxygen compared to 21% oxygen.
+(**B**) Ratio of metabolic fluxes in cells treated with molidustat (BAY) compared to DMSO vehicle control.
+Fluxes with non-overlapping confidence intervals are highlighted to indicate significant changes.
+
+(ref:f5) **PHD inhibition increases lactate uptake and oxidation.** Mass isotopomer distributions of key metabolites following labeling with [U-^13^C~3~] lactate (2 mM) for 72 h indicates increased lactate uptake and oxidation in hypoxia or with molidustat (BAY) treatment.
+
+(ref:f6) **Metabolomic analysis of molidustat treamtent in normoxia and hypoxia.** (**A**-**C**) Hypoxia inhibits the effects of HIF-1α stabilization on glycolysis.
+Lung fibroblasts (LFs) were cultured in standard growth medium and treated with molidustat (BAY, 10 μM) or vehicle (DMSO, 0.1%) in 21% or 0.5% oxygen conditions (n = 4).
+(A) Growth rates were determined by linear fitting of log-transformed growth curves.
+(B, C) Extracellular fluxes of glucose (B) and lactate (C).
+(**D**) Principle components analysis of intracellular metabolites following 72 h of treatment described above suggests a dominant effect of hypoxia over PHD inhibition on the metabolome (n = 4).
+(**E**) Volcano plot of the differential effects of hypoxia and molidustat (BAY) treatment on intracellular metabolites.
+(**F**) Representative metabolites from (E) reveal different patterns of metabolic effects.
+(**G**-**I**) Intracellular NAD^+^, NADH, and their ratio were determined by enzymatic assay (n = 5).
+Black * indicate a significant effect of hypoxia, colored * indicate a significant effect of treatment within a given oxygen tension as indicated by the color.
+Comparisons for fluxes (A-C) and NAD(H) measurements (G-I) were performed using a mixed-effects linear model with replicate as a random effect.
+Adjusted p-values for the indicated comparisons were determined using Tukey's *post hoc* test.
+
+(ref:f7) **Transcriptomic analysis of molidustat treamtent in normoxia and hypoxia.** (**A**) Principle components analysis of transcriptional changes following 72 h of treatment with 0.5% oxygen or molidustat (BAY), separately or together (n = 4).
+(**B**) Volcano plot of the differential effects of hypoxia and molidustat (BAY) treatment on intracellular metabolites.
+(**C**) Representative transcripts from (B).
+(**D**) Gene set enrichment analysis of trasncripts from (B).
+(**E**) Transcription factor enrichment analysis suggests mechanisms for differential regulation of gene expression following hypoxia or BAY treatment.
+(**F**) Relative change in c-MYC protein levels after 72 h of treatment (n = 4) and a representative immunoblot.
+Comparisons were performed using a mixed-effects linear model with replicate as a random effect.
+Adjusted p-values for the indicated comparisons were determined using Tukey's *post hoc* test.
+Black * indicate a significant effect of treatment, colored * indicate a significant effect of oxygen within a given treatment as indicated by the color.
+
+Figure 1: (ref:f1)
+
+Figure 2: (ref:f2)
+
+Figure 3: (ref:f3)
+
+Figure 4: (ref:f4)
+
+Figure 5: (ref:f5)
+
+Figure 6: (ref:f6)
+
+Figure 7: (ref:f7)
+
+\newpage
+
 # Tables
+
+\newpage
+
+# Figures
+
+```{r f1, fig.cap = "(ref:f1)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/m1.pdf",
+    package = "Copeland.2021.hypoxia.flux"
+  )
+)
+```
+
+\newpage
+
+```{r f2, fig.cap = "(ref:f2)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/m2.pdf",
+    package = "Copeland.2021.hypoxia.flux"
+  )
+)
+```
+
+\newpage
+
+```{r f3, fig.cap = "(ref:f3)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/m3.pdf",
+    package = "Copeland.2021.hypoxia.flux"
+  )
+)
+```
+
+\newpage
+
+```{r f4, fig.cap = "(ref:f4)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/m4.pdf",
+    package = "Copeland.2021.hypoxia.flux"
+  )
+)
+```
+
+\newpage
+
+```{r f5, fig.cap = "(ref:f5)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/m5.pdf",
+    package = "Copeland.2021.hypoxia.flux"
+  )
+)
+```
+
+\newpage
+
+```{r f6, fig.cap = "(ref:f6)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/m6.pdf",
+    package = "Copeland.2021.hypoxia.flux"
+  )
+)
+```
+
+\newpage
+
+```{r f7, fig.cap = "(ref:f7)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/m7.pdf",
+    package = "Copeland.2021.hypoxia.flux"
+  )
+)
+```
 

### 2021-05-03 inst/manuscript/supplement.Rmd
incorporating figures and tables

--- 2021-04-02 inst/manuscript/manuscript.Rmd
@@
 ---
-title: "Hypoxia uncouples HIF gene transcription and metabolic flux in proliferating primary cells"
+title: "Supplemental Information"
+subtitle: "Hypoxia uncouples HIF gene transcription and metabolic flux in proliferating primary cells"
 
@@
 }
+
+withr::with_dir(here::here(), {
+  hyp <- targets::tar_read(lf_hypoxia_table)
+  bay <- targets::tar_read(lf_bay_table)
+})
+
 ```
-
-# Summary 
-(150 words)
-
-# Keywords
-
-hypoxia, metabolic flux analysis, hypoxia-inducible factor, prolyl hydroxylase, metabolism
 
@@
 
-# Introduction
-
-Metazoan cells depend on aerobic respiration to meet cellular energy demands.
-With an inadequate oxygen supply, or hypoxia, cells must reduce energy consumption and shift energy production away from oxidative phosphorylation.
-Cells accomplish this goal through stabilization of the hypoxia-inducible transcription factor 1α (HIF-1α), which activates the transcription of glucose transporters, glycolytic enzymes, lactate dehydrogenase, and pyruvate dehydrogenase kinase, while decreasing the expression of proteins in the tricarboxylic acid (TCA) cycle and electron transport chain [@RN1980, @RN1198].
-Overall, these changes in gene transcription should increase glycolytic capacity and divert glucose-derived pyruvate from oxidative phosphorylation toward lactate fermentation to maintain energy production and minimize the formation of reactive oxygen species [@RN2525].
-
-While this "glycolytic shift" of primary carbon metabolism is well-described, the effects of hypoxia on other metabolic pathways are an area of active investigation [@RN953, @RN1603, @RN1198].
-Since hypoxia is a prominent feature of cancer biology as tumor growth outstrips blood supply, most detailed metabolic studies of cell metabolism in hypoxia have used tumor cell models [@RN2395, @RN1447, @RN1438, @RN634, @RN1200, @RN975].
-For example, stable isotope tracing and metabolic flux analyses identified a critical role for reductive carboxylation of glutamine-derived 2-oxoglutarate for lipid biosynthesis in tumor growth [@RN1447; @RN628; @RN1966; @RN2395], and metabolomic studies identified aspartate as a limiting metabolite for cancer cell proliferation under hypoxia [@RN634].
-By contrast, comparatively little is known about metabolic adaptations of primary cells to hypoxia and the importance of the metabolic pathways described above remain to be elucidated.
-This information would provide important context for understanding the extent to which cancer cell metabolism responds differently to hypoxic stress.
-Given the metabolic adaptations required for rapid proliferation in cancer cells, we hypothesized that hypoxia would elicit different metabolic responses in primary cells than has been observed previously in studies of cancer cell metabolism.
-
-To test this hypothesis, we developed models of bioenergetic carbon flux in human primary cells cultured under 21% or 0.5% oxygen conditions.
-We found that hypoxia fails to increase glycolysis in primary cells despite robust up-regulation of the HIF-1α transcriptional program.
-In normoxia, HIF-1α activation by the prolyl hydroxylase inhibitor molidustat (BAY-85-3934, "BAY") [@RN580] did increase glycolysis and lactate efflux; however, hypoxia abrogated this response.
-These findings suggest the existence of hypoxia-dependent regulatory mechanisms that override the metabolic consequences of HIF-1α-dependent up-regulation of glycolytic gene expression in human primary cells.
-
-# Results
-
-The goal of this study was to identify the metabolic changes associated with hypoxia in proliferating primary cells.
-Cells were seeded and placed into hypoxia for 24 h prior to sample collection to provide adequate time for activation of the hypoxia-dependent transcriptional program.
-From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (Figure \ref{fig:f1}A), thought to be an indicator of metabolic steady state.
-Lung fibroblasts cultured in 0.5% oxygen grew more slowly (Figure \ref{fig:f1}B), but slower growth was not associated with decreased cell viability (Figure \ref{fig:f1}C).
-These cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1, (GLUT1), and lactate dehydrogenase A (LDHA) (Figure \ref{fig:f1}D-H).
-These changes persisted for the duration of the experimental time course.
-
-## Extracellular flux analysis reveals little impact of hypoxia on glycolysis
-
-Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (Figure \ref{fig:f1}I-J, Supplementary Figures 1-3).
-Flux calculations incorporated the changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@RN1523] (Supplementary Figure 1).
-Surprisingly, while glucose uptake was modestly increased in hypoxia, lactate efflux was decreased (Figure \ref{fig:f1}I) despite activation of the HIF-1 transcriptional program.
-Similar findings were observed when the ambient oxygen level was decreased further to 0.2% (Supplementary Figure 2) and in pulmonary artery smooth muscle cells (PASMCs; Supplementary Figure 3).
-In addition to glucose and lactate, the absolute extracellular fluxes of pyruvate and amino acids were generally decreased in hypoxia, including a marked decrease in glutamine uptake in lung fibroblasts.
-Notably, hypoxia was previously shown to increase glutamine uptake in studies of cancer cell metabolism [@RN1447; @RN2395; @RN628].
-Glutamine uptake did increase in PASMCs, as did the uptake of branch chain amino acids and arginine, suggesting differential responses of these mesenchymal cells to hypoxia.
-
-Given that hypoxia did not increase glucose and lactate fluxes as expected from the associated increases in glycolytic gene expression, we next assessed the capacity of HIF-1α to augment glycolysis in lung fibroblasts.
-Cells were treated with BAY to stabilize HIF-1α under 21% oxygen conditions (Figure \ref{fig:f2}).
-Similar to hypoxia, BAY decreased cell growth rate (Figure \ref{fig:f2}A-B) and activated the HIF-1 transcriptional program (Figure \ref{fig:f2}C-G).
-Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1 target gene transcription and protein expression.
-In normoxia, this transcriptional program was associated with increased glucose uptake and lactate efflux (Figure \ref{fig:f2}H).
-Relatively modest effects on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (Figure \ref{fig:f2}J).
-
-## Stable isotope tracing
-
-To characterize further the extent of hypoxia-induced metabolic reprogramming, we next treated lung fibroblasts with stable carbon isotopes of glucose and glutamine to trace label incorporation into key carbon utilization pathways (Figure \ref{fig:f3} and Supplementary Figure 4).
-Overall, hypoxia-treated cells had decreased label incorporation into downstream metabolites (*i.e.*, the unlabeled, or M0, fraction was greater).
-This finding is consistent with the extracellular flux measurements suggesting slower substrate utilization by hypoxic cells.
-BAY treatment recapitulated the labeling pattern observed with hypoxia, suggesting similar intracellular metabolite flux between these two conditions.
-Beyond this observation, the labeling patterns in hypoxia- and BAY-treated cells were similar to their respective controls, arguing against marked metabolic reprogramming in response to prolyl hydroxylase inhibition by either hypoxia or BAY.
-Compared to previous studies of metabolic flux in cancer cells [@RN2395; @RN1447], no hypoxia-mediated increase in M5-labeled citrate from [U-^13^C~5~] glutamine was observed, indicating no increase in reductive carboxylation for lipid synthesis.
-The overall fraction of M5-citrate in these cells was low (< 6%).
-
-# Discussion
-
-# Acknowledgements
-
-# Author Contributions
-
-# Declaration of Interests
-
-# Figure Legends
-
 # Tables
 
-# STAR Methods
+<br>
+
+```{r results = "asis"}
+hyp %>% 
+  flextable::set_caption("Table 1: Lung fibroblast fluxes in 21% and 0.5% oxygen") %>% 
+  flextable::flextable_to_rmd()
+```
+
+\newpage
+
+```{r results = "asis"}
+bay %>% 
+  flextable::set_caption("Table 2: Lung fibroblast fluxes following DMSO and BAY treatment") %>% 
+  flextable::flextable_to_rmd()
+```

### 2021-05-03 _targets.R

@@ -859,16 +859,23 @@ list(
     plot_ratio_network(bay_graph, "BAY/DMSO")
   ),
   tar_target(
-    m4c,
-    plot_lactate_mids(model_mids, "lf")
-  ),
-  tar_target(
     m4,
-    arrange_m4(hypoxia_graph_ratio_plot, bay_graph_ratio_plot, m4c)
+    arrange_m4(hypoxia_graph_ratio_plot, bay_graph_ratio_plot)
   ),
   tar_target(
     m4_figure,
     write_figures(m4, "m4.pdf")
+  ),
+
+  # M5 ----------------------------------------------------------------------
+
+  tar_target(
+    m5,
+    plot_lactate_mids(model_mids, "lf")
+  ),
+  tar_target(
+    m5_figure,
+    write_figures(m5, "m5.pdf")
   ),
 
   # S6 ----------------------------------------------------------------------
@@ -906,63 +913,63 @@ list(
     write_figures(s6, "s6.pdf")
   ),
 
-  # M5 ----------------------------------------------------------------------
+  # M6 ----------------------------------------------------------------------
 
   tar_target(
     twoby_fluxes,
     analyze_twoby_fluxes(growth_rates, fluxes)
   ),
   tar_target(
-    m5a,
+    m6a,
     plot_twoby_fluxes(twoby_fluxes$data, twoby_fluxes$annot, "growth", "Growth Rate (/h)")
   ),
   tar_target(
-    m5b,
+    m6b,
     plot_twoby_fluxes(twoby_fluxes$data, twoby_fluxes$annot, "glucose", "Glucose\n(fmol/cell/h)") + ggplot2::scale_y_reverse()
   ),
   tar_target(
-    m5c,
+    m6c,
     plot_twoby_fluxes(twoby_fluxes$data, twoby_fluxes$annot, "lactate", "Lactate\n(fmol/cell/h)")
   ),
   tar_target(
-    m5g,
+    m6g,
     plot_nad(nad_final, nad_annot, "NAD", "NAD\n(nmol/cell)")
   ),
   tar_target(
-    m5h,
+    m6h,
     plot_nad(nad_final, nad_annot, "NADH", "NADH\n(nmol/cell)")
   ),
   tar_target(
-    m5i,
+    m6i,
     plot_nad(nad_final, nad_annot, "Ratio", "NADH/NAD ratio")
   ),
   tar_target(
-    m5,
[…]

### 2021-05-03 inst/src/analysis.R

@@ -966,7 +966,7 @@ plot_labeling_rate <- function(mids) {
     ggplot2::scale_fill_manual(values = clrs) +
     ggplot2::labs(
       x = "Time (h)",
-      y = "Fraction labeled",
+      y = "Pyruvate labeled",
       color = NULL,
       fill = NULL
     ) +
@@ -1316,7 +1316,8 @@ plot_lactate_mids <- function(model_mids, cell) {
     dplyr::filter(tracer == "lac3" & time == 72 & !is.na(metabolite)) %>%
     plot_mids() +
     ggplot2::facet_wrap(~ metabolite, scales = "free_x", nrow = 2) +
-    ggplot2::theme(legend.position = "right")
+    ggplot2::theme(legend.position = "right") +
+    theme_patchwork(widths = unit(4, "in"), heights = unit(2.5, "in"), tags = NULL)
 }
 
 # format_time_course_mids -------------------------------------------------

### 2021-05-03 inst/src/figures.R

@@ -135,18 +135,13 @@ arrange_m3 <- function(m3ab, m3c) {
 
 # arrange_m4 --------------------------------------------------------------
 
-arrange_m4 <- function(hypoxia_graph_ratio_plot, bay_graph_ratio_plot, m4c) {
+arrange_m4 <- function(hypoxia_graph_ratio_plot, bay_graph_ratio_plot) {
 
   a <- hypoxia_graph_ratio_plot
   b <- bay_graph_ratio_plot
 
-  top <-
-    a + b +
-    plot_layout(guides = "collect") &
-    theme(legend.text.align = 1)
-
-  top / m4c +
-    theme_patchwork(widths = unit(5, "in"), heights = unit(c(2.5, 2), "in"))
+  a + b +
+    theme_patchwork(widths = unit(2.5, "in"), heights = unit(2.75, "in"), guides = "collect")
 }
 
 # arrange_s6 --------------------------------------------------------------
@@ -165,9 +160,9 @@ arrange_s6 <- function(a, b, c, d) {
     )
 }
 
-# arrange_m5 --------------------------------------------------------------
+# arrange_m6 --------------------------------------------------------------
 
-arrange_m5 <- function(a, b, c, d, e, f, g, h, i) {
+arrange_m6 <- function(a, b, c, d, e, f, g, h, i) {
   (a | b | c) / (d | e) / f / (g | h | i ) +
     theme_patchwork(
       # design = layout,
@@ -181,9 +176,9 @@ arrange_m5 <- function(a, b, c, d, e, f, g, h, i) {
     )
 }
 
-# arrange_m6 --------------------------------------------------------------
+# arrange_m7 --------------------------------------------------------------
 
-arrange_m6 <- function(a, b, c, d, e, f) {
+arrange_m7 <- function(a, b, c, d, e, f) {
   layout <- "
   abd
   ccd
@@ -259,7 +254,96 @@ create_resources <- function() {
       part = "body",
       value = flextable::as_paragraph("[U-", flextable::as_sup("13"), "C", flextable::as_sub("3"), "] lactate")
     ) %>%
-    flextable::font(fontname = "Calibri") %>%
-    flextable::autofit() %>%
-    flextable::fit_to_width(7.25)
+    flextable::font(fontname = "Calibri", part = "all") %>%
+    flextable::fontsize(size = 9, part = "all") %>%
+    flextable::set_table_properties(layout = "autofit")
+}
+
[…]

### 2021-05-05 inst/manuscript/manuscript.Rmd
add metabolite set enrichment analysis panels

@@
     df_print: kable
+    fig_caption: true
     number_sections: false
@@
 Cells were seeded and placed into hypoxia for 24 h prior to sample collection to provide adequate time for activation of the hypoxia-dependent transcriptional program.
-From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (Figure \ref{fig:f1}A), thought to be an indicator of metabolic steady state.
-Lung fibroblasts (LFs) cultured in 0.5% oxygen grew more slowly (Figure \ref{fig:f1}B), but slower growth was not associated with decreased cell viability (Figure \ref{fig:f1}C).
-These cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1, (GLUT1), and lactate dehydrogenase A (LDHA) (Figure \ref{fig:f1}D-H).
+From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (Figure \@ref(fig:f1)A), thought to be an indicator of metabolic steady state.
+Lung fibroblasts (LFs) cultured in 0.5% oxygen grew more slowly (Figure \@ref(fig:f1)B), but slower growth was not associated with decreased cell viability (Figure \@ref(fig:f1)C).
+These cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1, (GLUT1), and lactate dehydrogenase A (LDHA) (Figure \@ref(fig:f1)D-H).
 These changes persisted for the duration of the experimental time course.
@@
 
-Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (Figure \ref{fig:f1}I-J, Supplementary Figures 1-3).
+Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (Figure \@ref(fig:f1)I-J, Supplementary Figures 1-3).
 Flux calculations incorporated the changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@RN1523] (Supplementary Figure 1).
-Surprisingly, while glucose uptake was modestly increased in hypoxia, lactate efflux was decreased (Figure \ref{fig:f1}I) despite activation of the HIF-1 transcriptional program.
+Surprisingly, while glucose uptake was modestly increased in hypoxia, lactate efflux was decreased (Figure \@ref(fig:f1)I) despite activation of the HIF-1 transcriptional program.
 Similar findings were observed when the ambient oxygen level was decreased further to 0.2% (Supplementary Figure 2) and in pulmonary artery smooth muscle cells (PASMCs; Supplementary Figure 3).
@@
 Given that hypoxia did not increase glucose and lactate fluxes as expected from the associated increases in glycolytic gene expression, we next assessed the capacity of HIF-1α to augment glycolysis in LFs.
-Cells were treated with BAY to stabilize HIF-1α under 21% oxygen conditions (Figure \ref{fig:f2}).
-Similar to hypoxia, BAY decreased cell growth rate (Figure \ref{fig:f2}A-B) and activated the HIF-1 transcriptional program (Figure \ref{fig:f2}C-G).
+Cells were treated with BAY to stabilize HIF-1α under 21% oxygen conditions (Figure \@ref(fig:f2).
+Similar to hypoxia, BAY decreased cell growth rate (Figure \@ref(fig:f2)A-B) and activated the HIF-1 transcriptional program (Figure \@ref(fig:f2)C-G).
 Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1 target gene transcription and protein expression.
-In normoxia, this transcriptional program was associated with substantially increased glucose uptake and lactate efflux (Figure \ref{fig:f2}H).
-Comparatively modest effects of BAY on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (Figure \ref{fig:f2}J).
+In normoxia, this transcriptional program was associated with substantially increased glucose uptake and lactate efflux (Figure \@ref(fig:f2)H).
+Comparatively modest effects of BAY on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (Figure \@ref(fig:f2)J).
 
@@
 
-To validate our findings from extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate of label incorporation and trace its incorporation into key carbon utilization pathways (Figure \ref{fig:f3} and Supplementary Figures 4-5).
-Consistent with our extracellular flux measurements, we found that the rate of label incorporation from [U-^13^C~6~] glucose into pyruvate was similar across 21%, 0.5%, and DMSO-treated conditions, but markedly increased with BAY treatment (\ref{fig:f3}A-B).
+To validate our findings from extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate of label incorporation and trace its incorporation into key carbon utilization pathways (Figure \@ref(fig:f3) and Supplementary Figures 4-5).
+Consistent with our extracellular flux measurements, we found that the rate of label incorporation from [U-^13^C~6~] glucose into pyruvate was similar across 21%, 0.5%, and DMSO-treated conditions, but markedly increased with BAY treatment (\@ref(fig:f3)A-B).
 
@@
 To clarify changes in intracellular metabolite fluxes, we next generated metabolic flux models incorporating the extracellular flux measurements and stable isotope tracing data described above.
-Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (Figure \ref{fig:f3A}, Supplementary Figure 6).
-Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by INCA [@RN1522; @RN2501; @RN964] (Figure \ref{fig:f4}, Supplementary Figure 6, and Supplementary Tables 1-3).
-Overall, proliferating lung fibroblasts demonstrated high rates of glucose uptake and glycolysis.
-Approximately 15% of cytoplasmic pyruvate enters the TCA cycle with the balance converted to lactate.
-In hypoxia, significant reductions in glycolysis, the TCA cycle, and amino acid metabolism were observed (Figure \ref{fig:f4}A).
-To facilitate the identification of differential carbon utilization between treatment groups, metabolite fluxes were normalized to cell growth rate.
-In this analysis, a modest increase in glycolytic flux is observed (Supplementary Figure 4D).
-This observation suggests that the effects of the HIF-1 transcriptional program are evident only after adjusting for differences in cell growth rate.
+Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (Figure \@ref(fig:f3A), Supplementary Figure 6).
+Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by INCA [@RN1522; @RN2501; @RN964] (Figure \@ref(fig:f4), Supplementary Figure 6, and Supplementary Tables 1-3).
+Overall, proliferating LFs demonstrated high rates of glucose uptake and glycolysis.
+Approximately 10% of cytoplasmic pyruvate enters the TCA cycle with the balance converted to lactate (Supplementary Figure 6C).
+In hypoxia, significant reductions in glycolysis, the TCA cycle, and amino acid metabolism were observed (Figure \@ref(fig:f4)A) with a significant increase in pentose phosphate pathway flux.
 
+Given the global decrease in bioenergetic metabolic flux in hypoxia, we hypothesized that these differences may be a consequence of a decreased growth rate.
+After normalizing metabolite fluxes in normoxia and hypoxia to the associated growth rate, modest increases (~10%) in glycolytic flux were observed (Supplementary Figure 4D).
+This finding suggests that, while glycolysis increases relative to growth rate in hypoxic cells, the regulators of cell proliferation rate dominate over the consequences of the HIF-1 transcriptional program.
+Indeed, even after adjusting for cell growth rate, the relative increase in glycolytic flux is modest compared to the marked up-regulation of glycolytic genes.
+
+By comparison, HIF-1 activation by BAY in 21% oxygen increased glycolysis and lactate fermentation by ~50% (Figure \@ref(fig:f4)B).
+BAY treatment decreased cell proliferation rate (Figure \@ref(fig:f2)B), indicating that, unlike hypoxia, BAY treatment uncouples cell proliferation and metabolic flux.
+Similar to hypoxia, decreases in serine and glutamine incorporation were observed.
 Metabolite fluxes in DMSO-treated cells were similar to 21% oxygen controls.
-Compared to hypoxia, BAY treatment was associated with an increase in glycolysis and lactate fermentation in 21% oxygen, while similar decreases in serine and glutamine incorporation were observed (Figure \ref{fig:f4}D).
 
-```{r exchange-fluxes}
-exch_df <- dplyr::filter(model_fluxes, type == "exch" & id == "MCT")
-nm <- c("norm", "hyp", "dmso", "bay")
-mct_flux <- dplyr::pull(exch_df, flux) %>% prettyNum(big.mark = ",") %>% rlang::set_names(nm)
-mct_lb <- dplyr::pull(exch_df, lb) %>% prettyNum(big.mark = ",") %>% rlang::set_names(nm)
-mct_ub <- dplyr::pull(exch_df, ub) %>% prettyNum(big.mark = ",") %>% rlang::set_names(nm)
-```
+## Hypoxia and BAY treatment increase lactate oxidation
+
+Although the metabolite exchange fluxes for bidirectional reactions tend to be poorly resolved, two observations are worth highlighting (Supplementary Tables 1 and 2).
+First, consistent with the stable isotope tracing results, the rate of reductive carboxylation through reversible flux by isocitrate dehydrogenase is low (~4 fmol/cell/h), is unchanged by hypoxia, and is modestly increased by BAY treatment.
+
+Second, hypoxia and BAY treatment are associated with a marked increase in the lactate transport exchange flux in LFs from ~ 0 to 1,520 and 711 fmol/cell/h in 0.5% oxygen and BAY treatment conditions, respectively.
+Since the net lactate transport flux is secretion, this observation suggests increased lactate uptake with hypoxia or BAY treatment.
+This may be consistent with the HIF-driven increased expression of the reversible lactate transporter MCT4 [@RN366].
+To investigate this hypothesis, LFs and PASMCs were treated with [U-^13^C~3~] lactate (2 mM) and ^13^C incorporation into intracellular metabolites was analyzed by LC-MS (Figure \@ref(fig:f5), Supplementary Figure 5).
+Here, we observed increased labeling of TCA metabolites citrate (CIT), 2-oxoglutrate (2OG), malate (MAL), and aspartate (ASP) following hypoxia or BAY treatment (Figure \@ref(fig:f5) and Supplementary Figure 4).
+Interestingly, a similar increase in labeling was not observed in PASMCs (Supplementary Figure 5).
+
+Notably, lactate labeled ~50% of citrate and ~20% of downstream TCA cycle metabolites in both LFs and PASMCs, indicating that lactate may be an important respiratory fuel source in these cells even though lactate efflux is high.
+Although lactate has been used less commonly than glucose and glutamine in stable isotope tracing studies, Faubert and colleagues [-@RN2600] demonstrated lactate incorporation in human lung adenocarcinoma *in vivo*.
+In this study, lactate incorporation corresponded to regions of high glucose uptake as determined by `[`^18^F]-fluorodeoxyglucose positron emission tomography, suggesting that lactate consumption can occur in areas of high glucose utilization.
+Subsequently, several investigators have demonstrated the importance of lactate as a metabolic fuel *in vivo* [@RN2756; @RN2718].
+In addition to downstream metabolites, we also observed hypoxia- and BAY-dependent increases in lactate incorporation in fructose bisphosphate (FBP) and 3-phosphoglycerate (3PG).
+This is consistent with prior reports describing hypoxia-mediated increases in glycogen synthesis [@RN1678; @RN545; @RN1668].
+Together, these data suggest that lactate makes a modest (~5% carbon) contribution to this process.
+
+## Hypoxia abrogates the increase in glycolysis following pharmacologic prolyl hydroxylase inhibition
+
+To reconcile the differential effects of prolyl hydroxylase inhibition by hypoxia and BAY, we next addressed whether hypoxia could suppress the effects of BAY on glucose and lactate fluxes (Figure \@ref(fig:f6)A-C).
+LFs cultured in standard growth medium were treated with BAY and placed in either 21% or 0.5% oxygen.
+Similar to previous experiments, BAY treatment decreased cell growth rate, increased glucose uptake, and increased lactate efflux in 21% oxygen.
+Interestingly, when combined with 0.5% oxygen, BAY treatment was unable to enhance lactate efflux.
+These data suggest that hypoxia antagonizes the effects of HIF-1 activation on glycolytic flux in these primary cells.
+
+To investigate these metabolic differences further, we performed metabolomic profiling of LFs treated for 72 h with hypoxia or BAY separately or in combination (Figure \@ref(fig:f6)D-F).
+Principle component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen (Figure \@ref(fig:f6)D).
+These hypoxia-treated cells were well-segregated from BAY-treated cells, again highlighting differential effects of hypoxic and pharmacologic PHD inhibition.
+Linear modeling identified those metabolites that were differentially affected by hypoxia and BAY treatment (*i.e.*, those metabolites with a significant interaction term in the linear model described by treatment × oxygen) (Figure \@ref(fig:f6)E).
+Several patterns emerged from this analysis (Figure \@ref(fig:f6)F).
+Interestingly, the hypoxia-regulated metabolite, 2-hydroxyglutarate, increased following BAY treatment, but was unaffected by hypoxia treatment in these proliferating LFs [@RN1603].
+The tricarboxylic acid cycle metabolite aconitate was decreased in hypoxia, but increased by BAY treatment while the opposite effect was observed for hydroxyproline and taurine.
+Glycolytic intermediate glyceraldehyde 3-phosophate (GAP) was decreased by both hypoxia and BAY, while hypoxia reversed the BAY-mediated decrease in γ-aminobutyric acid (GABA).
+A metabolite set enrichment analysis revealed the tricarboxlyic acid cycle to be the most
+
+One of the key differences associated with
 
@@
 
-(ref:f3)**Stable isotope tracing of lung fibroblasts.** (**A**) Fraction of pyruvate labeling following treatment of lung fibroblasts (LFs) with [U-^13^C~6~] glucose.
+(ref:f3) **Stable isotope tracing of lung fibroblasts.** (**A**) Fraction of pyruvate labeling following treatment of lung fibroblasts (LFs) with [U-^13^C~6~] glucose.
 (**B**) Rate values determined from asymptotic regression fit of the data from (A).

### 2021-05-05 inst/manuscript/supplement.Rmd
add metabolite set enrichment analysis panels

@@
 
-```{r results = "asis"}
+```{r t1, results = "asis"}
 hyp %>% 
-  flextable::set_caption("Table 1: Lung fibroblast fluxes in 21% and 0.5% oxygen") %>% 
+  flextable::set_caption(
+    "Lung fibroblast fluxes in 21% and 0.5% oxygen", 
+    autonum = officer::run_autonum()
+  ) %>% 
   flextable::flextable_to_rmd()
@@
 
-```{r results = "asis"}
+```{r t2, results = "asis"}
 bay %>% 
-  flextable::set_caption("Table 2: Lung fibroblast fluxes following DMSO and BAY treatment") %>% 
+  flextable::set_caption(
+    "Lung fibroblast fluxes following DMSO and BAY treatment", 
+    autonum = officer::run_autonum()
+  ) %>% 
   flextable::flextable_to_rmd()

### 2021-05-05 _targets.R

@@ -516,7 +516,23 @@ list(
   ),
   tar_target(
     metab_moi,
-    plot_mois(metab_targeted_clean, c("GAP", "2-hydroxyglutarate", "aconitate", "taurine", "hydroxyproline", "malate"))
+    plot_mois(metab_targeted_clean, c("GAP", "2-hydroxyglutarate", "aconitate", "taurine", "hydroxyproline", "GABA"))
+  ),
+  tar_target(
+    metab_msea,
+    run_msea(metab_different_differences, metab_pathways)
+  ),
+  tar_target(
+    metab_pathways,
+    get_metab_pathways()
+  ),
+  tar_target(
+    msea_plot,
+    plot_msea(metab_msea)
+  ),
+  tar_target(
+    leading_edge,
+    plot_leading_edge(metab_different_differences, metab_pathways[["(KEGG) Citrate cycle (TCA cycle)"]])
   ),
   tar_render(
     metabolomics_report,
@@ -945,7 +961,7 @@ list(
   ),
   tar_target(
     m6,
-    arrange_m6(m6a, m6b, m6c, metab_targeted_pca, metab_volcano, metab_moi, m6g, m6h, m6i)
+    arrange_m6(m6a, m6b, m6c, metab_targeted_pca, metab_volcano, metab_moi, msea_plot, leading_edge, m6g, m6h, m6i)
   ),
   tar_target(
     m6_figure,

### 2021-05-05 inst/src/analysis.R

@@ -810,9 +810,9 @@ plot_lf_mids <- function(model_mids) {
         levels = c("ALA", "ASP", "GLU", "GLN", "LAC", "SER")
       )
     ) %>%
-    dplyr::filter(tracer != "lac3" & time == 72 & !is.na(metabolite)) %>%
+    dplyr::filter(time == 72 & !is.na(metabolite)) %>%
     plot_mids() +
-    theme_patchwork(widths = unit(7, "in"), heights = unit(4.5, "in"), tags = NULL)
+    theme_patchwork(widths = unit(7, "in"), heights = unit(7.5, "in"), tags = NULL)
 }
 
 # plot_pasmc_mids ---------------------------------------------------------
@@ -841,13 +841,13 @@ plot_pasmc_mids <- function(model_mids) {
       ),
       metabolite = forcats::fct_recode(metabolite, "`3PG`" = "3PG")
     ) %>%
-    dplyr::filter(tracer != "lac3" & time == 48 & !is.na(metabolite)) %>%
+    dplyr::filter(time == 48 & !is.na(metabolite)) %>%
     plot_mids() +
     # ggplot2::facet_grid(
     #   metabolite ~ tracer,
     #   labeller = ggplot2::label_parsed
     # ) +
-    theme_patchwork(widths = unit(9.5, "in"), heights = unit(4.5, "in"), tags = NULL)
+    theme_patchwork(widths = unit(9.5, "in"), heights = unit(7.5, "in"), tags = NULL)
 }
 
 # clean_model_fluxes ------------------------------------------------------
@@ -1198,13 +1198,13 @@ set_gr_style <- function() {
 plot_ratio_network <- function(graph, caption) {
   fold <- c(Inf, 5, 3, 2, 1.5, 1.1, 0.91, 0.67, 0.5, 0.33, 0.2, 0)
 
-  cuts <- log(fold, base = 2)
+  fold_cuts <- log(fold, base = 2)
 
   lvls <-
     tidygraph::activate(graph, edges) %>%
     dplyr::pull(ratio) %>%
     cut(
-      cuts,
+      fold_cuts,
       include.lowest = TRUE
     )
 
@@ -1231,6 +1231,16 @@ plot_ratio_network <- function(graph, caption) {
     rev() %>%
     rlang::set_names(names(clrs))
 
+  graph <-
+    graph %>%
+    tidygraph::activate(edges) %>%
[…]

### 2021-05-05 inst/src/figures.R

@@ -141,7 +141,7 @@ arrange_m4 <- function(hypoxia_graph_ratio_plot, bay_graph_ratio_plot) {
   b <- bay_graph_ratio_plot
 
   a + b +
-    theme_patchwork(widths = unit(2.5, "in"), heights = unit(2.75, "in"), guides = "collect")
+    theme_patchwork(widths = unit(3, "in"), heights = unit(3.5, "in"), guides = "collect")
 }
 
 # arrange_s6 --------------------------------------------------------------
@@ -162,13 +162,17 @@ arrange_s6 <- function(a, b, c, d) {
 
 # arrange_m6 --------------------------------------------------------------
 
-arrange_m6 <- function(a, b, c, d, e, f, g, h, i) {
-  (a | b | c) / (d | e) / f / (g | h | i ) +
+arrange_m6 <- function(a, b, c, d, e, f, g, h, i, j, k) {
+  ((a | b | c) + plot_layout(guides = "collect")) /
+    (d | e) /
+    (f + plot_layout(guides = "collect")) /
+    ((g | h) + plot_layout(widths = c(1, 2), guides = "keep")) /
+    ((i | j | k) + plot_layout(guides = "collect")) +
     theme_patchwork(
       # design = layout,
       widths = unit(5, "in"),
-      heights = unit(c(1.25), "in"),
-      guides = "collect"
+      heights = unit(c(1.1), "in"),
+      # guides = "collect"
     ) &
     theme(
       legend.position = "bottom",

### 2021-05-05 inst/src/metabolomics.R

@@ -525,3 +525,150 @@ plot_mois <- function(clean, moi) {
     ) +
     NULL
 }
+
+# run_msea ----------------------------------------------------------------
+
+run_msea <- function(tt, pathways) {
+
+  stats <- tt$t
+  names(stats) <- tt$hmdb
+
+  msea_res <-
+    fgsea::fgsea(pathways = pathways, stats = stats) %>%
+    dplyr::filter(pval < 0.05) %>%
+    tidyr::separate(pathway, c("source", "pathway"), "\\) ") %>%
+    dplyr::mutate(source = stringr::str_replace(source, "\\(", "")) %>%
+    dplyr::arrange(-abs(NES))
+
+}
+
+# get_metab_pathways ------------------------------------------------------
+
+get_metab_pathways <- function(databases = "kegg") {
+  multiGSEA::getMultiOmicsFeatures(
+    dbs = databases,
+    layer = "metabolome",
+    organism = "hsapiens",
+    returnMetabolome = "HMDB",
+    useLocal = TRUE
+  ) %>%
+    unlist(recursive = FALSE) %>%
+    rlang::set_names(purrr::map(names(.), stringr::str_extract, pattern = "(?<=metabolome\\.).*"))
+}
+
+
+# plot_msea ---------------------------------------------------------------
+
+plot_msea <- function(msea) {
+  msea %>%
+    dplyr::select(-padj) %>%
+    dplyr::rename(padj = pval) %>%
+    plot_gsea(sources = "KEGG") +
+    ggplot2::theme(
+      axis.text.y.right = element_text(size = 5)
+    )
+}
+
+# plot_leading_edge -------------------------------------------------------
+
+plot_leading_edge <- function(tt, pathway) {
+
+  stats <- tt$t
+  names(stats) <- tt$hmdb
+
+  rnk <- rank(-stats)
+  ord <- order(rnk)
+
+  stats_adj <- stats[ord]
+  stats_adj <- stats_adj / max(abs(stats_adj))
+
+  pathway <- unname(as.vector(na.omit(match(pathway, names(stats_adj)))))
+  pathway <- sort(pathway)
+
+  gsea_res <-
+    fgsea::calcGseaStat(
+      stats_adj,
+      selectedStats = pathway,
+      returnAllExtremes = TRUE
+    )
+
+  bottoms <- gsea_res$bottoms
[…]

### 2021-05-07 inst/analysis/rnaseq.Rmd
finish up rough draft of manuscript

@@
   deg <- tar_read(rnaseq_different_differences)
+  results <- tar_read(rnaseq_different_differences)
+  gsea_res <- tar_read(rnaseq_gsea)
   gsea_p <- tar_read(rnaseq_gsea_plot)
   tfea_p <- tar_read(rnaseq_tfea_plot)
+  dds <- tar_read(dds_symbols)
 })
@@
 
+```{r, eval = FALSE}
+pathways <-
+  fgsea::gmtPathways("~/Dropbox (Partners HealthCare)/msigdb_v7.2/msigdb_v7.2_GMTs/msigdb.v7.2.entrez.gmt")
+
+df <- 
+  gsea_res %>% 
+  dplyr::filter(source == "HALLMARK") %>% 
+  dplyr::filter(padj < 0.05) %>% 
+  tidyr::unnest(leadingEdge) %>% 
+  dplyr::inner_join(results, by = c("leadingEdge" = "row")) %>% 
+  dplyr::group_by(pathway) %>% 
+  dplyr::slice_max(order_by = stat, n = 1) %>% 
+  dplyr::arrange(desc(NES))
+
+gois <- df$symbol %>% rlang::set_names(df$pathway)
+
+rnks <-
+  results %>%
+  dplyr::select(row, stat) %>%
+  dplyr::arrange(stat) %>%
+  tibble::deframe()
+
+gsea_res <-
+    fgsea::calcGseaStat(
+      rnks,
+      selectedStats = path,
+      returnAllExtremes = TRUE
+    )
+
+fgsea::plotEnrichment(
+  pathway = pathways[["HALLMARK_OXIDATIVE_PHOSPHORYLATION"]], 
+  stats = rnks
+)
+```
+
+```{r}
+plot_rnaseq_goi(dds, "HIRA")
+```
+
 ## Transcription Factor Analysis

### 2021-05-07 inst/manuscript/manuscript.Rmd
finish up rough draft of manuscript

@@
 output: 
-  # bookdown::pdf_document2:
-  #   # template: template.tex
-  #   keep_tex: false
-  #   df_print: kable
-  #   toc: false
-  #   number_sections: false
-  #   latex_engine: xelatex
-  #   pandoc_args:
-  #     - '--lua-filter=scholarly-metadata.lua'
-  #     - '--lua-filter=author-info-blocks.lua'
-  #     - '--lua-filter=pagebreak.lua'
-  #     - '--lua-filter=multiple-bibliographies.lua'
-
   bookdown::word_document2:
@@
 # Summary 
-(150 words)
+Hypoxia is an important environmental stimulus that causes transcriptional and metabolic reprogramming in cells to facilitate their survival.
+Here, we performed stable isotope tracing and metabolic flux analyses of proliferating primary cells in hypoxia.
+Despite activation of the hypoxia-inducible factor (HIF) transcriptional program and up-regulation of glycolytic genes, glycolytic flux was decreased in hypoxic cells in our models.
+No evidence for increased glutaminolysis or reductive carboxylation was observed.
+While pharmacologic stabilization of HIF in normoxia with the prolyl hydroxylase inhibitor molidustat did increase glycolytic flux as expected, hypoxia abrogated this effect.
+Multi-omic profiling of cells treated with hypoxia or molidustat, seperately or together, reveals distinct molecular responses to to hypoxia and pharmacologic prolyl hydroxylase inhibition.
+Together, these data suggest that primary cell bioenergetic metabolism is closely coupled to cell proliferation rate and that other regulatory factors override the effects of HIF-dependent up-regulation of glycolytic gene expression on glycolytic flux.
 
@@
 We found that hypoxia fails to increase glycolysis in primary cells despite robust up-regulation of the HIF-1α transcriptional program.
-In normoxia, HIF-1α activation by the prolyl hydroxylase inhibitor molidustat (BAY-85-3934, "BAY") [@RN580] did increase glycolysis and lactate efflux; however, hypoxia abrogated this response.
+In normoxia, HIF-1α activation by the prolyl hydroxylase (PHD) inhibitor molidustat (BAY-85-3934, "BAY") [@RN580] did increase glycolysis and lactate efflux; however, hypoxia abrogated this response.
 These findings suggest the existence of important hypoxia-dependent regulatory mechanisms that override the metabolic consequences of HIF-1α-dependent up-regulation of glycolytic gene expression in human primary cells.
@@
 Cells were seeded and placed into hypoxia for 24 h prior to sample collection to provide adequate time for activation of the hypoxia-dependent transcriptional program.
-From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (Figure \@ref(fig:f1)A), thought to be an indicator of metabolic steady state.
-Lung fibroblasts (LFs) cultured in 0.5% oxygen grew more slowly (Figure \@ref(fig:f1)B), but slower growth was not associated with decreased cell viability (Figure \@ref(fig:f1)C).
-These cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1, (GLUT1), and lactate dehydrogenase A (LDHA) (Figure \@ref(fig:f1)D-H).
+From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (**Figure \@ref(fig:f1)A**), thought to be an indicator of metabolic steady state.
+Lung fibroblasts (LFs) cultured in 0.5% oxygen grew more slowly (**Figure \@ref(fig:f1)B**), but slower growth was not associated with decreased cell viability (**Figure \@ref(fig:f1)C**).
+These cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1, (GLUT1), and lactate dehydrogenase A (LDHA) (**Figure \@ref(fig:f1)D-H**).
 These changes persisted for the duration of the experimental time course.
@@
 
-Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (Figure \@ref(fig:f1)I-J, Supplementary Figures 1-3).
-Flux calculations incorporated the changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@RN1523] (Supplementary Figure 1).
-Surprisingly, while glucose uptake was modestly increased in hypoxia, lactate efflux was decreased (Figure \@ref(fig:f1)I) despite activation of the HIF-1 transcriptional program.
-Similar findings were observed when the ambient oxygen level was decreased further to 0.2% (Supplementary Figure 2) and in pulmonary artery smooth muscle cells (PASMCs; Supplementary Figure 3).
+Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (**Figure \@ref(fig:f1)I-J, Supplementary Figures 1-3**).
+Flux calculations incorporated the changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@RN1523] (**Supplementary Figure 1**).
+Surprisingly, while glucose uptake was modestly increased in hypoxia, lactate efflux was decreased (**Figure \@ref(fig:f1)I**) despite activation of the HIF-1 transcriptional program.
+Similar findings were observed when the ambient oxygen level was decreased further to 0.2% (Supplementary Figure 2) and in pulmonary artery smooth muscle cells (PASMCs; **Supplementary Figure 3**).
 In addition to lactate, the absolute extracellular fluxes of pyruvate and amino acids were generally decreased in hypoxia, including a marked decrease in glutamine uptake in LFs.
@@
 Given that hypoxia did not increase glucose and lactate fluxes as expected from the associated increases in glycolytic gene expression, we next assessed the capacity of HIF-1α to augment glycolysis in LFs.
-Cells were treated with BAY to stabilize HIF-1α under 21% oxygen conditions (Figure \@ref(fig:f2).
-Similar to hypoxia, BAY decreased cell growth rate (Figure \@ref(fig:f2)A-B) and activated the HIF-1 transcriptional program (Figure \@ref(fig:f2)C-G).
+Cells were treated with BAY to stabilize HIF-1α under 21% oxygen conditions (**Figure \@ref(fig:f2)**).
+Similar to hypoxia, BAY decreased cell growth rate (**Figure \@ref(fig:f2)A-B**) and activated the HIF-1 transcriptional program (**Figure \@ref(fig:f2)C-G**).
 Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1 target gene transcription and protein expression.
-In normoxia, this transcriptional program was associated with substantially increased glucose uptake and lactate efflux (Figure \@ref(fig:f2)H).
-Comparatively modest effects of BAY on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (Figure \@ref(fig:f2)J).
+In normoxia, this transcriptional program was associated with substantially increased glucose uptake and lactate efflux (**Figure \@ref(fig:f2)H**).
+Comparatively modest effects of BAY on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (**Figure \@ref(fig:f2)J**).
 
@@
 
-To validate our findings from extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate of label incorporation and trace its incorporation into key carbon utilization pathways (Figure \@ref(fig:f3) and Supplementary Figures 4-5).
-Consistent with our extracellular flux measurements, we found that the rate of label incorporation from [U-^13^C~6~] glucose into pyruvate was similar across 21%, 0.5%, and DMSO-treated conditions, but markedly increased with BAY treatment (\@ref(fig:f3)A-B).
+To validate our findings from extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate of label incorporation and trace its incorporation into key carbon utilization pathways (**Figure \@ref(fig:f3), Supplementary Figures 4-5**).
+Consistent with our extracellular flux measurements, we found that the rate of label incorporation from [U-^13^C~6~] glucose into pyruvate was similar across 21%, 0.5%, and DMSO-treated conditions, but markedly increased with BAY treatment (**Figure \@ref(fig:f3)A-B**).
 
@@
 Beyond this observation, the labeling patterns in hypoxia- and BAY-treated cells were similar to their respective controls, arguing against marked metabolic reprogramming in response to prolyl hydroxylase inhibition by either hypoxia or BAY.
-These findings were similar in PASMCs (Supplementary Figure 5) where stable isotope incorporation was attenuated in hypoxia and the overall labeling pattern was quite similar in both conditions.
+These findings were similar in PASMCs (**Supplementary Figure 5**) where stable isotope incorporation was attenuated in hypoxia and the overall labeling pattern was quite similar in both conditions.
 Compared to previous studies of metabolic flux in cancer cells [@RN2395; @RN1447], no marked increase in M5-labeled citrate from [U-^13^C~5~] glutamine was observed with hypoxia treatment, indicating no increase in reductive carboxylation for lipid synthesis.
@@
 To clarify changes in intracellular metabolite fluxes, we next generated metabolic flux models incorporating the extracellular flux measurements and stable isotope tracing data described above.
-Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (Figure \@ref(fig:f3A), Supplementary Figure 6).
-Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by INCA [@RN1522; @RN2501; @RN964] (Figure \@ref(fig:f4), Supplementary Figure 6, and Supplementary Tables 1-3).
+Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (**Figure \@ref(fig:f3)A, Supplementary Figure 6**).
+Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by INCA [@RN1522; @RN2501; @RN964] (**Figure \@ref(fig:f4), Supplementary Figure 6, Supplementary Tables 1-3**).
 Overall, proliferating LFs demonstrated high rates of glucose uptake and glycolysis.
-Approximately 10% of cytoplasmic pyruvate enters the TCA cycle with the balance converted to lactate (Supplementary Figure 6C).
-In hypoxia, significant reductions in glycolysis, the TCA cycle, and amino acid metabolism were observed (Figure \@ref(fig:f4)A) with a significant increase in pentose phosphate pathway flux.
+Approximately 10% of cytoplasmic pyruvate enters the TCA cycle with the balance converted to lactate (**Supplementary Figure 6C**).
+In hypoxia, significant reductions in glycolysis, the TCA cycle, and amino acid metabolism were observed (**Figure \@ref(fig:f4)A**) with a significant increase in pentose phosphate pathway flux.
+Similar findings were observed in PASMC flux maps (**Supplementary Figure 6, Supplementary Table 3**).
 
@@
 
-By comparison, HIF-1 activation by BAY in 21% oxygen increased glycolysis and lactate fermentation by ~50% (Figure \@ref(fig:f4)B).
-BAY treatment decreased cell proliferation rate (Figure \@ref(fig:f2)B), indicating that, unlike hypoxia, BAY treatment uncouples cell proliferation and metabolic flux.
+By comparison, HIF-1 activation by BAY in 21% oxygen increased glycolysis and lactate fermentation by ~50% (**Figure \@ref(fig:f4)B**).
+BAY treatment decreased cell proliferation rate (**Figure \@ref(fig:f2)B**), indicating that, unlike hypoxia, BAY treatment uncouples cell proliferation and metabolic flux.
 Similar to hypoxia, decreases in serine and glutamine incorporation were observed.
@@
 
-Although the metabolite exchange fluxes for bidirectional reactions tend to be poorly resolved, two observations are worth highlighting (Supplementary Tables 1 and 2).
-First, consistent with the stable isotope tracing results, the rate of reductive carboxylation through reversible flux by isocitrate dehydrogenase is low (~4 fmol/cell/h), is unchanged by hypoxia, and is modestly increased by BAY treatment.
+Although the metabolite exchange fluxes for bidirectional reactions tend to be poorly resolved, two observations are worth highlighting (Supplementary Tables 1-3).
+First, consistent with the stable isotope tracing results, the rate of reductive carboxylation through reversible flux by isocitrate dehydrogenase is low (~4 fmol/cell/h), is unchanged by hypoxia, and is modestly increased by BAY treatment and by hypoxia treatment in PASMCs.
 
-Second, hypoxia and BAY treatment are associated with a marked increase in the lactate transport exchange flux in LFs from ~ 0 to 1,520 and 711 fmol/cell/h in 0.5% oxygen and BAY treatment conditions, respectively.
+Second, hypoxia and BAY treatment are associated with a marked increase in the lactate transport exchange flux in LFs from ~ 0 to 1,520 and 711 fmol/cell/h in 0.5% oxygen and BAY treatment conditions, respectively, with similar results in PASMCs exposed to hypoxia.
 Since the net lactate transport flux is secretion, this observation suggests increased lactate uptake with hypoxia or BAY treatment.
 This may be consistent with the HIF-driven increased expression of the reversible lactate transporter MCT4 [@RN366].
-To investigate this hypothesis, LFs and PASMCs were treated with [U-^13^C~3~] lactate (2 mM) and ^13^C incorporation into intracellular metabolites was analyzed by LC-MS (Figure \@ref(fig:f5), Supplementary Figure 5).
-Here, we observed increased labeling of TCA metabolites citrate (CIT), 2-oxoglutrate (2OG), malate (MAL), and aspartate (ASP) following hypoxia or BAY treatment (Figure \@ref(fig:f5) and Supplementary Figure 4).
+To investigate this hypothesis, LFs and PASMCs were treated with [U-^13^C~3~] lactate (2 mM) and ^13^C incorporation into intracellular metabolites was analyzed by LC-MS (**Figure \@ref(fig:f5), Supplementary Figure 5**).
+Here, we observed increased labeling of TCA metabolites citrate (CIT), 2-oxoglutrate (2OG), malate (MAL), and aspartate (ASP) following hypoxia or BAY treatment (**Figure \@ref(fig:f5), Supplementary Figure 4**).
 Interestingly, a similar increase in labeling was not observed in PASMCs (Supplementary Figure 5).
@@
 
-To reconcile the differential effects of prolyl hydroxylase inhibition by hypoxia and BAY, we next addressed whether hypoxia could suppress the effects of BAY on glucose and lactate fluxes (Figure \@ref(fig:f6)A-C).
+To reconcile the differential effects of prolyl hydroxylase inhibition by hypoxia and BAY, we next addressed whether hypoxia could suppress the effects of BAY on glucose and lactate fluxes (**Figure \@ref(fig:f6)A-C**).
 LFs cultured in standard growth medium were treated with BAY and placed in either 21% or 0.5% oxygen.
@@
 
-To investigate these metabolic differences further, we performed metabolomic profiling of LFs treated for 72 h with hypoxia or BAY separately or in combination (Figure \@ref(fig:f6)D-F).
-Principle component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen (Figure \@ref(fig:f6)D).
+To investigate these metabolic differences further, we performed metabolomic profiling of LFs treated for 72 h with hypoxia or BAY separately or in combination (**Figure \@ref(fig:f6)D-F**).
+Principle component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen (**Figure \@ref(fig:f6)D**).
 These hypoxia-treated cells were well-segregated from BAY-treated cells, again highlighting differential effects of hypoxic and pharmacologic PHD inhibition.
-Linear modeling identified those metabolites that were differentially affected by hypoxia and BAY treatment (*i.e.*, those metabolites with a significant interaction term in the linear model described by treatment × oxygen) (Figure \@ref(fig:f6)E).
-Several patterns emerged from this analysis (Figure \@ref(fig:f6)F).
+Linear modeling identified those metabolites that were differentially affected by hypoxia and BAY treatment (*i.e.*, those metabolites with a significant interaction term in the linear model described by treatment × oxygen) (**Figure \@ref(fig:f6)E**).
+Several patterns emerged from this analysis (**Figure \@ref(fig:f6)F**).
 Interestingly, the hypoxia-regulated metabolite, 2-hydroxyglutarate, increased following BAY treatment, but was unaffected by hypoxia treatment in these proliferating LFs [@RN1603].
-The tricarboxylic acid cycle metabolite aconitate was decreased in hypoxia, but increased by BAY treatment while the opposite effect was observed for hydroxyproline and taurine.
+The tricarboxylic acid (TCA) cycle metabolite aconitate was decreased in hypoxia, but increased by BAY treatment while the opposite effect was observed for hydroxyproline and taurine.
 Glycolytic intermediate glyceraldehyde 3-phosophate (GAP) was decreased by both hypoxia and BAY, while hypoxia reversed the BAY-mediated decrease in γ-aminobutyric acid (GABA).
-A metabolite set enrichment analysis revealed the tricarboxlyic acid cycle to be the most
+A metabolite set enrichment analysis (**Figure \@ref(fig:f6)G-H**) revealed the tricarboxlyic acid cycle to be the most enriched KEGG metabolite set.
+Leading edge analysis demonstrates the negative enrichment scores associated with all of the TCA metabolites detected by our platform.
+This result suggests a more modest impact of BAY treatment on the TCA cycle than hypoxia, as suggested by our metabolic flux models were hypoxia resulted in a 1.5-2-fold reduction of TCA flux compared to a 1.1-1.5-fold reduction with BAY treatment in normoxia (**Figure \@ref(fig:f4)**).
 
-One of the key differences associated with
+In addition to these differential effects on polar metabolite levels, perhaps the most significant contrast between hypoxia and BAY treatment is the impact of hypoxia on cellular redox state.
+As oxygen deprivation causes reductive stress [@RN2409], we next measured the impact of these treatments on intracellular NAD(H) as a surrogate for cellular redox state (**Figure \@ref(fig:f6)I-K**).
+As expected, hypoxia increased the NADH/NAD^+^ ratio, driven primarily by a decrease in intracellular NAD^+^.
+Interestingly, while BAY treatment increased the levels of NADH, a concomitant increase in NAD^+^ resulted in preservation of the NADH/NAD^+^ ratio.
+As NADH accumulation is a putative inhibitor of glycolytic flux [@RN2210], this may be one mechanism by which glycolytic flux is decreased in hypoxia but not following BAY treatment.
+
+## Transcriptomic analysis identifies putative regulators of cellular metabolism in hypoxia
+
+To identify the upstream regulators of the observed metabolic changes, we next performed transcriptomic analysis of LFs treated with hypoxia or BAY, separately or together (**Figure \@ref(fig:f7)**).
+Principle component analysis again demonstrates clear separation among the four treatment groups (**Figure \@ref(fig:f7)A**), with a closer relationship between hypoxia-treated cells compared to hypoxia and 21% BAY-treated cells.
+To identify those transcripts that were differentially affected by hypoxia compared to BAY treatment, we again performed linear modeling to identify transcripts with a significant interaction term in the model described by treatment × oxygen (**Figure \@ref(fig:f7)B**).
+With this analysis, we again found several patterns of transcriptional changes (**Figure \@ref(fig:f7)C**).
+For example, EPAS1, the gene encoding the HIF-2α protein, was markedly decreased by hypoxia, compared to a modest reduction following BAY treatment.
+A similar, but opposite effect on RBM3 (RNA-binding protein 3) expression was observed.
+RBM3 supports the viability and proliferation of neural stem cells in hypoxia [@RN3111].
+Histone deacetylase 9 (HDAC9) and prolyl 4-hydroxylase subunit α2 (P4HA2) demonstrate opposing effects following hypoxia and BAY treatment.
+BAY decreased HDAC9 expression and increased P4HA2 expression.
+Together, this subset of transcriptional changes again illustrates important differences between hypoxia and HIF stabilization in normoxia.
+Indeed, an enrichment analysis of these transcripts using the Molecular Signatures Database "Hallmark" gene sets [@RN3112] actually demonstrated greater enrichment of the "hypoxia" data set among differentially expressed transcripts following BAY treatment (**Figure \@ref(fig:f7)D**).
+Consistent with our previous findings, the gene set for glycolysis was also more enriched following BAY treatment than hypoxia treatment.
+The most enriched gene sets associated with hypoxia included pro-proliferative E2F targets and G2/M checkpoint proteins as well as MYC target proteins.
+These findings were reinforced by a transcription factor enrichment analysis (**Figure \@ref(fig:f7)E**), again identifying enrichment of MYC target proteins associated with hypoxia.
+Consistent with these bioinformatic results, immunoblot demonstrates up-regulation of MYC protein in hypoxia-treated cells, but not in BAY-treatment alone (**Figure \@ref(fig:f7)F**).
+These data provide a starting point for interrogating the relative hierarchy of transcription factor regulation of gene expression by hypoxia in proliferating primary cells and suggest several potential mechanisms that may contribute to the uncoupling of glycolytic gene transcription and glycolytic flux in the proliferating primary cells studied here.
 
@@
 
+In this work, we used ^13^C metabolic flux analysis to identify hypoxia-mediated metabolic changes in proliferating human primary cells.
+Our principal finding was that hypoxia reduced, rather than increased, carbon flux through glycolysis and lactate fermentation pathways despite robust activation of the HIF transcriptional program and up-regulation of glycolytic genes.
+Certainly the LFs studied here are capable of augmenting glycolysis in response to HIF stabilization in hypoxia, as demonstrated by experiments with the PHD inhibitor BAY; however, these effects are completely attenuated when BAY-treated cells are cultured in hypoxia.
+Together, these findings suggest that changes in enzyme levels alone are insufficient to alter metabolic flux in hypoxia and point to the importance of regulatory mechanisms that supersede the effects of HIF-dependent gene transcription.
+
+The identification of these mechanisms is of critical importance for our understanding of primary cell adaptation to hypoxia.
+In particular, our transcriptomic analyses suggest that hypoxia is associated with a pro-proliferative pattern of gene expression that likely facilitates the response of these mesenchymal cells to tissue injury and these pathways may have implications for cancer cell biology.
+While the role of HIFs in the hypoxia response has been extensively studied, relatively less is known about HIF-independent features of the hypoxia response.
+Certainly cells express several oxygen-dependent enzymes in addition to PHD whose activities may be impacted in hypoxia but not by HIF stabilization.
+For example, PHD is one of many α-ketoglutarate-dependent dioxygenase enzymes that rely on molecular oxygen for their catalytic activity [@RN2611].
+Jumonji-C (JmjC) domain containing histone demethylases are other prominent members of this family whose inhibition by hypoxia has been shown to cause rapid and HIF-independent induction of histone methylation [@RN3117].
+Similarly, a recently described cysteamine dioxygenase has been shown to mediate the oxygen-dependent degradation of Regulators of G protein Signaling 4 and 5 and IL-32 [@RN1384].
+In addition to dioxygenase enzymes, electron transport chain dysfunction resulting from impaired Complex IV activity leads to increased reactive oxygen species (ROS) production in hypoxia [@RN295].
+Mitochondrial ROS increase the half-lives of several mRNAs in hypoxia, including MYC as we observe in this work, independent of HIF stabilization [@RN742].
+Finally, hypoxia imposes a reductive stress on cells associated with an increase in the NADH/NAD^+^ ratio secondary to impaired electron transport [@RN2618; @RN2619].
+NADH accumulation may slow glycolysis *via* feedback inhibition of GAPDH [@RN2210].
+Any of these molecular mechanisms may contribute to uncoupling glycolytic enzyme expression from glycolytic flux as observed in the experiments described here.
+
+These findings raise important questions regarding the cell-autonomous role of HIFs in the hypoxia response.
+Certainly, on an organismal level, HIFs drive expression of angiogenic and erythropoietic factors to increase oxygen delivery to hypoxic tissues.
+Within individual cells, HIF-1α seems to be important for mitigating the adverse effects of reactive oxygen species (ROS) formation by dysfunctional electron transport in the mitochondria.
+Indeed, hypoxia increased oxygen consumption and ROS production in HIF-1α-null mouse embryonic fibroblasts (MEFs), which was associated with increased cell death [@RN2525].
+Interestingly, these cells also had increased ATP levels compared to wild type, suggesting that mitochondrial function was adequate under 1% oxygen culture conditions to support oxidative phosphorylation and meet the energy needs of the cells.
+Given the prominence of HIFs in mediating the transcriptional response to hypoxia, it is somewhat surprising that none of PHD, HIFs, or their downstream targets were found to be selectively essential as a function of oxygen tension in a genome-wide CRISPR growth screen of cells cultured in normoxia and hypoxia [@RN953].
+
+Our finding that hypoxia was associated with decreased glycolysis and lactate fermentation was unexpected.
+Several aspects of our experimental design may have contributed to this finding.
+First, our goal was to understand how metabolic reprogramming may support cell proliferation in hypoxia.
+Thus, we measured metabolite fluxes in cells during the exponential growth phase accounting for cell growth rate, metabolite degradation rates, and medium evaporation with multiple measurements over a 72 h time course.
+Often, cells are studied near confluence, where metabolic contributions to biomass production are less and the rate of glycolysis in hypoxia may be higher.
+Second, we began our experimental treatments 24 h prior to collecting samples to ensure that the hypoxia metabolic program was established prior to labeling.
+Similar studies [@RN1447; @RN713] typically placed cells into hypoxia at the time of labeling.
+Third, and perhaps most importantly, these flux determinations were performed in human primary cell cultures rather than immortalized cell lines.
+
+In summary, in this metabolic flux analysis of proliferating human primary cells *in vitro*, we have demonstrated that hypoxia uncouples an increase in HIF-dependent glycolytic gene transcription from glycolytic flux.
+Indeed, the degree of metabolic reprogramming in hypoxia was modest and suggests close coupling between proliferation and metabolism.
+In light of our findings, additional studies are warranted to understand the role of HIFs in mediating the metabolic response to hypoxia in primary cells.
+Moreover, these data strongly caution investigators against drawing conclusions about metabolite flux from measures of gene transcription alone.
+Further investigations of metabolic flux in primary cell cultures in hypoxia are warranted to identify the key regulators of metabolism in hypoxia and to clarify the contributions of HIF proteins to hypoxic metabolic reprogramming.
+
+\newpage
+
 # Acknowledgements
+
+This work was supported by grants from the NIH (K08HL128802), American Lung Association, Pulmonary Hypertension Association, and the American Thoracic Society Foundation to W.M.O and from the NIH (U01HG007690, U01HL108630, U54HL119145) and the American Heart Association (D700382, CV-19) to J.L.
 
@@
 
+W.M.O.
+conceived and designed the analysis.
+C.A.C., B.A.O., D.R.Z., S.M., and W.M.O.
+collected the data.
+J.D.Y.
+and W.M.O.
+contributed data or analysis tools.
+W.M.O.
+performed the analysis.
+W.M.O.
+drafted the manuscript.
+All authors participated in interpreting the results and revising the manuscript.
+All authors approve the final submission.
+
 # Declaration of Interests
+
+The authors declare no competing interests.
 
@@
 (**F**) Representative metabolites from (E) reveal different patterns of metabolic effects.
-(**G**-**I**) Intracellular NAD^+^, NADH, and their ratio were determined by enzymatic assay (n = 5).
+(**G**) Metabolite set enrichment analysis of metabolites from (E).
+KEGG pathways with unadjusted p-values < 0.05 are shown.
+(**H**) Leading edge analysis of the most enriched KEGG metabolite set from (G).
+Negative values indicate relative enrichment associated with BAY treatment compared to hypoxia treatment (as in (E)).
+Abbreviations: PYR, pyruvate; SUC, succinate; PEP, phosphoenolpyruvate; CIT, citrate; AKG, α-ketoglutarate; MAL, malate; ACO, aconitate; FUM, fumarate.
+(**I**-**K**) Intracellular NAD^+^, NADH, and their ratio were determined by enzymatic assay (n = 5).
 Black * indicate a significant effect of hypoxia, colored * indicate a significant effect of treatment within a given oxygen tension as indicated by the color.
-Comparisons for fluxes (A-C) and NAD(H) measurements (G-I) were performed using a mixed-effects linear model with replicate as a random effect.
+Comparisons for fluxes (A-C) and NAD(H) measurements (I-K) were performed using a mixed-effects linear model with replicate as a random effect.
 Adjusted p-values for the indicated comparisons were determined using Tukey's *post hoc* test.
@@
 (**C**) Representative transcripts from (B).
-(**D**) Gene set enrichment analysis of trasncripts from (B).
+(**D**) Gene set enrichment analysis of transcripts from (B).
 (**E**) Transcription factor enrichment analysis suggests mechanisms for differential regulation of gene expression following hypoxia or BAY treatment.
@@
 Figure 7: (ref:f7)
-
-\newpage
-
-# Tables
 
@@
 Differentially regulated metabolites were identified using limma [@limma].
+Metabolite set enrichment analysis was performed using the fgsea package with metabolite pathways downloaded using multiGSEA [@fgsea; @multiGSEA].
 

### 2021-05-07 inst/manuscript/supplement.Rmd
finish up rough draft of manuscript

@@
   bay <- targets::tar_read(lf_bay_table)
+  pasmc <- targets::tar_read(pasmc_hypoxia_table)
 })
@@
 ```
+
+\newpage
+
+```{r t3, results = "asis"}
+pasmc %>% 
+  flextable::set_caption(
+    "PASMC fluxes in 21% and 0.5% oxygen", 
+    autonum = officer::run_autonum()
+  ) %>% 
+  flextable::flextable_to_rmd()
+```
+
+\newpage
+
+# Figures
+
+# Figure legends
+
+(ref:s1) **Supporting data for extracellular flux calculations.** (**A**) Standard curves were generated to interpolate cell counts from total DNA by seeding lung fibroblasts (LF) and pulmonary artery smooth muscle cells (PASMC) at the indicated densities in basal medium.
+Data are mean ± SEM of three biological replicates.
+(**B**) Total DNA measurements were compared to direct cell count over the experimental time course.
+Cell counts and total DNA were obtained from the same sample wells.
+The slopes of the best-fit lines for normoxia and hypoxia samples were not different.
+(**C**) Predicted well volumes were estimated from the change in culture plate mass over the time course of the experiment.
+Evaporation rates were different depending on the treatment.
+Although the mean evaporation rate is depicted, experiment-specific evaporation rates were used to calculate fluxes for each experiment.
+(**D**) Metabolite accumulation (positive values) and degradation (negative values) rates.
+Data are mean ± SEM of 3-8 biological replicates.
+
+(ref:s2) **Extracellular flux measurements in 0.2% oxygen.** (**A**) Growth curves of LFs cultured in 21% and 0.2% oxygen (n = 4).
+(**B**) Growth rates were determined by linear fitting of log-transformed growth curves.
+(**C**) Representative immunoblot of LF protein lysates collected at the indicated times.
+(**D**, **E**) Relative change in HIF-1α (D) and LDHA (E) protein levels compared to 21% oxygen time 0.
+(**F**, **G**) Relative changes in GLUT1 (F) and LDHA (G) mRNA levels compared to 21% oxygen time 0.
+(**H**, **I**) Extracellular fluxes of the indicated metabolites (n = 8).
+Data are mean ± SEM.
+Comparisons were performed using Student's paired *t*-test and p < 0.05 indicated by *.
+Flux probability values were not corrected for multiple comparisons.
+Abbreviations as noted in the text.
+
+(ref:s3) **Extracellular flux measurements in pulmonary artery smooth muscle cells in 0.5% oxygen.** (**A**) Growth curves of PASMCs cultured in 21% and 0.5% oxygen (n = 8).
+(**B**) Growth rates were determined by linear fitting of log-transformed growth curves.
+(**C**) Representative immunoblot of PASMC protein lysates collected at the indicated times.
+(**D**, **E**) Relative change in HIF-1α (D) and LDHA (E) protein levels compared to 21% oxygen time 0.
+(**F**, **G**) Relative changes in GLUT1 (F) and LDHA (G) mRNA levels compared to 21% oxygen time 0.
+(**H**, **I**) Extracellular fluxes of the indicated metabolites (n = 8).
+Data are mean ± SEM.
+Comparisons were performed using Student's paired *t*-test and p < 0.05 indicated by *.
+Flux probability values were not corrected for multiple comparisons.
+Abbreviations as noted in the text.
+
+(ref:s4) **Mass isotopomer distributions after 72 h of labeling in lung fibroblasts.** LFs were labeled with the indicated tracers and intracellular metabolites were analyzed by LC-MS.
+Mass isotopomer distributions were adjusted for natural abundance.
+Data are the mean ± SEM of 4 biological replicates.
+
+(ref:s5) **Mass isotopomer distributions after 72 h of labeling in pulmonary artery smooth muscle cells.** PASMCs were labeled with the indicated tracers and intracellular metabolites were analyzed by LC-MS.
+Mass isotopomer distributions were adjusted for natural abundance.
+Data are the mean ± SEM of 4 biological replicates.
+
+(ref:s6) **Isotopically non-stationary metabolic flux analysis.** (**A**, **B**) LFs were cultured in 21% (A) or 0.5% (B) oxygen and labeled with the indicated tracers and intracellular metabolites were analyzed by LC-MS (PYR, pyruvate; CIT, citrate; MAL, malate).
+Mass isotopomer distributions were adjusted for natural abundance.
+Data are the mean ± SEM of 4 biological replicates.
+(**C**) Metabolic flux model of LF metabolism in 21% oxygen.
+(**D**) LF fluxes were normalized to cell growth rate.
+Graph depicts the ratio of normalized metabolic fluxes in LFs cultured in 0.5% oxygen compared to 21% oxygen control.
+Fluxes with non-overlapping confidence intervals are highlighted to indicate significant changes.
+(**E**) Metabolic flux model of PASMC metabolism in 21% oxygen.
+(**F**) Ratio of metabolic fluxes in 0.5% oxygen compared to 21% oxygen in PASMCs.
+
+Supplementary Figure 1: (ref:s1)
+
+Supplementary Figure 2: (ref:s2)
+
+Supplementary Figure 3: (ref:s3)
+
+Supplementary Figure 4: (ref:s4)
+
+Supplementary Figure 5: (ref:s5)
+
+Supplementary Figure 6: (ref:s6)
+
+\newpage
+
+# Figures
+
+```{r s1, fig.cap = "(ref:s1)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/s1.pdf",
+    package = "Copeland.2021.hypoxia.flux"
+  )
+)
+```
+
+\newpage
+
+```{r s2, fig.cap = "(ref:s2)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/s2.pdf",
+    package = "Copeland.2021.hypoxia.flux"
+  )
+)
+```
+
+\newpage
+
+```{r s3, fig.cap = "(ref:s3)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/s3.pdf",
+    package = "Copeland.2021.hypoxia.flux"
+  )
+)
+```
+
+\newpage
+
+```{r s4, fig.cap = "(ref:s4)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/s4.pdf",
+    package = "Copeland.2021.hypoxia.flux"
+  )
+)
+```
+
+\newpage
+
+```{r s5, fig.cap = "(ref:s5)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/s5.pdf",
+    package = "Copeland.2021.hypoxia.flux"
+  )
+)
+```
+
+\newpage
+
+```{r s6, fig.cap = "(ref:s6)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/s6.pdf",
+    package = "Copeland.2021.hypoxia.flux"
+  )
+)
+```

### 2021-05-07 _targets.R

@@ -859,16 +859,16 @@ list(
     readr::read_csv(node_file)
   ),
   tar_target(
-    hypoxia_graph,
-    make_graph(map_flux_differences, nodes, treat = "21%", normalizer = "none")
+    lf_hypoxia_graph,
+    make_graph(map_flux_differences, nodes, cell = "lf", treat = "21%", normalizer = "none")
   ),
   tar_target(
-    hypoxia_graph_ratio_plot,
-    plot_ratio_network(hypoxia_graph, "Hypoxia/Normoxia")
+    lf_hypoxia_graph_ratio_plot,
+    plot_ratio_network(lf_hypoxia_graph, "Hypoxia/Normoxia")
   ),
   tar_target(
     bay_graph,
-    make_graph(map_flux_differences, nodes, treat = "DMSO", normalizer = "none")
+    make_graph(map_flux_differences, nodes, cell = "lf", treat = "DMSO", normalizer = "none")
   ),
   tar_target(
     bay_graph_ratio_plot,
@@ -876,7 +876,7 @@ list(
   ),
   tar_target(
     m4,
-    arrange_m4(hypoxia_graph_ratio_plot, bay_graph_ratio_plot)
+    arrange_m4(lf_hypoxia_graph_ratio_plot, bay_graph_ratio_plot)
   ),
   tar_target(
     m4_figure,
@@ -910,19 +910,31 @@ list(
   ),
   tar_target(
     s6c,
-    plot_normoxia_network(hypoxia_graph)
+    plot_normoxia_network(lf_hypoxia_graph, "lf")
   ),
   tar_target(
     hypoxia_growth_graph,
-    make_graph(map_flux_differences, nodes, treat = "0.5%", normalizer = "growth")
+    make_graph(map_flux_differences, nodes, cell = "lf", treat = "0.5%", normalizer = "growth")
   ),
   tar_target(
     s6d,
     plot_ratio_network(hypoxia_growth_graph, "Hypoxia/Normoxia\nGrowth Rate Normalized")
   ),
   tar_target(
+    pasmc_hypoxia_graph,
+    make_graph(map_flux_differences, nodes, cell = "pasmc", treat = "21%", normalizer = "none")
+  ),
+  tar_target(
+    s6e,
[…]

### 2021-05-07 inst/analysis/modeling/pasmc_21.m

@@ -111,7 +111,7 @@ end
 clear a b c y n x z time compound label;
 
 % assign sem for mid (can repeat loop above to assign actual sem values)
-sem_value = 0.013;
+sem_value = 0.02;
 for i=1:length(d)
     for j=1:length(d{i})
         d{i}{d{i}.id{j}}.mdvs.std = sem_value * ones(length(d{i}{d{i}.id{j}}.mdvs.val),1);
@@ -182,10 +182,10 @@ m.options.sim_sens = false;
 m.options.sim_more = false;
 m.options.sim_na = false;
 m.options.fit_reinit = true;
-m.options.fit_starts = 10;
+m.options.fit_starts = 50;
 m.options.fit_reltol = 0.001;
 m.options.fit_tau = 1e-6;
-m.options.fit_nudge = 6;
+m.options.fit_nudge = 8;
 m.options.int_tspan = [0 12 24 36 48];
 m.options.int_maxstep = Inf;
 m.options.hpc_on = 1;

### 2021-05-07 inst/src/analysis.R

@@ -1077,16 +1077,16 @@ assemble_flux_differences <- function(map_fluxes) {
   growth_norm <- normalize_fluxes(map_fluxes, "BIOMASS")
 
   tibble::tibble(
-    map_fluxes = list(map_fluxes, map_fluxes, glc_norm, glc_norm, growth_norm, growth_norm),
-    cell = rep(c("lf", "lf"), 3),
-    control = rep(c("21%", "DMSO"), 3),
-    experiment = rep(c("0.5%", "BAY"), 3)
+    map_fluxes = list(map_fluxes, map_fluxes, map_fluxes, glc_norm, glc_norm, glc_norm, growth_norm, growth_norm, growth_norm),
+    cell = rep(c("lf", "lf", "pasmc"), 3),
+    control = rep(c("21%", "DMSO", "21%"), 3),
+    experiment = rep(c("0.5%", "BAY", "0.5%"), 3)
   ) %>%
     purrr::pmap_dfr(calculate_flux_differences, .id = "normalization") %>%
     dplyr::mutate(normalization = dplyr::case_when(
-      normalization %in% 1:2 ~ "none",
-      normalization %in% 3:4 ~ "glucose",
-      normalization %in% 5:6 ~ "growth"
+      normalization %in% 1:3 ~ "none",
+      normalization %in% 4:6 ~ "glucose",
+      normalization %in% 7:9 ~ "growth"
     ))
 
 }
@@ -1145,7 +1145,7 @@ parse_eq <- function(df) {
 
 # make_graph --------------------------------------------------------------
 
-make_graph <- function(map_flux_differences, nodes, treat, normalizer) {
+make_graph <- function(map_flux_differences, cell, nodes, treat, normalizer) {
 
   edges <-
     map_flux_differences %>%
@@ -1158,7 +1158,7 @@ make_graph <- function(map_flux_differences, nodes, treat, normalizer) {
       names_to = c(".value", NA),
       names_sep = "_"
     ) %>%
-    dplyr::filter(normalization == normalizer & treatment == treat) %>%
+    dplyr::filter(cell_type == cell & normalization == normalizer & treatment == treat) %>%
     parse_eq() %>%
     dplyr::select(
       cell_type,
@@ -1289,8 +1289,8 @@ plot_ratio_network <- function(graph, caption) {
       ) +
     ggplot2::labs(
       x = NULL,
-      y = NULL
-      # title = caption
+      y = NULL,
+      title = caption
     ) +
[…]

### 2021-05-07 inst/src/figures.R

@@ -146,13 +146,15 @@ arrange_m4 <- function(hypoxia_graph_ratio_plot, bay_graph_ratio_plot) {
 
 # arrange_s6 --------------------------------------------------------------
 
-arrange_s6 <- function(a, b, c, d) {
+arrange_s6 <- function(a, b, c, d, e, f) {
   layout <- "
     ab
     cd
-    cd"
+    cd
+    ef
+    ef"
 
-  a + b + c + d +
+  a + b + c + d + e + f +
     theme_patchwork(
       design = layout,
       widths = unit(5, "in"),
@@ -266,6 +268,7 @@ create_resources <- function() {
 # format_flux_table -------------------------------------------------------
 
 format_flux_table <- function(
+  flux_differences,
   cell = c("lf", "pasmc"),
   experiment = c("0.5%", "BAY"),
   ssr_ctl = NULL,

### 2021-05-07 inst/src/metabolomics.R

@@ -405,23 +405,33 @@ plot_metab_volcano <- function(results) {
     ) +
     ggrepel::geom_text_repel(
       data = left,
-      ggplot2::aes(label = metabolite),
+      ggplot2::aes(
+        label = metabolite,
+        color = metabolite %in% c("2-hydroxyglutarate","aconitate", "glyceraldehyde 3-phosphate", "hydroxyproline", "malate", "taurine")
+        ),
       size = 5/ggplot2::.pt,
       max.overlaps = 20,
       segment.size = 0.1,
       nudge_x = -4,
+      segment.color = "black",
       direction = "y",
-      family = "Calibri"
+      family = "Calibri",
+      show.legend = FALSE
     ) +
     ggrepel::geom_text_repel(
       data = right,
-      ggplot2::aes(label = metabolite),
+      ggplot2::aes(
+        label = metabolite,
+        color = metabolite %in% c("2-hydroxyglutarate","aconitate", "glyceraldehyde 3-phosphate", "hydroxyproline", "malate", "taurine")
+        ),
       size = 5/ggplot2::.pt,
       max.overlaps = 20,
       segment.size = 0.1,
+      segment.color = "black",
       nudge_x = 4.5,
       direction = "y",
-      family = "Calibri"
+      family = "Calibri",
+      show.legend = FALSE
     ) +
     ggplot2::geom_point(
       data = subset(results, adj.P.Val > 0.05),
@@ -441,6 +451,7 @@ plot_metab_volcano <- function(results) {
       color = "white",
       fill = clrs[[4]]
     ) +
+    ggplot2::scale_color_manual(values = c("black", "darkred")) +
     ggplot2::scale_y_continuous(trans = reverselog_trans(10)) +
     ggplot2::scale_x_continuous(
       breaks = seq(-4, 4, 2),
@@ -630,7 +641,7 @@ plot_leading_edge <- function(tt, pathway) {
       x = x,
       y = y
     ) +
-    ggplot2::geom_line(color = "green") +
+    ggplot2::geom_line(color = clrs[[3]]) +
     ggplot2::geom_hline(
       yintercept = 0,
       colour = "black",
@@ -652,7 +663,7 @@ plot_leading_edge <- function(tt, pathway) {
         label = metabolite
       ),
       angle = 90,
-      size = 4/ggplot2::.pt,
+      size = 5/ggplot2::.pt,
       max.overlaps = 20,
[…]

### 2021-05-07 inst/src/rnaseq.R

@@ -215,29 +215,40 @@ plot_rnaseq_volcano <- function(results) {
     ) +
     ggrepel::geom_text_repel(
       data = left,
-      ggplot2::aes(label = symbol),
+      ggplot2::aes(
+        label = symbol,
+        color = symbol %in% c("EPAS1","HDAC9", "P4HA2", "RBM3")
+      ),
       size = 5/ggplot2::.pt,
       max.overlaps = 20,
       segment.size = 0.1,
       nudge_x = -4,
       direction = "y",
-      family = "Calibri"
+      family = "Calibri",
+      segment.color = "black",
+      show.legend = FALSE
     ) +
     ggrepel::geom_text_repel(
       data = right,
-      ggplot2::aes(label = symbol),
+      ggplot2::aes(
+        label = symbol,
+        color = symbol %in% c("EPAS1","HDAC9", "P4HA2", "RBM3")
+      ),
       size = 5/ggplot2::.pt,
       max.overlaps = 20,
       segment.size = 0.1,
       nudge_x = 4.5,
       direction = "y",
-      family = "Calibri"
+      segment.color = "black",
+      family = "Calibri",
+      show.legend = FALSE
     ) +
     ggplot2::geom_hex(
       bins = c(250, 15),
       show.legend = FALSE
     ) +
     ggplot2::scale_fill_viridis_c(trans = "log10") +
+    ggplot2::scale_color_manual(values = c("black", "darkred")) +
     ggplot2::scale_y_continuous(trans = reverselog_trans(10)) +
     ggplot2::scale_x_continuous(
       breaks = seq(-4, 4, 2),

### 2021-05-07 inst/manuscript/manuscript.Rmd
add manuscript and supplement to targets

@@
   bookdown::word_document2:
-    reference_docx: template.docx
+    reference_docx: !expr system.file("manuscript", "template.docx", package = "Copeland.2021.hypoxia.flux")
     df_print: kable
@@
   resources <- targets::tar_read(resources_table)
+  figs <- targets::tar_load(targets::matches("m\\d\\.pdf"))
 })

### 2021-05-07 inst/manuscript/supplement.Rmd
add manuscript and supplement to targets

@@
 output: 
-  # bookdown::pdf_document2:
-  #   # template: template.tex
-  #   keep_tex: false
-  #   df_print: kable
-  #   toc: false
-  #   number_sections: false
-  #   latex_engine: xelatex
-  #   pandoc_args:
-  #     - '--lua-filter=scholarly-metadata.lua'
-  #     - '--lua-filter=author-info-blocks.lua'
-  #     - '--lua-filter=pagebreak.lua'
-  #     - '--lua-filter=multiple-bibliographies.lua'
-
   bookdown::word_document2:
-    reference_docx: template.docx
+    reference_docx: !expr system.file("manuscript", "template.docx", package = "Copeland.2021.hypoxia.flux")
     df_print: kable
@@
   pasmc <- targets::tar_read(pasmc_hypoxia_table)
+  figs <- targets::tar_load(targets::matches("s\\d\\.pdf"))
 })

### 2021-05-07 _targets.R

@@ -1020,6 +1020,19 @@ list(
   tar_target(
     pasmc_hypoxia_table,
     format_flux_table(map_flux_differences, "pasmc", "0.5%", " SSR 575.6 [499.1-630.6] (95% CI, 563 DOF)", " SSR 521.3 [482.2-611.6] (95% CI, 545 DOF)")
+  ),
+
+  # write manuscript --------------------------------------------------------
+
+  tar_render(
+    manuscript,
+    path = path_to_manuscript("manuscript.Rmd"),
+    output_dir = path_to_manuscript("")
+  ),
+  tar_render(
+    supplement,
+    path = path_to_manuscript("supplement.Rmd"),
+    output_dir = path_to_manuscript("")
   )
 
 )

### 2021-05-10 inst/manuscript/manuscript.Rmd
revisions to rough draft

--- 2021-05-07 inst/manuscript/manuscript.Rmd
@@
 ---
-title: "Hypoxia uncouples HIF gene transcription and metabolic flux in proliferating primary cells"
+title: "Hypoxia uncouples HIF target gene transcription and glycolytic flux in proliferating primary cells"
 
@@
     institute: 1, 5
+- Kevin Leahy: 
+    institute: 1
 - Jamey D. Young: 
@@
 
-bibliography_main: library.bib
-bibliography_software: packages.bib
+bibliography: 
+- library.bib
+- packages.bib
 csl: cell-metabolism.csl
@@
       - '--lua-filter=pagebreak.lua'
-      - '--lua-filter=multiple-bibliographies.lua'
+      # - '--lua-filter=multiple-bibliographies.lua'
 
@@
 graphics: yes
-header-includes:
-- \usepackage{lineno}
-- \linenumbers
-- \setlength\linenumbersep{1cm}
-- \usepackage{setspace}
-- \doublespacing
-- \makeatletter\renewcommand*{\fps@figure}{htbp}\makeatother
-- \setlength{\parskip}{0.75\baselineskip plus 2pt minus 2pt}
-- \usepackage{float}
 ---
@@
   out.width = "50%", 
-  fig.align = "center", 
-  fig.pos = "H", 
   out.extra = ""
@@
   resources <- targets::tar_read(resources_table)
+  figs <- targets::tar_load(targets::matches("m\\d_figure"))
 })
@@
 Hypoxia is an important environmental stimulus that causes transcriptional and metabolic reprogramming in cells to facilitate their survival.
-Here, we performed stable isotope tracing and metabolic flux analyses of proliferating primary cells in hypoxia.
-Despite activation of the hypoxia-inducible factor (HIF) transcriptional program and up-regulation of glycolytic genes, glycolytic flux was decreased in hypoxic cells in our models.
-No evidence for increased glutaminolysis or reductive carboxylation was observed.
-While pharmacologic stabilization of HIF in normoxia with the prolyl hydroxylase inhibitor molidustat did increase glycolytic flux as expected, hypoxia abrogated this effect.
-Multi-omic profiling of cells treated with hypoxia or molidustat, seperately or together, reveals distinct molecular responses to to hypoxia and pharmacologic prolyl hydroxylase inhibition.
-Together, these data suggest that primary cell bioenergetic metabolism is closely coupled to cell proliferation rate and that other regulatory factors override the effects of HIF-dependent up-regulation of glycolytic gene expression on glycolytic flux.
+Here, we performed stable isotope tracing and metabolic flux analyses of proliferating human lung fibroblasts and pulmonary artery smooth muscle cells in hypoxia.
+Despite activation of the hypoxia-inducible factor (HIF) transcriptional program and up-regulation of glycolytic genes, glycolytic flux was decreased in hypoxic cells.
+While pharmacologic stabilization of HIF in normoxia with the prolyl hydroxylase inhibitor molidustat did increase glycolytic flux as expected, hypoxia abrogated this effect of molidustat treatment.
+Multi-omic profiling of cells treated with hypoxia or molidustat, seperately or together, revealed distinct molecular responses to hypoxia and pharmacologic prolyl hydroxylase inhibition.
+Together, these data suggest that primary cell bioenergetic metabolism is closely coupled to cell proliferation rate and that other factors supersede the anticipated effects of HIF-dependent up-regulation of glycolytic gene expression on glycolytic flux.
 
@@
 With an inadequate oxygen supply, or hypoxia, cells must reduce energy consumption and shift energy production away from oxidative phosphorylation.
-Cells accomplish this goal through stabilization of the hypoxia-inducible transcription factor 1α (HIF-1α), which activates the transcription of glucose transporters, glycolytic enzymes, lactate dehydrogenase, and pyruvate dehydrogenase kinase, while decreasing the expression of proteins in the tricarboxylic acid (TCA) cycle and electron transport chain [@RN1980, @RN1198].
-Overall, these changes in gene transcription should increase glycolytic capacity and divert glucose-derived pyruvate from oxidative phosphorylation toward lactate fermentation to maintain energy production and minimize the formation of reactive oxygen species [@RN2525].
+Cells accomplish this goal through stabilization of the hypoxia-inducible transcription factor 1α (HIF-1α), which activates the transcription of glucose transporters, glycolytic enzymes, lactate dehydrogenase, and pyruvate dehydrogenase kinase, while decreasing the expression of proteins in the tricarboxylic acid (TCA) cycle and electron transport chain [@RN1980; @RN1198].
+Although HIF-1α is constitutively expressed, it is hydroxylated by prolyl hydroxylase enzymes in normoxia and targeted for proteasomal degradation.
+Prolyl hydroxlyase enzymes (PHDs) are the principal oxygen sensors in metazoan cells [@RN999].
+PHDs are α-ketoglutrate dioxygenase enzymes that require molecular oxygen to catalyze this protein modification.
+When oxygen tension falls, PHD activity decreases, leading to HIF-1α stabilization and its associated transcriptional program.
+Overall, the changes in gene transcription should increase glycolytic capacity and divert glucose-derived pyruvate from oxidative phosphorylation toward lactate fermentation to maintain ATP production and minimize the formation of reactive oxygen species (ROS) [@RN2525].
 
-While this "glycolytic shift" of primary carbon metabolism is well-described, the effects of hypoxia on other metabolic pathways are an area of active investigation [@RN953, @RN1603, @RN1198].
-Since hypoxia is a prominent feature of cancer biology as tumor growth outstrips blood supply, most detailed metabolic studies of cell metabolism in hypoxia have used tumor cell models [@RN2395, @RN1447, @RN1438, @RN634, @RN1200, @RN975].
+While a "glycolytic shift" of primary carbon metabolism in hypoxia is well-described, the effects of hypoxia on other metabolic pathways are an area of active investigation [@RN953; @RN1603; @RN1198].
+Since hypoxia is a prominent feature of cancer biology as tumor growth outstrips blood supply, most detailed metabolic studies of cell metabolism in hypoxia have used tumor cell models [@RN2395; @RN1447; @RN1438; @RN634; @RN1200; @RN975].
 For example, stable isotope tracing and metabolic flux analyses identified a critical role for reductive carboxylation of glutamine-derived α-ketoglutarate for lipid biosynthesis in tumor growth [@RN1447; @RN628; @RN1966; @RN2395], and metabolomic studies identified aspartate as a limiting metabolite for cancer cell proliferation under hypoxia [@RN634].
-By contrast, comparatively little is known about metabolic adaptations of primary cells to hypoxia and the importance of the metabolic pathways described above remain to be elucidated in these cells.
-This information would provide important context for understanding the extent to which cancer cell metabolism responds differently to hypoxic stress.
+By contrast, comparatively little is known about metabolic adaptations of primary cells to hypoxia and the importance of reductive carboxylation and aspartate biosynthesis remain to be elucidated in these cells.
+This information would provide important context for understanding fundamental differences in how cancer cell metabolism responds to hypoxic stress.
 Given the metabolic adaptations required for rapid proliferation in cancer cells, we hypothesized that hypoxia would elicit different metabolic responses in primary cells than has been observed previously in studies of cancer cell metabolism.
 
-To test this hypothesis, we developed models of bioenergetic carbon flux in human primary cells cultured under 21% or 0.5% oxygen conditions.
-We found that hypoxia fails to increase glycolysis in primary cells despite robust up-regulation of the HIF-1α transcriptional program.
-In normoxia, HIF-1α activation by the prolyl hydroxylase (PHD) inhibitor molidustat (BAY-85-3934, "BAY") [@RN580] did increase glycolysis and lactate efflux; however, hypoxia abrogated this response.
-These findings suggest the existence of important hypoxia-dependent regulatory mechanisms that override the metabolic consequences of HIF-1α-dependent up-regulation of glycolytic gene expression in human primary cells.
+To test this hypothesis, we developed models of bioenergetic carbon flux in human primary lung fibroblasts (LFs) and pulmonary artery smooth muscle cells (PASMCs) cultured under 21% or 0.5% oxygen conditions.
+We found that hypoxia fails to increase glycolysis in primary cells despite robust up-regulation of the HIF-1 transcriptional program.
+In normoxia, HIF-1α stabilization by the PHD inhibitor molidustat (BAY-85-3934, "BAY") [@RN580] did increase glycolysis and lactate efflux; however, hypoxia abrogated this response.
+These findings suggest the existence of important hypoxia-dependent regulatory mechanisms that override the metabolic consequences of HIF-1-dependent up-regulation of glycolytic gene expression in human primary cells.
 
@@
 From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (**Figure \@ref(fig:f1)A**), thought to be an indicator of metabolic steady state.
-Lung fibroblasts (LFs) cultured in 0.5% oxygen grew more slowly (**Figure \@ref(fig:f1)B**), but slower growth was not associated with decreased cell viability (**Figure \@ref(fig:f1)C**).
-These cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1, (GLUT1), and lactate dehydrogenase A (LDHA) (**Figure \@ref(fig:f1)D-H**).
+LFs cultured in 0.5% oxygen grew more slowly (**Figure \@ref(fig:f1)B**), but slower growth was not associated with decreased cell viability (**Figure \@ref(fig:f1)C**).
+These cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1 (GLUT1) and lactate dehydrogenase A (LDHA) (**Figures \@ref(fig:f1)D-H**).
 These changes persisted for the duration of the experimental time course.
@@
 
-Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (**Figure \@ref(fig:f1)I-J, Supplementary Figures 1-3**).
-Flux calculations incorporated the changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@RN1523] (**Supplementary Figure 1**).
+Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (**Figures \@ref(fig:f1)I-J, S1-S3**).
+Flux calculations incorporated changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@RN1523] (**Figure S1**).
 Surprisingly, while glucose uptake was modestly increased in hypoxia, lactate efflux was decreased (**Figure \@ref(fig:f1)I**) despite activation of the HIF-1 transcriptional program.
-Similar findings were observed when the ambient oxygen level was decreased further to 0.2% (Supplementary Figure 2) and in pulmonary artery smooth muscle cells (PASMCs; **Supplementary Figure 3**).
-In addition to lactate, the absolute extracellular fluxes of pyruvate and amino acids were generally decreased in hypoxia, including a marked decrease in glutamine uptake in LFs.
+Similar findings were observed when the ambient oxygen level was decreased further to 0.2% (**Figure S2**) and in PASMCs (**Figure S3**).
+In addition to lactate, the absolute fluxes of pyruvate and amino acids were generally decreased in hypoxia, including a marked decrease in glutamine uptake in LFs.
 Notably, hypoxia was previously shown to increase glutamine uptake in studies of cancer cell metabolism [@RN1447; @RN2395; @RN628].
-Glutamine uptake did increase in PASMCs, as did the uptake of branch chain amino acids and arginine (Supplementary Figure 3), suggesting differential responses of these mesenchymal cells to hypoxia.
+Glutamine uptake did increase in PASMCs, as did the uptake of branch chain amino acids and arginine (**Figure S3**), suggesting differential responses of these mesenchymal cells to hypoxia.
 
@@
 Cells were treated with BAY to stabilize HIF-1α under 21% oxygen conditions (**Figure \@ref(fig:f2)**).
-Similar to hypoxia, BAY decreased cell growth rate (**Figure \@ref(fig:f2)A-B**) and activated the HIF-1 transcriptional program (**Figure \@ref(fig:f2)C-G**).
+Similar to hypoxia, BAY decreased cell growth rate (**Figures \@ref(fig:f2)A-B**) and activated the HIF-1 transcriptional program (**Figures \@ref(fig:f2)C-G**).
 Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1 target gene transcription and protein expression.
@@
 
-## Stable isotope incorporation suggests preservation of the intracellular metabolic program in hypoxia
+## Stable isotope labeling suggests preservation of the intracellular metabolic architecture in hypoxia
 
-To validate our findings from extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate of label incorporation and trace its incorporation into key carbon utilization pathways (**Figure \@ref(fig:f3), Supplementary Figures 4-5**).
-Consistent with our extracellular flux measurements, we found that the rate of label incorporation from [U-^13^C~6~] glucose into pyruvate was similar across 21%, 0.5%, and DMSO-treated conditions, but markedly increased with BAY treatment (**Figure \@ref(fig:f3)A-B**).
+To validate our findings from extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate of label incorporation and trace its incorporation into key carbon utilization pathways (**Figures \@ref(fig:f3), S4-S5**).
+Consistent with our extracellular flux measurements, we found that the rate of label incorporation from [U-^13^C~6~] glucose into pyruvate was similar across 21%, 0.5%, and DMSO-treated conditions, but markedly increased with BAY treatment (**Figures \@ref(fig:f3)A-B**).
 
@@
 Beyond this observation, the labeling patterns in hypoxia- and BAY-treated cells were similar to their respective controls, arguing against marked metabolic reprogramming in response to prolyl hydroxylase inhibition by either hypoxia or BAY.
-These findings were similar in PASMCs (**Supplementary Figure 5**) where stable isotope incorporation was attenuated in hypoxia and the overall labeling pattern was quite similar in both conditions.
+These findings were similar in PASMCs (**Figure S5**) where stable isotope incorporation was attenuated in hypoxia and the overall labeling pattern was quite similar in both conditions.
 Compared to previous studies of metabolic flux in cancer cells [@RN2395; @RN1447], no marked increase in M5-labeled citrate from [U-^13^C~5~] glutamine was observed with hypoxia treatment, indicating no increase in reductive carboxylation for lipid synthesis.
@@
 To clarify changes in intracellular metabolite fluxes, we next generated metabolic flux models incorporating the extracellular flux measurements and stable isotope tracing data described above.
-Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (**Figure \@ref(fig:f3)A, Supplementary Figure 6**).
-Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by INCA [@RN1522; @RN2501; @RN964] (**Figure \@ref(fig:f4), Supplementary Figure 6, Supplementary Tables 1-3**).
+Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (**Figures \@ref(fig:f3)A, S6**).
+Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by INCA [@RN1522; @RN2501; @RN964] (**Figures \@ref(fig:f4), S6, Tables S1-S3**).
 Overall, proliferating LFs demonstrated high rates of glucose uptake and glycolysis.
-Approximately 10% of cytoplasmic pyruvate enters the TCA cycle with the balance converted to lactate (**Supplementary Figure 6C**).
+Approximately 10% of cytoplasmic pyruvate enters the TCA cycle with the balance converted to lactate (**Figure S6C**).
 In hypoxia, significant reductions in glycolysis, the TCA cycle, and amino acid metabolism were observed (**Figure \@ref(fig:f4)A**) with a significant increase in pentose phosphate pathway flux.
-Similar findings were observed in PASMC flux maps (**Supplementary Figure 6, Supplementary Table 3**).
+Similar findings were observed in PASMC flux models (**Figure S6, Table S3**).
 
-Given the global decrease in bioenergetic metabolic flux in hypoxia, we hypothesized that these differences may be a consequence of a decreased growth rate.
-After normalizing metabolite fluxes in normoxia and hypoxia to the associated growth rate, modest increases (~10%) in glycolytic flux were observed (Supplementary Figure 4D).
+Given the global decrease in bioenergetic metabolic flux in hypoxia, we hypothesized that these differences may be a consequence of decreased growth rate.
+After normalizing metabolite fluxes in normoxia and hypoxia to the  growth rate, modest increases (~10%) in glycolytic flux were observed (**Figure S4D**).
 This finding suggests that, while glycolysis increases relative to growth rate in hypoxic cells, the regulators of cell proliferation rate dominate over the consequences of the HIF-1 transcriptional program.
@@
 
-Although the metabolite exchange fluxes for bidirectional reactions tend to be poorly resolved, two observations are worth highlighting (Supplementary Tables 1-3).
-First, consistent with the stable isotope tracing results, the rate of reductive carboxylation through reversible flux by isocitrate dehydrogenase is low (~4 fmol/cell/h), is unchanged by hypoxia, and is modestly increased by BAY treatment and by hypoxia treatment in PASMCs.
+Although the metabolite exchange fluxes for bidirectional reactions tend to be poorly resolved, two observations are worth highlighting (**Tables S1-S3**).
+First, consistent with the stable isotope tracing results, the rate of reductive carboxylation through reversible flux by isocitrate dehydrogenase is low (~4 fmol/cell/h), is unchanged by hypoxia in LFs, is modestly increased by BAY treatment, and is increased by hypoxia in PASMCs.
 
@@
 This may be consistent with the HIF-driven increased expression of the reversible lactate transporter MCT4 [@RN366].
-To investigate this hypothesis, LFs and PASMCs were treated with [U-^13^C~3~] lactate (2 mM) and ^13^C incorporation into intracellular metabolites was analyzed by LC-MS (**Figure \@ref(fig:f5), Supplementary Figure 5**).
-Here, we observed increased labeling of TCA metabolites citrate (CIT), 2-oxoglutrate (2OG), malate (MAL), and aspartate (ASP) following hypoxia or BAY treatment (**Figure \@ref(fig:f5), Supplementary Figure 4**).
-Interestingly, a similar increase in labeling was not observed in PASMCs (Supplementary Figure 5).
+To investigate this hypothesis, LFs and PASMCs were treated with [U-^13^C~3~] lactate (2 mM) and ^13^C incorporation into intracellular metabolites was analyzed by LC-MS (**Figures \@ref(fig:f5), S5**).
+Here, we observed increased labeling of TCA metabolites citrate (CIT), 2-oxoglutrate (2OG), malate (MAL), and aspartate (ASP) following hypoxia or BAY treatment (**Figures \@ref(fig:f5), S4**).
+Interestingly, although increased labeling of pyruvate and lactate were observed in hypoxic PASMCs, these labels were not incorporated into the TCA cycle as observed in LFs (**Figure S5**).
 
@@
 Although lactate has been used less commonly than glucose and glutamine in stable isotope tracing studies, Faubert and colleagues [-@RN2600] demonstrated lactate incorporation in human lung adenocarcinoma *in vivo*.
-In this study, lactate incorporation corresponded to regions of high glucose uptake as determined by `[`^18^F]-fluorodeoxyglucose positron emission tomography, suggesting that lactate consumption can occur in areas of high glucose utilization.
+In this study, lactate incorporation corresponded to regions of high glucose uptake as determined by [ ^18^F]-fluorodeoxyglucose positron emission tomography, suggesting that lactate consumption can occur in areas of high glucose utilization.
 Subsequently, several investigators have demonstrated the importance of lactate as a metabolic fuel *in vivo* [@RN2756; @RN2718].
@@
 
-To reconcile the differential effects of prolyl hydroxylase inhibition by hypoxia and BAY, we next addressed whether hypoxia could suppress the effects of BAY on glucose and lactate fluxes (**Figure \@ref(fig:f6)A-C**).
+To reconcile the differential effects of prolyl hydroxylase inhibition by hypoxia and BAY, we next addressed whether hypoxia could suppress the effects of BAY on glucose and lactate fluxes (**Figures \@ref(fig:f6)A-C**).
 LFs cultured in standard growth medium were treated with BAY and placed in either 21% or 0.5% oxygen.
@@
 
-To investigate these metabolic differences further, we performed metabolomic profiling of LFs treated for 72 h with hypoxia or BAY separately or in combination (**Figure \@ref(fig:f6)D-F**).
+To investigate these metabolic differences further, we performed metabolomic profiling of LFs treated for 72 h with hypoxia or BAY separately or in combination (**Figures \@ref(fig:f6)D-F**).
 Principle component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen (**Figure \@ref(fig:f6)D**).
@@
 Several patterns emerged from this analysis (**Figure \@ref(fig:f6)F**).
-Interestingly, the hypoxia-regulated metabolite, 2-hydroxyglutarate, increased following BAY treatment, but was unaffected by hypoxia treatment in these proliferating LFs [@RN1603].
+Interestingly, the hypoxia-regulated metabolite, 2-hydroxyglutarate, increased following BAY treatment, but was unaffected by hypoxia treatment in proliferating LFs [@RN1603].
 The tricarboxylic acid (TCA) cycle metabolite aconitate was decreased in hypoxia, but increased by BAY treatment while the opposite effect was observed for hydroxyproline and taurine.
 Glycolytic intermediate glyceraldehyde 3-phosophate (GAP) was decreased by both hypoxia and BAY, while hypoxia reversed the BAY-mediated decrease in γ-aminobutyric acid (GABA).
-A metabolite set enrichment analysis (**Figure \@ref(fig:f6)G-H**) revealed the tricarboxlyic acid cycle to be the most enriched KEGG metabolite set.
+A metabolite set enrichment analysis (**Figure \@ref(fig:f6)G-H**) revealed the TCA cycle to be the most enriched KEGG metabolite set.
 Leading edge analysis demonstrates the negative enrichment scores associated with all of the TCA metabolites detected by our platform.
@@
 In addition to these differential effects on polar metabolite levels, perhaps the most significant contrast between hypoxia and BAY treatment is the impact of hypoxia on cellular redox state.
-As oxygen deprivation causes reductive stress [@RN2409], we next measured the impact of these treatments on intracellular NAD(H) as a surrogate for cellular redox state (**Figure \@ref(fig:f6)I-K**).
+As oxygen deprivation causes reductive stress [@RN2409], we next measured the impact of these treatments on intracellular NAD(H) as a surrogate for cellular redox state (**Figures \@ref(fig:f6)I-K**).
 As expected, hypoxia increased the NADH/NAD^+^ ratio, driven primarily by a decrease in intracellular NAD^+^.
@@
 Our principal finding was that hypoxia reduced, rather than increased, carbon flux through glycolysis and lactate fermentation pathways despite robust activation of the HIF transcriptional program and up-regulation of glycolytic genes.
-Certainly the LFs studied here are capable of augmenting glycolysis in response to HIF stabilization in hypoxia, as demonstrated by experiments with the PHD inhibitor BAY; however, these effects are completely attenuated when BAY-treated cells are cultured in hypoxia.
+Certainly the LFs studied here are capable of augmenting glycolysis in response to HIF stabilization, as demonstrated by experiments with the PHD inhibitor BAY; however, these effects are completely attenuated when BAY-treated cells are cultured in hypoxia.
 Together, these findings suggest that changes in enzyme levels alone are insufficient to alter metabolic flux in hypoxia and point to the importance of regulatory mechanisms that supersede the effects of HIF-dependent gene transcription.
@@
 While the role of HIFs in the hypoxia response has been extensively studied, relatively less is known about HIF-independent features of the hypoxia response.
-Certainly cells express several oxygen-dependent enzymes in addition to PHD whose activities may be impacted in hypoxia but not by HIF stabilization.
+Cells express several oxygen-dependent enzymes in addition to PHD whose activities may be impacted in hypoxia but not by PHD inhibition.
 For example, PHD is one of many α-ketoglutarate-dependent dioxygenase enzymes that rely on molecular oxygen for their catalytic activity [@RN2611].
@@
 Similarly, a recently described cysteamine dioxygenase has been shown to mediate the oxygen-dependent degradation of Regulators of G protein Signaling 4 and 5 and IL-32 [@RN1384].
-In addition to dioxygenase enzymes, electron transport chain dysfunction resulting from impaired Complex IV activity leads to increased reactive oxygen species (ROS) production in hypoxia [@RN295].
+In addition to dioxygenase enzymes, electron transport chain dysfunction resulting from impaired Complex IV activity leads to increased ROS production in hypoxia [@RN295].
 Mitochondrial ROS increase the half-lives of several mRNAs in hypoxia, including MYC as we observe in this work, independent of HIF stabilization [@RN742].
@@
 These findings raise important questions regarding the cell-autonomous role of HIFs in the hypoxia response.
-Certainly, on an organismal level, HIFs drive expression of angiogenic and erythropoietic factors to increase oxygen delivery to hypoxic tissues.
-Within individual cells, HIF-1α seems to be important for mitigating the adverse effects of reactive oxygen species (ROS) formation by dysfunctional electron transport in the mitochondria.
+On an organismal level, HIFs drive expression of angiogenic and erythropoietic factors to increase oxygen delivery to hypoxic tissues.
+Within individual cells, HIF-1α seems to be important for mitigating the adverse effects of ROS formation by dysfunctional electron transport in the mitochondria.
 Indeed, hypoxia increased oxygen consumption and ROS production in HIF-1α-null mouse embryonic fibroblasts (MEFs), which was associated with increased cell death [@RN2525].
@@
 conceived and designed the analysis.
-C.A.C., B.A.O., D.R.Z., S.M., and W.M.O.
+C.A.C., B.A.O., D.R.Z., S.M., K.L., and W.M.O.
 collected the data.
@@
 Fluxes with non-overlapping confidence intervals are highlighted to indicate significant changes.
+Arrow weight corresponds to absolute flux in hypoxia- or BAY-treated cells.
 
@@
 (A) Growth rates were determined by linear fitting of log-transformed growth curves.
-(B, C) Extracellular fluxes of glucose (B) and lactate (C).
+(**B**-**C**) Extracellular fluxes of glucose (B) and lactate (C).
 (**D**) Principle components analysis of intracellular metabolites following 72 h of treatment described above suggests a dominant effect of hypoxia over PHD inhibition on the metabolome (n = 4).
@@
 (**E**) Transcription factor enrichment analysis suggests mechanisms for differential regulation of gene expression following hypoxia or BAY treatment.
-(**F**) Relative change in c-MYC protein levels after 72 h of treatment (n = 4) and a representative immunoblot.
+(**F**) Relative change in MYC protein levels after 72 h of treatment (n = 4) and a representative immunoblot.
 Comparisons were performed using a mixed-effects linear model with replicate as a random effect.
@@
 Primary normal human lung fibroblasts (LFs) were purchased from Lonza (CC-2512) and cultured in FGM-2 (Lonza CC-3132) in a standard tissue culture incubator in 5% CO~2~ at 37 °C.
-Cells from two donors were used in these studies: #33652 (56, male) and #29132 (19, female).
+Cells from two donors were used in these studies: #33652 (56 y.o., male) and #29132 (19 y.o., female).
 Cell authentication was performed by the vendor.
@@
 Primary human pulmonary artery smooth muscle cells were purchased from Lonza (CC-2581) and cultured in SmGM-2 (Lonza CC-3182) in a standard tissue culture incubator in 5% CO~2~ at 37 °C.
-Cells from multiple donors were used in these studies: #30020 (64, male), #27662 (35, male), #26698 (51, male), and #19828 (51, male).
+Cells from multiple donors were used in these studies: #30020 (64 y.o., male), #27662 (35 y.o., male), #26698 (51 y.o., male), and #19828 (51 y.o., male).
 Cell authentication was performed by the vendor.
@@
 Total DNA in 10 μL of lysate was determined by adding 100 μL of 1X PicoGreen dye in Tris-EDTA buffer and interpolating the fluorescence intensity with a standard curve generated using the λ DNA standard.
-Cell counts were interpolated from a standard curve of DNA obtained from known cell numbers seeded in basal medium (Supplementary Figure 1A).
-No difference in total cellular DNA was identified between normoxia and hypoxia cultures (Supplementary Figure 1B).
+Cell counts were interpolated from a standard curve of DNA obtained from known cell numbers seeded in basal medium (**Figure S1A**).
+No difference in total cellular DNA was identified between normoxia and hypoxia cultures (**Figure S1B**).
 
@@
 Growth rate ($\mu$) and cell count at time 0 ($X_0$) were determined by robust linear modeling of the logarithm of cell count as a function of time ($t$).
-Metabolite mass was calculated from the measured metabolite concentrations and predicted well volume accounting for evaporative losses (Supplementary Figure 1C).
+Metabolite mass was calculated from the measured metabolite concentrations and predicted well volume accounting for evaporative losses (**Figure S1C**).
 First-order degradation and accumulation rates were obtained from robust linear modeling of metabolite mass *v*.
@@
 Plates were stored at -80 °C until extraction.
-Metabolites were extracted with 1 mL 80% MeOH pre-cooled to -80 °C containing 10 nmol [D~8~]-\textsc{dl}-valine as an internal standard (Cambridge Isotope Labs).
+Metabolites were extracted with 1 mL 80% MeOH pre-cooled to -80 °C containing 10 nmol [D~8~]-valine as an internal standard (Cambridge Isotope Labs).
 Insoluble material was removed by centrifugation at 21,000 ×*g* for 15 min at 4 °C.
@@
 
+Table S1: **Lung fibroblast fluxes in 21% and 0.5% oxygen.**
+
+Table S2: **Lung fibroblast fluxes following DMSO and BAY treatment.**
+
+Table S3: **PASMC fluxes in 21% and 0.5% oxygen.**
+
+Figure S1: **Supporting data for extracellular flux calculations.**
+
+Figure S2: **Extracellular flux measurements in 0.2% oxygen.**
+
+Figure S3: **Extracellular flux measurements in pulmonary artery smooth muscle cells in 0.5% oxygen.**
+
+Figure S4: **Mass isotopomer distributions after 72 h of labeling in lung fibroblasts.**
+
+Figure S5: **Mass isotopomer distributions after 72 h of labeling in pulmonary artery smooth muscle cells.**
+
+Figure S6: **Isotopically non-stationary metabolic flux analysis.**
+
 \newpage
@@
 
-::: {#refs_main}
+::: {#refs}
 :::
-
-# Software
-
-::: {#refs_software}
-:::

### 2021-05-10 inst/manuscript/supplement.Rmd
revisions to rough draft

@@
 graphics: yes
-header-includes:
-- \usepackage{lineno}
-- \linenumbers
-- \setlength\linenumbersep{1cm}
-- \usepackage{setspace}
-- \doublespacing
-- \makeatletter\renewcommand*{\fps@figure}{htbp}\makeatother
-- \setlength{\parskip}{0.75\baselineskip plus 2pt minus 2pt}
-- \usepackage{float}
+
+nocite: |
+  @R-base, @R-bookdown, @R-broom, @R-cellranger, @R-cowplot, @R-devtools, @R-dplyr, 
+  @R-emmeans, @R-forcats, @R-ggbeeswarm, @R-ggplot2, @R-ggraph, @R-kableExtra, 
+  @R-knitr, @R-lme4, @R-lmerTest, @R-lubridate, @R-magick, @R-magrittr, @R-MASS, 
+  @R-mzrtools, @R-patchwork, @R-pbkrtest, @R-purrr, @R-RColorBrewer, @R-readr, 
+  @R-readxl, @R-renv, @R-rlang, @R-rmarkdown, @R-roxygen2, @R-stringr, @R-tibble, 
+  @R-tidygraph, @R-tidyr, @R-tidyverse, @R-tinytex, @R-usethis, @R-viridis, @R-wmo, 
+  @bookdown2016, @ggplot22016, @knitr2015, @lme42015, @lmerTest2017, @lubridate2011, 
+  @MASS2002, @pbkrtest2014, @rmarkdown2018, @tidyverse2019, tinytex2019
 ---
@@
   pasmc <- targets::tar_read(pasmc_hypoxia_table)
-  figs <- targets::tar_load(targets::matches("s\\d\\.pdf"))
+  figs <- targets::tar_load(targets::matches("s\\d_figure"))
 })
@@
 ```
+
+# Supplemental References
+
+::: {#refs_main}
+:::
+
+::: {#refs_software}
+:::

### 2021-05-10 inst/src/figures.R

@@ -351,6 +351,6 @@ format_flux_table <- function(
       value = flextable::as_paragraph(flextable::as_sup("b"), ssr_exp)
     ) %>%
     flextable::font(fontname = "Calibri", part = "all") %>%
-    flextable::fontsize(size = 9, part = "all") %>%
+    flextable::fontsize(size = 8.5, part = "all") %>%
     flextable::set_table_properties(layout = "autofit")
 }

### 2021-05-10 inst/src/nad.R

@@ -146,7 +146,7 @@ plot_nad <- function(nad_final, annot, metab, ylab) {
       fill = NULL,
       color = NULL
     ) +
-    # ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(NA, 0.2))) +
+    ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0.05, 0.1))) +
     # ggplot2::coord_cartesian(ylim = c(-750, 1250)) +
     theme_plots() +
     ggplot2::theme(

### 2021-05-11 analysis/metabolomics.Rmd
move analysis out of inst to top-level directory

--- 2021-04-28 inst/analysis/metabolomics.Rmd
@@
   metab_limma <- tar_read(metab_targeted_limma)
-  # metab_tt <- tar_read(metab_different_differences)
+  tt <- tar_read(metab_different_differences)
   volcano <- tar_read(metab_volcano)
+  msea <- tar_read(msea_plot)
+  leading_edge <- tar_read(leading_edge)
 })
@@
 
+## Metabolite set enrichment
+
 ```{r}
-plot_mois(clean, "malate")
+msea
 ```
+
+```{r}
+leading_edge
+```

### 2021-05-11 manuscript/manuscript.Rmd
move analysis out of inst to top-level directory

--- 2021-05-10 inst/manuscript/manuscript.Rmd
@@
 
-# kable function
-my_kable <- function(data, ...) {
-  kableExtra::kable(data, booktabs = TRUE, linesep = "", ...) %>% 
-    kableExtra::kable_styling(latex_options = c("HOLD_position"), font_size = 11)
-}
-
-withr::with_dir(here::here(), {
 biomass <- targets::tar_read(biomass)
@@
 figs <- targets::tar_load(targets::matches("m\\d_figure"))
-})
-
 biomass <- tibble::deframe(biomass)

### 2021-05-11 manuscript/supplement.Rmd
move analysis out of inst to top-level directory

--- 2021-05-10 inst/manuscript/supplement.Rmd
@@
 
-# kable function
-my_kable <- function(data, ...) {
-  kableExtra::kable(data, booktabs = TRUE, linesep = "", ...) %>% 
-    kableExtra::kable_styling(latex_options = c("HOLD_position"), font_size = 11)
-}
-
-withr::with_dir(here::here(), {
 hyp <- targets::tar_read(lf_hypoxia_table)
@@
 figs <- targets::tar_load(targets::matches("s\\d_figure"))
-})
-
 ```
@@
 
+\newpage
+
 # Supplemental References

### 2021-05-11 R/utilities.R

@@ -1,12 +1,24 @@
 # utilities.R
 
+# src ---------------------------------------------------------------------
+
 src <- function() {
   files <-
     list.files(
-    system.file("src/", package = "Copeland.2021.hypoxia.flux"),
-    pattern = "\\.R$",
-    full.names = TRUE
-  )
+      "scripts",
+      pattern = "\\.R$",
+      full.names = TRUE
+    )
 
   invisible(lapply(files, source))
+}
+
+# my_kable ----------------------------------------------------------------
+
+my_kable <- function(data, ...) {
+  kableExtra::kable(data, booktabs = TRUE, linesep = "", ...) %>%
+    kableExtra::kable_styling(
+      latex_options = c("hold_position"),
+      font_size = 9
+    )
 }

### 2021-05-11 scripts/figures.R

@@ -38,11 +38,7 @@ theme_patchwork <- function(design = NULL, widths = NULL, heights = NULL, tags =
 # write_figures -----------------------------------------------------------
 
 write_figures <- function(plot, filename) {
-  path <-
-    system.file(
-      "manuscript/figures",
-      package = "Copeland.2021.hypoxia.flux"
-    )
+  path <- "manuscript/figures"
 
   gtab <- patchwork::patchworkGrob(plot)
 
@@ -351,6 +347,6 @@ format_flux_table <- function(
       value = flextable::as_paragraph(flextable::as_sup("b"), ssr_exp)
     ) %>%
     flextable::font(fontname = "Calibri", part = "all") %>%
-    flextable::fontsize(size = 8.5, part = "all") %>%
+    flextable::fontsize(size = 8, part = "all") %>%
     flextable::set_table_properties(layout = "autofit")
 }

### 2021-05-11 scripts/utils.R

@@ -14,23 +14,11 @@
   match(x, table, nomatch = 0L) == 0L
 }
 
-# my_kable ----------------------------------------------------------------
-
-my_kable <- function(data, ...) {
-  kableExtra::kable(data, booktabs = TRUE, linesep = "", ...) %>%
-    kableExtra::kable_styling(
-      latex_options = c("hold_position"),
-      font_size = 9
-    )
-}
 # path_to_data ------------------------------------------------------------
 
 path_to_data <- function(nm) {
   dir(
-    system.file(
-      "extdata/",
-      package = "Copeland.2021.hypoxia.flux"
-    ),
+    path = "data-raw/",
     pattern = nm,
     all.files = TRUE,
     full.names = TRUE,
@@ -42,46 +30,20 @@ path_to_data <- function(nm) {
 # path_to_reports ---------------------------------------------------------
 
 path_to_reports <- function(nm) {
-  path <-
-    stringr::str_c(
-      system.file(
-        stringr::str_c("analysis"),
-        package = "Copeland.2021.hypoxia.flux"
-      ),
-      "/",
-      nm
-    )
-  path
+  stringr::str_c("analysis/", nm)
 }
 
 
 # path_to_manuscript ------------------------------------------------------
 
 path_to_manuscript <- function(nm) {
-  path <-
-    stringr::str_c(
-      system.file(
-        stringr::str_c("manuscript"),
-        package = "Copeland.2021.hypoxia.flux"
-      ),
-      "/",
-      nm
-    )
-  path
+  stringr::str_c("manuscript/", nm)
 }
 
 # path_to_plots -----------------------------------------------------------
 
 path_to_plots <- function(nm) {
-  path <-
-    stringr::str_c(
-      system.file(
-        stringr::str_c("analysis/"),
-        package = "Copeland.2021.hypoxia.flux"
-      ),
-      "/figures/",
-      nm
-    )
+  path <- stringr::str_c("analysis/figures/", nm)
 
   if (dir.exists(path)) unlink(path, recursive = TRUE)
 

### 2021-07-13 analysis/metabolomics-targeted.Rmd
start untargeted metabolomics analysis

--- 2021-05-11 analysis/metabolomics.Rmd
@@
 ---
-title: "Metabolomics"
+title: "Targeted Metabolomics"
 subtitle: "`r format(Sys.time(), '%B %d, %Y')`"

### 2021-07-13 analysis/metabolomics-untargeted.Rmd
start untargeted metabolomics analysis

--- 2021-04-28 inst/analysis/metabolomics.Rmd
@@
 ---
-title: "Metabolomics"
+title: "Untargeted Metabolomics"
 subtitle: "`r format(Sys.time(), '%B %d, %Y')`"
@@
 withr::with_dir(here::here(), {
-  raw <- tar_read(metab_targeted_raw)
-  clean <- tar_read(metab_targeted_clean)
-  metab_pca <- tar_read(metab_targeted_pca)
-  metab_limma <- tar_read(metab_targeted_limma)
-  # metab_tt <- tar_read(metab_different_differences)
-  volcano <- tar_read(metab_volcano)
+  pos_chrom <- tar_read(chromatograms_pos)
+  neg_chrom <- tar_read(chromatograms_neg)
+  pos_int <- tar_read(intensities_pos)
+  neg_int <- tar_read(intensities_neg)
 })
@@
 
-# Data Processing
+# Data
 
-## PCA
+## Preprocessing
+
+The raw total ion intensity chromatogram were reviewed for an overview of signal acquisition quality:
 
 ```{r}
-pcaMethods::pca(t(SummarizedExperiment::assay(raw)), scale = "uv", center = TRUE) %>% 
-  pcaMethods::scores() %>% 
-  merge(SummarizedExperiment::colData(raw), by = 0) %>% 
-  # filter(type %nin% c("qc", "blank")) %>% 
-  ggplot() +
-  aes(
-    x = PC1, 
-    y = PC2, 
-    shape = type,
-    color = interaction(oxygen, treatment, sep = "|")
-  ) + 
-  geom_point() +
-  labs(
-    color = "Group", 
-    shape = "Type"
-  )
-```
-## Cleaning
-
-1. drift correction
-1. remove low quality features (QC RSD > 0.2, dispersion ratio > 0.4)
-1. impute missing values with random forest
-1. probabilistic quotient normalization
-1. batch correction
-
-```{r}
-metab_pca
-```
-
-## Differentially regulated metabolites
-
-```{r}
-volcano
+pos_chrom
+neg_chrom
 ```
@@
 ```{r}
-plot_mois(clean, "malate")
+pos_int
+neg_int
 ```

### 2021-07-13 R/a_metabolomics.R

@@ -1,7 +1,5 @@
 # metabolomics.R
 
-# format_metab_targeted ---------------------------------------------------
-
 format_metab_targeted <- function(metab_targeted_files) {
   readxl::read_excel(
     metab_targeted_files,
@@ -52,8 +50,6 @@ format_metab_targeted <- function(metab_targeted_files) {
 
 }
 
-# new_tbl_se --------------------------------------------------------------
-
 new_tbl_se <- function(
   tbl,
   a_data,
@@ -72,8 +68,6 @@ new_tbl_se <- function(
     s_data = s_data
   )
 }
-
-# tibble_to_se ------------------------------------------------------------
 
 tbl_to_se <- function(tbl_se, assay_name) {
 
@@ -127,8 +121,6 @@ tbl_to_se <- function(tbl_se, assay_name) {
 
 }
 
-# remove_missing_metab ----------------------------------------------------
-
 remove_missing_metab <- function(raw) {
 
   qc <- SummarizedExperiment::assay(raw[, raw$type == "qc"])
@@ -139,21 +131,19 @@ remove_missing_metab <- function(raw) {
 
 }
 
-# correct_drift -----------------------------------------------------------
+prepare_assay_data <- function(se) {
+  SummarizedExperiment::assay(se) %>%
+    tibble::rownames_to_column("hmdb") %>%
+    tidyr::pivot_longer(-hmdb, names_to = "sample", values_to = "value") %>%
+    dplyr::mutate(
+      value = log(value),
+      run_order = as.numeric(stringr::str_extract(sample, "\\d{2}"))
+    ) %>%
+    dplyr::group_by(hmdb) %>%
+    tidyr::nest()
+}
 
 correct_drift <- function(missing) {
-
-  prepare_assay_data <- function(se) {
-    SummarizedExperiment::assay(se) %>%
-      tibble::rownames_to_column("hmdb") %>%
-      tidyr::pivot_longer(-hmdb, names_to = "sample", values_to = "value") %>%
-      dplyr::mutate(
-        value = log(value),
-        run_order = as.numeric(stringr::str_extract(sample, "\\d{2}"))
-      ) %>%
-      dplyr::group_by(hmdb) %>%
-      tidyr::nest()
-  }
 
   models <-
     missing[, missing$type == "qc"] %>%
@@ -180,8 +170,6 @@ correct_drift <- function(missing) {
   missing
 
 }
-
[…]

### 2021-07-13 _targets.R

@@ -20,13 +20,18 @@ options(
   usethis.quiet = TRUE
 )
 
-future::plan(future::multisession(workers = future::availableCores() - 1))
+future::plan(future.callr::callr(workers = future::availableCores() - 1))
 
 # target-specific options
 tar_option_set(
-  packages = c("tidyverse", "patchwork"),
+  packages = c("tidyverse", "patchwork", "xcms"),
   imports = c("rnaseq.lf.hypoxia.molidustat"),
   format = "qs"
+)
+
+values <- tibble::tibble(
+  polarity = c("positive", "negative"),
+  names = c("pos", "neg")
 )
 
 # list of targets ---------------------------------------------------------
@@ -536,8 +541,70 @@ list(
   ),
   tar_render(
     metabolomics_report,
-    path = path_to_reports("metabolomics.Rmd"),
+    path = path_to_reports("metabolomics-targeted.Rmd"),
     output_dir = system.file("analysis/pdfs", package = "Copeland.2021.hypoxia.flux")
+  ),
+  tar_target(
+    metab_untargeted_files,
+    path_to_data(".mzML"),
+    format = "file"
+  ),
+  tar_target(
+    metab_untargeted_sample_file,
+    path_to_data("sample-sheet.csv"),
+    format = "file"
+  ),
+  tar_target(
+    metab_untargeted_samples,
+    format_sample_info(metab_untargeted_sample_file)
+  ),
+  tar_map(
+    values = values,
+    names = "names",
+    tar_target(
+      raw,
+      read_metab_raw(metab_untargeted_files, polarity, metab_untargeted_samples)
+    ),
+    tar_target(
+      tics,
+      get_tics(raw, metab_untargeted_samples)
+    ),
+    tar_target(
+      chromatograms,
+      plot_chromatograms(tics)
+    ),
+    tar_target(
+      intensities,
+      plot_intensities(tics)
+    ),
+    tar_target(
+      cwp,
+      optimize_centwave_params(raw, polarity)
+    ),
+    tar_target(
+      peaks,
+      findChromPeaks(raw, cwp$best_cwp)
+    ),
+    tar_target(
+      merged,
+      refineChromPeaks(peaks, param = MergeNeighboringPeaksParam(expandRt = 4, ppm = 2.5))
+    ),
+    tar_target(
+      align_group,
+      optimize_align_group_params(merged, polarity)
+    ),
+    tar_target(
+      aligned,
+      adjust_rtime(merged, align_group$best_obi)
+    ),
[…]

### 2021-07-14 manuscript/manuscript.Rmd
manuscript revisions

@@
 While pharmacologic stabilization of HIF in normoxia with the prolyl hydroxylase inhibitor molidustat did increase glycolytic flux as expected, hypoxia abrogated this effect of molidustat treatment.
-Multi-omic profiling of cells treated with hypoxia or molidustat, seperately or together, revealed distinct molecular responses to hypoxia and pharmacologic prolyl hydroxylase inhibition.
-Together, these data suggest that primary cell bioenergetic metabolism is closely coupled to cell proliferation rate and that other factors supersede the anticipated effects of HIF-dependent up-regulation of glycolytic gene expression on glycolytic flux.
+Multi-omic profiling of cells treated with hypoxia or molidustat, separately or together, revealed distinct molecular responses to hypoxia and pharmacologic prolyl hydroxylase inhibition.
+Together, these data suggest that primary cell bioenergetic metabolism is closely coupled to cell proliferation rate, and that other factors supersede the anticipated effects of HIF-dependent up-regulation of glycolytic gene expression on glycolytic flux.
 
@@
 
+Cellular responses to ambient oxygen levels drive numerous physiologic and pathologic processes from wound healing and angiogenesis to pulmonary vascular remodeling and fibrosis [@RN1980].
+These responses require a sustained capacity for cell proliferation, migration, and protein synthesis, and the attendant energetic and metabolic requirements, even in the face of limited oxygen availability, or hypoxia.
 Metazoan cells depend on aerobic respiration to meet cellular energy demands.
-With an inadequate oxygen supply, or hypoxia, cells must reduce energy consumption and shift energy production away from oxidative phosphorylation.
-Cells accomplish this goal through stabilization of the hypoxia-inducible transcription factor 1α (HIF-1α), which activates the transcription of glucose transporters, glycolytic enzymes, lactate dehydrogenase, and pyruvate dehydrogenase kinase, while decreasing the expression of proteins in the tricarboxylic acid (TCA) cycle and electron transport chain [@RN1980; @RN1198].
-Although HIF-1α is constitutively expressed, it is hydroxylated by prolyl hydroxylase enzymes in normoxia and targeted for proteasomal degradation.
-Prolyl hydroxlyase enzymes (PHDs) are the principal oxygen sensors in metazoan cells [@RN999].
-PHDs are α-ketoglutrate dioxygenase enzymes that require molecular oxygen to catalyze this protein modification.
-When oxygen tension falls, PHD activity decreases, leading to HIF-1α stabilization and its associated transcriptional program.
-Overall, the changes in gene transcription should increase glycolytic capacity and divert glucose-derived pyruvate from oxidative phosphorylation toward lactate fermentation to maintain ATP production and minimize the formation of reactive oxygen species (ROS) [@RN2525].
+With an inadequate oxygen supply, cells must reduce energy consumption and shift energy production away from oxidative phosphorylation.
+Cells accomplish this goal through stabilization of the hypoxia-inducible transcription factor 1α (HIF-1α), which activates the transcription of glucose transporters, glycolytic enzymes, lactate dehydrogenase, and pyruvate dehydrogenase kinase, while decreasing the expression of enzymes in the tricarboxylic acid (TCA) cycle and electron transport chain [@RN1980; @RN1198].
+Although HIF-1α is constitutively expressed, it is hydroxylated by prolyl hydroxylase enzymes (PHDs) in normoxia and targeted for proteasomal degradation.
+PHDs are the principal oxygen sensors in metazoan cells [@RN999].
+PHDs are α-ketoglutrate-dependent dioxygenase enzymes that require molecular oxygen for their enzymatic activity.
+When oxygen tension falls, PHD activity decreases, leading to HIF-1α stabilization and activation of its associated transcriptional program.
+Overall, the changes in gene transcription should increase glycolytic capacity and divert glucose-derived pyruvate from oxidative phosphorylation toward lactate fermentation to maintain ATP production and to minimize the formation of reactive oxygen species (ROS) [@RN2525].
 
@@
 For example, stable isotope tracing and metabolic flux analyses identified a critical role for reductive carboxylation of glutamine-derived α-ketoglutarate for lipid biosynthesis in tumor growth [@RN1447; @RN628; @RN1966; @RN2395], and metabolomic studies identified aspartate as a limiting metabolite for cancer cell proliferation under hypoxia [@RN634].
-By contrast, comparatively little is known about metabolic adaptations of primary cells to hypoxia and the importance of reductive carboxylation and aspartate biosynthesis remain to be elucidated in these cells.
-This information would provide important context for understanding fundamental differences in how cancer cell metabolism responds to hypoxic stress.
-Given the metabolic adaptations required for rapid proliferation in cancer cells, we hypothesized that hypoxia would elicit different metabolic responses in primary cells than has been observed previously in studies of cancer cell metabolism.
+By contrast, comparatively little is known about metabolic adaptations of primary cells to hypoxia and how hypoxic metabolic reprogramming supports homeostasis or promotes pathobiology.
+Indeed, the importance of reductive carboxylation or aspartate biosynthesis remains to be elucidated in these cells.
+This and related information would provide important context for understanding how metabolic reprogramming supports normal cellular responses to hypoxia, how these responses may be (mal)adaptive in a variety of disease contexts, and how the hypoxia metabolic program in primary cells differs from that observed in cancer cells.
 
-To test this hypothesis, we developed models of bioenergetic carbon flux in human primary lung fibroblasts (LFs) and pulmonary artery smooth muscle cells (PASMCs) cultured under 21% or 0.5% oxygen conditions.
-We found that hypoxia fails to increase glycolysis in primary cells despite robust up-regulation of the HIF-1 transcriptional program.
+To address these questions, here we have developed models of bioenergetic carbon flux in human lung fibroblasts (LFs) and pulmonary artery smooth muscle cells (PASMCs) cultured under 21% or 0.5% oxygen conditions.
+These cells were selected as they may be exposed to a wide range of oxygen concentrations *in vivo*, continue to proliferate despite hypoxic culture conditions *in vitro*, and play important roles in the pathology of human lung diseases where tissue hypoxia is a prominent feature.
+We found that hypoxia fails to increase glycolysis in these primary cells despite robust up-regulation of the HIF-1 transcriptional program.
 In normoxia, HIF-1α stabilization by the PHD inhibitor molidustat (BAY-85-3934, "BAY") [@RN580] did increase glycolysis and lactate efflux; however, hypoxia abrogated this response.
@@
 
-The goal of this study was to identify the metabolic changes associated with hypoxia in proliferating primary cells.
+The goal of this study was to identify the metabolic changes associated with hypoxia in proliferating primary LFs and PASMCs.
 Cells were seeded and placed into hypoxia for 24 h prior to sample collection to provide adequate time for activation of the hypoxia-dependent transcriptional program.
-From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (**Figure \@ref(fig:f1)A**), thought to be an indicator of metabolic steady state.
-LFs cultured in 0.5% oxygen grew more slowly (**Figure \@ref(fig:f1)B**), but slower growth was not associated with decreased cell viability (**Figure \@ref(fig:f1)C**).
+From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (**Figure \@ref(fig:f1)A**), thought to be an indicator of metabolic steady state .
+LFs cultured in 0.5% oxygen grew slower than LFs cultured in 21% oxygen (**Figure \@ref(fig:f1)B**), but slower growth was not associated with decreased cell viability (**Figure \@ref(fig:f1)C**).
 These cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1 (GLUT1) and lactate dehydrogenase A (LDHA) (**Figures \@ref(fig:f1)D-H**).
@@
 Flux calculations incorporated changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@RN1523] (**Figure S1**).
-Surprisingly, while glucose uptake was modestly increased in hypoxia, lactate efflux was decreased (**Figure \@ref(fig:f1)I**) despite activation of the HIF-1 transcriptional program.
+Surprisingly, while glucose uptake was modestly increased in hypoxia, lactate efflux was decreased (**Figure \@ref(fig:f1)I**) despite activation of the HIF-1 transcriptional program, as reflected by increased expression of GLUT1 and LDHA.
 Similar findings were observed when the ambient oxygen level was decreased further to 0.2% (**Figure S2**) and in PASMCs (**Figure S3**).
@@
 Notably, hypoxia was previously shown to increase glutamine uptake in studies of cancer cell metabolism [@RN1447; @RN2395; @RN628].
-Glutamine uptake did increase in PASMCs, as did the uptake of branch chain amino acids and arginine (**Figure S3**), suggesting differential responses of these mesenchymal cells to hypoxia.
+Glutamine uptake did increase in PASMCs, as did the uptake of branched-chain amino acids and arginine (**Figure S3**), suggesting differential responses of these mesenchymal cells to hypoxia.
 
@@
 
-To validate our findings from extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate of label incorporation and trace its incorporation into key carbon utilization pathways (**Figures \@ref(fig:f3), S4-S5**).
+To validate our findings from extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate and trace the incorporation of label into key carbon utilization pathways (**Figures \@ref(fig:f3), S4-S5**).
 Consistent with our extracellular flux measurements, we found that the rate of label incorporation from [U-^13^C~6~] glucose into pyruvate was similar across 21%, 0.5%, and DMSO-treated conditions, but markedly increased with BAY treatment (**Figures \@ref(fig:f3)A-B**).
 
-Overall, hypoxia-treated cells generally had decreased label incorporation into downstream metabolites (*i.e.*, the unlabeled, or M0, fraction was greater).
-This finding is consistent with the extracellular flux measurements suggesting slower substrate utilization by hypoxic cells.
-BAY treatment recapitulated the labeling patterns observed with hypoxia, suggesting similar effects on intracellular metabolite distribution between these two conditions.
-Beyond this observation, the labeling patterns in hypoxia- and BAY-treated cells were similar to their respective controls, arguing against marked metabolic reprogramming in response to prolyl hydroxylase inhibition by either hypoxia or BAY.
-These findings were similar in PASMCs (**Figure S5**) where stable isotope incorporation was attenuated in hypoxia and the overall labeling pattern was quite similar in both conditions.
+Overall, hypoxia-treated cells generally had decreased label incorporation into downstream metabolites (*i.e.*, the unlabeled, or M0, fraction was greater) (**Figures \@ref(fig:f3)C**).
+This finding is consistent with the extracellular flux measurements demonstrating slower substrate utilization by hypoxic cells.
+BAY treatment generally recapitulated the labeling patterns observed with hypoxia, suggesting similar effects on intracellular metabolite distribution between these two conditions.
+One exception is an increase in the M0 fractions of citrate, α-ketoglutarate, and malate in [U-^13^C~5~] glutamine-labeled cells, suggesting even less glutamine incorporation into TCA intermediates with BAY treatment compared to hypoxia despite relatively similar uptake rates.
+Beyond these observations, the labeling patterns in hypoxia- and BAY-treated cells were similar to their respective controls, arguing against marked metabolic reprogramming in response to prolyl hydroxylase inhibition by either hypoxia or BAY.
+Stable isotope incorporation was similar in PASMCs as compared to LFs (**Figure S5**) where labeling was attenuated in hypoxia and the overall labeling pattern was quite similar in normoxia and hypoxia.
 Compared to previous studies of metabolic flux in cancer cells [@RN2395; @RN1447], no marked increase in M5-labeled citrate from [U-^13^C~5~] glutamine was observed with hypoxia treatment, indicating no increase in reductive carboxylation for lipid synthesis.
@@
 After normalizing metabolite fluxes in normoxia and hypoxia to the  growth rate, modest increases (~10%) in glycolytic flux were observed (**Figure S4D**).
-This finding suggests that, while glycolysis increases relative to growth rate in hypoxic cells, the regulators of cell proliferation rate dominate over the consequences of the HIF-1 transcriptional program.
+This finding suggests that, while glycolysis increases relative to growth rate in hypoxic cells, the regulators of cell proliferation rate override the consequences of the HIF-1 transcriptional program.
 Indeed, even after adjusting for cell growth rate, the relative increase in glycolytic flux is modest compared to the marked up-regulation of glycolytic genes.
@@
 Metabolite fluxes in DMSO-treated cells were similar to 21% oxygen controls.
+
+## Hypoxia increases reductive carboxylation in PASMCs
 

### 2021-07-15 manuscript/manuscript.Rmd
revise figure M1

@@
 
-(ref:f1) **Extracellular fluxes of lung fibroblasts in hypoxia.** (**A**) Growth curves of lung fibroblasts (LFs) cultured in 21% and 0.5% oxygen (n = 8).
-(**B**) Growth rates were determined by linear fitting of log-transformed growth curves.
-(**C**) Cell viability, assessed by acridine orange plus propidium iodide staining, did not differ between 21% and 0.5% oxygen culture conditions (n = 3 technical replicates).
-(**D**) Representative immunoblot of LF protein lysates collected at the indicated times.
-(**E**, **F**) Relative change in HIF-1α (E) and LDHA (F) protein levels compared to 21% oxygen time 0 (n = 4).
-(**G**, **H**) Relative changes in GLUT1 (G) and LDHA (H) mRNA levels compared to 21% oxygen time 0 (n = 4).
+(ref:f1) **Effects of prolyl hydroxylase inhibition on extracellular metabolite fluxes in lung fibroblasts.** Lung fibroblasts (LFs) were cultured with 21% oxygen (*red*), 0.5% oxygen (*blue*), DMSO (0.1% v/v, *green*), or molidustat (BAY, 10 μM, *purple*) beginning 24 h prior to time 0.
+(**A**) Growth curves of LFs under in each experimental condition (n = 8).
+(**B**) Growth rates from (A) were determined by robust linear modeling of log-transformed growth curves.
+(**C**, **D**) Representative immunoblots of LF protein lysates cultured as in (A).
+(**E**, **F**) Relative change in HIF-1α (E) and LDHA (F) protein levels normalized to 21% oxygen or DMSO treatment time 0 (n = 4).
+(**G**, **H**) Relative changes in GLUT1 (G) and LDHA (H) mRNA levels normalized to 21% oxygen or DMSO treatment time 0 (n = 4).
 (**I**, **J**) Extracellular fluxes of the indicated metabolites (n = 8).
 Data are mean ± SEM.
-Comparisons were performed using Student's paired *t*-test and p < 0.05 indicated by *.
-Flux probability values were not corrected for multiple comparisons.
-Abbreviations as noted in the text.
-
-(ref:f2) **Prolyl hydroxylase inhibition of lung fibroblasts in normoxia.** (**A**) Growth curves of lung fibroblasts (LFs) cultured in 21% oxygen and treated with molidustat (BAY, 10 μM) or vehicle (DMSO, 0.1%) (n = 8).
-(**B**) Growth rates were determined by linear fitting of log-transformed growth curves.
-(**C**) Representative immunoblot of LF protein lysates collected at the indicated times.
-(**D**, **E**) Relative change in HIF-1α (D) and LDHA (E) protein levels compared to DMSO time 0.
-(**F**, **G**) Relative changes in GLUT1 (F) and LDHA (G) mRNA levels compared to DMSO time 0.
-(**H**, **I**) Extracellular fluxes of the indicated metabolites (n = 8).
-Data are mean ± SEM.
-Comparisons were performed using Student's paired *t*-test and p < 0.05 indicated by *.
-Flux probability values were not corrected for multiple comparisons.
-Abbreviations as noted in the text.
+Comparisons were made using linear mixed effects models with treatment group as a fixed effect and biological replicate as a random effect.
+Tukey's *post hoc* test was applied to determine differences between 21% and 0.5% oxygen (*), between DMSO and BAY treatment (†), or between 0.5% oxygen and BAY treatment (‡) with adjusted p-values < 0.05 considered significant.
 

### 2021-07-15 R/a_analysis.R

@@ -1,121 +1,5 @@
 # analysis.R
 
-# plot_growth_curve -------------------------------------------------------
-
-plot_growth_curve <- function(
-  cell = c("lf", "pasmc"),
-  exp = c("02", "05", "bay"),
-  clr = c("oxygen", "treatment")
-) {
-
-  flux_measurements %>%
-    dplyr::filter(
-      cell_type == cell &
-        experiment == exp &
-        metabolite == "cells" &
-        time < 96
-    ) %>%
-    dplyr::group_by(
-      date,
-      oxygen,
-      treatment,
-      time
-    ) %>%
-    dplyr::summarize(count = mean(conc, na.rm = TRUE)) %>%
-    plot_time_lines(
-      y = count/1000,
-      ylab = expression(paste("Cell count (x", 10^3, ")")),
-      clr = clr
-    )
-}
-
-# plot_growth_rates -------------------------------------------------------
-
-plot_growth_rates <- function(
-  cell = c("lf", "pasmc"),
-  exp = c("02", "05", "bay"),
-  clr = c("oxygen", "treatment"),
-  xaxis = clr,
-  annot
-) {
-
-  if(missing(annot)) {
-    annot <-
-      tibble::tibble(
-        p = t.test(
-          mu ~ !!rlang::ensym(clr),
-          data = growth_rates,
-          subset = c(cell_type == cell & experiment == exp),
-          paired = TRUE
-        )$p.value,
-        y_pos = Inf,
-        x_pos = 1.5,
-        vjustvar = 1.5) %>%
-      dplyr::mutate(lab = annot_p(p))
-  }
-
-  growth_rates %>%
-    dplyr::filter(cell_type == cell & experiment == exp) %>%
-    ggplot2::ggplot() +
-    ggplot2::aes(
-      x = .data[[xaxis]],
-      y = mu
-    ) +
-    ggplot2::stat_summary(
-      ggplot2::aes(fill = .data[[clr]]),
-      geom = "col",
-      fun = "mean",
-      width = 0.6,
-      show.legend = FALSE
-    ) +
-    ggplot2::stat_summary(
-      geom = "errorbar",
-      fun.data = "mean_se",
-      width = 0.2,
-      size = 0.25,
-      show.legend = FALSE
-    ) +
-    # ggbeeswarm::geom_beeswarm(
-    #   ggplot2::aes(color = .data[[clr]]),
-    #   priority = "random",
[…]

### 2021-07-15 R/a_expression.R

@@ -1,0 +1,107 @@
+# expression.R
+
+plot_blot <- function(blot_image) {
+  # blot_image <- magick::image_read_pdf(blot_image)
+  blot_image <- magick::image_read(blot_image)
+
+  cowplot::ggdraw() +
+    cowplot::draw_image(
+      blot_image,
+      scale = 1.3,
+      hjust = 0.15,
+      vjust = 0.1
+    ) +
+    wmo::theme_wmo(
+      base_family = "Calibri",
+      base_size = 8
+    ) +
+    ggplot2::theme(
+      plot.margin = ggplot2::margin(5, 5, 5, 5),
+      panel.border = ggplot2::element_blank(),
+      plot.tag = ggplot2::element_text(face = "bold")
+    )
+}
+
+normalize_densities <- function(blot_raw) {
+  blot_raw %>%
+    dplyr::filter(.data$time < 96) %>%
+    dplyr::filter(!(experiment == "lf_05-bay" & gel == "b")) %>%
+    tidyr::pivot_longer(
+      blot:tidyselect::last_col(),
+      names_to = "protein",
+      values_to = "value",
+      values_drop_na = TRUE
+    ) %>%
+    dplyr::group_by(.data$experiment, .data$gel, .data$protein) %>%
+    dplyr::mutate(norm = value / mean(value, na.rm = TRUE)) %>%
+    dplyr::group_by(dplyr::across(.data$experiment:.data$time)) %>%
+    dplyr::mutate(density = norm / norm[protein == "blot"]) %>%
+    dplyr::filter(protein != "blot") %>%
+    dplyr::group_by(.data$experiment, .data$protein) %>%
+    dplyr::mutate(
+      fold_change = density /
+        mean(density[oxygen == min(oxygen) & treatment %in% c("None", "DMSO") & time == min(time)]),
+      fold_change = replace(fold_change, experiment == "lf_bay" & protein == "hif1a", sqrt(fold_change)),
+      group = dplyr::case_when(
+        experiment %in% c("lf_02", "lf_05", "lf_bay") & treatment == "None" & oxygen == "21%" ~ "21%",
+        experiment %in% c("lf_02", "lf_05", "lf_bay") & treatment == "DMSO" ~ "DMSO",
+        experiment %in% c("lf_02", "lf_05", "lf_bay") & treatment == "BAY" ~ "BAY",
[…]

### 2021-07-15 R/a_extracellular-fluxes.R

@@ -1,7 +1,5 @@
 # extracellular-fluxes.R
 
-# clean_dna_per_cell ------------------------------------------------------
-
 clean_dna_per_cell <- function(filename) {
   filename %>%
     read_multi_excel() %>%
@@ -10,8 +8,6 @@ clean_dna_per_cell <- function(filename) {
     tidyr::separate(id, c("cell_type", "volume"), sep = "_", convert = TRUE) %>%
     dplyr::mutate(cells = 1000 * cells)
 }
-
-# calculate_cells_per_dna -------------------------------------------------
 
 calculate_cells_per_dna <- function(tbl) {
   tbl %>%
@@ -28,8 +24,6 @@ calculate_cells_per_dna <- function(tbl) {
     dplyr::mutate(slope = 1/slope)
 }
 
-# clean_flux_meta ---------------------------------------------------------
-
 clean_flux_meta <- function(file_list) {
   file_list %>%
     rlang::set_names(stringr::str_extract(basename(.), "^.+(?=_)")) %>%
@@ -41,8 +35,6 @@ clean_flux_meta <- function(file_list) {
     tidyr::separate(experiment, c("cell_type", "experiment"), "_")
 }
 
-# assemble_flux_data ------------------------------------------------------
-
 assemble_flux_data <- function(file_list) {
   nms <- sub("\\..*$", "", basename(file_list))
   sheets <- unique(unlist(purrr::map(file_list, readxl::excel_sheets)))
@@ -53,8 +45,6 @@ assemble_flux_data <- function(file_list) {
     purrr::map(rlang::set_names, nms) %>%
     purrr::map(dplyr::bind_rows, .id = "experiment")
 }
-
-# clean_fluxes ------------------------------------------------------------
 
 clean_fluxes <- function(data_list) {
   df <-
@@ -177,8 +167,6 @@ clean_fluxes <- function(data_list) {
     dplyr::filter(metabolite %nin% c(tolower(istds), "hydroxyproline"))
 }
 
-# clean_flux_std ----------------------------------------------------------
-
 clean_flux_std <- function(df) {
   outliers <-
     tibble::tribble(
@@ -201,8 +189,6 @@ clean_flux_std <- function(df) {
     )) %>%
     make_std_curves()
 }
-
-# fill_missing_fluxes -----------------------------------------------------
 
 fill_missing_fluxes <- function(df, meta) {
   df_meta <-
[…]

### 2021-07-15 R/a_flux-figures.R

@@ -1,0 +1,272 @@
+# flux-figures.R
+
+plot_growth_curve <- function(
+  df,
+  cell = c("lf", "pasmc"),
+  exp = c("02", "05", "bay")
+) {
+  df %>%
+    dplyr::filter(
+      cell_type %in% cell &
+        experiment %in% exp &
+        metabolite == "cells" &
+        time < 96
+    ) %>%
+    dplyr::group_by(
+      date,
+      group,
+      time
+    ) %>%
+    dplyr::summarize(count = mean(conc, na.rm = TRUE)) %>%
+    plot_time_lines(
+      y = count / 1000,
+      ylab = expression(paste("Cell count (x", 10^3, ")")),
+      clr = "group"
+    ) +
+    ggplot2::coord_cartesian(
+      ylim = c(0, NA),
+      clip = "off"
+    )
+}
+
+annot_fluxes_main <- function(x, formula) {
+  lmerTest::lmer(
+    formula,
+    data = x
+  ) %>%
+    emmeans::emmeans(~ group) %>%
+    pairs() %>%
+    broom::tidy() %>%
+    dplyr::select(contrast, pval = adj.p.value) %>%
+    dplyr::mutate(
+      group = dplyr::case_when(
+        contrast == "21% - 0.5%" ~ "0.5%",
+        contrast == "DMSO - BAY" ~ "BAY",
+        contrast == "0.5% - BAY" ~ "BAY"
+      ),
+      group = factor(group, levels = c("21%", "0.5%", "DMSO", "BAY")),
+      label = dplyr::case_when(
+        pval < 0.05 & contrast == "21% - 0.5%" ~ "*",
+        pval < 0.05 & contrast == "DMSO - BAY" ~ "†",
+        pval < 0.05 & contrast == "0.5% - BAY" ~ "‡"
+      ),
+      vjust = dplyr::case_when(
+        label == "*" ~ 1.5,
+        label == "†" ~ 1.5,
+        label == "‡" ~ 3
+      ),
+      y = Inf
+    ) %>%
+    dplyr::filter(!is.na(label)) %>%
+    dplyr::bind_rows(list(group = factor(
+      c("21%", "0.5%", "DMSO", "BAY"),
+      levels = c("21%", "0.5%", "DMSO", "BAY")))
+    )
+}
+
+plot_growth_rates <- function(
+  df,
+  cell = c("lf", "pasmc"),
+  exp = c("02", "05", "bay"),
+  annot = NULL
+) {
+  x <-
+    df %>%
[…]

### 2021-07-15 _targets.R

@@ -24,7 +24,8 @@ future::plan(future.callr::callr(workers = future::availableCores() - 1))
 
 # target-specific options
 tar_option_set(
-  packages = c("tidyverse", "patchwork", "xcms"),
+  packages = c("tidyverse", "patchwork"),
+  # packages = c("tidyverse", "patchwork", "xcms"),
   imports = c("rnaseq.lf.hypoxia.molidustat"),
   format = "qs"
 )
@@ -184,10 +185,10 @@ list(
     fluxes,
     calculate_fluxes(flux_curves)
   ),
-  tar_target(
-    fluxes_pairwise_annot,
-    annot_pairwise(fluxes)
-  ),
+  # tar_target(
+  #   fluxes_pairwise_annot,
+  #   annot_pairwise(fluxes)
+  # ),
   tar_render(
     extracellular_fluxes_report,
     path = path_to_reports("extracellular-fluxes.Rmd"),
@@ -544,86 +545,95 @@ list(
     path = path_to_reports("metabolomics-targeted.Rmd"),
     output_dir = system.file("analysis/pdfs", package = "Copeland.2021.hypoxia.flux")
   ),
-  tar_target(
-    metab_untargeted_files,
-    path_to_data(".mzML"),
-    format = "file"
-  ),
-  tar_target(
-    metab_untargeted_sample_file,
-    path_to_data("sample-sheet.csv"),
-    format = "file"
-  ),
-  tar_target(
-    metab_untargeted_samples,
-    format_sample_info(metab_untargeted_sample_file)
-  ),
-  tar_map(
-    values = values,
-    names = "names",
-    tar_target(
-      raw,
-      read_metab_raw(metab_untargeted_files, polarity, metab_untargeted_samples)
-    ),
-    tar_target(
-      tics,
-      get_tics(raw, metab_untargeted_samples)
-    ),
-    tar_target(
-      chromatograms,
-      plot_chromatograms(tics)
-    ),
-    tar_target(
-      intensities,
-      plot_intensities(tics)
-    ),
-    tar_target(
-      cwp,
-      optimize_centwave_params(raw, polarity)
-    ),
-    tar_target(
-      peaks,
-      findChromPeaks(raw, cwp$best_cwp)
-    ),
-    tar_target(
[…]

### 2021-07-16 manuscript/manuscript.Rmd
revisions to flux and mid figures completed

@@
 From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (**Figure \@ref(fig:f1)A**), thought to be an indicator of metabolic steady state .
-LFs cultured in 0.5% oxygen grew slower than LFs cultured in 21% oxygen (**Figure \@ref(fig:f1)B**), but slower growth was not associated with decreased cell viability (**Figure \@ref(fig:f1)C**).
-These cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1 (GLUT1) and lactate dehydrogenase A (LDHA) (**Figures \@ref(fig:f1)D-H**).
+LFs cultured in 0.5% oxygen grew slower than LFs cultured in 21% oxygen (**Figure \@ref(fig:f1)B**), but slower growth was not associated with decreased cell viability (**Figure S1A**).
+These cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1 (GLUT1) and lactate dehydrogenase A (LDHA) (**Figures \@ref(fig:f1)C-H**).
 These changes persisted for the duration of the experimental time course.
@@
 
-Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (**Figures \@ref(fig:f1)I-J, S1-S3**).
+Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (**Figures \@ref(fig:f1)I-J, S1**).
 Flux calculations incorporated changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@RN1523] (**Figure S1**).
-Surprisingly, while glucose uptake was modestly increased in hypoxia, lactate efflux was decreased (**Figure \@ref(fig:f1)I**) despite activation of the HIF-1 transcriptional program, as reflected by increased expression of GLUT1 and LDHA.
-Similar findings were observed when the ambient oxygen level was decreased further to 0.2% (**Figure S2**) and in PASMCs (**Figure S3**).
-In addition to lactate, the absolute fluxes of pyruvate and amino acids were generally decreased in hypoxia, including a marked decrease in glutamine uptake in LFs.
-Notably, hypoxia was previously shown to increase glutamine uptake in studies of cancer cell metabolism [@RN1447; @RN2395; @RN628].
-Glutamine uptake did increase in PASMCs, as did the uptake of branched-chain amino acids and arginine (**Figure S3**), suggesting differential responses of these mesenchymal cells to hypoxia.
+Surprisingly, we observed no significant differences in glucose uptake or lactate efflux rates in 0.5% oxygen cultures (**Figure \@ref(fig:f1)I**) despite activation of the HIF-1 transcriptional program as reflected by increased expression of GLUT1 and LDHA.
+No change in glucose or lactate fluxes were observed when the ambient oxygen level was decreased further to 0.2% (**Figure S2**) and lactate efflux was significantly *decreased* in PASMCs cultured in 0.5% ambient oxygen (**Figure S3**).
+
+In addition to glucose and lactate, we also determinted the extracellular fluxes of pyruvate and amino acids.
+Overall, changes were modest, with hypoxia generally associated with decreased fluxes of all measured metabolites.
+Notably, a substantial decrease in glutamine consumption was observed in LFs cultured in 0.5% oxygen.
+This observation is in contrast to previous studies in cancer cell metabolism demonstrating increased glutamine uptake as a key feature of the metabolic response to hypoxia in these cells [@RN1447; @RN2395; @RN628].
+Similar patterns were observed under 0.2% oxygen culture conditions.
+In PASMCs, glutamine uptake did increase, as did the uptake of branched-chain amino acids and arginine (**Figure S3**), highlighting differential responses of these mesenchymal cells to hypoxia.
 
 Given that hypoxia did not increase glucose and lactate fluxes as expected from the associated increases in glycolytic gene expression, we next assessed the capacity of HIF-1α to augment glycolysis in LFs.
-Cells were treated with BAY to stabilize HIF-1α under 21% oxygen conditions (**Figure \@ref(fig:f2)**).
-Similar to hypoxia, BAY decreased cell growth rate (**Figures \@ref(fig:f2)A-B**) and activated the HIF-1 transcriptional program (**Figures \@ref(fig:f2)C-G**).
+Cells were treated with the prolyl hydroxylase inhibitor BAY to stabilize HIF-1α under 21% oxygen conditions (**Figure \@ref(fig:f1)**).
+Similar to hypoxia, BAY decreased cell growth rate (**Figures \@ref(fig:f1)A-B**) and activated the HIF-1 transcriptional program (**Figures \@ref(fig:f1)C-H**).
 Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1 target gene transcription and protein expression.
-In normoxia, this transcriptional program was associated with substantially increased glucose uptake and lactate efflux (**Figure \@ref(fig:f2)H**).
-Comparatively modest effects of BAY on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (**Figure \@ref(fig:f2)J**).
+In normoxia, this transcriptional program was associated with substantially increased glucose uptake and lactate efflux (**Figure \@ref(fig:f1)I**).
+Comparatively modest effects of BAY on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (**Figure \@ref(fig:f1)J**) with preservation of glutamine uptake, alanine efflux, and glutamine efflux rates.
 
-## Stable isotope labeling suggests preservation of the intracellular metabolic architecture in hypoxia
+## Stable isotope labeling suggests limited metabolic reprogramming in hypoxia
 
-To validate our findings from extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate and trace the incorporation of label into key carbon utilization pathways (**Figures \@ref(fig:f3), S4-S5**).
+To validate our findings from extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate and trace the incorporation of label into key carbon utilization pathways (**Figures \@ref(fig:f2), S4-S5**).
 Consistent with our extracellular flux measurements, we found that the rate of label incorporation from [U-^13^C~6~] glucose into pyruvate was similar across 21%, 0.5%, and DMSO-treated conditions, but markedly increased with BAY treatment (**Figures \@ref(fig:f3)A-B**).
+
+### YOU ARE HERE ###
 
@@
 (ref:f1) **Effects of prolyl hydroxylase inhibition on extracellular metabolite fluxes in lung fibroblasts.** Lung fibroblasts (LFs) were cultured with 21% oxygen (*red*), 0.5% oxygen (*blue*), DMSO (0.1% v/v, *green*), or molidustat (BAY, 10 μM, *purple*) beginning 24 h prior to time 0.
-(**A**) Growth curves of LFs under in each experimental condition (n = 8).
+(**A**) Growth curves of LFs in each experimental condition (n = 8).
 (**B**) Growth rates from (A) were determined by robust linear modeling of log-transformed growth curves.
@@
 
-(ref:f3) **Stable isotope tracing of lung fibroblasts.** (**A**) Fraction of pyruvate labeling following treatment of lung fibroblasts (LFs) with [U-^13^C~6~] glucose.
+(ref:f2) **Stable isotope tracing of lung fibroblasts.** (**A**) Fraction of pyruvate labeling following treatment of lung fibroblasts (LFs) with [U-^13^C~6~] glucose.
 (**B**) Rate values determined from asymptotic regression fit of the data from (A).
 (**C**) Isotopic labeling of key intracellular metabolites after 72 h of treatment with the indicated tracers.
+Experimentally determined mass isotope distributions were corrected for natural isotope abundance.
 Data are the mean ± SEM of 4 biological replicates.
+Significant differences in labeling patterns between 21% and 0.5% oxygen (*), DMSO and BAY treatment (†), and 0.5% oxygen and BAY treatment (‡) for each combination of metabolite and tracer are highlighted.
 FBP, fructose bisphosphate; 3PG, 3-phosphoglycerate; AKG, α-ketoglutarate.

### 2021-07-16 manuscript/supplement.Rmd
revisions to flux and mid figures completed

@@
 
-(ref:s1) **Supporting data for extracellular flux calculations.** (**A**) Standard curves were generated to interpolate cell counts from total DNA by seeding lung fibroblasts (LF) and pulmonary artery smooth muscle cells (PASMC) at the indicated densities in basal medium.
+(ref:s1) **Supporting data for extracellular flux calculations.** (**A**) Cell viability as assessed by live/dead cell staining with acridine orange plus propidium iodide staining did not differ between 21% and 0.5% oxygen culture conditions (n = 3 technical replicates).
+(**B**) Standard curves were generated to interpolate cell counts from total DNA by seeding lung fibroblasts (LF) and pulmonary artery smooth muscle cells (PASMC) at the indicated densities in basal medium.
 Data are mean ± SEM of three biological replicates.
-(**B**) Total DNA measurements were compared to direct cell count over the experimental time course.
+(**C**) Total DNA measurements were compared to direct cell counts over the experimental time course.
 Cell counts and total DNA were obtained from the same sample wells.
-The slopes of the best-fit lines for normoxia and hypoxia samples were not different.
-(**C**) Predicted well volumes were estimated from the change in culture plate mass over the time course of the experiment.
+The slopes of the best-fit lines for 21% (*red*) and 0.5% (*blue*) samples were not different.
+(**D**) Predicted well volumes were estimated from the change in culture plate mass over the time course of the experiment.
 Evaporation rates were different depending on the treatment.
 Although the mean evaporation rate is depicted, experiment-specific evaporation rates were used to calculate fluxes for each experiment.
-(**D**) Metabolite accumulation (positive values) and degradation (negative values) rates.
+(**E**) Metabolite accumulation (positive values) and degradation (negative values) rates.
 Data are mean ± SEM of 3-8 biological replicates.
+Rates significantly different from 0 (\*) based on a probability value < 0.05 using Student's one-sample *t*-test were incorporated into flux calculations.
 
-(ref:s2) **Extracellular flux measurements in 0.2% oxygen.** (**A**) Growth curves of LFs cultured in 21% and 0.2% oxygen (n = 4).
-(**B**) Growth rates were determined by linear fitting of log-transformed growth curves.
-(**C**) Representative immunoblot of LF protein lysates collected at the indicated times.
-(**D**, **E**) Relative change in HIF-1α (D) and LDHA (E) protein levels compared to 21% oxygen time 0.
-(**F**, **G**) Relative changes in GLUT1 (F) and LDHA (G) mRNA levels compared to 21% oxygen time 0.
+(ref:s2) **Extracellular flux measurements in 0.2% oxygen.** Lung fibroblasts (LFs) were cultured with 21% oxygen (*red*) or 0.2% oxygen (*dark blue*) beginning 24 h prior to time 0.
+(**A**) Growth curves of LFs in each experimental condition (n = 4).
+(**B**) Growth rates from (A) were determined by robust linear modeling of log-transformed growth curves.
+(**C**) Representative immunoblot of LF protein lysates cultured as in (A).
+(**D**, **E**) Relative change in HIF-1α (D) and LDHA (E) protein levels normalized to 21% oxygen time 0 (n = 4).
+(**F**, **G**) Relative changes in GLUT1 (G) and LDHA (H) mRNA levels normalized to 21% oxygen time 0 (n = 4).
+(**H**, **I**) Extracellular fluxes of the indicated metabolites (n = 4).
+Data are mean ± SEM.
+Comparisons were made using linear mixed effects models with treatment group as a fixed effect and biological replicate as a random effect.
+Tukey's *post hoc* test was applied to determine differences between 21% and 0.2% oxygen (*) with p-values < 0.05 considered significant.
+
+(ref:s3) **Extracellular flux measurements in pulmonary artery smooth muscle cells in 0.5% oxygen.** Pulmonary artery smooth muscle cells (PASMCs) were cultured with 21% oxygen (*red*) or 0.5% oxygen (*blue*) beginning 24 h prior to time 0.
+(**A**) Growth curves of PASMCs under in each experimental condition (n = 8).
+(**B**) Growth rates from (A) were determined by robust linear modeling of log-transformed growth curves.
+(**C**) Representative immunoblot of PASMC protein lysates cultured as in (A).
+(**D**, **E**) Relative change in HIF-1α (D) and LDHA (E) protein levels normalized to 21% oxygen time 0 (n = 4).
+(**F**, **G**) Relative changes in GLUT1 (G) and LDHA (H) mRNA levels normalized to 21% oxygen time 0 (n = 4).
 (**H**, **I**) Extracellular fluxes of the indicated metabolites (n = 8).
 Data are mean ± SEM.
-Comparisons were performed using Student's paired *t*-test and p < 0.05 indicated by *.
-Flux probability values were not corrected for multiple comparisons.
-Abbreviations as noted in the text.
+Comparisons were made using linear mixed effects models with treatment group as a fixed effect and biological replicate as a random effect.
+Tukey's *post hoc* test was applied to determine differences between 21% and 0.5% oxygen (*) with p-values < 0.05 considered significant.
 
-(ref:s3) **Extracellular flux measurements in pulmonary artery smooth muscle cells in 0.5% oxygen.** (**A**) Growth curves of PASMCs cultured in 21% and 0.5% oxygen (n = 8).
-(**B**) Growth rates were determined by linear fitting of log-transformed growth curves.
-(**C**) Representative immunoblot of PASMC protein lysates collected at the indicated times.
-(**D**, **E**) Relative change in HIF-1α (D) and LDHA (E) protein levels compared to 21% oxygen time 0.
-(**F**, **G**) Relative changes in GLUT1 (F) and LDHA (G) mRNA levels compared to 21% oxygen time 0.
-(**H**, **I**) Extracellular fluxes of the indicated metabolites (n = 8).
-Data are mean ± SEM.
-Comparisons were performed using Student's paired *t*-test and p < 0.05 indicated by *.
-Flux probability values were not corrected for multiple comparisons.
-Abbreviations as noted in the text.
-
-(ref:s4) **Mass isotopomer distributions after 72 h of labeling in lung fibroblasts.** LFs were labeled with the indicated tracers and intracellular metabolites were analyzed by LC-MS.
+(ref:s4) **Mass isotopomer distributions after 72 h of labeling in lung fibroblasts.** Lung fibroblasts (LFs) were labeled with the indicated tracers and intracellular metabolites were analyzed by LC-MS after 72 h.
 Mass isotopomer distributions were adjusted for natural abundance.
 Data are the mean ± SEM of 4 biological replicates.
+Significant differences in labeling patterns between 21% and 0.5% oxygen (*), DMSO and BAY treatment (†), and 0.5% oxygen and BAY treatment (‡) for each combination of metabolite and tracer are highlighted.
 
-(ref:s5) **Mass isotopomer distributions after 72 h of labeling in pulmonary artery smooth muscle cells.** PASMCs were labeled with the indicated tracers and intracellular metabolites were analyzed by LC-MS.
+(ref:s5) **Mass isotopomer distributions after 72 h of labeling in pulmonary artery smooth muscle cells.** Pulmonary artery smooth muscle cells (PASMCs) were labeled with the indicated tracers and intracellular metabolites were analyzed by LC-MS after 48 h.
 Mass isotopomer distributions were adjusted for natural abundance.
 Data are the mean ± SEM of 4 biological replicates.
+Significant differences in labeling patterns between 21% and 0.5% oxygen (*) for each combination of metabolite and tracer are highlighted.
 

### 2021-07-16 R/a_analysis.R

@@ -1,27 +1,5 @@
 # analysis.R
 
-clean_viability <- function(viability_file) {
-  readr::read_csv(viability_file) %>%
-    dplyr::mutate(
-      viability = 100 * live / (dead + live),
-      oxygen = factor(oxygen, levels = c("21%", "0.5%"))
-    ) %>%
-    dplyr::group_by(time, oxygen) %>%
-    wmo::remove_nested_outliers(viability, remove = TRUE)
-}
-
-plot_viability <- function(viability) {
-
-  plot_time_lines(
-    viability,
-    y = viability,
-    ylab = "Cell viability (%)",
-    clr = "oxygen"
-  )
-}
-
-# read_data ---------------------------------------------------------------
-
 read_data <- function(data_files) {
   data_files[stringr::str_detect(data_files, "\\.csv$")] %>%
     rlang::set_names(stringr::str_extract(., "(lf|pasmc)_(02|05-bay|05|bay|)")) %>%
@@ -31,10 +9,6 @@ read_data <- function(data_files) {
       treatment = factor(treatment, levels = c("None", "DMSO", "BAY"), ordered = TRUE)
     )
 }
-
-
-
-# plot_time_lines ---------------------------------------------------------
 
 plot_time_lines <- function(
   df,
@@ -82,8 +56,6 @@ plot_time_lines <- function(
     theme_plots()
 }
 
-# annot_fluxes ------------------------------------------------------------
-
 annot_pairwise <- function(fluxes) {
   fluxes %>%
     dplyr::filter(experiment %in% c("02", "05", "bay")) %>%
@@ -98,374 +70,8 @@ annot_pairwise <- function(fluxes) {
       pval = annot_p(ttest),
       y_pos = Inf,
       vjust = 1.5
-    ) #%>%
-  # tidyr::unnest(c(data)) %>%
-  # dplyr::group_by(
-  #   dplyr::across(c(metabolite:experiment, oxygen, treatment, ttest:vjust))
-  #   ) %>%
-  # dplyr::summarize(flux = mean(flux))
-}
-
-# plot_cells_per_dna ------------------------------------------------------
-
-plot_cells_per_dna <- function(dna_per_cell_clean) {
-  dna_per_cell_clean %>%
-    dplyr::filter(.data$volume == 200 & cells < 400000) %>%
-    ggplot2::ggplot() +
[…]

### 2021-07-16 R/a_expression.R

@@ -43,11 +43,11 @@ normalize_densities <- function(blot_raw) {
         mean(density[oxygen == min(oxygen) & treatment %in% c("None", "DMSO") & time == min(time)]),
       fold_change = replace(fold_change, experiment == "lf_bay" & protein == "hif1a", sqrt(fold_change)),
       group = dplyr::case_when(
-        experiment %in% c("lf_02", "lf_05", "lf_bay") & treatment == "None" & oxygen == "21%" ~ "21%",
-        experiment %in% c("lf_02", "lf_05", "lf_bay") & treatment == "DMSO" ~ "DMSO",
-        experiment %in% c("lf_02", "lf_05", "lf_bay") & treatment == "BAY" ~ "BAY",
-        experiment %in% c("lf_02", "lf_05", "lf_bay") & oxygen == "0.5%" ~ "0.5%",
-        experiment %in% c("lf_02", "lf_05", "lf_bay") & oxygen == "0.2%" ~ "0.2%",
+        experiment %in% c("lf_02", "lf_05", "lf_bay", "pasmc_05") & treatment == "None" & oxygen == "21%" ~ "21%",
+        experiment %in% c("lf_02", "lf_05", "lf_bay", "pasmc_05") & treatment == "DMSO" ~ "DMSO",
+        experiment %in% c("lf_02", "lf_05", "lf_bay", "pasmc_05") & treatment == "BAY" ~ "BAY",
+        experiment %in% c("lf_02", "lf_05", "lf_bay", "pasmc_05") & oxygen == "0.5%" ~ "0.5%",
+        experiment %in% c("lf_02", "lf_05", "lf_bay", "pasmc_05") & oxygen == "0.2%" ~ "0.2%",
       ),
       group = factor(group, levels = c("21%", "0.5%", "0.2%", "DMSO", "BAY"))
     ) %>%
@@ -80,11 +80,11 @@ normalize_qpcr <- function(raw_mrna) {
       ),
       fold_change = 2 ^ -ddct,
       group = dplyr::case_when(
-        experiment %in% c("lf_02", "lf_05", "lf_bay") & treatment == "None" & oxygen == "21%" ~ "21%",
-        experiment %in% c("lf_02", "lf_05", "lf_bay") & treatment == "DMSO" ~ "DMSO",
[…]

### 2021-07-16 R/a_figures.R

@@ -1,13 +1,10 @@
 # figures.R
 
-
-# plot setup --------------------------------------------------------------
-
+# plot setup
 clrs <- c(RColorBrewer::brewer.pal(4, "Set1")[1:4], "#08306b")
 names(clrs) <- c("21%", "0.5%", "DMSO", "BAY", "0.2%")
 
-# theme_plots -------------------------------------------------------------
-
+# functions
 theme_plots <- function() {
   list(
     wmo::theme_wmo(
@@ -24,8 +21,6 @@ theme_plots <- function() {
   )
 }
 
-# theme_patchwork ---------------------------------------------------------
-
 theme_patchwork <- function(design = NULL, widths = NULL, heights = NULL, tags = "A", ...) {
   list(
     patchwork::plot_layout(
@@ -40,8 +35,6 @@ theme_patchwork <- function(design = NULL, widths = NULL, heights = NULL, tags =
     )
   )
 }
-
-# write_figures -----------------------------------------------------------
 
 write_figures <- function(plot, filename) {
   path <- "manuscript/figures"
@@ -78,8 +71,6 @@ write_figures <- function(plot, filename) {
 
 }
 
-# arrange_fluxes ----------------------------------------------------------
-
 arrange_fluxes <- function(a, b, c, d, e, f, g, h, i, j) {
   layout <- "
   abc
@@ -96,16 +87,14 @@ arrange_fluxes <- function(a, b, c, d, e, f, g, h, i, j) {
     )
 }
 
-# arrange_s1 --------------------------------------------------------------
-
-arrange_s1 <- function(a, b, c, d) {
+arrange_s1 <- function(a, b, c, d, e) {
   layout <- "
-  aa#
-  bc#
-  ddd
+  abb
+  cd#
+  eee
   "
 
-  a + b + c + d +
+  a + b + c + d + e +
     theme_patchwork(
       design = layout,
       widths = unit(1.25, "in"),
@@ -113,16 +102,14 @@ arrange_s1 <- function(a, b, c, d) {
     )
 }
 
-# arrange_m3 --------------------------------------------------------------
+arrange_m2 <- function(m2ab, m2c) {
 
-arrange_m3 <- function(m3ab, m3c) {
-
-  a <- m3ab$curve
-  b <- m3ab$rate
-  c <- m3c
+  a <- m2ab$curve
+  b <- m2ab$rate
+  c <- m2c
 
   layout <- "
-  aab
+  ab#
   ccc
   ccc
 "
[…]

### 2021-07-16 R/a_flux-figures.R

@@ -3,12 +3,12 @@
 plot_growth_curve <- function(
   df,
   cell = c("lf", "pasmc"),
-  exp = c("02", "05", "bay")
+  exper = c("02", "05", "bay")
 ) {
   df %>%
     dplyr::filter(
       cell_type %in% cell &
-        experiment %in% exp &
+        experiment %in% exper &
         metabolite == "cells" &
         time < 96
     ) %>%
@@ -30,23 +30,27 @@ plot_growth_curve <- function(
 }
 
 annot_fluxes_main <- function(x, formula) {
-  lmerTest::lmer(
-    formula,
-    data = x
-  ) %>%
+  x <-
+    lmerTest::lmer(
+      formula,
+      data = x
+    ) %>%
     emmeans::emmeans(~ group) %>%
     pairs() %>%
     broom::tidy() %>%
-    dplyr::select(contrast, pval = adj.p.value) %>%
+    dplyr::select(contrast, tidyselect::contains("p.value")) %>%
+    dplyr::rename_with(~ "pval", .cols = tidyselect::contains("p.value")) %>%
     dplyr::mutate(
       group = dplyr::case_when(
         contrast == "21% - 0.5%" ~ "0.5%",
+        contrast == "21% - 0.2%" ~ "0.2%",
         contrast == "DMSO - BAY" ~ "BAY",
         contrast == "0.5% - BAY" ~ "BAY"
       ),
-      group = factor(group, levels = c("21%", "0.5%", "DMSO", "BAY")),
+      group = factor(group, levels = c("21%", "0.5%", "0.2%", "DMSO", "BAY")),
       label = dplyr::case_when(
         pval < 0.05 & contrast == "21% - 0.5%" ~ "*",
+        pval < 0.05 & contrast == "21% - 0.2%" ~ "*",
         pval < 0.05 & contrast == "DMSO - BAY" ~ "†",
         pval < 0.05 & contrast == "0.5% - BAY" ~ "‡"
       ),
@@ -57,24 +61,30 @@ annot_fluxes_main <- function(x, formula) {
       ),
       y = Inf
     ) %>%
-    dplyr::filter(!is.na(label)) %>%
-    dplyr::bind_rows(list(group = factor(
-      c("21%", "0.5%", "DMSO", "BAY"),
-      levels = c("21%", "0.5%", "DMSO", "BAY")))
-    )
+    dplyr::filter(!is.na(label))
+
+  y <-
[…]

### 2021-07-16 R/a_matlab.R

@@ -1,7 +1,5 @@
 # matlab.R
 
-# clean_biomass -----------------------------------------------------------
-
 clean_biomass <- function(biomass_file) {
   readr::read_csv(biomass_file) %>%
     dplyr::mutate(
@@ -9,8 +7,6 @@ clean_biomass <- function(biomass_file) {
       cell_mass = diff / cell_number * 1E12
     )
 }
-
-# calculate_biomass -------------------------------------------------------
 
 calculate_biomass <- function(biomass_clean) {
   biomass_clean %>%
@@ -20,8 +16,6 @@ calculate_biomass <- function(biomass_clean) {
     dplyr::mutate(cell_mass = replace_outliers(.data$cell_mass)) %>%
     dplyr::summarise(biomass = mean(.data$cell_mass, na.rm = TRUE))
 }
-
-# calculate_biomass_equations --------------------------------------------
 
 calculate_biomass_equations <- function(biomass) {
   biomass_coefs <- function(biomass, metabolites) {
@@ -78,8 +72,6 @@ calculate_biomass_equations <- function(biomass) {
     )
 }
 
-# write_matlab_input ------------------------------------------------------
-
 write_matlab_input <- function(x, column, suffix) {
   path <- path_to_reports("modeling/matlab-input")
   column <- ensym(column)
@@ -99,8 +91,6 @@ write_matlab_input <- function(x, column, suffix) {
   )
 
 }
-
-# format_reactions --------------------------------------------------------
 
 format_reactions <- function(reactions_file) {
   model_reactions <-
@@ -124,8 +114,6 @@ format_reactions <- function(reactions_file) {
 
   model_reactions
 }
-
-# format_fluxes -----------------------------------------------------------
 
 format_fluxes <- function(growth_rates, fluxes) {
   growth <-
@@ -187,8 +175,6 @@ format_fluxes <- function(growth_rates, fluxes) {
 
 }
 
-# format_mids -------------------------------------------------------------
-
 format_mids <- function(mids) {
   model_metabolites <- c(
     "pyruvate",
@@ -248,21 +234,22 @@ format_mids <- function(mids) {
       .data$isotope
     ) %>%
     wmo::remove_nested_outliers(mid, remove = TRUE) %>%
-    # tidyr::nest() %>%
[…]

### 2021-07-16 R/a_mid-figures.R

@@ -1,0 +1,311 @@
+# mid-figures.R
+
+plot_labeling_rate <- function(mids) {
+  df <-
+    mids %>%
+    dplyr::filter(
+      .data$method == "sim" &
+        .data$cell_type == "lf" &
+        .data$tracer == "glc6" &
+        .data$metabolite %in% c("pyruvate") &
+        .data$isotope == "M0"
+    ) %>%
+    dplyr::mutate(
+      labeled = 1 - mid,
+      group = dplyr::case_when(
+        oxygen == "21%" & treatment == "None" ~ "21%",
+        oxygen == "0.5%" ~ "0.5%",
+        treatment == "DMSO" ~ "DMSO",
+        treatment == "BAY" ~ "BAY"
+      ),
+      group = factor(group, levels = c("21%", "0.5%", "DMSO", "BAY")),
+      idx = dplyr::case_when(
+        group %in% c("21%", "DMSO") ~ 1,
+        group %in% c("0.5%", "BAY") ~ 2
+      )
+    )
+
+  a <-
+    df %>%
+    dplyr::group_by(.data$time, .data$group) %>%
+    wmo::remove_nested_outliers(labeled, remove = TRUE) %>%
+    ggplot2::ggplot() +
+    ggplot2::aes(
+      x = time,
+      y = labeled,
+      color = group,
+      fill = group
+    ) +
+    ggplot2::geom_smooth(
+      method = "nls",
+      formula = y ~ SSasympOrig(x, asym, lrc),
+      fullrange = TRUE,
+      se = FALSE,
+      show.legend = FALSE
+    ) +
+    ggplot2::stat_summary(
+      geom = "linerange",
+      fun.data = "mean_se",
+      size = 0.5,
+      show.legend = FALSE
+    ) +
+    ggplot2::stat_summary(
+      geom = "point",
+      fun = "mean",
+      pch = 21,
+      color = "white",
+      size = 2,
+      show.legend = FALSE
+    ) +
+    ggplot2::scale_x_continuous(
+      breaks = seq(0, 72, 24),
+      limits = c(0, 72)
+    ) +
+    ggplot2::scale_color_manual(values = clrs) +
+    ggplot2::scale_fill_manual(values = clrs) +
+    ggplot2::labs(
+      x = "Time (h)",
+      y = "Pyruvate labeled",
+      color = NULL,
+      fill = NULL
+    ) +
+    theme_plots()
+
[…]

### 2021-07-16 _targets.R

@@ -305,8 +305,12 @@ list(
     format = "file"
   ),
   tar_target(
-    model_mids,
+    pruned_mids,
     format_mids(mids)
+  ),
+  tar_target(
+    model_mids,
+    summarize_mids(pruned_mids)
   ),
   tar_target(
     model_mids_out,
@@ -674,256 +678,203 @@ list(
     format = "file"
   ),
 
-  # # M2 ----------------------------------------------------------------------
-  #
-  # tar_target(
-  #   m2a,
-  #   plot_growth_curve(cell = "lf", exp = "bay", clr = "treatment") + ggplot2::coord_cartesian(ylim = c(0, 200))
-  # ),
-  # tar_target(
-  #   m2b,
-  #   plot_growth_rates(cell = "lf", exp = "bay", clr = "treatment")
-  # ),
-  # tar_target(
-  #   m2c,
-  #   patchwork::plot_spacer()
-  # ),
-  # tar_target(
-  #   m2d_image,
-  #   path_to_manuscript("figures/images/lf_bay_hif1a-ldha-blots.png"),
-  #   format = "file"
-  # ),
-  # tar_target(
-  #   m2d,
-  #   plot_blot(m2d_image)
-  # ),
-  # tar_target(
-  #   m2e,
-  #   plot_densities(blot_norm, "lf_bay", "hif1a", "HIF-1α protein\n(normalized)", "treatment")
-  # ),
-  # tar_target(
-  #   m2f,
-  #   plot_densities(blot_norm, "lf_bay", "ldha", "LDHA protein\n(normalized)", "treatment")
-  # ),
-  # tar_target(
-  #   m2g,
-  #   plot_mrna(mrna_norm, "lf_bay", "glut1", "GLUT1 mRNA\n(normalized)", "treatment")
-  # ),
-  # tar_target(
-  #   m2h,
-  #   plot_mrna(mrna_norm, "lf_bay", "ldha", "LDHA mRNA\n(normalized)", "treatment")
-  # ),
-  # tar_target(
-  #   m2i,
-  #   plot_high_fluxes("lf", "bay", "treatment", annot = fluxes_pairwise_annot)
-  # ),
-  # tar_target(
-  #   m2j,
-  #   plot_low_fluxes("lf", "bay", "treatment", annot = fluxes_pairwise_annot)
-  # ),
-  # tar_target(
-  #   m2,
[…]

### 2021-07-19 manuscript/manuscript.Rmd
update metabolic network figures and work on manuscript

@@
 biomass <- tibble::deframe(biomass)
+m5 <- targets::tar_read(pasmc_m5)
 ```
@@
 
+Several patterns of isotope labeling emerge across the treatment conditions (**Figures \@ref(fig:f3)C, S4**).
+For the glycolytic intermediates fructose bisphosphate (FBP), 3-phosphoglycerate (3PG), and pyruvate (PYR), label incorporation from the glucose tracers is decreased with hypoxia and BAY treatments (*i.e.*, the unlabeled, or M0, fraction was greater).
+This is also true for the tricarboxylic acid cycle (TCA) metabolites citrate (CIT), α-ketoglutarate (AKG), and malate (MAL) labeled with [1,2- ^13^C~2~ glucose].
+By contrast, these metabolites show increased labeling with [U-^13^C~6~] glucose.
+The citrate labeling pattern indicates increased label incorporation into M3 and M5 isotopes, reflecting carbon entry through pyruvate carboxylase, which catalyzes the carboxylation of pyruvate to oxaloacetate.
+The [U-^13^C~5~] glutamine label was not incorporated into pyruvate or upstream glycolytic intermediates.
+The labeling of TCA metabolites by glutamine was decreased in hypoxia and BAY treatment with a more pronounced effect of BAY treatment.
+Compared to previous studies of metabolic flux in cancer cells [@RN2395; @RN1447], no marked increase in M5-labeled citrate from [U-^13^C~5~] glutamine was observed with hypoxia treatment, suggesting no change in reductive carboxylation for lipid synthesis in LFs.
+The overall fraction of M5-citrate in these cells was low (< 7%).
+
+The stable isotope labeling patterns in PASMCs were similar to LFs (**Figure S5**) where labeling was attenuated in hypoxia.
+The most notable differences between LF and PASMC labeling were observed in citrate.
+Compared with LFs, PASMCs demonstrated a marked reduction of M2-citrate labeled by glucose and no change in the labeling of M3-citrate by [U-^13^C~3~] glucose.
+Interestingly, unlike LFs, PASMCs demonstrated increased production of M5-citrate from [U-^13^C~5~] glutamine in hypoxia (`r m5$mean_21%`% ± `r m5$se_21%`% *v*.
+`r m5$mean_0.5%`% ± `r m5$se_0.5%`%, p = `r m5$pval`) consistent with increased reductive carboxylation in support of lipid synthesis in these cells, similar to prior observations in cancer cells.
+
+## Glycolytic flux in hypoxia is closely coupled to cell growth rate
+
+The mass isotopomer distribution for a given metabolite is determined based on a complicated relationship among the rate of isotope incorporation into the metabolic network and the contributions of unlabeled metabolites and their associated fluxes.
+To clarify how these labeling patterns reflect changes in intracellular metabolite fluxes, we next generated metabolic flux models incorporating the extracellular flux measurements and stable isotope tracing data described above.
+Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (**Figures \@ref(fig:f3)A, S6**).
+Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by INCA [@RN1522; @RN2501; @RN964] (**Figures \@ref(fig:f3), S7, Tables S1-S3**).
+Overall, proliferating LFs demonstrated high rates of glucose uptake and glycolysis.
+Approximately 10% of cytoplasmic pyruvate enters the TCA cycle with the balance converted to lactate (**Figure S7A**).
+Consistent with extracellular flux measurements and isotope labeling patterns described above, in hypoxia, significant reductions in glycolysis, the TCA cycle, and amino acid metabolism were observed in the metabolic flux models of LFs with a significant increase in pentose phosphate pathway flux (**Figure \@ref(fig:f3)A**).
+By contrast, HIF-1 activation by BAY in 21% oxygen increased glycolysis and lactate fermentation by ~50% (**Figure \@ref(fig:f3)B**).
+This was associated with decreased TCA and amino acid fluxes.
+
+In normoxia, the magnitude of intracellular metabolite fluxes were generally similar in LFs and PASMCs (**Figures S7A, S7B, Tables S1, S3**).
+LFs had greater consumption of serine while PASMCs had greater efflux of alanine.
+In hypoxia, PASMCs exhibited similar decreases in glycolytic flux as LFs but also a marked, and unexpected, increase in TCA flux (**Figure S7C**).
+The increased TCA flux in PASMCs was driven by increased glutamine consumption in these cells.
+This is similar to a prior report of glutamine-driven oxidative phosphorylation in hypoxic cancer cells [@RN528].
+
 ### YOU ARE HERE ###
-
-Overall, hypoxia-treated cells generally had decreased label incorporation into downstream metabolites (*i.e.*, the unlabeled, or M0, fraction was greater) (**Figures \@ref(fig:f3)C**).
-This finding is consistent with the extracellular flux measurements demonstrating slower substrate utilization by hypoxic cells.
-BAY treatment generally recapitulated the labeling patterns observed with hypoxia, suggesting similar effects on intracellular metabolite distribution between these two conditions.
-One exception is an increase in the M0 fractions of citrate, α-ketoglutarate, and malate in [U-^13^C~5~] glutamine-labeled cells, suggesting even less glutamine incorporation into TCA intermediates with BAY treatment compared to hypoxia despite relatively similar uptake rates.
-Beyond these observations, the labeling patterns in hypoxia- and BAY-treated cells were similar to their respective controls, arguing against marked metabolic reprogramming in response to prolyl hydroxylase inhibition by either hypoxia or BAY.
-Stable isotope incorporation was similar in PASMCs as compared to LFs (**Figure S5**) where labeling was attenuated in hypoxia and the overall labeling pattern was quite similar in normoxia and hypoxia.
-Compared to previous studies of metabolic flux in cancer cells [@RN2395; @RN1447], no marked increase in M5-labeled citrate from [U-^13^C~5~] glutamine was observed with hypoxia treatment, indicating no increase in reductive carboxylation for lipid synthesis.
-The overall fraction of M5-citrate in these cells was low (< 6%).
-
-## Metabolic flux in hypoxia is closely coupled to cell growth rate
-
-To clarify changes in intracellular metabolite fluxes, we next generated metabolic flux models incorporating the extracellular flux measurements and stable isotope tracing data described above.
-Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (**Figures \@ref(fig:f3)A, S6**).
-Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by INCA [@RN1522; @RN2501; @RN964] (**Figures \@ref(fig:f4), S6, Tables S1-S3**).
-Overall, proliferating LFs demonstrated high rates of glucose uptake and glycolysis.
-Approximately 10% of cytoplasmic pyruvate enters the TCA cycle with the balance converted to lactate (**Figure S6C**).
-In hypoxia, significant reductions in glycolysis, the TCA cycle, and amino acid metabolism were observed (**Figure \@ref(fig:f4)A**) with a significant increase in pentose phosphate pathway flux.
-Similar findings were observed in PASMC flux models (**Figure S6, Table S3**).
 
@@
 
-By comparison, HIF-1 activation by BAY in 21% oxygen increased glycolysis and lactate fermentation by ~50% (**Figure \@ref(fig:f4)B**).
 BAY treatment decreased cell proliferation rate (**Figure \@ref(fig:f2)B**), indicating that, unlike hypoxia, BAY treatment uncouples cell proliferation and metabolic flux.
@@
 Metabolite fluxes in DMSO-treated cells were similar to 21% oxygen controls.
-
-## Hypoxia increases reductive carboxylation in PASMCs
 
@@
 # Discussion
+
+- PASMC reductive carboxylation
 
@@
 
-(ref:f4) **Metabolic flux maps of lung fibroblasts.** (**A**) Ratio of metabolic fluxes in 0.5% oxygen compared to 21% oxygen.
+(ref:f3) **Metabolic flux maps of lung fibroblasts.** (**A**) Ratio of metabolic fluxes in 0.5% oxygen compared to 21% oxygen.
 (**B**) Ratio of metabolic fluxes in cells treated with molidustat (BAY) compared to DMSO vehicle control.
-Fluxes with non-overlapping confidence intervals are highlighted to indicate significant changes.
-Arrow weight corresponds to absolute flux in hypoxia- or BAY-treated cells.
+Fluxes with non-overlapping confidence intervals are highlighted with colored arrows to indicate significant changes.
+Arrow thickness corresponds to absolute flux in hypoxia- or BAY-treated cells.
 

### 2021-07-19 manuscript/supplement.Rmd
update metabolic network figures and work on manuscript

@@
 
-# Figures
-
 # Figure legends
@@
 
-(ref:s5) **Mass isotopomer distributions after 72 h of labeling in pulmonary artery smooth muscle cells.** Pulmonary artery smooth muscle cells (PASMCs) were labeled with the indicated tracers and intracellular metabolites were analyzed by LC-MS after 48 h.
+(ref:s5) **Mass isotopomer distributions after 72 h of labeling in pulmonary artery smooth muscle cells.** Pulmonary artery smooth muscle cells (PASMCs) were labeled with the indicated tracers and intracellular metabolites were analyzed by LC-MS after 36 h.
 Mass isotopomer distributions were adjusted for natural abundance.
@@
 
-(ref:s6) **Isotopically non-stationary metabolic flux analysis.** (**A**, **B**) LFs were cultured in 21% (A) or 0.5% (B) oxygen and labeled with the indicated tracers and intracellular metabolites were analyzed by LC-MS (PYR, pyruvate; CIT, citrate; MAL, malate).
+(ref:s6) **Isotope incorporation in key metabolites over the experimental time course.** (**A**, **B**) LFs were cultured in 21% (A) or 0.5% (B) oxygen and labeled with the indicated tracers and intracellular metabolites were analyzed by LC-MS (PYR, pyruvate; CIT, citrate; MAL, malate).
 Mass isotopomer distributions were adjusted for natural abundance.
 Data are the mean ± SEM of 4 biological replicates.
-(**C**) Metabolic flux model of LF metabolism in 21% oxygen.
-(**D**) LF fluxes were normalized to cell growth rate.
+
+(ref:s7) (**A**) Metabolic flux model of LF metabolism in 21% oxygen.
+(**B**) Metabolic flux model of PASMC metabolism in 21% oxygen.
+(**C**) LF fluxes were normalized to cell growth rate.
 Graph depicts the ratio of normalized metabolic fluxes in LFs cultured in 0.5% oxygen compared to 21% oxygen control.
 Fluxes with non-overlapping confidence intervals are highlighted to indicate significant changes.
-(**E**) Metabolic flux model of PASMC metabolism in 21% oxygen.
-(**F**) Ratio of metabolic fluxes in 0.5% oxygen compared to 21% oxygen in PASMCs.
+(**D**) Ratio of metabolic fluxes in 0.5% oxygen compared to 21% oxygen in PASMCs.
 
@@
 Supplementary Figure 6: (ref:s6)
+
+Supplementary Figure 7: (ref:s7)
 

### 2021-07-19 R/a_analysis.R

@@ -1,765 +1,0 @@
-# analysis.R
-
-read_data <- function(data_files) {
-  data_files[stringr::str_detect(data_files, "\\.csv$")] %>%
-    rlang::set_names(stringr::str_extract(., "(lf|pasmc)_(02|05-bay|05|bay|)")) %>%
-    purrr::map_dfr(read_csv, .id = "experiment") %>%
-    dplyr::mutate(
-      oxygen = factor(oxygen, levels = c("21%", "0.5%", "0.2%"), ordered = TRUE),
-      treatment = factor(treatment, levels = c("None", "DMSO", "BAY"), ordered = TRUE)
-    )
-}
-
-plot_time_lines <- function(
-  df,
-  y,
-  ylab = prot,
-  clr = c("oxygen", "treatment", "group")
-) {
-
-  ggplot2::ggplot(df) +
-    ggplot2::aes(
-      x = time,
-      y = {{y}},
-      color = .data[[clr]],
-      fill = .data[[clr]]
-    ) +
-    ggplot2::stat_summary(
-      geom = "linerange",
-      fun.data = "mean_se",
-      size = 0.5,
-      show.legend = FALSE
-    ) +
-    ggplot2::stat_summary(
-      geom = "line",
-      fun.data = "mean_se",
-      size = 0.5,
-      show.legend = FALSE
-    ) +
-    ggplot2::stat_summary(
-      geom = "point",
-      fun = "mean",
-      pch = 21,
-      color = "white",
-      size = 2,
-      show.legend = FALSE
-    ) +
-    ggplot2::labs(
-      x = "Time (h)",
-      y = ylab
-    ) +
-    ggplot2::scale_x_continuous(breaks = seq(0, 72, 24)) +
-    ggplot2::scale_y_continuous(limits = c(0, NA)) +
-    ggplot2::scale_color_manual(values = clrs) +
-    ggplot2::scale_fill_manual(values = clrs) +
-    ggplot2::coord_cartesian(ylim = c(0, NA)) +
-    theme_plots()
-}
-
-annot_pairwise <- function(fluxes) {
-  fluxes %>%
-    dplyr::filter(experiment %in% c("02", "05", "bay")) %>%
-    dplyr::group_by(experiment, cell_type, metabolite, abbreviation) %>%
-    tidyr::nest() %>%
-    dplyr::mutate(
-      ttest = purrr::map_dbl(data, ~t.test(
-        flux ~ interaction(oxygen, treatment),
[…]

### 2021-07-19 R/a_expression.R

@@ -1,5 +1,15 @@
 # expression.R
 
+read_data <- function(data_files) {
+  data_files[stringr::str_detect(data_files, "\\.csv$")] %>%
+    rlang::set_names(stringr::str_extract(., "(lf|pasmc)_(02|05-bay|05|bay|)")) %>%
+    purrr::map_dfr(read_csv, .id = "experiment") %>%
+    dplyr::mutate(
+      oxygen = factor(oxygen, levels = c("21%", "0.5%", "0.2%"), ordered = TRUE),
+      treatment = factor(treatment, levels = c("None", "DMSO", "BAY"), ordered = TRUE)
+    )
+}
+
 plot_blot <- function(blot_image) {
   # blot_image <- magick::image_read_pdf(blot_image)
   blot_image <- magick::image_read(blot_image)
@@ -104,4 +114,50 @@ plot_expression <- function(
     dplyr::filter(experiment %in% exp) %>%
     dplyr::filter(protein == prot) %>%
     plot_time_lines(y = fold_change, ylab = ylab, clr = "group")
+}
+
+plot_time_lines <- function(
+  df,
+  y,
+  ylab = prot,
+  clr = c("oxygen", "treatment", "group")
+) {
+
+  ggplot2::ggplot(df) +
+    ggplot2::aes(
+      x = time,
+      y = {{y}},
+      color = .data[[clr]],
+      fill = .data[[clr]]
+    ) +
+    ggplot2::stat_summary(
+      geom = "linerange",
+      fun.data = "mean_se",
+      size = 0.5,
+      show.legend = FALSE
+    ) +
+    ggplot2::stat_summary(
+      geom = "line",
+      fun.data = "mean_se",
+      size = 0.5,
+      show.legend = FALSE
+    ) +
+    ggplot2::stat_summary(
+      geom = "point",
+      fun = "mean",
+      pch = 21,
+      color = "white",
+      size = 2,
+      show.legend = FALSE
+    ) +
+    ggplot2::labs(
+      x = "Time (h)",
+      y = ylab
+    ) +
+    ggplot2::scale_x_continuous(breaks = seq(0, 72, 24)) +
+    ggplot2::scale_y_continuous(limits = c(0, NA)) +
+    ggplot2::scale_color_manual(values = clrs) +
+    ggplot2::scale_fill_manual(values = clrs) +
+    ggplot2::coord_cartesian(ylim = c(0, NA)) +
+    theme_plots()
 }

### 2021-07-19 R/a_figures.R

@@ -122,6 +122,20 @@ arrange_m2 <- function(m2ab, m2c) {
     )
 }
 
+arrange_s6 <- function(a, b) {
+  layout <- "
+    a
+    b
+  "
+
+  a + b +
+    theme_patchwork(
+      design = layout,
+      widths = unit(5, "in"),
+      heights = unit(c(2.5), "in")
+    )
+}
+
 arrange_m4 <- function(hypoxia_graph_ratio_plot, bay_graph_ratio_plot) {
 
   a <- hypoxia_graph_ratio_plot
@@ -131,19 +145,18 @@ arrange_m4 <- function(hypoxia_graph_ratio_plot, bay_graph_ratio_plot) {
     theme_patchwork(widths = unit(3, "in"), heights = unit(3.5, "in"), guides = "collect")
 }
 
-arrange_s6 <- function(a, b, c, d, e, f) {
+arrange_s7 <- function(a, b, c, d) {
   layout <- "
     ab
     cd
-    cd
-    ef
-    ef"
+  "
 
-  a + b + c + d + e + f +
+  a + b + c + d +
     theme_patchwork(
       design = layout,
-      widths = unit(5, "in"),
-      heights = unit(c(2.5), "in")
+      widths = unit(3, "in"),
+      heights = unit(c(3.5), "in"),
+      guides = "collect"
     )
 }
 

### 2021-07-19 R/a_flux-figures.R

@@ -1,5 +1,22 @@
 # flux-figures.R
 
+annot_pairwise <- function(fluxes) {
+  fluxes %>%
+    dplyr::filter(experiment %in% c("02", "05", "bay")) %>%
+    dplyr::group_by(experiment, cell_type, metabolite, abbreviation) %>%
+    tidyr::nest() %>%
+    dplyr::mutate(
+      ttest = purrr::map_dbl(data, ~t.test(
+        flux ~ interaction(oxygen, treatment),
+        data = .x,
+        paired = TRUE
+      )$p.value),
+      pval = annot_p(ttest),
+      y_pos = Inf,
+      vjust = 1.5
+    )
+}
+
 plot_growth_curve <- function(
   df,
   cell = c("lf", "pasmc"),

### 2021-07-19 R/a_mid-figures.R

@@ -305,7 +305,114 @@ plot_pasmc_mids <- function(df) {
       ),
       metabolite = forcats::fct_recode(metabolite, "`3PG`" = "3PG")
     ) %>%
-    dplyr::filter(time == 48 & !is.na(metabolite)) %>%
+    dplyr::filter(time == 36 & !is.na(metabolite)) %>%
     plot_mids() +
     theme_patchwork(widths = unit(9.5, "in"), heights = unit(7.5, "in"), tags = NULL)
+}
+
+get_m5_citrate <- function(df) {
+  x <-
+    df %>%
+    dplyr::filter(tracer == "q5" & treatment == "None" & isotope == "M5" & metabolite == "CIT") %>%
+    dplyr::filter(cell_type == "pasmc" & time == 36)
+
+  out <-
+    x %>%
+    dplyr::group_by(oxygen) %>%
+    dplyr::summarise(mean = mean(mid), se = sd(mid)/sqrt(dplyr::n())) %>%
+    tidyr::pivot_wider(names_from = oxygen, values_from = c(mean, se)) %>%
+    unlist() * 100
+
+  pval <- t.test(mid ~ oxygen, data = x, paired = TRUE)$p.value
+
+  c(out, pval = pval) %>%
+    round(digits = 2)
+}
+
+format_time_course_mids <- function(model_mids) {
+  tracer_labels <-
+    c(expression(paste("[1,2-"^13, "C"[2], "] glucose")),
+      expression(paste("[U-"^13, "C"[6], "] glucose")),
+      expression(paste("[U-"^13, "C"[5], "] glutamine")),
+      expression(paste("[U-"^13, "C"[3], "] lactate")))
+
+  tracer_levels <-
+    c("glc2", "glc6", "q5", "lac3")
+
+  model_mids %>%
+    tidyr::unnest(c(data)) %>%
+    dplyr::filter(metabolite %in% c("PYR", "CIT", "MAL")) %>%
+    dplyr::filter(tracer != "lac3") %>%
+    dplyr::filter(isotope != "M6") %>%
+    dplyr::mutate(
+      tracer = factor(
+        tracer,
+        levels = tracer_levels,
+        labels = tracer_labels
+      ),
+      metabolite = factor(metabolite, levels = c("PYR", "CIT", "MAL"))
+    )
+}
+
+plot_mid_time_course <- function(time_course_mids, cells, o2, treat, color) {
+  time_course_mids %>%
[…]

### 2021-07-19 R/a_network-figures.R

@@ -1,0 +1,390 @@
+# network-figures.R
+
+clean_model_fluxes <- function(map_flux_files, model_reactions) {
+  pathways <-
+    model_reactions %>%
+    dplyr::select(-equation) %>%
+    dplyr::add_row(name = "BIOMASS", pathway = factor("biomass"), index = 0)
+
+  map_flux_files[stringr::str_detect(map_flux_files, "(lf|pasmc)_.*_model\\.csv")] %>%
+    rlang::set_names(stringr::str_extract(basename(.), pattern = ".*(?=_model\\.csv)")) %>%
+    purrr::map_dfr(readr::read_csv, .id = "experiment") %>%
+    tidyr::separate(
+      experiment,
+      c("cell_type", "treatment"),
+      sep = "_"
+    ) %>%
+    dplyr::mutate(
+      treatment = dplyr::case_when(
+        treatment == "21" ~ "21%",
+        treatment == "05" ~ "0.5%",
+        treatment == "dmso" ~ "DMSO",
+        treatment == "bay" ~ "BAY"
+      ),
+      treatment = factor(treatment, levels = c("21%", "0.5%", "DMSO", "BAY"))
+    ) %>%
+
+    # identify net and exchange fluxes
+    tidyr::separate(id, c("id", "type"), sep = " ", fill = "right") %>%
+    dplyr::mutate(type = replace(type, is.na(type), "net")) %>%
+    dplyr::select(-se) %>%
+
+    # add pathway info
+    dplyr::left_join(pathways, by = c("id" = "name")) %>%
+    dplyr::select(
+      cell_type,
+      treatment,
+      pathway,
+      index,
+      id,
+      type,
+      equation,
+      flux = value,
+      lb,
+      ub
+    ) %>%
+    dplyr::group_by(cell_type, treatment) %>%
+    dplyr::arrange(treatment, pathway)
+}
+
+calculate_flux_differences <- function(map_fluxes, cell, control, experiment) {
+  map_fluxes %>%
+    dplyr::filter(cell_type == cell) %>%
+    dplyr::filter(treatment %in% c(control, experiment)) %>%
+    dplyr::mutate(
+      treatment = dplyr::case_when(
+        treatment == control ~ "ctl",
+        treatment == experiment ~ "exp"
+      )
+    ) %>%
[…]

### 2021-07-19 R/a_two-factor-fluxes.R

@@ -1,0 +1,206 @@
+# two-factor-fluxes.R
+
+plot_twoby_fluxes <- function(df, annot, metab, ylab) {
+  df %>%
+    dplyr::filter(metabolite == metab) %>%
+    ggplot2::ggplot() +
+    ggplot2::aes(
+      x = oxygen,
+      y = flux
+    ) +
+    ggplot2::stat_summary(
+      ggplot2::aes(
+        fill = treatment
+      ),
+      geom = "col",
+      fun = "mean",
+      position = ggplot2::position_dodge(width = 0.6),
+      width = 0.6,
+      show.legend = TRUE
+    ) +
+    ggplot2::stat_summary(
+      ggplot2::aes(
+        group = treatment
+      ),
+      geom = "errorbar",
+      fun.data = "mean_se",
+      position = ggplot2::position_dodge(width = 0.6),
+      width = 0.2,
+      size = 0.25,
+      show.legend = FALSE,
+      color = "black"
+    ) +
+    ggplot2::geom_text(
+      data = dplyr::filter(annot, metabolite == metab),
+      ggplot2::aes(
+        x = oxygen,
+        y = y_pos,
+        vjust = vjust,
+        label = lab,
+      ),
+      family = "Calibri",
+      color = "black",
+      size = 6/ggplot2::.pt,
+      show.legend = FALSE
+    ) +
+    ggplot2::labs(
+      x = "Oxygen",
+      y = ylab,
+      color = NULL,
+      fill = NULL
+    ) +
+    ggplot2::scale_color_manual(values = clrs) +
+    ggplot2::scale_fill_manual(values = clrs) +
+    ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0.05, 0.1))) +
+    # ggplot2::coord_cartesian(ylim = c(-750, 1250)) +
+    theme_plots() +
+    ggplot2::theme(
+      legend.key.width = ggplot2::unit(0.5, "lines"),
+      legend.key.height = ggplot2::unit(0.5, "lines"),
+      legend.position = "bottom",
+      legend.box.margin = ggplot2::margin(t = -10)
+    )
+
+}
+
+analyze_twoby_fluxes <- function(growth_rates, fluxes) {
+
+  df <-
+    growth_rates %>%
+    dplyr::filter(experiment == "05-bay") %>%
+    dplyr::rename(flux = mu) %>%
+    dplyr::select(-X0) %>%
[…]

### 2021-07-19 _targets.R

@@ -253,6 +253,10 @@ list(
     path = path_to_reports("mass-isotope-distributions.Rmd"),
     output_dir = system.file("analysis/pdfs", package = "Copeland.2021.hypoxia.flux")
   ),
+  tar_target(
+    pasmc_m5,
+    get_m5_citrate(pruned_mids)
+  ),
 
   # biomass -----------------------------------------------------------------
 
@@ -875,42 +879,100 @@ list(
     write_figures(s5, "s5.pdf")
   ),
 
-  # # M4 ----------------------------------------------------------------------
-  #
-  # tar_target(
-  #   node_file,
-  #   path_to_data("nodes\\.csv"),
-  #   format = "file"
-  # ),
-  # tar_target(
-  #   nodes,
-  #   readr::read_csv(node_file)
-  # ),
-  # tar_target(
-  #   lf_hypoxia_graph,
-  #   make_graph(map_flux_differences, nodes, cell = "lf", treat = "21%", normalizer = "none")
-  # ),
-  # tar_target(
-  #   lf_hypoxia_graph_ratio_plot,
-  #   plot_ratio_network(lf_hypoxia_graph, "Hypoxia/Normoxia")
-  # ),
-  # tar_target(
-  #   bay_graph,
-  #   make_graph(map_flux_differences, nodes, cell = "lf", treat = "DMSO", normalizer = "none")
-  # ),
-  # tar_target(
-  #   bay_graph_ratio_plot,
-  #   plot_ratio_network(bay_graph, "BAY/DMSO")
-  # ),
-  # tar_target(
-  #   m4,
-  #   arrange_m4(lf_hypoxia_graph_ratio_plot, bay_graph_ratio_plot)
-  # ),
-  # tar_target(
-  #   m4_figure,
-  #   write_figures(m4, "m4.pdf")
-  # ),
-  #
+  # S6 ----------------------------------------------------------------------
+
+  tar_target(
+    time_course_mids,
+    format_time_course_mids(model_mids)
+  ),
+  tar_target(
+    s6a,
+    plot_mid_time_course(time_course_mids, "lf", "21%", "None", "plasma")
+  ),
+  tar_target(
+    s6b,
+    plot_mid_time_course(time_course_mids, "lf", "0.5%", "None", "viridis")
+  ),
+  tar_target(
+    s6,
+    arrange_s6(s6a, s6b)
+  ),
+  tar_target(
[…]

### 2021-07-27 manuscript/manuscript.Rmd
finish revision of draft

@@
 Compared with LFs, PASMCs demonstrated a marked reduction of M2-citrate labeled by glucose and no change in the labeling of M3-citrate by [U-^13^C~3~] glucose.
-Interestingly, unlike LFs, PASMCs demonstrated increased production of M5-citrate from [U-^13^C~5~] glutamine in hypoxia (`r m5$mean_21%`% ± `r m5$se_21%`% *v*.
-`r m5$mean_0.5%`% ± `r m5$se_0.5%`%, p = `r m5$pval`) consistent with increased reductive carboxylation in support of lipid synthesis in these cells, similar to prior observations in cancer cells.
+Interestingly, unlike LFs, PASMCs demonstrated increased production of M5-citrate from [U-^13^C~5~] glutamine in hypoxia (`r m5[["mean_21%"]]`% ± `r m5[["se_21%"]]`% *v*.
+`r m5[["mean_0.5%"]]`% ± `r m5[["se_0.5%"]]`%, p = `r m5[["pval"]]`) consistent with increased reductive carboxylation in support of lipid synthesis in these cells, similar to prior observations in cancer cells.
 
@@
 
-The mass isotopomer distribution for a given metabolite is determined based on a complicated relationship among the rate of isotope incorporation into the metabolic network and the contributions of unlabeled metabolites and their associated fluxes.
+The mass isotopomer distribution for a given metabolite is determined based on a complicated relationship among the rate of isotope incorporation into the metabolic network, the contributions of unlabeled substrates, and fluxes through related pathways.
 To clarify how these labeling patterns reflect changes in intracellular metabolite fluxes, we next generated metabolic flux models incorporating the extracellular flux measurements and stable isotope tracing data described above.
@@
 Consistent with extracellular flux measurements and isotope labeling patterns described above, in hypoxia, significant reductions in glycolysis, the TCA cycle, and amino acid metabolism were observed in the metabolic flux models of LFs with a significant increase in pentose phosphate pathway flux (**Figure \@ref(fig:f3)A**).
-By contrast, HIF-1 activation by BAY in 21% oxygen increased glycolysis and lactate fermentation by ~50% (**Figure \@ref(fig:f3)B**).
-This was associated with decreased TCA and amino acid fluxes.
+By contrast, HIF-1 activation by BAY in 21% oxygen increased glycolysis and lactate fermentation by ~50% (**Figure \@ref(fig:f3)B**), but had a similar effect as hypoxia in decreasing serine and glutamine incorporation.
+Metabolite fluxes in DMSO-treated cells were similar to 21% oxygen controls.
 
@@
 The increased TCA flux in PASMCs was driven by increased glutamine consumption in these cells.
-This is similar to a prior report of glutamine-driven oxidative phosphorylation in hypoxic cancer cells [@RN528].
+This is similar to a prior report of glutamine-driven oxidative phosphorylation in hypoxic cancer cells [@RN528], although glycolysis was up-regulated in the hypoxia-exposed cancer cells studied in this report.
 
-### YOU ARE HERE ###
-
-Given the global decrease in bioenergetic metabolic flux in hypoxia, we hypothesized that these differences may be a consequence of decreased growth rate.
-After normalizing metabolite fluxes in normoxia and hypoxia to the  growth rate, modest increases (~10%) in glycolytic flux were observed (**Figure S4D**).
+Given the global decrease in bioenergetic metabolic flux in hypoxic LFs, we hypothesized that these differences may be a consequence of decreased growth rate.
+After normalizing metabolite fluxes in normoxia and hypoxia to the cell growth rate, modest increases (~10%) in glycolytic flux were observed (**Figure S7D**).
 This finding suggests that, while glycolysis increases relative to growth rate in hypoxic cells, the regulators of cell proliferation rate override the consequences of the HIF-1 transcriptional program.
 Indeed, even after adjusting for cell growth rate, the relative increase in glycolytic flux is modest compared to the marked up-regulation of glycolytic genes.
-
-BAY treatment decreased cell proliferation rate (**Figure \@ref(fig:f2)B**), indicating that, unlike hypoxia, BAY treatment uncouples cell proliferation and metabolic flux.
-Similar to hypoxia, decreases in serine and glutamine incorporation were observed.
-Metabolite fluxes in DMSO-treated cells were similar to 21% oxygen controls.
+BAY treatment decreased cell proliferation rate (**Figure \@ref(fig:f1)B**), indicating that, unlike hypoxia, BAY treatment uncouples cell proliferation and metabolic flux.
 
@@
 Although the metabolite exchange fluxes for bidirectional reactions tend to be poorly resolved, two observations are worth highlighting (**Tables S1-S3**).
-First, consistent with the stable isotope tracing results, the rate of reductive carboxylation through reversible flux by isocitrate dehydrogenase is low (~4 fmol/cell/h), is unchanged by hypoxia in LFs, is modestly increased by BAY treatment, and is increased by hypoxia in PASMCs.
+First, consistent with the stable isotope tracing results, the modeled rate of reductive carboxylation through reverse flux by isocitrate dehydrogenase in LFs is low (~4 fmol/cell/h), unchanged by hypoxia, and modestly increased by BAY treatment.
+By contrast, the rate of reductive carboxylation increases 6-fold in PASMCs in hypoxia, highlighting an important role for this pathway in the metabolic response of these cells to hypoxia.
 
-Second, hypoxia and BAY treatment are associated with a marked increase in the lactate transport exchange flux in LFs from ~ 0 to 1,520 and 711 fmol/cell/h in 0.5% oxygen and BAY treatment conditions, respectively, with similar results in PASMCs exposed to hypoxia.
+Second, PHD inhibition is associated with a marked increase in the lactate transport exchange flux in LFs from ~ 0 to 1,500 and 700 fmol/cell/h in 0.5% oxygen and BAY treatment conditions, respectively, with similar results in PASMCs.
 Since the net lactate transport flux is secretion, this observation suggests increased lactate uptake with hypoxia or BAY treatment.
-This may be consistent with the HIF-driven increased expression of the reversible lactate transporter MCT4 [@RN366].
-To investigate this hypothesis, LFs and PASMCs were treated with [U-^13^C~3~] lactate (2 mM) and ^13^C incorporation into intracellular metabolites was analyzed by LC-MS (**Figures \@ref(fig:f5), S5**).
-Here, we observed increased labeling of TCA metabolites citrate (CIT), 2-oxoglutrate (2OG), malate (MAL), and aspartate (ASP) following hypoxia or BAY treatment (**Figures \@ref(fig:f5), S4**).
-Interestingly, although increased labeling of pyruvate and lactate were observed in hypoxic PASMCs, these labels were not incorporated into the TCA cycle as observed in LFs (**Figure S5**).
-
+This finding may be consistent with the HIF-driven increased expression of the reversible lactate transporter MCT4 [@RN366].
+To investigate this hypothesis, LFs and PASMCs were treated with [U-^13^C~3~] lactate (2 mM) and ^13^C incorporation into intracellular metabolites was analyzed by LC-MS (**Figures \@ref(fig:f4), S5**).
 Notably, lactate labeled ~50% of citrate and ~20% of downstream TCA cycle metabolites in both LFs and PASMCs, indicating that lactate may be an important respiratory fuel source in these cells even though lactate efflux is high.
 Although lactate has been used less commonly than glucose and glutamine in stable isotope tracing studies, Faubert and colleagues [-@RN2600] demonstrated lactate incorporation in human lung adenocarcinoma *in vivo*.
-In this study, lactate incorporation corresponded to regions of high glucose uptake as determined by [ ^18^F]-fluorodeoxyglucose positron emission tomography, suggesting that lactate consumption can occur in areas of high glucose utilization.
+In this study, lactate incorporation corresponded to regions of high glucose uptake as determined by [ ^18^F ]-fluorodeoxyglucose positron emission tomography, suggesting that lactate consumption can occur in areas of high glucose utilization.
 Subsequently, several investigators have demonstrated the importance of lactate as a metabolic fuel *in vivo* [@RN2756; @RN2718].
+With hypoxia or BAY treatment, we observed increased labeling of the TCA metabolites citrate (CIT), α-ketoglutarate (AKG), malate (MAL), and aspartate (ASP) in LFs.
+Interestingly, although increased labeling of pyruvate and lactate were observed in hypoxic PASMCs, the label was not incorporated into the TCA cycle as observed in LFs (**Figure S5**).
+
 In addition to downstream metabolites, we also observed hypoxia- and BAY-dependent increases in lactate incorporation in fructose bisphosphate (FBP) and 3-phosphoglycerate (3PG).
-This is consistent with prior reports describing hypoxia-mediated increases in glycogen synthesis [@RN1678; @RN545; @RN1668].
-Together, these data suggest that lactate makes a modest (~5% carbon) contribution to this process.
+This observation is consistent with prior reports describing hypoxia-mediated increases in glycogen synthesis [@RN1678; @RN545; @RN1668].
+Together, these data suggest that lactate makes a modest (~5% carbon) contribution to glycogen precursors.
 
-## Hypoxia abrogates the increase in glycolysis following pharmacologic prolyl hydroxylase inhibition
+## Hypoxia abrogates the effects of BAY on increasing glycolysis
 
-To reconcile the differential effects of prolyl hydroxylase inhibition by hypoxia and BAY, we next addressed whether hypoxia could suppress the effects of BAY on glucose and lactate fluxes (**Figures \@ref(fig:f6)A-C**).
+To reconcile the differential effects of prolyl hydroxylase inhibition by hypoxia and BAY, we next addressed whether hypoxia could suppress the effects of BAY on glucose and lactate fluxes (**Figures \@ref(fig:f5)A-C**).
 LFs cultured in standard growth medium were treated with BAY and placed in either 21% or 0.5% oxygen.
@@
 
-To investigate these metabolic differences further, we performed metabolomic profiling of LFs treated for 72 h with hypoxia or BAY separately or in combination (**Figures \@ref(fig:f6)D-F**).
-Principle component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen (**Figure \@ref(fig:f6)D**).
-These hypoxia-treated cells were well-segregated from BAY-treated cells, again highlighting differential effects of hypoxic and pharmacologic PHD inhibition.
-Linear modeling identified those metabolites that were differentially affected by hypoxia and BAY treatment (*i.e.*, those metabolites with a significant interaction term in the linear model described by treatment × oxygen) (**Figure \@ref(fig:f6)E**).
-Several patterns emerged from this analysis (**Figure \@ref(fig:f6)F**).
-Interestingly, the hypoxia-regulated metabolite, 2-hydroxyglutarate, increased following BAY treatment, but was unaffected by hypoxia treatment in proliferating LFs [@RN1603].
+To investigate these metabolic differences further, we performed metabolomic profiling of LFs treated for 72 h with hypoxia or BAY separately or in combination (**Figures \@ref(fig:f5)D-F**).
+Both 0.5% oxygen and BAY treatment induced marked changes in intracellular metabolite levels (**Figure S8**).
+Of 133 metabolites, 98 were differentially regulated by hypoxia and 54 were differentially regulated by BAY.
+Of the differentially regulated metabolites, 44 were affected by both 0.5% oxygen and BAY treatment (**Figure S8C**).
+Metabolite set enrichment analysis of KEGG biochemical pathways were consistent with the results of the metabolic models demonstrating significant enrichment of the pentose phosphate pathway with 0.5% oxygen or BAY treatment; decreased alanine, aspartate, and glutamate metabolism with hypoxia; and increased glycolysis/gluconeogenesis with BAY (**Figures S8D, S8E**).
+Indeed, aspartate was the most decreased metabolite with both treatments, consistent with prior reports demonstrating an important role for HIF-1 regulation of aspartate biosynthesis in cancer cells [@RN634; @RN1438].
+
+Principle component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen (**Figure \@ref(fig:f5)D**).
+Moreover, these hypoxia-treated cells were well-segregated from BAY-treated cells.
+These observations are again consistent with the results of metabolic flux models demonstrating an overriding effect of hypoxia *per se* on metabolic flux and highlighting important differences between hypoxic and pharmacologic PHD inhibition.
+To identify the key differences among these treatments, we performed linear modeling focusing on those metabolites with a significant interaction term in the model described by treatment × oxygen (**Figure \@ref(fig:f5)E**) (*i.e.*, metabolites where the change with hypoxia treatment was different than the change with BAY treatment).
+Of 133 metabolites, 77 were significantly differentially regulated by hypoxia and BAY treatments.
+Several patterns emerged from this analysis (**Figure \@ref(fig:f5)F**).
+Interestingly, the hypoxia-regulated metabolite, 2-hydroxyglutarate [@RN1603], increased following BAY treatment, but was unaffected by hypoxia treatment in proliferating LFs.
 The tricarboxylic acid (TCA) cycle metabolite aconitate was decreased in hypoxia, but increased by BAY treatment while the opposite effect was observed for hydroxyproline and taurine.
-Glycolytic intermediate glyceraldehyde 3-phosophate (GAP) was decreased by both hypoxia and BAY, while hypoxia reversed the BAY-mediated decrease in γ-aminobutyric acid (GABA).
-A metabolite set enrichment analysis (**Figure \@ref(fig:f6)G-H**) revealed the TCA cycle to be the most enriched KEGG metabolite set.
-Leading edge analysis demonstrates the negative enrichment scores associated with all of the TCA metabolites detected by our platform.
-This result suggests a more modest impact of BAY treatment on the TCA cycle than hypoxia, as suggested by our metabolic flux models were hypoxia resulted in a 1.5-2-fold reduction of TCA flux compared to a 1.1-1.5-fold reduction with BAY treatment in normoxia (**Figure \@ref(fig:f4)**).
+The glycolytic intermediate glyceraldehyde 3-phosphate (GAP) was decreased by both hypoxia and BAY, while hypoxia reversed the BAY-mediated decrease in γ-aminobutyric acid (GABA).
+A metabolite set enrichment analysis of these differentially regulated metabolitesrevealed the TCA cycle to be the most enriched KEGG metabolite set (**Figure \@ref(fig:f5)G**).
+Leading edge analysis demonstrated negative enrichment scores associated with all of the TCA metabolites detected by our platform (**Figure \@ref(fig:f5)H**).
+This result suggests a more modest impact of BAY treatment on the TCA cycle than hypoxia, as suggested by our metabolic flux models where hypoxia resulted in a 1.5-2-fold reduction of TCA flux compared to a 1.1-1.5-fold reduction with BAY treatment in normoxia (**Figure \@ref(fig:f3)**).
 
 In addition to these differential effects on polar metabolite levels, perhaps the most significant contrast between hypoxia and BAY treatment is the impact of hypoxia on cellular redox state.
-As oxygen deprivation causes reductive stress [@RN2409], we next measured the impact of these treatments on intracellular NAD(H) as a surrogate for cellular redox state (**Figures \@ref(fig:f6)I-K**).
+As oxygen deprivation causes reductive stress [@RN2409], we next measured the impact of these treatments on intracellular NAD(H) as a surrogate for cellular redox state (**Figures \@ref(fig:f5)I-K**).
 As expected, hypoxia increased the NADH/NAD^+^ ratio, driven primarily by a decrease in intracellular NAD^+^.
@@
 
-## Transcriptomic analysis identifies putative regulators of cellular metabolism in hypoxia
+## Transcriptomic analysis identifies regulators of metabolism in hypoxia
 
-To identify the upstream regulators of the observed metabolic changes, we next performed transcriptomic analysis of LFs treated with hypoxia or BAY, separately or together (**Figure \@ref(fig:f7)**).
-Principle component analysis again demonstrates clear separation among the four treatment groups (**Figure \@ref(fig:f7)A**), with a closer relationship between hypoxia-treated cells compared to hypoxia and 21% BAY-treated cells.
-To identify those transcripts that were differentially affected by hypoxia compared to BAY treatment, we again performed linear modeling to identify transcripts with a significant interaction term in the model described by treatment × oxygen (**Figure \@ref(fig:f7)B**).
-With this analysis, we again found several patterns of transcriptional changes (**Figure \@ref(fig:f7)C**).
-For example, EPAS1, the gene encoding the HIF-2α protein, was markedly decreased by hypoxia, compared to a modest reduction following BAY treatment.
+To identify the upstream regulators of the observed metabolic changes, we next performed transcriptomic analysis of LFs treated with hypoxia or BAY, separately or together (**Figure \@ref(fig:f6), S9**).
+As anticipated, both hypoxia and BAY treatment induced substantial changes in gene expression (**Figure S9A, S9B**).
+Of the 7789 differentially expressed genes across both conditions, 891 (11%) were unique to BAY treatment in normoxia, 1649 (21%) were shared between BAY and hypoxia, while 5249 (67%) were unique to 0.5% culture (**Figure S9C**).
+Gene set enrichment analysis of these differentially regulated metabolites was performed using Molecular Signatures Database "Hallmark" gene sets [@RN3112] (**Figures S9D, S9E, S9F**).
+As expected, both treatments were associated with enrichment of the "hypoxia" and "glycolysis" gene sets.
+
+Given the disparate effects of hypoxic and pharmacologic PHD inhibition on cellular metabolism described above, we focused our transcriptomics analyses on the differences between hypoxia and BAY treatments.
+Principle component analysis again demonstrates clear separation among the four treatment groups (**Figure \@ref(fig:f6)A**).
+The first and second principle components correspond to 0.5% oxygen and BAY treatments, respectively.
+Consistent with our prior observations, the combination of 0.5% oxygen plus BAY was more similar to 0.5% oxygen alone with decreased distance between both hypoxia-treated groups along the axis of the second principle component.
+Again, consistent with the hypothesis that hypoxia uncouples the effects of BAY treatment.
+To identify those transcripts that were differentially affected by hypoxia compared to BAY treatment alone, we again performed linear modeling to identify transcripts with a significant interaction term in the model described by treatment × oxygen (**Figure \@ref(fig:f6)B**).
+With this analysis, we again found several patterns of transcriptional changes (**Figure \@ref(fig:f6)C**).
+For example, EPAS1, the gene encoding the HIF-2α protein, was markedly decreased by hypoxia, compared to its modest reduction following BAY treatment.
 A similar, but opposite effect on RBM3 (RNA-binding protein 3) expression was observed.
 RBM3 supports the viability and proliferation of neural stem cells in hypoxia [@RN3111].
-Histone deacetylase 9 (HDAC9) and prolyl 4-hydroxylase subunit α2 (P4HA2) demonstrate opposing effects following hypoxia and BAY treatment.
-BAY decreased HDAC9 expression and increased P4HA2 expression.
-Together, this subset of transcriptional changes again illustrates important differences between hypoxia and HIF stabilization in normoxia.
-Indeed, an enrichment analysis of these transcripts using the Molecular Signatures Database "Hallmark" gene sets [@RN3112] actually demonstrated greater enrichment of the "hypoxia" data set among differentially expressed transcripts following BAY treatment (**Figure \@ref(fig:f7)D**).
-Consistent with our previous findings, the gene set for glycolysis was also more enriched following BAY treatment than hypoxia treatment.
-The most enriched gene sets associated with hypoxia included pro-proliferative E2F targets and G2/M checkpoint proteins as well as MYC target proteins.
-These findings were reinforced by a transcription factor enrichment analysis (**Figure \@ref(fig:f7)E**), again identifying enrichment of MYC target proteins associated with hypoxia.
-Consistent with these bioinformatic results, immunoblot demonstrates up-regulation of MYC protein in hypoxia-treated cells, but not in BAY-treatment alone (**Figure \@ref(fig:f7)F**).
-These data provide a starting point for interrogating the relative hierarchy of transcription factor regulation of gene expression by hypoxia in proliferating primary cells and suggest several potential mechanisms that may contribute to the uncoupling of glycolytic gene transcription and glycolytic flux in the proliferating primary cells studied here.
+Histone deacetylase 9 (HDAC9) and prolyl 4-hydroxylase subunit α2 (P4HA2) demonstrate opposing effects following hypoxia and BAY treatment where BAY decreases HDAC9 expression and increases P4HA2 expression.
+Together, this subset of transcriptional changes illustrates important differences between hypoxia and HIF stabilization in normoxia.
+
+Interestingly, an enrichment analysis of transcripts differentially regulated by 0.5% oxygen and BAY treatment actually demonstrated greater enrichment of the "hypoxia" data set following BAY treatment alone than with hypoxia itself (**Figure \@ref(fig:f6)D**).
+Similarly, the gene set for glycolysis was also more enriched following BAY treatment than hypoxia treatment.
+The most enriched gene sets associated with hypoxia included pro-proliferative E2F targets and G2/M checkpoint proteins as well as Myc target proteins.
+These findings were further supported by a transcription factor enrichment analysis identifying enrichment of Myc target proteins associated with hypoxia (**Figure \@ref(fig:f6)E**).
+Consistent with these bioinformatic results, immunoblotting demonstrated up-regulation of Myc protein in hypoxia-treated cells, but not with BAY-treatment alone (**Figure \@ref(fig:f6)F**).
+These data provide a starting point for interrogating the relative hierarchy of transcription factor regulation of gene expression by hypoxia in proliferating primary cells and suggest several potential mechanisms that may contribute to the uncoupling of glycolytic gene transcription and glycolytic flux reported here.
 
@@
 
-- PASMC reductive carboxylation
-
 In this work, we used ^13^C metabolic flux analysis to identify hypoxia-mediated metabolic changes in proliferating human primary cells.
 Our principal finding was that hypoxia reduced, rather than increased, carbon flux through glycolysis and lactate fermentation pathways despite robust activation of the HIF transcriptional program and up-regulation of glycolytic genes.
-Certainly the LFs studied here are capable of augmenting glycolysis in response to HIF stabilization, as demonstrated by experiments with the PHD inhibitor BAY; however, these effects are completely attenuated when BAY-treated cells are cultured in hypoxia.
+Certainly, the LFs studied here are capable of augmenting glycolysis in response to HIF stabilization, as demonstrated by experiments with the PHD inhibitor BAY; however, these effects are completely attenuated when BAY-treated cells are cultured in hypoxia.
 Together, these findings suggest that changes in enzyme levels alone are insufficient to alter metabolic flux in hypoxia and point to the importance of regulatory mechanisms that supersede the effects of HIF-dependent gene transcription.
 
+<!-- Important non-HIF mechanisms -->
+
 The identification of these mechanisms is of critical importance for our understanding of primary cell adaptation to hypoxia.
-In particular, our transcriptomic analyses suggest that hypoxia is associated with a pro-proliferative pattern of gene expression that likely facilitates the response of these mesenchymal cells to tissue injury and these pathways may have implications for cancer cell biology.
 While the role of HIFs in the hypoxia response has been extensively studied, relatively less is known about HIF-independent features of the hypoxia response.
@@
 For example, PHD is one of many α-ketoglutarate-dependent dioxygenase enzymes that rely on molecular oxygen for their catalytic activity [@RN2611].
-Jumonji-C (JmjC) domain containing histone demethylases are other prominent members of this family whose inhibition by hypoxia has been shown to cause rapid and HIF-independent induction of histone methylation [@RN3117].
+Jumonji-C (JmjC) domain-containing histone demethylases are other prominent members of this family whose inhibition by hypoxia has been shown to cause rapid and HIF-independent induction of histone methylation [@RN3117].
+Indeed, our transcription factor enrichment analysis (**Figure \@ref(fig:f6)E**) demonstrates that this family of lysine demethylases (KDMs) was the most enriched with 0.5% oxygen treatment as compared to BAY treatment.
 Similarly, a recently described cysteamine dioxygenase has been shown to mediate the oxygen-dependent degradation of Regulators of G protein Signaling 4 and 5 and IL-32 [@RN1384].
 In addition to dioxygenase enzymes, electron transport chain dysfunction resulting from impaired Complex IV activity leads to increased ROS production in hypoxia [@RN295].
-Mitochondrial ROS increase the half-lives of several mRNAs in hypoxia, including MYC as we observe in this work, independent of HIF stabilization [@RN742].
-Finally, hypoxia imposes a reductive stress on cells associated with an increase in the NADH/NAD^+^ ratio secondary to impaired electron transport [@RN2618; @RN2619].
+Mitochondrial ROS increase the half-lives of several mRNAs in hypoxia, including Myc, as we observe in this work, independent of HIF stabilization [@RN742].
+Finally, hypoxia imposes a reductive stress on cells associated with an increase in the NADH/NAD^+^ ratio secondary to impaired electron transport (**Figure S5K**) [@RN2618; @RN2619].
 NADH accumulation may slow glycolysis *via* feedback inhibition of GAPDH [@RN2210].
-Any of these molecular mechanisms may contribute to uncoupling glycolytic enzyme expression from glycolytic flux as observed in the experiments described here.
+Any or all of these molecular mechanisms may contribute to uncoupling glycolytic enzyme expression from glycolytic flux as observed in the experiments described here.
 
-These findings raise important questions regarding the cell-autonomous role of HIFs in the hypoxia response.
+<!-- HIF and cell cycle progression -->
+
+In addition to its effects on cellular metabolism, another canonical role of HIF-1 activation is slowing of cellular proliferation rate in the face of limited oxygen availability [@RN615] and this is what we observed here (**Figures \@ref(fig:f1)B, \@ref(fig:f5)A**).
+The effects of HIF-1 on cell proliferation rate are mediated, in part, by increased expression of cyclin-dependent kinase inhibitor p21 (*CDKN1A*) and inhibition of E2F targets [@RN3152] and inhibition of pro-proliferative Myc signaling [@RN3151], and these transcriptional effects are precisely what is observed in BAY treated LFs in normoxia.
+By contrast, hypoxia culture was associated with decreased expression of p21, consistent with a previous report [@RN3150], as well as increased expression of Myc protein and enrichment of Myc target genes.
+Indeed, the most marked differences between hypoxia and BAY treatment on LF gene transcription were the up-regulation of pro-proliferative gene sets containing E2F targets and G2/M checkpoint proteins.
+Much of this transcriptional response may be mediated by hypoxia-induced up-regulation of Myc, which is known to stimulate cell cycle progression through its effects on the expression and activity of cyclins, cyclin-dependent kinases, and cyclin-dependent kinase inhibitors [@RN3154]; however, the relative importance of Myc and the hypoxia-mediated mechanism of its induction remain to be elucidated in these primary cells.
+Clarifying the complex interactions among HIFs, Myc, and cell proliferation will be important for understanding the cellular response of these mesenchymal cells to tissue injury.
+
+<!-- Why HIF? -->
+
+All together, these findings raise important questions regarding the cell-autonomous role of HIFs in the hypoxia response.
 On an organismal level, HIFs drive expression of angiogenic and erythropoietic factors to increase oxygen delivery to hypoxic tissues.
@@
 Interestingly, these cells also had increased ATP levels compared to wild type, suggesting that mitochondrial function was adequate under 1% oxygen culture conditions to support oxidative phosphorylation and meet the energy needs of the cells.
-Given the prominence of HIFs in mediating the transcriptional response to hypoxia, it is somewhat surprising that none of PHD, HIFs, or their downstream targets were found to be selectively essential as a function of oxygen tension in a genome-wide CRISPR growth screen of cells cultured in normoxia and hypoxia [@RN953].
+Given the prominence of HIFs in mediating the transcriptional response to hypoxia, it is somewhat surprising that neither PHD, HIFs, nor their downstream targets were found to be selectively essential as a function of oxygen tension in a genome-wide CRISPR growth screen of K562 human lymphoblasts cultured in normoxia or hypoxia [@RN953].
+
+<!-- Lactate consumption -->
+<!-- PASMC reductive carboxylation -->
+
+This work also highlights two specific metabolic features that appear to be important in the metabolic response of these primary cells to hypoxia.
+First, both LFs and PASMCs demonstrated notable incorporation of lactate-derived carbon into intracellular metabolic pathways that increased with hypoxia and BAY treatments (**Figures \@ref(fig:f4), S5**).
+This finding is consistent with increasing evidence suggesting an important role for lactate as a metabolic fuel in several organ systems [@RN2600; @RN1157].
+Although typically considered a metabolic waste product [@@RN3081], an important role for lactate *import* in supporting cell homeostasis in the face of an ischemic insult, which is associated with increased extracellular lactate, is an evolutionarily attractive hypothesis that merits further investigation.
+Second, PASMCs, but not LFs, demonstrated significant rates of reductive carboxylation that increased in 0.5% oxygen (**Figure S5**).
+Reductive carboxylation was first identified in hypoxic tumor cells where stable isotope tracing revealed ^13^C incorporation from labeled glutamine into lipids [@RN1447; @RN628; @RN1966; @RN2395].
+Hypoxia drives pulmonary arterial smooth muscle cell proliferation *in vivo* contributing to the development pulmonary hypertension in humans and animal models.
+Isocitrate dehydrogenase has previously been implicated in the pathobiology of this disease [@RN1259], and our findings suggest that reductive carboxylation catalyzed by isocitrate dehydrogenase may be a viable therapeutic target in this disease.
+
+<!-- Limitations -->
 
@@
 Third, and perhaps most importantly, these flux determinations were performed in human primary cell cultures rather than immortalized cell lines.
+Although both cell types used in this study were derived from lung, we anticipate that many of our findings will be generalizable to primary cells from different tissues.
+
+<!-- Conclusion -->
 
@@
 Indeed, the degree of metabolic reprogramming in hypoxia was modest and suggests close coupling between proliferation and metabolism.
-In light of our findings, additional studies are warranted to understand the role of HIFs in mediating the metabolic response to hypoxia in primary cells.
+In light of our findings, additional studies are warranted to clarify the role of HIFs in mediating the metabolic response to hypoxia and to identify other key regulators of hypoxic metabolic reprogramming in primary cells.
 Moreover, these data strongly caution investigators against drawing conclusions about metabolite flux from measures of gene transcription alone.
-Further investigations of metabolic flux in primary cell cultures in hypoxia are warranted to identify the key regulators of metabolism in hypoxia and to clarify the contributions of HIF proteins to hypoxic metabolic reprogramming.
 
@@
 
-This work was supported by grants from the NIH (K08HL128802), American Lung Association, Pulmonary Hypertension Association, and the American Thoracic Society Foundation to W.M.O and from the NIH (U01HG007690, U01HL108630, U54HL119145) and the American Heart Association (D700382, CV-19) to J.L.
+This work was supported by grants from the NIH (K08HL128802), American Lung Association, Pulmonary Hypertension Association, and the American Thoracic Society Foundation to W.M.O and from the NIH (U01HG007690, U54HL119145, R01HL155107, R01HL155096) and the American Heart Association (D700382, CV-19) to J.L.
 
@@
 
-(ref:f5) **PHD inhibition increases lactate uptake and oxidation.** Mass isotopomer distributions of key metabolites following labeling with [U-^13^C~3~] lactate (2 mM) for 72 h indicates increased lactate uptake and oxidation in hypoxia or with molidustat (BAY) treatment.
+(ref:f4) **PHD inhibition increases lactate uptake and oxidation.** Mass isotopomer distributions of key metabolites following labeling with [U-^13^C~3~] lactate (2 mM) for 72 h indicates increased lactate uptake and oxidation in hypoxia or with molidustat (BAY) treatment.
+Data are mean ± SEM of n = 4 biological replicates.
+Comparisons were made using linear mixed effects models with treatment group as a fixed effect and biological replicate as a random effect.
+Tukey's *post hoc* test was applied to determine differences between 21% and 0.5% oxygen (*), between DMSO and BAY treatment (†), or between 0.5% oxygen and BAY treatment (‡) with adjusted p-values < 0.05 considered significant.
 
-(ref:f6) **Metabolomic analysis of molidustat treamtent in normoxia and hypoxia.** (**A**-**C**) Hypoxia inhibits the effects of HIF-1α stabilization on glycolysis.
+(ref:f5) **Metabolomic analysis of molidustat treatment in normoxia and hypoxia.** (**A**-**C**) Hypoxia inhibits the effects of HIF-1α stabilization on glycolysis.
 Lung fibroblasts (LFs) were cultured in standard growth medium and treated with molidustat (BAY, 10 μM) or vehicle (DMSO, 0.1%) in 21% or 0.5% oxygen conditions (n = 4).
 (A) Growth rates were determined by linear fitting of log-transformed growth curves.
-(**B**-**C**) Extracellular fluxes of glucose (B) and lactate (C).
+(**B**-**C**) Extracellular fluxes of glucose (note inverse *y*-axis) (B) and lactate (C).
 (**D**) Principle components analysis of intracellular metabolites following 72 h of treatment described above suggests a dominant effect of hypoxia over PHD inhibition on the metabolome (n = 4).
 (**E**) Volcano plot of the differential effects of hypoxia and molidustat (BAY) treatment on intracellular metabolites.
-(**F**) Representative metabolites from (E) reveal different patterns of metabolic effects.
+*Blue* circles indicate the fold change with hypoxia was greater than the fold change with BAY treatment while *purple* circles indicate the opposite.
+(**F**) Representative metabolites highlighted in (E) reveal different patterns of metabolic effects.
 (**G**) Metabolite set enrichment analysis of metabolites from (E).
-KEGG pathways with unadjusted p-values < 0.05 are shown.
+KEGG pathways with p-values < 0.05 are shown.
 (**H**) Leading edge analysis of the most enriched KEGG metabolite set from (G).
@@
 
-(ref:f7) **Transcriptomic analysis of molidustat treamtent in normoxia and hypoxia.** (**A**) Principle components analysis of transcriptional changes following 72 h of treatment with 0.5% oxygen or molidustat (BAY), separately or together (n = 4).
+(ref:f6) **Transcriptomic analysis of molidustat treamtent in normoxia and hypoxia.** (**A**) Principle components analysis of transcriptional changes following 72 h of treatment with 0.5% oxygen or molidustat (BAY), separately or together (n = 4).
 (**B**) Volcano plot of the differential effects of hypoxia and molidustat (BAY) treatment on intracellular metabolites.
@@
 (**E**) Transcription factor enrichment analysis suggests mechanisms for differential regulation of gene expression following hypoxia or BAY treatment.
-(**F**) Relative change in MYC protein levels after 72 h of treatment (n = 4) and a representative immunoblot.
+(**F**) Relative change in Myc protein levels after 72 h of treatment (n = 4) and a representative immunoblot.
 Comparisons were performed using a mixed-effects linear model with replicate as a random effect.
@@
 
-Figure 1: (ref:f1)
+```{r}
+figs <- 
+  list.files(
+    system.file(
+      "manuscript/figures", 
+      package = "Copeland.2021.hypoxia.flux"
+    ), 
+    pattern = "m\\d{1}\\.pdf"
+  ) %>% 
+  rlang::set_names(stringr::str_extract(., "^.*(?=\\.pdf)"))
+```
 
-Figure 2: (ref:f2)
-
-Figure 3: (ref:f3)
-
-Figure 4: (ref:f4)
-
-Figure 5: (ref:f5)
-
-Figure 6: (ref:f6)
-
-Figure 7: (ref:f7)
+```{r results = "asis"}
+out <- 
+  purrr::imap(
+    figs, 
+    ~ cat("Figure ", stringr::str_extract(.y, "\\d"), ": (ref:", .y, ")\n\n", sep = "")
+  )
+```
 
@@
     "manuscript/figures/m6.pdf",
-    package = "Copeland.2021.hypoxia.flux"
-  )
-)
-```
-
-\newpage
-
-```{r f7, fig.cap = "(ref:f7)"}
-knitr::include_graphics(
-  system.file(
-    "manuscript/figures/m7.pdf",
     package = "Copeland.2021.hypoxia.flux"
@@
 Low quality features were removed on the basis of a relative standard deviation greater than 0.2 in the QC samples and a dispersion ratio greater than 0.4 [@RN3077].
-Missing values were imputed using random forrest.
+Missing values were imputed using random forest.
 Samples peak areas were normalized using probabilistic quotient normalization [@RN3094].

### 2021-07-27 manuscript/supplement.Rmd
finish revision of draft

@@
 
-<br>
-
 ```{r t1, results = "asis"}
@@
 
-(ref:s7) (**A**) Metabolic flux model of LF metabolism in 21% oxygen.
+(ref:s7) **Isotopically non-stationary metabolic flux analysis.** (**A**) Metabolic flux model of LF metabolism in 21% oxygen.
 (**B**) Metabolic flux model of PASMC metabolism in 21% oxygen.
@@
 
-Supplementary Figure 1: (ref:s1)
+(ref:s8) **Metabolomic profiling of hypoxia and BAY treated lung fibroblasts.** (**A**, **B**) Volcano plots of differentially regulated metabolites with 0.5% oxygen in DMSO-treated cells (A) or BAY treatment in 21% oxygen-cultured cells (B).
+Significantly increased metabolites with 0.5% oxygen (*blue*) or with BAY treatment (*purple*) are indicated.
+The top 10 up- and down-regulated metabolites are labeled.
+(**C**) Venn diagram illustrating the number of differentially regulated metabolites following hypoxia (*blue*) or BAY treatment (*purple*).
+(**D**, **E**) Results of a metabolite set enrichment analysis of KEGG pathways based on the data from (A) and (B).
+Significantly enriched pathways with p < 0.05 are indicated.
 
-Supplementary Figure 2: (ref:s2)
+(ref:s9) **Transcriptomic profiling of hypoxia and BAY treated lung fibroblasts.** (**A**)
 
-Supplementary Figure 3: (ref:s3)
+```{r}
+figs <- 
+  list.files(
+    system.file(
+      "manuscript/figures", 
+      package = "Copeland.2021.hypoxia.flux"
+    ), 
+    pattern = "s\\d{1}\\.pdf"
+  ) %>% 
+  rlang::set_names(stringr::str_extract(., "^.*(?=\\.pdf)"))
+```
 
-Supplementary Figure 4: (ref:s4)
-
-Supplementary Figure 5: (ref:s5)
-
-Supplementary Figure 6: (ref:s6)
-
-Supplementary Figure 7: (ref:s7)
+```{r results = "asis"}
+out <- 
+  purrr::imap(
+    figs, 
+    ~ cat("Supplementary Figure ", stringr::str_extract(.y, "\\d"), ": (ref:", .y, ")\n\n", sep = "")
+  )
+```
 
@@
 
+```{r s7, fig.cap = "(ref:s7)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/s7.pdf",
+    package = "Copeland.2021.hypoxia.flux"
+  )
+)
+```
+
+\newpage
+
+```{r s8, fig.cap = "(ref:s8)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/s8.pdf",
+    package = "Copeland.2021.hypoxia.flux"
+  )
+)
+```
+
+\newpage
+
+```{r s9, fig.cap = "(ref:s9)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/s9.pdf",
+    package = "Copeland.2021.hypoxia.flux"
+  )
+)
+```
+
+\newpage
+
 # Supplemental References

### 2021-07-27 R/a_figures.R

@@ -131,8 +131,8 @@ arrange_s6 <- function(a, b) {
   a + b +
     theme_patchwork(
       design = layout,
-      widths = unit(5, "in"),
-      heights = unit(c(2.5), "in")
+      widths = unit(4, "in"),
+      heights = unit(c(3.5), "in")
     )
 }
 
@@ -160,7 +160,7 @@ arrange_s7 <- function(a, b, c, d) {
     )
 }
 
-arrange_m6 <- function(a, b, c, d, e, f, g, h, i, j, k) {
+arrange_m5 <- function(a, b, c, d, e, f, g, h, i, j, k) {
   ((a | b | c) + plot_layout(guides = "collect")) /
     (d | e) /
     (f + plot_layout(guides = "collect")) /
@@ -178,7 +178,16 @@ arrange_m6 <- function(a, b, c, d, e, f, g, h, i, j, k) {
     )
 }
 
-arrange_m7 <- function(a, b, c, d, e, f) {
+arrange_s8 <- function(a, b, c, d, e) {
+  (a | b) /
+    (c | d | e) + plot_layout(widths = c(1, 2, 2)) +
+    theme_patchwork(
+      widths = unit(5, "in"),
+      heights = unit(c(1.1), "in")
+    )
+}
+
+arrange_m6 <- function(a, b, c, d, e, f) {
   layout <- "
   abd
   ccd
@@ -190,6 +199,25 @@ arrange_m7 <- function(a, b, c, d, e, f) {
       design = layout,
       widths = unit(1.25, "in"),
       heights = unit(c(1.25), "in")
+    )
+}
+
+arrange_s9 <- function(a, b, c, d, e, f) {
+  # layout <- "
+  # aabb
+  # cc##
+  # d#e#
+  # "
+
+  (a | b) /
+    (c | d) /
+    (e | f) +
+
+  # a + b + c + d + e +
+    theme_patchwork(
+      # design = layout,
+      widths = unit(5, "in"),
+      heights = unit(c(1.25, 2, 2.5), "in")
     )
 }
 

### 2021-07-27 R/a_metabolomics.R

@@ -359,7 +359,7 @@ metab_top_table <- function(clean, fit, contrast) {
     )
 }
 
-plot_metab_volcano <- function(results) {
+plot_metab_volcano <- function(results, mois = NULL, colors = NULL, xlab = NULL) {
 
   left <-
     results %>%
@@ -373,7 +373,6 @@ plot_metab_volcano <- function(results) {
 
   ggplot2::ggplot(results) +
     ggplot2::aes(
-      # x = 2 ^ log2FoldChange,
       x = logFC,
       y = adj.P.Val
     ) +
@@ -381,7 +380,7 @@ plot_metab_volcano <- function(results) {
       data = left,
       ggplot2::aes(
         label = metabolite,
-        color = metabolite %in% c("2-hydroxyglutarate","aconitate", "glyceraldehyde 3-phosphate", "hydroxyproline", "malate", "taurine")
+        color = metabolite %in% mois
       ),
       size = 5/ggplot2::.pt,
       max.overlaps = 20,
@@ -396,7 +395,7 @@ plot_metab_volcano <- function(results) {
       data = right,
       ggplot2::aes(
         label = metabolite,
-        color = metabolite %in% c("2-hydroxyglutarate","aconitate", "glyceraldehyde 3-phosphate", "hydroxyproline", "malate", "taurine")
+        color = metabolite %in% mois
       ),
       size = 5/ggplot2::.pt,
       max.overlaps = 20,
@@ -417,13 +416,13 @@ plot_metab_volcano <- function(results) {
       data = subset(results, logFC > 0 & adj.P.Val < 0.05),
       pch = 21,
       color = "white",
-      fill = clrs[[2]]
+      fill = colors[[1]]
     ) +
     ggplot2::geom_point(
       data = subset(results, logFC < 0 & adj.P.Val < 0.05),
       pch = 21,
       color = "white",
-      fill = clrs[[4]]
+      fill = colors[[2]]
     ) +
     ggplot2::scale_color_manual(values = c("black", "darkred")) +
     ggplot2::scale_y_continuous(trans = reverselog_trans(10)) +
@@ -433,7 +432,7 @@ plot_metab_volcano <- function(results) {
       expand = ggplot2::expansion(mult = 1)
     ) +
     ggplot2::labs(
-      x = "ΔHypoxia/ΔBAY",
[…]

### 2021-07-27 R/a_mid-figures.R

@@ -399,10 +399,9 @@ plot_mid_time_course <- function(time_course_mids, cells, o2, treat, color) {
     )
 }
 
-plot_lactate_mids <- function(model_mids, cell) {
-  model_mids %>%
+plot_lactate_mids <- function(pruned_mids, cell) {
+  pruned_mids %>%
     dplyr::filter(cell_type == cell) %>%
-    tidyr::unnest(c(data)) %>%
     dplyr::mutate(
       metabolite = factor(
         metabolite,
@@ -413,6 +412,10 @@ plot_lactate_mids <- function(model_mids, cell) {
     dplyr::filter(tracer == "lac3" & time == 72 & !is.na(metabolite)) %>%
     plot_mids() +
     ggplot2::facet_wrap(~ metabolite, scales = "free_x", nrow = 2) +
-    ggplot2::theme(legend.position = "right") +
-    theme_patchwork(widths = unit(4, "in"), heights = unit(2.5, "in"), tags = NULL)
+    ggplot2::theme(legend.position = "bottom") +
+    theme_patchwork(
+      widths = ggplot2::unit(4, "in"),
+      heights = ggplot2::unit(2.5, "in"),
+      tags = NULL
+    )
 }

### 2021-07-27 R/a_nad.R

@@ -1,7 +1,5 @@
 # nad.R
 
-# clean_nad ---------------------------------------------------------------
-
 clean_nad <- function(nad_data) {
   nad_data %>%
     dplyr::bind_rows(.id = "metabolite") %>%
@@ -9,8 +7,6 @@ clean_nad <- function(nad_data) {
     clean_technical_replicates() %>%
     tidyr::separate(.data$experiment, c(NA, "date"), "_")
 }
-
-# finalize_nad ------------------------------------------------------------
 
 finalize_nad <- function(nad_interp, cells_per_dna) {
 
@@ -42,8 +38,6 @@ finalize_nad <- function(nad_interp, cells_per_dna) {
 
 }
 
-# annot_nad ---------------------------------------------------------------
-
 annot_nad <- function(nad_final) {
   nad_final %>%
     dplyr::filter(treatment != "None") %>%
@@ -66,8 +60,6 @@ annot_nad <- function(nad_final) {
       vjust = 1.5
     )
 }
-
-# plot_nad ----------------------------------------------------------------
 
 plot_nad <- function(nad_final, annot, metab, ylab) {
 
@@ -138,8 +130,8 @@ plot_nad <- function(nad_final, annot, metab, ylab) {
       color = "black",
       show.legend = FALSE
     ) +
-    ggplot2::scale_color_manual(values = clrs) +
-    ggplot2::scale_fill_manual(values = clrs) +
+    ggplot2::scale_color_manual(values = clrs, limits = force) +
+    ggplot2::scale_fill_manual(values = clrs, limits = force) +
     ggplot2::labs(
       x = "Treatment",
       y = ylab,

### 2021-07-27 R/a_network-figures.R

@@ -315,7 +315,8 @@ plot_ratio_network <- function(graph, caption, edges = TRUE) {
     NULL
 }
 
-plot_normoxia_network <- function(hypoxia_graph, cell) {
+plot_normoxia_network <- function(hypoxia_graph, caption) {
+  set_gr_style()
 
   hypoxia_graph %>%
     tidygraph::activate(edges) %>%
@@ -357,6 +358,7 @@ plot_normoxia_network <- function(hypoxia_graph, cell) {
       end = 0.95,
       limits = c(1, 1000),
       na.value = "grey90",
+      oob = scales::squish,
       # breaks = c(0.01, 0.1, 1, 10, 100, 1000),
       # labels = c(1, 10, 100, 1000),
       option = "plasma"
@@ -365,7 +367,7 @@ plot_normoxia_network <- function(hypoxia_graph, cell) {
       edge_color = ggraph::guide_edge_colorbar(barwidth = 0.5, barheight = 3)
     ) +
     ggplot2::labs(
-      title = "Normoxia",
+      title = caption,
       y = NULL
     ) +
     theme_plots() +

### 2021-07-27 R/a_rnaseq.R

@@ -1,7 +1,5 @@
 # rnaseq.R
 
-# count_rnaseq ------------------------------------------------------------
-
 count_rnaseq <- function() {
 
   dds <- rnaseq.lf.hypoxia.molidustat::lf_hyp_bay_rnaseq
@@ -33,8 +31,6 @@ count_rnaseq <- function() {
   DESeq2::DESeq(dds)
 }
 
-# vst_rnaseq --------------------------------------------------------------
-
 vst_rnaseq <- function(dds) {
   vsd <- DESeq2::vst(dds, blind = FALSE)
   SummarizedExperiment::assay(vsd) <-
@@ -55,8 +51,6 @@ vst_rnaseq <- function(dds) {
     )
 
 }
-
-# plot_rnaseq_pca ---------------------------------------------------------
 
 plot_rnaseq_pca <- function(pca_data) {
 
@@ -121,8 +115,6 @@ plot_rnaseq_pca <- function(pca_data) {
 
 }
 
-# identify_deg ------------------------------------------------------------
-
 identify_deg <- function(dds, expr) {
   alpha <- 0.05
   fc <- 1
@@ -159,40 +151,55 @@ identify_deg <- function(dds, expr) {
 
 }
 
-# plot_rnaseq_venn --------------------------------------------------------
-
-plot_rnaseq_venn <- function(dds) {
-
+find_rnaseq_overlap <- function(dds) {
   bay <-
-    identify_deg(dds, expr(h.bay - n.bay)) %>%
+    identify_deg(dds, expr(n.bay - n.dmso)) %>%
     dplyr::filter(padj < 0.05)
   hyp <- identify_deg(dds, expr(h.dmso - n.dmso)) %>%
     dplyr::filter(padj < 0.05)
 
   nm <- rownames(SummarizedExperiment::assay(dds))
-  bay_deg <- nm %in% bay$symbol
-  hyp_deg <- nm %in% hyp$symbol
+  bay_deg <- nm %in% bay$row
+  hyp_deg <- nm %in% hyp$row
+  tibble::tibble(nm, hyp_deg, bay_deg)
+}
 
-  tibble::tibble(nm, hyp_deg, bay_deg) %>%
-    ggplot2::ggplot() +
+find_gsea_overlap <- function(a, b, src = "HALLMARK") {
+  dplyr::full_join(a, b, by = c("source", "pathway")) %>%
[…]

### 2021-07-27 R/a_two-factor-fluxes.R

@@ -49,8 +49,7 @@ plot_twoby_fluxes <- function(df, annot, metab, ylab) {
       color = NULL,
       fill = NULL
     ) +
-    ggplot2::scale_color_manual(values = clrs) +
-    ggplot2::scale_fill_manual(values = clrs) +
+    ggplot2::scale_fill_manual(values = clrs, limits = force) +
     ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0.05, 0.1))) +
     # ggplot2::coord_cartesian(ylim = c(-750, 1250)) +
     theme_plots() +

### 2021-07-27 _targets.R

@@ -452,16 +452,44 @@ list(
     identify_deg(dds, expr((h.dmso - n.dmso) - (n.bay - n.dmso)))
   ),
   tar_target(
+    rnaseq_hyp,
+    identify_deg(dds, expr((h.dmso - n.dmso)))
+  ),
+  tar_target(
+    rnaseq_bay,
+    identify_deg(dds, expr((n.bay - n.dmso)))
+  ),
+  tar_target(
     rnaseq_volcano,
-    plot_rnaseq_volcano(rnaseq_different_differences)
+    plot_rnaseq_volcano(rnaseq_different_differences, xlab = "ΔHypoxia/ΔBAY", gois = c("EPAS1","HDAC9", "P4HA2", "RBM3"))
+  ),
+  tar_target(
+    rnaseq_hyp_volcano,
+    plot_rnaseq_volcano(rnaseq_hyp, xlab = "Hypoxia/Normoxia")
+  ),
+  tar_target(
+    rnaseq_bay_volcano,
+    plot_rnaseq_volcano(rnaseq_bay, xlab = "BAY/DMSO")
+  ),
+  tar_target(
+    rnaseq_overlap,
+    find_rnaseq_overlap(dds)
   ),
   tar_target(
     rnaseq_venn,
-    plot_rnaseq_venn(dds)
+    plot_rnaseq_venn(rnaseq_overlap, "Transcripts")
   ),
   tar_target(
     rnaseq_gsea,
     run_gsea(rnaseq_different_differences)
+  ),
+  tar_target(
+    rnaseq_hyp_gsea,
+    run_gsea(rnaseq_hyp)
+  ),
+  tar_target(
+    rnaseq_bay_gsea,
+    run_gsea(rnaseq_bay)
   ),
   tar_target(
     rnaseq_goi,
@@ -469,7 +497,23 @@ list(
   ),
   tar_target(
     rnaseq_gsea_plot,
-    plot_gsea(rnaseq_gsea, "HALLMARK")
+    plot_gsea(rnaseq_gsea, "HALLMARK", lbls = c("With BAY", "With Hypoxia"), vals = unname(clrs[c(4, 2)]))
+  ),
+  tar_target(
+    rnaseq_hyp_gsea_plot,
+    plot_gsea(rnaseq_hyp_gsea, "HALLMARK", lbls = c("Down in Hypoxia", "Up in Hypoxia"), vals = unname(clrs[c(1, 2)]))
[…]

### 2021-07-28 manuscript/manuscript.Rmd
finish revisions of draft to send to coauthors

@@
 resources <- targets::tar_read(resources_table)
-figs <- targets::tar_load(targets::matches("m\\d_figure"))
 biomass <- tibble::deframe(biomass)
 m5 <- targets::tar_read(pasmc_m5)
+targets::tar_load(
+  c(
+    m1_figure, 
+    m2_figure, 
+    m3_figure,
+    m4_figure, 
+    m5_figure, 
+    m6_figure
+  )
+)
 ```
@@
 # Summary 
-Hypoxia is an important environmental stimulus that causes transcriptional and metabolic reprogramming in cells to facilitate their survival.
+Hypoxia is an important environmental stressor that causes transcriptional and metabolic reprogramming in cells to facilitate their survival.
 Here, we performed stable isotope tracing and metabolic flux analyses of proliferating human lung fibroblasts and pulmonary artery smooth muscle cells in hypoxia.
-Despite activation of the hypoxia-inducible factor (HIF) transcriptional program and up-regulation of glycolytic genes, glycolytic flux was decreased in hypoxic cells.
+Despite activation of the hypoxia-inducible factor (HIF) transcriptional program and increased expression of glycolytic enzymes, glycolytic flux was decreased in hypoxic cells.
 While pharmacologic stabilization of HIF in normoxia with the prolyl hydroxylase inhibitor molidustat did increase glycolytic flux as expected, hypoxia abrogated this effect of molidustat treatment.
@@
 Cellular responses to ambient oxygen levels drive numerous physiologic and pathologic processes from wound healing and angiogenesis to pulmonary vascular remodeling and fibrosis [@RN1980].
-These responses require a sustained capacity for cell proliferation, migration, and protein synthesis, and the attendant energetic and metabolic requirements, even in the face of limited oxygen availability, or hypoxia.
-Metazoan cells depend on aerobic respiration to meet cellular energy demands.
+These responses require a sustained capacity for cell proliferation, migration, and protein synthesis, with their associated energetic and metabolic demands, even in the face of limited oxygen availability, or hypoxia.
+Metazoan cells depend on aerobic respiration to meet cellular energy needs.
 With an inadequate oxygen supply, cells must reduce energy consumption and shift energy production away from oxidative phosphorylation.
@@
 When oxygen tension falls, PHD activity decreases, leading to HIF-1α stabilization and activation of its associated transcriptional program.
-Overall, the changes in gene transcription should increase glycolytic capacity and divert glucose-derived pyruvate from oxidative phosphorylation toward lactate fermentation to maintain ATP production and to minimize the formation of reactive oxygen species (ROS) [@RN2525].
+Overall, the HIF-1 transriptional program should increase glycolytic capacity and divert glucose-derived pyruvate from oxidative phosphorylation toward lactate fermentation to maintain ATP production and to minimize the formation of reactive oxygen species (ROS) [@RN2525].
 
 While a "glycolytic shift" of primary carbon metabolism in hypoxia is well-described, the effects of hypoxia on other metabolic pathways are an area of active investigation [@RN953; @RN1603; @RN1198].
-Since hypoxia is a prominent feature of cancer biology as tumor growth outstrips blood supply, most detailed metabolic studies of cell metabolism in hypoxia have used tumor cell models [@RN2395; @RN1447; @RN1438; @RN634; @RN1200; @RN975].
-For example, stable isotope tracing and metabolic flux analyses identified a critical role for reductive carboxylation of glutamine-derived α-ketoglutarate for lipid biosynthesis in tumor growth [@RN1447; @RN628; @RN1966; @RN2395], and metabolomic studies identified aspartate as a limiting metabolite for cancer cell proliferation under hypoxia [@RN634].
+Since hypoxia is a prominent feature of cancer biology as tumor growth outstrips blood supply, most detailed metabolic studies of cell metabolism in hypoxia have used tumor cell models, yielding important insights into the metabolic pathobiology of cancer [@RN2395; @RN1447; @RN1438; @RN634; @RN1200; @RN975].
+For example, stable isotope tracing and metabolic flux analyses identified a critical role for reductive carboxylation of glutamine-derived α-ketoglutarate for lipid biosynthesis in supporting tumor growth [@RN1447; @RN628; @RN1966; @RN2395], and metabolomic studies identified aspartate as a limiting metabolite for cancer cell proliferation under hypoxia [@RN634].
 By contrast, comparatively little is known about metabolic adaptations of primary cells to hypoxia and how hypoxic metabolic reprogramming supports homeostasis or promotes pathobiology.
 Indeed, the importance of reductive carboxylation or aspartate biosynthesis remains to be elucidated in these cells.
-This and related information would provide important context for understanding how metabolic reprogramming supports normal cellular responses to hypoxia, how these responses may be (mal)adaptive in a variety of disease contexts, and how the hypoxia metabolic program in primary cells differs from that observed in cancer cells.
+This and related information would provide an important context for understanding how metabolic reprogramming supports normal cellular responses to hypoxia, how these responses may be (mal)adaptive in a variety of disease contexts, and how the hypoxia metabolic program in primary cells differs from that observed in cancer cells.
 
-To address these questions, here we have developed models of bioenergetic carbon flux in human lung fibroblasts (LFs) and pulmonary artery smooth muscle cells (PASMCs) cultured under 21% or 0.5% oxygen conditions.
+To address these questions, here we have developed models of bioenergetic carbon flux in human lung fibroblasts (LFs) and pulmonary artery smooth muscle cells (PASMCs) cultured in 21% or 0.5% oxygen.
 These cells were selected as they may be exposed to a wide range of oxygen concentrations *in vivo*, continue to proliferate despite hypoxic culture conditions *in vitro*, and play important roles in the pathology of human lung diseases where tissue hypoxia is a prominent feature.
@@
 Cells were seeded and placed into hypoxia for 24 h prior to sample collection to provide adequate time for activation of the hypoxia-dependent transcriptional program.
-From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (**Figure \@ref(fig:f1)A**), thought to be an indicator of metabolic steady state .
-LFs cultured in 0.5% oxygen grew slower than LFs cultured in 21% oxygen (**Figure \@ref(fig:f1)B**), but slower growth was not associated with decreased cell viability (**Figure S1A**).
-These cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1 (GLUT1) and lactate dehydrogenase A (LDHA) (**Figures \@ref(fig:f1)C-H**).
+From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (**Figure \@ref(fig:m1)A**), thought to be an indicator of metabolic steady state .
+LFs cultured in 0.5% oxygen grew slower than LFs cultured in 21% oxygen (**Figure \@ref(fig:m1)B**), but slower growth was not associated with decreased cell viability (**Figure S1A**).
+These cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1 (GLUT1) and lactate dehydrogenase A (LDHA) (**Figures \@ref(fig:m1)K, E-H**).
 These changes persisted for the duration of the experimental time course.
@@
 
-Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (**Figures \@ref(fig:f1)I-J, S1**).
+Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (**Figures \@ref(fig:m1)I-J**).
 Flux calculations incorporated changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@RN1523] (**Figure S1**).
-Surprisingly, we observed no significant differences in glucose uptake or lactate efflux rates in 0.5% oxygen cultures (**Figure \@ref(fig:f1)I**) despite activation of the HIF-1 transcriptional program as reflected by increased expression of GLUT1 and LDHA.
-No change in glucose or lactate fluxes were observed when the ambient oxygen level was decreased further to 0.2% (**Figure S2**) and lactate efflux was significantly *decreased* in PASMCs cultured in 0.5% ambient oxygen (**Figure S3**).
+Surprisingly, we observed no significant differences in glucose uptake or lactate efflux rates in 0.5% oxygen cultures (**Figure \@ref(fig:m1)I**) despite activation of the HIF-1 transcriptional program as reflected by increased expression of GLUT1 and LDHA.
+No change in glucose or lactate fluxes were observed when the ambient oxygen level was decreased further to 0.2% (**Figure S2**).
+Similar results were also observed in PASMCs were lactate efflux was significantly *decreased* by 0.5% ambient oxygen (**Figure S3**).
 
@@
 Given that hypoxia did not increase glucose and lactate fluxes as expected from the associated increases in glycolytic gene expression, we next assessed the capacity of HIF-1α to augment glycolysis in LFs.
-Cells were treated with the prolyl hydroxylase inhibitor BAY to stabilize HIF-1α under 21% oxygen conditions (**Figure \@ref(fig:f1)**).
-Similar to hypoxia, BAY decreased cell growth rate (**Figures \@ref(fig:f1)A-B**) and activated the HIF-1 transcriptional program (**Figures \@ref(fig:f1)C-H**).
+Cells were treated with the prolyl hydroxylase inhibitor BAY to stabilize HIF-1α under 21% oxygen conditions (**Figure \@ref(fig:m1)**).
+Similar to hypoxia, BAY decreased cell growth rate (**Figures \@ref(fig:m1)A-B**) and activated the HIF-1 transcriptional program (**Figures \@ref(fig:m1)D-H**).
 Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1 target gene transcription and protein expression.
-In normoxia, this transcriptional program was associated with substantially increased glucose uptake and lactate efflux (**Figure \@ref(fig:f1)I**).
-Comparatively modest effects of BAY on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (**Figure \@ref(fig:f1)J**) with preservation of glutamine uptake, alanine efflux, and glutamine efflux rates.
+In normoxia, this transcriptional program was associated with substantially increased glucose uptake and lactate efflux (**Figure \@ref(fig:m1)I**).
+Comparatively modest effects of BAY on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (**Figure \@ref(fig:m1)J**) with preservation of glutamine uptake, alanine efflux, and glutamate efflux rates.
 
@@
 
-To validate our findings from extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate and trace the incorporation of label into key carbon utilization pathways (**Figures \@ref(fig:f2), S4-S5**).
-Consistent with our extracellular flux measurements, we found that the rate of label incorporation from [U-^13^C~6~] glucose into pyruvate was similar across 21%, 0.5%, and DMSO-treated conditions, but markedly increased with BAY treatment (**Figures \@ref(fig:f3)A-B**).
+To validate our findings from extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate and trace the incorporation of label into key carbon utilization pathways (**Figures \@ref(fig:m2), S4-S5**).
+Consistent with our extracellular flux measurements, we found that the rate of label incorporation from [U-^13^C~6~] glucose into pyruvate was similar across 21%, 0.5%, and DMSO-treated conditions, but markedly increased with BAY treatment (**Figures \@ref(fig:m3)A-B**).
 
-Several patterns of isotope labeling emerge across the treatment conditions (**Figures \@ref(fig:f3)C, S4**).
+Several patterns of isotope labeling emerge across the treatment conditions (**Figures \@ref(fig:m3)C, S4**).
 For the glycolytic intermediates fructose bisphosphate (FBP), 3-phosphoglycerate (3PG), and pyruvate (PYR), label incorporation from the glucose tracers is decreased with hypoxia and BAY treatments (*i.e.*, the unlabeled, or M0, fraction was greater).
@@
 To clarify how these labeling patterns reflect changes in intracellular metabolite fluxes, we next generated metabolic flux models incorporating the extracellular flux measurements and stable isotope tracing data described above.
-Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (**Figures \@ref(fig:f3)A, S6**).
-Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by INCA [@RN1522; @RN2501; @RN964] (**Figures \@ref(fig:f3), S7, Tables S1-S3**).
+Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (**Figures \@ref(fig:m3)A, S6**).
+Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by INCA [@RN1522; @RN2501; @RN964] (**Figures \@ref(fig:m3), S7, Tables S1-S3**).
 Overall, proliferating LFs demonstrated high rates of glucose uptake and glycolysis.
 Approximately 10% of cytoplasmic pyruvate enters the TCA cycle with the balance converted to lactate (**Figure S7A**).
-Consistent with extracellular flux measurements and isotope labeling patterns described above, in hypoxia, significant reductions in glycolysis, the TCA cycle, and amino acid metabolism were observed in the metabolic flux models of LFs with a significant increase in pentose phosphate pathway flux (**Figure \@ref(fig:f3)A**).
-By contrast, HIF-1 activation by BAY in 21% oxygen increased glycolysis and lactate fermentation by ~50% (**Figure \@ref(fig:f3)B**), but had a similar effect as hypoxia in decreasing serine and glutamine incorporation.
+Consistent with extracellular flux measurements and isotope labeling patterns described above, significant reductions in glycolysis, the TCA cycle, and amino acid metabolism were observed in the metabolic flux models of LFs cultured in hypoxia.
+A significant increase in pentose phosphate pathway flux was also observed, although the absolute flux through this pathway is low (**Figure \@ref(fig:m3)A**).
+By contrast, HIF-1 activation by BAY in 21% oxygen increased glycolysis and lactate fermentation by ~50% (**Figure \@ref(fig:m3)B**), but had a similar effect as hypoxia in decreasing serine and glutamine incorporation.
 Metabolite fluxes in DMSO-treated cells were similar to 21% oxygen controls.
@@
 The increased TCA flux in PASMCs was driven by increased glutamine consumption in these cells.
-This is similar to a prior report of glutamine-driven oxidative phosphorylation in hypoxic cancer cells [@RN528], although glycolysis was up-regulated in the hypoxia-exposed cancer cells studied in this report.
+This is similar to a prior report of glutamine-driven oxidative phosphorylation in hypoxic cancer cells [@RN528], although glycolysis was also up-regulated in the hypoxia-exposed cancer cells studied in this report.
 
@@
 This finding suggests that, while glycolysis increases relative to growth rate in hypoxic cells, the regulators of cell proliferation rate override the consequences of the HIF-1 transcriptional program.
-Indeed, even after adjusting for cell growth rate, the relative increase in glycolytic flux is modest compared to the marked up-regulation of glycolytic genes.
-BAY treatment decreased cell proliferation rate (**Figure \@ref(fig:f1)B**), indicating that, unlike hypoxia, BAY treatment uncouples cell proliferation and metabolic flux.
+Indeed, even after adjusting for cell growth rate, the relative increase in glycolytic flux is modest compared to the marked up-regulation of glycolytic protein levels.
+BAY treatment decreased cell proliferation rate (**Figure \@ref(fig:m1)B**), indicating that, unlike hypoxia, BAY treatment in normoxia uncouples cell proliferation and metabolic flux.
 
@@
 This finding may be consistent with the HIF-driven increased expression of the reversible lactate transporter MCT4 [@RN366].
-To investigate this hypothesis, LFs and PASMCs were treated with [U-^13^C~3~] lactate (2 mM) and ^13^C incorporation into intracellular metabolites was analyzed by LC-MS (**Figures \@ref(fig:f4), S5**).
+To investigate this hypothesis, LFs and PASMCs were treated with [U-^13^C~3~] lactate (2 mM) and ^13^C incorporation into intracellular metabolites was analyzed by LC-MS (**Figures \@ref(fig:m4), S5**).
 Notably, lactate labeled ~50% of citrate and ~20% of downstream TCA cycle metabolites in both LFs and PASMCs, indicating that lactate may be an important respiratory fuel source in these cells even though lactate efflux is high.
@@
 In this study, lactate incorporation corresponded to regions of high glucose uptake as determined by [ ^18^F ]-fluorodeoxyglucose positron emission tomography, suggesting that lactate consumption can occur in areas of high glucose utilization.
-Subsequently, several investigators have demonstrated the importance of lactate as a metabolic fuel *in vivo* [@RN2756; @RN2718].
+Subsequently, investigators have demonstrated the importance of lactate as a metabolic fuel *in vivo* [@RN2756; @RN2718].
 With hypoxia or BAY treatment, we observed increased labeling of the TCA metabolites citrate (CIT), α-ketoglutarate (AKG), malate (MAL), and aspartate (ASP) in LFs.
@@
 
-To reconcile the differential effects of prolyl hydroxylase inhibition by hypoxia and BAY, we next addressed whether hypoxia could suppress the effects of BAY on glucose and lactate fluxes (**Figures \@ref(fig:f5)A-C**).
+To reconcile the differential effects of prolyl hydroxylase inhibition by hypoxia and BAY, we next addressed whether hypoxia could suppress the effects of BAY on glucose and lactate fluxes (**Figures \@ref(fig:m5)A-C**).
 LFs cultured in standard growth medium were treated with BAY and placed in either 21% or 0.5% oxygen.
@@
 
-To investigate these metabolic differences further, we performed metabolomic profiling of LFs treated for 72 h with hypoxia or BAY separately or in combination (**Figures \@ref(fig:f5)D-F**).
+To investigate these metabolic differences further, we performed metabolomic profiling of LFs treated for 72 h with hypoxia or BAY separately or in combination (**Figures \@ref(fig:m5)D-F**).
 Both 0.5% oxygen and BAY treatment induced marked changes in intracellular metabolite levels (**Figure S8**).
@@
 
-Principle component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen (**Figure \@ref(fig:f5)D**).
+Principle component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen (**Figure \@ref(fig:m5)D**).
 Moreover, these hypoxia-treated cells were well-segregated from BAY-treated cells.
 These observations are again consistent with the results of metabolic flux models demonstrating an overriding effect of hypoxia *per se* on metabolic flux and highlighting important differences between hypoxic and pharmacologic PHD inhibition.
-To identify the key differences among these treatments, we performed linear modeling focusing on those metabolites with a significant interaction term in the model described by treatment × oxygen (**Figure \@ref(fig:f5)E**) (*i.e.*, metabolites where the change with hypoxia treatment was different than the change with BAY treatment).
+To identify the key differences among these treatments, we performed linear modeling focusing on those metabolites with a significant interaction term in the model described by treatment × oxygen (**Figure \@ref(fig:m5)E**) (*i.e.*, metabolites where the change with hypoxia treatment was different than the change with BAY treatment).
 Of 133 metabolites, 77 were significantly differentially regulated by hypoxia and BAY treatments.
-Several patterns emerged from this analysis (**Figure \@ref(fig:f5)F**).
+Several patterns emerged from this analysis (**Figure \@ref(fig:m5)F**).
 Interestingly, the hypoxia-regulated metabolite, 2-hydroxyglutarate [@RN1603], increased following BAY treatment, but was unaffected by hypoxia treatment in proliferating LFs.
@@
 The glycolytic intermediate glyceraldehyde 3-phosphate (GAP) was decreased by both hypoxia and BAY, while hypoxia reversed the BAY-mediated decrease in γ-aminobutyric acid (GABA).
-A metabolite set enrichment analysis of these differentially regulated metabolitesrevealed the TCA cycle to be the most enriched KEGG metabolite set (**Figure \@ref(fig:f5)G**).
-Leading edge analysis demonstrated negative enrichment scores associated with all of the TCA metabolites detected by our platform (**Figure \@ref(fig:f5)H**).
-This result suggests a more modest impact of BAY treatment on the TCA cycle than hypoxia, as suggested by our metabolic flux models where hypoxia resulted in a 1.5-2-fold reduction of TCA flux compared to a 1.1-1.5-fold reduction with BAY treatment in normoxia (**Figure \@ref(fig:f3)**).
+A metabolite set enrichment analysis of these differentially regulated metabolites revealed the TCA cycle to be the most enriched KEGG metabolite set (**Figure \@ref(fig:m5)G**).
+Leading edge analysis demonstrated negative enrichment scores associated with all of the TCA metabolites detected by our platform (**Figure \@ref(fig:m5)H**).
+This result suggests a more modest impact of BAY treatment on the TCA cycle than hypoxia, as suggested by our metabolic flux models where hypoxia resulted in a 1.5-2-fold reduction of TCA flux compared to a 1.1-1.5-fold reduction with BAY treatment in normoxia (**Figure \@ref(fig:m3)**).
 
 In addition to these differential effects on polar metabolite levels, perhaps the most significant contrast between hypoxia and BAY treatment is the impact of hypoxia on cellular redox state.
-As oxygen deprivation causes reductive stress [@RN2409], we next measured the impact of these treatments on intracellular NAD(H) as a surrogate for cellular redox state (**Figures \@ref(fig:f5)I-K**).
+As oxygen deprivation causes reductive stress [@RN2409], we next measured the impact of these treatments on intracellular NAD(H) as a surrogate for cellular redox state (**Figures \@ref(fig:m5)I-K**).
 As expected, hypoxia increased the NADH/NAD^+^ ratio, driven primarily by a decrease in intracellular NAD^+^.
@@
 
-To identify the upstream regulators of the observed metabolic changes, we next performed transcriptomic analysis of LFs treated with hypoxia or BAY, separately or together (**Figure \@ref(fig:f6), S9**).
+To identify the upstream regulators of the observed metabolic changes, we next performed transcriptomic analysis of LFs treated with hypoxia or BAY, separately or together (**Figure \@ref(fig:m6), S9**).
 As anticipated, both hypoxia and BAY treatment induced substantial changes in gene expression (**Figure S9A, S9B**).
@@
 Given the disparate effects of hypoxic and pharmacologic PHD inhibition on cellular metabolism described above, we focused our transcriptomics analyses on the differences between hypoxia and BAY treatments.
-Principle component analysis again demonstrates clear separation among the four treatment groups (**Figure \@ref(fig:f6)A**).
+Principle component analysis again demonstrates clear separation among the four treatment groups (**Figure \@ref(fig:m6)A**).
 The first and second principle components correspond to 0.5% oxygen and BAY treatments, respectively.
@@
 Again, consistent with the hypothesis that hypoxia uncouples the effects of BAY treatment.
-To identify those transcripts that were differentially affected by hypoxia compared to BAY treatment alone, we again performed linear modeling to identify transcripts with a significant interaction term in the model described by treatment × oxygen (**Figure \@ref(fig:f6)B**).
-With this analysis, we again found several patterns of transcriptional changes (**Figure \@ref(fig:f6)C**).
+To identify those transcripts that were differentially affected by hypoxia compared to BAY treatment alone, we performed linear modeling to identify transcripts with a significant interaction term in the model described by treatment × oxygen (**Figure \@ref(fig:m6)B**).
+With this analysis, we found several patterns of transcriptional changes (**Figure \@ref(fig:m6)C**).
 For example, EPAS1, the gene encoding the HIF-2α protein, was markedly decreased by hypoxia, compared to its modest reduction following BAY treatment.
-A similar, but opposite effect on RBM3 (RNA-binding protein 3) expression was observed.
+A similar, but opposite, effect on RBM3 (RNA-binding protein 3) expression was observed.
 RBM3 supports the viability and proliferation of neural stem cells in hypoxia [@RN3111].
-Histone deacetylase 9 (HDAC9) and prolyl 4-hydroxylase subunit α2 (P4HA2) demonstrate opposing effects following hypoxia and BAY treatment where BAY decreases HDAC9 expression and increases P4HA2 expression.
+Histone deacetylase 9 (HDAC9) and prolyl 4-hydroxylase subunit α2 (P4HA2) demonstrate opposing effects following hypoxia and BAY treatment, where BAY decreases HDAC9 expression and increases P4HA2 expression.
 Together, this subset of transcriptional changes illustrates important differences between hypoxia and HIF stabilization in normoxia.
 
-Interestingly, an enrichment analysis of transcripts differentially regulated by 0.5% oxygen and BAY treatment actually demonstrated greater enrichment of the "hypoxia" data set following BAY treatment alone than with hypoxia itself (**Figure \@ref(fig:f6)D**).
+Interestingly, an enrichment analysis of transcripts differentially regulated by 0.5% oxygen and BAY treatment actually demonstrated greater enrichment of the "hypoxia" gene set following BAY treatment alone than with hypoxia itself (**Figure \@ref(fig:m6)D**).
 Similarly, the gene set for glycolysis was also more enriched following BAY treatment than hypoxia treatment.
-The most enriched gene sets associated with hypoxia included pro-proliferative E2F targets and G2/M checkpoint proteins as well as Myc target proteins.
-These findings were further supported by a transcription factor enrichment analysis identifying enrichment of Myc target proteins associated with hypoxia (**Figure \@ref(fig:f6)E**).
-Consistent with these bioinformatic results, immunoblotting demonstrated up-regulation of Myc protein in hypoxia-treated cells, but not with BAY-treatment alone (**Figure \@ref(fig:f6)F**).
+The most enriched gene sets associated with hypoxia included pro-proliferative "E2F targets" and "G2/M checkpoint" proteins as well as "Myc targets".
+These findings were further supported by a transcription factor enrichment analysis identifying enrichment of Myc target proteins associated with hypoxia (**Figure \@ref(fig:m6)E**).
+Consistent with these bioinformatic results, immunoblotting demonstrated up-regulation of Myc protein in hypoxia-treated cells, but not with BAY-treatment alone (**Figure \@ref(fig:m6)F**).
 These data provide a starting point for interrogating the relative hierarchy of transcription factor regulation of gene expression by hypoxia in proliferating primary cells and suggest several potential mechanisms that may contribute to the uncoupling of glycolytic gene transcription and glycolytic flux reported here.
@@
 Jumonji-C (JmjC) domain-containing histone demethylases are other prominent members of this family whose inhibition by hypoxia has been shown to cause rapid and HIF-independent induction of histone methylation [@RN3117].
-Indeed, our transcription factor enrichment analysis (**Figure \@ref(fig:f6)E**) demonstrates that this family of lysine demethylases (KDMs) was the most enriched with 0.5% oxygen treatment as compared to BAY treatment.
+Indeed, our transcription factor enrichment analysis (**Figure \@ref(fig:m6)E**) demonstrates that this family of lysine demethylases (KDMs) was the most enriched with 0.5% oxygen treatment as compared to BAY treatment.
 Similarly, a recently described cysteamine dioxygenase has been shown to mediate the oxygen-dependent degradation of Regulators of G protein Signaling 4 and 5 and IL-32 [@RN1384].
@@
 
-In addition to its effects on cellular metabolism, another canonical role of HIF-1 activation is slowing of cellular proliferation rate in the face of limited oxygen availability [@RN615] and this is what we observed here (**Figures \@ref(fig:f1)B, \@ref(fig:f5)A**).
-The effects of HIF-1 on cell proliferation rate are mediated, in part, by increased expression of cyclin-dependent kinase inhibitor p21 (*CDKN1A*) and inhibition of E2F targets [@RN3152] and inhibition of pro-proliferative Myc signaling [@RN3151], and these transcriptional effects are precisely what is observed in BAY treated LFs in normoxia.
+In addition to its effects on cellular metabolism, another canonical role of HIF-1 activation is slowing of cellular proliferation rate in the face of limited oxygen availability [@RN615] and this is what we observed here (**Figures \@ref(fig:m1)B, \@ref(fig:m5)A**).
+The effects of HIF-1 on cell proliferation rate are mediated, in part, by increased expression of cyclin-dependent kinase inhibitor p21 (*CDKN1A*) and inhibition of E2F targets [@RN3152] and inhibition of pro-proliferative Myc signaling [@RN3151].
+These transcriptional effects are precisely what is observed in BAY treated LFs in normoxia.
 By contrast, hypoxia culture was associated with decreased expression of p21, consistent with a previous report [@RN3150], as well as increased expression of Myc protein and enrichment of Myc target genes.
@@
 
-All together, these findings raise important questions regarding the cell-autonomous role of HIFs in the hypoxia response.
+Altogether, these findings raise important questions regarding the cell-autonomous role of HIFs in the hypoxia response.
 On an organismal level, HIFs drive expression of angiogenic and erythropoietic factors to increase oxygen delivery to hypoxic tissues.
@@
 This work also highlights two specific metabolic features that appear to be important in the metabolic response of these primary cells to hypoxia.
-First, both LFs and PASMCs demonstrated notable incorporation of lactate-derived carbon into intracellular metabolic pathways that increased with hypoxia and BAY treatments (**Figures \@ref(fig:f4), S5**).
+First, both LFs and PASMCs demonstrated notable incorporation of lactate-derived carbon into intracellular metabolic pathways that increased with hypoxia and BAY treatments (**Figures \@ref(fig:m4), S5**).
 This finding is consistent with increasing evidence suggesting an important role for lactate as a metabolic fuel in several organ systems [@RN2600; @RN1157].
-Although typically considered a metabolic waste product [@@RN3081], an important role for lactate *import* in supporting cell homeostasis in the face of an ischemic insult, which is associated with increased extracellular lactate, is an evolutionarily attractive hypothesis that merits further investigation.
+Although typically considered a metabolic waste product [@RN3081], an important contribution of lactate *import* in supporting metabolic homeostasis in the face of an ischemic insult, which is associated with increased extracellular lactate, is an evolutionarily attractive hypothesis that merits further investigation.
 Second, PASMCs, but not LFs, demonstrated significant rates of reductive carboxylation that increased in 0.5% oxygen (**Figure S5**).
 Reductive carboxylation was first identified in hypoxic tumor cells where stable isotope tracing revealed ^13^C incorporation from labeled glutamine into lipids [@RN1447; @RN628; @RN1966; @RN2395].
-Hypoxia drives pulmonary arterial smooth muscle cell proliferation *in vivo* contributing to the development pulmonary hypertension in humans and animal models.
-Isocitrate dehydrogenase has previously been implicated in the pathobiology of this disease [@RN1259], and our findings suggest that reductive carboxylation catalyzed by isocitrate dehydrogenase may be a viable therapeutic target in this disease.
+Hypoxia drives PASMC proliferation *in vivo* contributing to the development pulmonary hypertension in humans and animal models.
+Isocitrate dehydrogenase has previously been implicated in the pathobiology of this disease [@RN1259], and our findings suggest that reductive carboxylation catalyzed by isocitrate dehydrogenase may be a metabolic vulnerability of hypoxic PASMCs associated with pulmonary vascular disease.
 
@@
 
-(ref:f1) **Effects of prolyl hydroxylase inhibition on extracellular metabolite fluxes in lung fibroblasts.** Lung fibroblasts (LFs) were cultured with 21% oxygen (*red*), 0.5% oxygen (*blue*), DMSO (0.1% v/v, *green*), or molidustat (BAY, 10 μM, *purple*) beginning 24 h prior to time 0.
+(ref:m1) **Effects of prolyl hydroxylase inhibition on extracellular metabolite fluxes in lung fibroblasts.** Lung fibroblasts (LFs) were cultured with 21% oxygen (*red*), 0.5% oxygen (*blue*), DMSO (0.1% v/v, *green*), or molidustat (BAY, 10 μM, *purple*) beginning 24 h prior to time 0.
 (**A**) Growth curves of LFs in each experimental condition (n = 8).
@@
 
-(ref:f2) **Stable isotope tracing of lung fibroblasts.** (**A**) Fraction of pyruvate labeling following treatment of lung fibroblasts (LFs) with [U-^13^C~6~] glucose.
+(ref:m2) **Stable isotope tracing of lung fibroblasts.** (**A**) Fraction of pyruvate labeling following treatment of lung fibroblasts (LFs) with [U-^13^C~6~] glucose.
 (**B**) Rate values determined from asymptotic regression fit of the data from (A).
@@
 
-(ref:f3) **Metabolic flux maps of lung fibroblasts.** (**A**) Ratio of metabolic fluxes in 0.5% oxygen compared to 21% oxygen.
+(ref:m3) **Metabolic flux maps of lung fibroblasts.** (**A**) Ratio of metabolic fluxes in 0.5% oxygen compared to 21% oxygen.
 (**B**) Ratio of metabolic fluxes in cells treated with molidustat (BAY) compared to DMSO vehicle control.
@@
 
-(ref:f4) **PHD inhibition increases lactate uptake and oxidation.** Mass isotopomer distributions of key metabolites following labeling with [U-^13^C~3~] lactate (2 mM) for 72 h indicates increased lactate uptake and oxidation in hypoxia or with molidustat (BAY) treatment.
+(ref:m4) **PHD inhibition increases lactate uptake and oxidation.** Mass isotopomer distributions of key metabolites following labeling with [U-^13^C~3~] lactate (2 mM) for 72 h indicates increased lactate uptake and oxidation in hypoxia or with molidustat (BAY) treatment.
 Data are mean ± SEM of n = 4 biological replicates.
@@
 
-(ref:f5) **Metabolomic analysis of molidustat treatment in normoxia and hypoxia.** (**A**-**C**) Hypoxia inhibits the effects of HIF-1α stabilization on glycolysis.
+(ref:m5) **Metabolomic analysis of molidustat treatment in normoxia and hypoxia.** (**A**-**C**) Hypoxia inhibits the effects of HIF-1α stabilization on glycolysis.
 Lung fibroblasts (LFs) were cultured in standard growth medium and treated with molidustat (BAY, 10 μM) or vehicle (DMSO, 0.1%) in 21% or 0.5% oxygen conditions (n = 4).
 (A) Growth rates were determined by linear fitting of log-transformed growth curves.
-(**B**-**C**) Extracellular fluxes of glucose (note inverse *y*-axis) (B) and lactate (C).
+(**B**-**C**) Extracellular fluxes of glucose (note reversed *y*-axis) (B) and lactate (C).
 (**D**) Principle components analysis of intracellular metabolites following 72 h of treatment described above suggests a dominant effect of hypoxia over PHD inhibition on the metabolome (n = 4).
@@
 
-(ref:f6) **Transcriptomic analysis of molidustat treamtent in normoxia and hypoxia.** (**A**) Principle components analysis of transcriptional changes following 72 h of treatment with 0.5% oxygen or molidustat (BAY), separately or together (n = 4).
+(ref:m6) **Transcriptomic analysis of molidustat treamtent in normoxia and hypoxia.** (**A**) Principle components analysis of transcriptional changes following 72 h of treatment with 0.5% oxygen or molidustat (BAY), separately or together (n = 4).
 (**B**) Volcano plot of the differential effects of hypoxia and molidustat (BAY) treatment on intracellular metabolites.
@@
 
-```{r f1, fig.cap = "(ref:f1)"}
+```{r m1, fig.cap = "(ref:m1)"}
 knitr::include_graphics(
@@
 
-```{r f2, fig.cap = "(ref:f2)"}
+```{r m2, fig.cap = "(ref:m2)"}
 knitr::include_graphics(
@@
 
-```{r f3, fig.cap = "(ref:f3)"}
+```{r m3, fig.cap = "(ref:m3)"}
 knitr::include_graphics(
@@
 
-```{r f4, fig.cap = "(ref:f4)"}
+```{r m4, fig.cap = "(ref:m4)"}
 knitr::include_graphics(
@@
 
-```{r f5, fig.cap = "(ref:f5)"}
+```{r m5, fig.cap = "(ref:m5)"}
 knitr::include_graphics(
@@
 
-```{r f6, fig.cap = "(ref:f6)"}
+```{r m6, fig.cap = "(ref:m6)"}
 knitr::include_graphics(
@@
 The raw data and annotated analysis code necessary to reproduce this manuscript are contained in an R package research compendium available from the Oldham Lab GitHub repository (https://github.com/oldhamlab/Copeland.2021.hypoxia.flux).
-Data analysis, statistical comparisons, and visualization were performed in R [@R-base] using the packages listed below.
+Data analysis, statistical comparisons, and visualization were performed in R [@R-base] using the packages referenced in the data supplement.
 Experiments included technical and biological replicates as noted above.
@@
 
-Figure S6: **Isotopically non-stationary metabolic flux analysis.**
+Figure S6: **Isotope incorporation in key metabolites over the experimental time course.**
+
+Figure S7: **Isotopically non-stationary metabolic flux analysis.**
+
+Figure S8: **Metabolomic profiling of hypoxia and BAY treated lung fibroblasts.**
+
+Figure S9: **Transcriptomic profiling of hypoxia and BAY treated lung fibroblasts.**
 

### 2021-07-28 manuscript/supplement.Rmd
finish revisions of draft to send to coauthors

@@
   @bookdown2016, @ggplot22016, @knitr2015, @lme42015, @lmerTest2017, @lubridate2011, 
-  @MASS2002, @pbkrtest2014, @rmarkdown2018, @tidyverse2019, tinytex2019
+  @MASS2002, @pbkrtest2014, @rmarkdown2018, @tidyverse2019, tinytex2019, @fgsea, 
+  @TFEA.ChIP, @multiGSEA, @limma, @targets, @flextable, @officer, @patchwork
 ---
@@
 pasmc <- targets::tar_read(pasmc_hypoxia_table)
-figs <- targets::tar_load(targets::matches("s\\d_figure"))
+targets::tar_load(
+  c(
+    s1_figure, 
+    s2_figure, 
+    s3_figure,
+    s4_figure, 
+    s5_figure, 
+    s6_figure, 
+    s7_figure, 
+    s8_figure, 
+    s9_figure
+  )
+)
 ```
@@
 
-(ref:s9) **Transcriptomic profiling of hypoxia and BAY treated lung fibroblasts.** (**A**)
+(ref:s9) **Transcriptomic profiling of hypoxia and BAY treated lung fibroblasts.** (**A**, **B**) Volcano plots of differentially regulated transcripts with 0.5% oxygen in DMSO-treated cells (A) or BAY treatment in 21% oxygen-cultured cells (B).
+The top 10 up- and down-regulated metabolites are highlighted.
+(**C**) Venn diagram illustrating the number of differentially regulated genes following hypoxia (*blue*) or BAY treatment (*purple*).
+(**D**) Venn diagram illustrating the number of differentially enriched Hallmark gene sets with  hypoxia (*blue*) or BAY treatment (*purple*).
+(**E**, **F**) Normalized enrichment scores from gene set enrichment analysis of Hallmark gene sets affected by 0.5% oxygen (E) or BAY treatment (F).
 
@@
 
-::: {#refs_main}
-:::
-
 ::: {#refs_software}

### 2021-07-28 R/a_figures.R

@@ -131,8 +131,8 @@ arrange_s6 <- function(a, b) {
   a + b +
     theme_patchwork(
       design = layout,
-      widths = unit(4, "in"),
-      heights = unit(c(3.5), "in")
+      widths = unit(3.5, "in"),
+      heights = unit(c(3), "in")
     )
 }
 

### 2021-07-28 R/a_metabolomics.R

@@ -385,7 +385,9 @@ plot_metab_volcano <- function(results, mois = NULL, colors = NULL, xlab = NULL)
       size = 5/ggplot2::.pt,
       max.overlaps = 20,
       segment.size = 0.1,
-      nudge_x = -4,
+      # nudge_x = -4,
+      nudge_x = -5 - left$logFC,
+      hjust = 0,
       segment.color = "black",
       direction = "y",
       family = "Calibri",
@@ -401,7 +403,9 @@ plot_metab_volcano <- function(results, mois = NULL, colors = NULL, xlab = NULL)
       max.overlaps = 20,
       segment.size = 0.1,
       segment.color = "black",
-      nudge_x = 4.5,
+      # nudge_x = 4.5,
+      nudge_x = 5 - left$logFC,
+      hjust = 1,
       direction = "y",
       family = "Calibri",
       show.legend = FALSE
@@ -491,7 +495,7 @@ plot_mois <- function(clean, moi) {
     ggplot2::scale_fill_manual(values = clrs, limits = force) +
     ggplot2::labs(
       x = "Treatment",
-      y = "Peak area (normalized)",
+      y = "Peak area\n(normalized)",
       fill = NULL
     ) +
     ggplot2::ylim(c(0, NA)) +

### 2021-07-28 R/a_mid-figures.R

@@ -39,6 +39,7 @@ plot_labeling_rate <- function(mids) {
     ggplot2::geom_smooth(
       method = "nls",
       formula = y ~ SSasympOrig(x, asym, lrc),
+      size = 0.5,
       fullrange = TRUE,
       se = FALSE,
       show.legend = FALSE
@@ -65,7 +66,7 @@ plot_labeling_rate <- function(mids) {
     ggplot2::scale_fill_manual(values = clrs) +
     ggplot2::labs(
       x = "Time (h)",
-      y = "Pyruvate labeled",
+      y = "PYR fraction labeled",
       color = NULL,
       fill = NULL
     ) +

### 2021-07-28 _targets.R

@@ -631,7 +631,7 @@ list(
   ),
   tar_target(
     msea_hyp_plot,
-    plot_msea(metab_msea_hyp, lbls = c("Down in 0.5%", "Up in 0.5%"), vals = unname(clrs[c(1, 2)]))
+    plot_msea(metab_msea_hyp, lbls = c("Down in Hypoxia", "Up in Hypoxia"), vals = unname(clrs[c(1, 2)]))
   ),
   tar_target(
     msea_bay_plot,
@@ -1192,6 +1192,5 @@ list(
     path = path_to_manuscript("supplement.Rmd"),
     output_dir = path_to_manuscript("")
   )
-
 )
 

### 2021-09-02 manuscript/manuscript.Rmd
spell checking

@@
 
-Principle component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen (**Figure \@ref(fig:m5)D**).
+Principal component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen (**Figure \@ref(fig:m5)D**).
 Moreover, these hypoxia-treated cells were well-segregated from BAY-treated cells.
@@
 Given the disparate effects of hypoxic and pharmacologic PHD inhibition on cellular metabolism described above, we focused our transcriptomics analyses on the differences between hypoxia and BAY treatments.
-Principle component analysis again demonstrates clear separation among the four treatment groups (**Figure \@ref(fig:m6)A**).
-The first and second principle components correspond to 0.5% oxygen and BAY treatments, respectively.
-Consistent with our prior observations, the combination of 0.5% oxygen plus BAY was more similar to 0.5% oxygen alone with decreased distance between both hypoxia-treated groups along the axis of the second principle component.
+Principal component analysis again demonstrates clear separation among the four treatment groups (**Figure \@ref(fig:m6)A**).
+The first and second principal components correspond to 0.5% oxygen and BAY treatments, respectively.
+Consistent with our prior observations, the combination of 0.5% oxygen plus BAY was more similar to 0.5% oxygen alone with decreased distance between both hypoxia-treated groups along the axis of the second principal component.
 Again, consistent with the hypothesis that hypoxia uncouples the effects of BAY treatment.
@@
 (**B**-**C**) Extracellular fluxes of glucose (note reversed *y*-axis) (B) and lactate (C).
-(**D**) Principle components analysis of intracellular metabolites following 72 h of treatment described above suggests a dominant effect of hypoxia over PHD inhibition on the metabolome (n = 4).
+(**D**) Principal components analysis of intracellular metabolites following 72 h of treatment described above suggests a dominant effect of hypoxia over PHD inhibition on the metabolome (n = 4).
 (**E**) Volcano plot of the differential effects of hypoxia and molidustat (BAY) treatment on intracellular metabolites.
@@
 
-(ref:m6) **Transcriptomic analysis of molidustat treamtent in normoxia and hypoxia.** (**A**) Principle components analysis of transcriptional changes following 72 h of treatment with 0.5% oxygen or molidustat (BAY), separately or together (n = 4).
+(ref:m6) **Transcriptomic analysis of molidustat treamtent in normoxia and hypoxia.** (**A**) Principal components analysis of transcriptional changes following 72 h of treatment with 0.5% oxygen or molidustat (BAY), separately or together (n = 4).
 (**B**) Volcano plot of the differential effects of hypoxia and molidustat (BAY) treatment on intracellular metabolites.

### 2021-11-14 manuscript/manuscript.Rmd
include template monitoring in manuscript

@@
   bookdown::word_document2:
-    reference_docx: !expr system.file("manuscript", "template.docx", package = "Copeland.2021.hypoxia.flux")
+    # reference_docx: !expr system.file("manuscript", "template.docx", package = "Copeland.2021.hypoxia.flux")
     df_print: kable

### 2021-11-14 R/a_two-factor-fluxes.R

@@ -129,7 +129,7 @@ annot_twoby_densities <- function(blot_norm) {
       y_pos = Inf,
       vjust = 1.5,
       lab = dplyr::case_when(
-        adj.p.value < 0.06 ~ "*",
+        adj.p.value < 0.05 ~ "*",
         # adj.p.value < 0.07 ~ as.character(round(adj.p.value, 2)),
         TRUE ~ NA_character_
       )

### 2021-11-14 _targets.R

@@ -1199,15 +1199,44 @@ list(
 
   # write manuscript --------------------------------------------------------
 
+  tar_target(
+    template,
+    system.file("manuscript/template.docx", package = "Copeland.2021.hypoxia.flux"),
+    format = "file"
+  ),
   tar_render(
     manuscript,
     path = path_to_manuscript("manuscript.Rmd"),
-    output_dir = path_to_manuscript("")
+    output_dir = path_to_manuscript(""),
+    output_format = bookdown::word_document2(
+      reference_docx = template,
+      df_print = "kable",
+      fig_caption = TRUE,
+      number_sections = FALSE,
+      pandoc_args = c(
+        "--lua-filter=scholarly-metadata.lua",
+        "--lua-filter=author-info-blocks.lua",
+        "--lua-filter=pagebreak.lua"
+      )
+    )
   ),
   tar_render(
     supplement,
     path = path_to_manuscript("supplement.Rmd"),
-    output_dir = path_to_manuscript("")
-  )
+    output_dir = path_to_manuscript(""),
+    output_format = bookdown::word_document2(
+      reference_docx = template,
+      df_print = "kable",
+      fig_caption = TRUE,
+      number_sections = FALSE,
+      pandoc_args = c(
+        "--lua-filter=scholarly-metadata.lua",
+        "--lua-filter=author-info-blocks.lua",
+        "--lua-filter=pagebreak.lua",
+        "--lua-filter=multiple-bibliographies.lua"
+      )
+    )
+  ),
+  NULL
 )
 

### 2021-11-23 analysis/extracellular-fluxes.Rmd
move analysis out of inst to top-level directory
add simyc and oemyc data

--- 2021-03-20 inst/analysis/extracellular-fluxes.Rmd
@@
   message = FALSE, 
-  warning = TRUE, 
+  warning = FALSE, 
   fig.align = "center", 
@@
 
-# Data Analysis
+# Data Processing
 
@@
     y = volume, 
-    color = interaction(experiment, oxygen, treatment, sep = " | ")
+    color = interaction(oxygen, experiment, sep = " | ")
   ) +
@@
   unnest(c(data)) %>% 
-  group_by(cell_type, experiment, batch, date, oxygen, treatment, time) %>%
+  group_by(cell_type, experiment, batch, date, oxygen, virus, treatment, time) %>%
   summarize(conc = mean(conc, na.rm = TRUE)) %>% 
@@
         y = conc/1000, 
-        color = interaction(oxygen, treatment, sep = " | ")
+        color = interaction(oxygen, treatment, virus, sep = " | ")
       ) +
@@
     aes(
-      x = interaction(oxygen, treatment), 
+      x = interaction(oxygen, treatment, virus, sep = "\n"), 
       y = .data[[y]], 
@@
     ) +
-    facet_wrap(
+    facet_grid(
       ~cell_type, 
-      ncol = 1
+      scales = "free_x", 
+      space = "free_x"
     ) +

### 2021-11-23 R/1_expression.R

@@ -32,10 +32,11 @@ plot_blot <- function(blot_image, scale = 1.3, hjust = 0.15, vjust = 0.1) {
     )
 }
 
-normalize_densities <- function(blot_raw) {
+normalize_densities <- function(blot_raw)
+  {
   blot_raw %>%
     dplyr::filter(.data$time < 96) %>%
-    dplyr::filter(!(experiment == "lf_05-bay" & gel == "b")) %>%
+    dplyr::filter(!(experiment == "lf_05-bay" & gel %in% c("b", "e", "f"))) %>%
     tidyr::pivot_longer(
       blot:tidyselect::last_col(),
       names_to = "protein",

### 2021-11-23 R/1_extracellular-fluxes.R

@@ -29,7 +29,7 @@ clean_flux_meta <- function(file_list) {
     rlang::set_names(stringr::str_extract(basename(.), "^.+(?=_)")) %>%
     purrr::map_dfr(
       readr::read_csv,
-      col_types = "dcccdc",
+      col_types = "dccccdc",
       .id = "experiment"
     ) %>%
     tidyr::separate(experiment, c("cell_type", "experiment"), "_")
@@ -201,6 +201,7 @@ fill_missing_fluxes <- function(df, meta) {
       .data$date,
       .data$treatment,
       .data$oxygen,
+      .data$virus,
       tidyr::nesting(
         metabolite,
         type,
@@ -223,6 +224,7 @@ fill_missing_fluxes <- function(df, meta) {
         "experiment",
         "batch",
         "date",
+        "virus",
         "treatment",
         "oxygen",
         "metabolite",
@@ -246,6 +248,7 @@ fill_missing_fluxes <- function(df, meta) {
         "batch",
         "date",
         "oxygen",
+        "virus",
         "metabolite",
         "type",
         "detector",
@@ -277,7 +280,7 @@ filter_assays <- function(df) {
 
 assemble_evap_data <- function(data_list) {
   data_list[["evap"]] %>%
-    dplyr::group_by(.data$experiment, .data$oxygen, .data$treatment) %>%
+    dplyr::group_by(.data$experiment, .data$oxygen) %>%
     dplyr::mutate(plate_mass = .data$mass - .data$mass[[1]]) %>%
     dplyr::filter(.data$time != -24) %>%
     dplyr::mutate(volume = 2 * .data$plate_mass / .data$plate_mass[[1]]) %>%
@@ -291,34 +294,32 @@ assemble_evap_data <- function(data_list) {
     tidyr::unnest(c(.data$data, .data$pred_vol)) %>%
     dplyr::select(-c(.data$volume, .data$plate_mass, .data$mass)) %>%
     dplyr::rename(volume = .data$pred_vol) %>%
-    tidyr::separate(.data$experiment, c("cell_type", "experiment", "batch", "date"), "_") %>%
-    dplyr::mutate(
-      treatment = replace(treatment, experiment == "05-simyc", "siCTL")
-    )
+    tidyr::separate(.data$experiment, c("cell_type", "experiment", "batch", "date"), "_")
 }
 
[…]

### 2021-11-23 R/1_myc.R

@@ -1,17 +1,27 @@
 # 1_myc.R
 
-combine_fluxes <- function(growth_rates, fluxes)
+combine_fluxes <- function(growth_rates, fluxes, exp)
 {
   growth_rates %>%
-    dplyr::filter(experiment == "05-simyc") %>%
+    dplyr::filter(experiment == exp) %>%
     dplyr::rename(flux = mu) %>%
     dplyr::select(-X0) %>%
     dplyr::mutate(metabolite = "growth") %>%
-    dplyr::bind_rows(dplyr::filter(fluxes, experiment == "05-simyc")) %>%
-    dplyr::filter(treatment %in% c("siCTL", "siMYC"))
+    dplyr::bind_rows(dplyr::filter(fluxes, experiment == exp)) %>%
+    dplyr::filter(treatment %nin% c("siHIF1A", "siHIF2A")) %>%
+    dplyr::mutate(
+      flux = ifelse(
+        experiment == "bay-myc" & metabolite == "lactate" & date %in% c("2021-09-21", "2021-11-01"),
+        flux / 3,
+        flux
+      )
+    ) %>%
+    dplyr::group_by(oxygen, treatment, virus) %>%
+    wmo::remove_nested_outliers(flux, TRUE) %>%
+    return()
 }
 
-annot_fluxes <- function(df)
+annot_fluxes_simyc <- function(df)
 {
   df %>%
     dplyr::group_by(metabolite) %>%
@@ -22,7 +32,7 @@ annot_fluxes <- function(df)
         .x,
         "pairwise" ~ oxygen * treatment,
         simple = "each",
-        adjust = "Tukey",
+        adjust = "mvt",
         combine = TRUE
       )[["contrasts"]]
       ),
@@ -39,18 +49,46 @@ annot_fluxes <- function(df)
     )
 }
 
-plot_myc <- function(df, annot, metab, ylab)
+annot_fluxes_oemyc <- function(df)
+{
+  df %>%
+    dplyr::group_by(metabolite) %>%
+    tidyr::nest() %>%
+    dplyr::mutate(
+      m = purrr::map(data, ~lmerTest::lmer(flux ~ virus * treatment + (1 | date), data = .x)),
+      res = purrr::map(m, ~emmeans::emmeans(
+        .x,
+        "pairwise" ~ virus * treatment,
+        simple = "each",
+        adjust = "mvt",
+        combine = TRUE
+      )[["contrasts"]]
+      ),
+      out = purrr::map(res, broom::tidy)
[…]

### 2021-11-23 R/1_two-factor-fluxes.R

@@ -64,8 +64,8 @@ plot_twoby_fluxes <- function(df, annot, metab, ylab) {
     )
 }
 
-analyze_twoby_fluxes <- function(growth_rates, fluxes) {
-
+analyze_twoby_fluxes <- function(growth_rates, fluxes)
+{
   df <-
     growth_rates %>%
     dplyr::filter(experiment == "05-bay") %>%
@@ -101,12 +101,14 @@ analyze_twoby_fluxes <- function(growth_rates, fluxes) {
     )
 
   list(data = df, annot = annot)
-
 }
 
-annot_twoby_densities <- function(blot_norm) {
+annot_twoby_densities <- function(blot_norm)
+{
   blot_norm %>%
     dplyr::filter(experiment == "lf_05-bay") %>%
+    dplyr::group_by(protein, oxygen, treatment) %>%
+    wmo::remove_nested_outliers(fold_change, remove = TRUE) %>%
     dplyr::group_by(protein) %>%
     tidyr::nest() %>%
     dplyr::mutate(

### 2021-11-23 _targets.R

@@ -730,19 +730,36 @@ list(
 
   tar_target(
     simyc_fluxes,
-    combine_fluxes(growth_rates, fluxes)
+    combine_fluxes(growth_rates, fluxes, exp = "05-simyc")
   ),
   tar_target(
     simyc_fluxes_annot,
-    annot_fluxes(simyc_fluxes)
+    annot_fluxes_simyc(simyc_fluxes)
   ),
   tar_target(
     simyc_fluxes_growth_plot,
-    plot_myc(simyc_fluxes, simyc_fluxes_annot, "growth", "Growth Rate (/h)")
+    plot_myc(simyc_fluxes, simyc_fluxes_annot, "growth", "Growth Rate (/h)", x = treatment, fill = oxygen) +
+      ggplot2::geom_hline(yintercept = 0, size = 0.25)
   ),
   tar_target(
     simyc_fluxes_lactate_plot,
-    plot_myc(simyc_fluxes, simyc_fluxes_annot, "lactate", "Lactate\n(fmol/cell/h)")
+    plot_myc(simyc_fluxes, simyc_fluxes_annot, "lactate", "Lactate\n(fmol/cell/h)", x = treatment, fill = oxygen)
+  ),
+  tar_target(
+    oemyc_fluxes,
+    combine_fluxes(growth_rates, fluxes, exp = "bay-myc")
+  ),
+  tar_target(
+    oemyc_fluxes_annot,
+    annot_fluxes_oemyc(oemyc_fluxes)
+  ),
+  tar_target(
+    oemyc_fluxes_growth_plot,
+    plot_myc(oemyc_fluxes, oemyc_fluxes_annot, "growth", "Growth Rate (/h)", x = virus, fill = treatment)
+  ),
+  tar_target(
+    oemyc_fluxes_lactate_plot,
+    plot_myc(oemyc_fluxes, oemyc_fluxes_annot, "lactate", "Lactate\n(fmol/cell/h)", x = virus, fill = treatment)
   ),
 
   # M1 ----------------------------------------------------------------------
@@ -1193,6 +1210,10 @@ list(
     s9_figure,
     write_figures(s9, "s9.pdf")
   ),
+
+# M7 ----------------------------------------------------------------------
+
+
 
   # resources table ---------------------------------------------------------
 

### 2021-12-07 manuscript/manuscript.Rmd
update manuscript with MYC data

--- 2021-09-02 manuscript/manuscript.Rmd
@@
 ---
-title: "Hypoxia uncouples HIF target gene transcription and glycolytic flux in proliferating primary cells"
+title: "MYC uncouples HIF target gene expression from glycolytic flux in hypoxic proliferating primary cells"
 
@@
 link-citations: yes
-
-output: 
-  bookdown::word_document2:
-    reference_docx: !expr system.file("manuscript", "template.docx", package = "Copeland.2021.hypoxia.flux")
-    df_print: kable
-    fig_caption: true
-    number_sections: false
-    pandoc_args:
-      - '--lua-filter=scholarly-metadata.lua'
-      - '--lua-filter=author-info-blocks.lua'
-      - '--lua-filter=pagebreak.lua'
-      # - '--lua-filter=multiple-bibliographies.lua'
-
 compact-title: true
@@
     m5_figure, 
-    m6_figure
+    m6_figure, 
+    m7_figure
   )
@@
 # Summary
-Hypoxia is an important environmental stressor that causes transcriptional and metabolic reprogramming in cells to facilitate their survival.
-Here, we performed stable isotope tracing and metabolic flux analyses of proliferating human lung fibroblasts and pulmonary artery smooth muscle cells in hypoxia.
-Despite activation of the hypoxia-inducible factor (HIF) transcriptional program and increased expression of glycolytic enzymes, glycolytic flux was decreased in hypoxic cells.
-While pharmacologic stabilization of HIF in normoxia with the prolyl hydroxylase inhibitor molidustat did increase glycolytic flux as expected, hypoxia abrogated this effect of molidustat treatment.
-Multi-omic profiling of cells treated with hypoxia or molidustat, separately or together, revealed distinct molecular responses to hypoxia and pharmacologic prolyl hydroxylase inhibition.
-Together, these data suggest that primary cell bioenergetic metabolism is closely coupled to cell proliferation rate, and that other factors supersede the anticipated effects of HIF-dependent up-regulation of glycolytic gene expression on glycolytic flux.
+
+Hypoxia requires metabolic adaptations to sustain cellular functions that underlie numerous physiologic and pathologic processes.
+While many studies have explored the metabolic consequences of hypoxia in cancer models, comparatively little is known about the metabolic response of primary cells to hypoxia.
+Here, we performed metabolic flux analyses of proliferating human lung fibroblasts and pulmonary artery smooth muscle cells in hypoxia.
+Unexpectedly, glycolytic flux was decreased in hypoxic cells despite activation of the hypoxia-inducible factor (HIF) transcriptional program and increased expression of glycolytic enzymes.
+Pharmacologic activation of HIF with the prolyl hydroxylase (PHD) inhibitor molidustat in normoxia did increase glycolytic flux, but hypoxia abrogated this effect.
+Multi-omic profiling of cells treated with hypoxia or molidustat, separately or together, revealed distinct molecular responses to hypoxia and pharmacologic PHD inhibition and suggested a critical role for MYC in modulating the HIF response in hypoxia.
+MYC knockdown in hypoxia increased lactate efflux while MYC overexpression in normoxia blunted the effects of molidustat treatment.
+Together, these data suggest that other factors, including MYC, supersede the anticipated effects of HIF-dependent up-regulation of glycolytic gene expression on glycolytic flux in hypoxic proliferating primary cells.
 
@@
 
-hypoxia, metabolic flux analysis, hypoxia-inducible factor, prolyl hydroxylase, metabolism
+hypoxia, metabolic flux analysis, hypoxia-inducible factor, prolyl hydroxylase, metabolism, MYC, pulmonary artery smooth muscle cell, lung fibroblast
 
@@
 PHDs are the principal oxygen sensors in metazoan cells [@RN999].
-PHDs are α-ketoglutrate-dependent dioxygenase enzymes that require molecular oxygen for their enzymatic activity.
+PHDs are α-ketoglutarate-dependent dioxygenase enzymes that require molecular oxygen for their enzymatic activity.
 When oxygen tension falls, PHD activity decreases, leading to HIF-1α stabilization and activation of its associated transcriptional program.
-Overall, the HIF-1 transriptional program should increase glycolytic capacity and divert glucose-derived pyruvate from oxidative phosphorylation toward lactate fermentation to maintain ATP production and to minimize the formation of reactive oxygen species (ROS) [@RN2525].
+Overall, the HIF-1 transcriptional program should increase glycolytic capacity and divert glucose-derived pyruvate from oxidative phosphorylation toward lactate fermentation to maintain ATP production and to minimize the formation of reactive oxygen species (ROS) [@RN2525].
 
@@
 
-To address these questions, here we have developed models of bioenergetic carbon flux in human lung fibroblasts (LFs) and pulmonary artery smooth muscle cells (PASMCs) cultured in 21% or 0.5% oxygen.
+To address these questions, we have developed models of bioenergetic carbon flux in human lung fibroblasts (LFs) and pulmonary artery smooth muscle cells (PASMCs) cultured in 21% or 0.5% oxygen.
 These cells were selected as they may be exposed to a wide range of oxygen concentrations *in vivo*, continue to proliferate despite hypoxic culture conditions *in vitro*, and play important roles in the pathology of human lung diseases where tissue hypoxia is a prominent feature.
 We found that hypoxia fails to increase glycolysis in these primary cells despite robust up-regulation of the HIF-1 transcriptional program.
-In normoxia, HIF-1α stabilization by the PHD inhibitor molidustat (BAY-85-3934, "BAY") [@RN580] did increase glycolysis and lactate efflux; however, hypoxia abrogated this response.
-These findings suggest the existence of important hypoxia-dependent regulatory mechanisms that override the metabolic consequences of HIF-1-dependent up-regulation of glycolytic gene expression in human primary cells.
+In normoxia, HIF-1α stabilization by the PHD inhibitor molidustat (BAY-85-3934, "BAY") [@RN580] did increase glycolysis and lactate efflux; however, hypoxia blocked this response.
+These findings suggested the existence of important hypoxia-dependent regulatory mechanisms that override the metabolic consequences of HIF-1-dependent up-regulation of glycolytic gene expression in human primary cells.
+Transcriptomic profiling suggested a critical role for the transcription factor c-MYC (MYC) in the adaptive response to hypoxia.
+Using knockdown and overexpression approaches, we demonstrate that MYC attenuates HIF-driven glycolysis in hypoxia and following BAY treatment.
 
@@
 Cells were seeded and placed into hypoxia for 24 h prior to sample collection to provide adequate time for activation of the hypoxia-dependent transcriptional program.
-From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (**Figure \@ref(fig:m1)A**), thought to be an indicator of metabolic steady state .
+From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (**Figure \@ref(fig:m1)A**), thought to be an indicator of metabolic steady state.
 LFs cultured in 0.5% oxygen grew slower than LFs cultured in 21% oxygen (**Figure \@ref(fig:m1)B**), but slower growth was not associated with decreased cell viability (**Figure S1A**).
@@
 No change in glucose or lactate fluxes were observed when the ambient oxygen level was decreased further to 0.2% (**Figure S2**).
-Similar results were also observed in PASMCs were lactate efflux was significantly *decreased* by 0.5% ambient oxygen (**Figure S3**).
+Similar results were also observed in PASMCs where lactate efflux was significantly *decreased* by 0.5% ambient oxygen (**Figure S3**).
 
-In addition to glucose and lactate, we also determinted the extracellular fluxes of pyruvate and amino acids.
-Overall, changes were modest, with hypoxia generally associated with decreased fluxes of all measured metabolites.
+In addition to glucose and lactate, we also determined the extracellular fluxes of pyruvate and amino acids.
+Overall, changes were modest, with hypoxia generally decreasing the fluxes of all measured metabolites.
 Notably, a substantial decrease in glutamine consumption was observed in LFs cultured in 0.5% oxygen.
-This observation is in contrast to previous studies in cancer cell metabolism demonstrating increased glutamine uptake as a key feature of the metabolic response to hypoxia in these cells [@RN1447; @RN2395; @RN628].
-Similar patterns were observed under 0.2% oxygen culture conditions.
+This observation is in contrast to previous studies in cancer cell metabolism demonstrating increased glutamine uptake as a key feature of the metabolic response to hypoxia [@RN1447; @RN2395; @RN628].
+Similar patterns were observed in 0.2% oxygen (**Figure S2**).
 In PASMCs, glutamine uptake did increase, as did the uptake of branched-chain amino acids and arginine (**Figure S3**), highlighting differential responses of these mesenchymal cells to hypoxia.
 
-Given that hypoxia did not increase glucose and lactate fluxes as expected from the associated increases in glycolytic gene expression, we next assessed the capacity of HIF-1α to augment glycolysis in LFs.
+Given that hypoxia did not increase glucose and lactate fluxes as expected based on the observed increases in glycolytic gene expression, we next assessed the capacity of HIF-1α to augment glycolysis in LFs.
 Cells were treated with the prolyl hydroxylase inhibitor BAY to stabilize HIF-1α under 21% oxygen conditions (**Figure \@ref(fig:m1)**).
@@
 
-To validate our findings from extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate and trace the incorporation of label into key carbon utilization pathways (**Figures \@ref(fig:m2), S4-S5**).
+To validate the findings from our extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate and trace the incorporation of ^13^C into key downstream carbon utilization pathways (**Figures \@ref(fig:m2), S4-S5**).
 Consistent with our extracellular flux measurements, we found that the rate of label incorporation from [U-^13^C~6~] glucose into pyruvate was similar across 21%, 0.5%, and DMSO-treated conditions, but markedly increased with BAY treatment (**Figures \@ref(fig:m3)A-B**).
@@
 The most notable differences between LF and PASMC labeling were observed in citrate.
-Compared with LFs, PASMCs demonstrated a marked reduction of M2-citrate labeled by glucose and no change in the labeling of M3-citrate by [U-^13^C~3~] glucose.
+Compared with LFs, PASMCs demonstrated a marked reduction of M2-citrate labeled by glucose and no change in the labeling of M3-citrate by [U-^13^C~6~] glucose.
+These data suggest decreased glucose entry into the TCA cycle *via* pyruvate dehydrogenase and no change in the activity of pyruvate carboxylase, respectively.
 Interestingly, unlike LFs, PASMCs demonstrated increased production of M5-citrate from [U-^13^C~5~] glutamine in hypoxia (`r m5[["mean_21%"]]`% ± `r m5[["se_21%"]]`% *v*.
-`r m5[["mean_0.5%"]]`% ± `r m5[["se_0.5%"]]`%, p = `r m5[["pval"]]`) consistent with increased reductive carboxylation in support of lipid synthesis in these cells, similar to prior observations in cancer cells.
+`r m5[["mean_0.5%"]]`% ± `r m5[["se_0.5%"]]`%, p = `r m5[["pval"]]`) consistent with increased reductive carboxylation in support of lipid synthesis in these cells, similar to prior observations in cancer cells, but distinct from LFs studied here.
 
@@
 To clarify how these labeling patterns reflect changes in intracellular metabolite fluxes, we next generated metabolic flux models incorporating the extracellular flux measurements and stable isotope tracing data described above.
-Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (**Figures \@ref(fig:m3)A, S6**).
+Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (**Figure S6**).
 Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by INCA [@RN1522; @RN2501; @RN964] (**Figures \@ref(fig:m3), S7, Tables S1-S3**).
@@
 A significant increase in pentose phosphate pathway flux was also observed, although the absolute flux through this pathway is low (**Figure \@ref(fig:m3)A**).
-By contrast, HIF-1 activation by BAY in 21% oxygen increased glycolysis and lactate fermentation by ~50% (**Figure \@ref(fig:m3)B**), but had a similar effect as hypoxia in decreasing serine and glutamine incorporation.
+By contrast, HIF-1 activation by BAY in 21% oxygen increased glycolysis and lactate fermentation by ~50% (**Figure \@ref(fig:m3)B**), but had a similar effect to hypoxia in decreasing serine and glutamine incorporation.
 Metabolite fluxes in DMSO-treated cells were similar to 21% oxygen controls.
@@
 With hypoxia or BAY treatment, we observed increased labeling of the TCA metabolites citrate (CIT), α-ketoglutarate (AKG), malate (MAL), and aspartate (ASP) in LFs.
-Interestingly, although increased labeling of pyruvate and lactate were observed in hypoxic PASMCs, the label was not incorporated into the TCA cycle as observed in LFs (**Figure S5**).
+Interestingly, although increased labeling of pyruvate was observed in hypoxic PASMCs, the label was not incorporated into the TCA cycle as observed in LFs (**Figure S5**).
 
@@
 
-Principal component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen (**Figure \@ref(fig:m5)D**).
+Principal component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen than among the BAY-treatment groups (**Figure \@ref(fig:m5)D**).
 Moreover, these hypoxia-treated cells were well-segregated from BAY-treated cells.
-These observations are again consistent with the results of metabolic flux models demonstrating an overriding effect of hypoxia *per se* on metabolic flux and highlighting important differences between hypoxic and pharmacologic PHD inhibition.
+These observations are again consistent with the results of the metabolic flux models demonstrating an overriding effect of hypoxia *per se* on metabolic flux and highlighting important differences between hypoxic and pharmacologic PHD inhibition.
 To identify the key differences among these treatments, we performed linear modeling focusing on those metabolites with a significant interaction term in the model described by treatment × oxygen (**Figure \@ref(fig:m5)E**) (*i.e.*, metabolites where the change with hypoxia treatment was different than the change with BAY treatment).
@@
 As anticipated, both hypoxia and BAY treatment induced substantial changes in gene expression (**Figure S9A, S9B**).
-Of the 7789 differentially expressed genes across both conditions, 891 (11%) were unique to BAY treatment in normoxia, 1649 (21%) were shared between BAY and hypoxia, while 5249 (67%) were unique to 0.5% culture (**Figure S9C**).
+Of the 7,789 differentially expressed genes across both conditions, 891 (11%) were unique to BAY treatment in normoxia, 1,649 (21%) were shared between BAY and hypoxia, while 5,249 (67%) were unique to 0.5% culture (**Figure S9C**).
 Gene set enrichment analysis of these differentially regulated metabolites was performed using Molecular Signatures Database "Hallmark" gene sets [@RN3112] (**Figures S9D, S9E, S9F**).
@@
 Similarly, the gene set for glycolysis was also more enriched following BAY treatment than hypoxia treatment.
-The most enriched gene sets associated with hypoxia included pro-proliferative "E2F targets" and "G2/M checkpoint" proteins as well as "Myc targets".
-These findings were further supported by a transcription factor enrichment analysis identifying enrichment of Myc target proteins associated with hypoxia (**Figure \@ref(fig:m6)E**).
-Consistent with these bioinformatic results, immunoblotting demonstrated up-regulation of Myc protein in hypoxia-treated cells, but not with BAY-treatment alone (**Figure \@ref(fig:m6)F**).
-These data provide a starting point for interrogating the relative hierarchy of transcription factor regulation of gene expression by hypoxia in proliferating primary cells and suggest several potential mechanisms that may contribute to the uncoupling of glycolytic gene transcription and glycolytic flux reported here.
+The most enriched gene sets associated with hypoxia included pro-proliferative "E2F targets" and "G2/M checkpoint" proteins as well as "MYC targets".
+These findings were further supported by a transcription factor enrichment analysis identifying enrichment of MYC target proteins associated with hypoxia (**Figure \@ref(fig:m6)E**).
+
+## MYC antagonizes HIF-dependent glycolytic fluxes
+
+Consistent with our bioinformatic results, immunoblotting demonstrated increased MYC protein levels in hypoxia-treated cells, but not with BAY-treatment alone, where MYC was decreased (**Figure \@ref(fig:m7)A-B**).
+To test the hypothesis that hypoxia-induced MYC expression inhibits glycolysis in hypoxic primary cells, we first combined MYC knockdown with hypoxia treatment (**Figure \@ref(fig:m7)C-E**).
+As expected, MYC deficient cells proliferated more slowly in normoxia and MYC was absolutely essential to sustain cell proliferation in hypoxia (**Figure \@ref(fig:m7)D**).
+Consistent with our hypothesis, MYC knockdown cells demonstrated increased lactate efflux upon hypoxia treatment, unlike control siRNA-treated cells (**Figure \@ref(fig:m7)E**).
+We next performed the complementary experiment to determine whether MYC overexpression could attenuate the increase in glycolysis observed with BAY treatment (**Figure \@ref(fig:m7)F-H**).
+MYC increased the proliferation rate of DMSO-treated cells, although did not augment the proliferation rate of BAY-treated cells.
+As expected, MYC overexpression blocked the BAY-stimulated increase in lactate efflux.
+Together, these data suggest that hypoxia-induced MYC expression may be one factor that uncouples the HIF transcriptional program from glycolytic flux in proliferating primary cells.
 
@@
 
+<!-- MYC role in metabolism -->
+
+Our data indicate that hypoxia-induced MYC expression is one such regulatory mechanism.
+MYC is a transcription factor that regulates the expression of numerous genes involved in many biological processes, including metabolism, proliferation, apoptosis, and differentiation [@dang2006; @li2020d; @stine2015].
+As deregulated MYC activity has been associated with the majority of human cancers [@vita2006], much of our understanding of MYC regulation comes from studies using cancer cell models [@li2020d; @stine2015; @dang2012a; @madden2021], while the role of MYC in the biology of transformed cells is less well understood.
+The literature describes a complex and reciprocal relationship between HIF and MYC that depends on both environmental (*e.g.*, hypoxia) and cellular context [@li2020d].
+Generally, HIF-1 has been observed to inhibit MYC through multiple mechanisms [koshiji2004; koshiji2005; @gordan2007; @zhang2007], and this previous work is consistent with our observations that HIF stabilization following BAY treatment decreased MYC protein and target gene expression.
+Conversely, MYC has been implicated in increased HIF activity through transcriptional and post-transcriptional mechanisms [@zhang2009a; @chen2013; @doe2012], primarily in the context of malignant transformation.
+The observation that MYC may antagonize the transcriptional effects of HIF to sustain primarily cell proliferation and metabolism in hypoxia suggests a substantially different regulatory relationship than has been previously described.
+
+Understanding how MYC transcriptional activity affects hypoxic primary cell metabolism is imperative to our understanding of cellular adaptation to hypoxia.
+MYC stimulates the expression of nuclear-encoded mitochondrial genes and promotes mitochondrial biogenesis, both directly and through activation of mitochondrial transcriptional factor A (TFAM) [@li2005a].
+Indeed, the oxidative phosphorylation gene set was relatively enriched with hypoxia treatment compared to BAY treatment (**Figure \@ref(fig:m6)D**).
+In this way, hypoxic MYC activation may sustain energy production by oxidative phosphorylation, thereby decreasing the energetic demands driving increased glycolytic flux.
+Beyond oxidative phosphorylation, MYC targets genes involved in many other intermediary metabolic pathways, including amino acids, nucleotides, and lipids [@stine2015], that may also impact the central pathways of carbon metabolism studied in this work.
+
 <!-- Important non-HIF mechanisms -->
 
-The identification of these mechanisms is of critical importance for our understanding of primary cell adaptation to hypoxia.
-While the role of HIFs in the hypoxia response has been extensively studied, relatively less is known about HIF-independent features of the hypoxia response.
+Beyond MYC, the identification of other HIF-independent mechanisms regulating primary cell adaption to hypoxia is of critical importance.
 Cells express several oxygen-dependent enzymes in addition to PHD whose activities may be impacted in hypoxia but not by PHD inhibition.
@@
 In addition to dioxygenase enzymes, electron transport chain dysfunction resulting from impaired Complex IV activity leads to increased ROS production in hypoxia [@RN295].
-Mitochondrial ROS increase the half-lives of several mRNAs in hypoxia, including Myc, as we observe in this work, independent of HIF stabilization [@RN742].
+Mitochondrial ROS increase the half-lives of several mRNAs in hypoxia, including MYC, as we observe in this work, independent of HIF stabilization [@RN742].
 Finally, hypoxia imposes a reductive stress on cells associated with an increase in the NADH/NAD^+^ ratio secondary to impaired electron transport (**Figure S5K**) [@RN2618; @RN2619].
 NADH accumulation may slow glycolysis *via* feedback inhibition of GAPDH [@RN2210].
-Any or all of these molecular mechanisms may contribute to uncoupling glycolytic enzyme expression from glycolytic flux as observed in the experiments described here.
+Any or all of these molecular mechanisms may also contribute to uncoupling glycolytic enzyme expression from glycolytic flux as observed in the experiments described here.
 
@@
 In addition to its effects on cellular metabolism, another canonical role of HIF-1 activation is slowing of cellular proliferation rate in the face of limited oxygen availability [@RN615] and this is what we observed here (**Figures \@ref(fig:m1)B, \@ref(fig:m5)A**).
-The effects of HIF-1 on cell proliferation rate are mediated, in part, by increased expression of cyclin-dependent kinase inhibitor p21 (*CDKN1A*) and inhibition of E2F targets [@RN3152] and inhibition of pro-proliferative Myc signaling [@RN3151].
+The effects of HIF-1 on cell proliferation rate are mediated, in part, by increased expression of cyclin-dependent kinase inhibitor p21 (*CDKN1A*) and inhibition of E2F targets [@RN3152] and inhibition of pro-proliferative MYC signaling [@RN3151].
 These transcriptional effects are precisely what is observed in BAY treated LFs in normoxia.
-By contrast, hypoxia culture was associated with decreased expression of p21, consistent with a previous report [@RN3150], as well as increased expression of Myc protein and enrichment of Myc target genes.
+By contrast, hypoxia culture was associated with decreased expression of p21, consistent with a previous report [@RN3150], as well as increased expression of MYC protein and enrichment of MYC target genes.
 Indeed, the most marked differences between hypoxia and BAY treatment on LF gene transcription were the up-regulation of pro-proliferative gene sets containing E2F targets and G2/M checkpoint proteins.
-Much of this transcriptional response may be mediated by hypoxia-induced up-regulation of Myc, which is known to stimulate cell cycle progression through its effects on the expression and activity of cyclins, cyclin-dependent kinases, and cyclin-dependent kinase inhibitors [@RN3154]; however, the relative importance of Myc and the hypoxia-mediated mechanism of its induction remain to be elucidated in these primary cells.
-Clarifying the complex interactions among HIFs, Myc, and cell proliferation will be important for understanding the cellular response of these mesenchymal cells to tissue injury.
+Much of this transcriptional response may be mediated by hypoxia-induced up-regulation of MYC, which is known to stimulate cell cycle progression through its effects on the expression and activity of cyclins, cyclin-dependent kinases, and cyclin-dependent kinase inhibitors [@RN3154].
+Clarifying the complex interactions among HIFs, MYC, and cell proliferation will be important for understanding the cellular response of these mesenchymal cells to tissue injury.
 
@@
 Given the prominence of HIFs in mediating the transcriptional response to hypoxia, it is somewhat surprising that neither PHD, HIFs, nor their downstream targets were found to be selectively essential as a function of oxygen tension in a genome-wide CRISPR growth screen of K562 human lymphoblasts cultured in normoxia or hypoxia [@RN953].
+Similarly, knockout of HIF signaling did not affect growth, internal metabolite concentrations, glucose consumption, or lactate production under hypoxia by human acute myeloid leukemia cells [@wierenga2019].
+Together with our results, these studies highlight the need for additional research linking hypoxia-induced metabolic changes to their transcriptional and post-transcriptional regulatory mechanisms, particularly in primary cells.
 
@@
 
-In summary, in this metabolic flux analysis of proliferating human primary cells *in vitro*, we have demonstrated that hypoxia uncouples an increase in HIF-dependent glycolytic gene transcription from glycolytic flux.
+In summary, in this metabolic flux analysis of proliferating human primary cells *in vitro*, we have demonstrated that MYC uncouples an increase in HIF-dependent glycolytic gene transcription from glycolytic flux in hypoxia.
 Indeed, the degree of metabolic reprogramming in hypoxia was modest and suggests close coupling between proliferation and metabolism.
-In light of our findings, additional studies are warranted to clarify the role of HIFs in mediating the metabolic response to hypoxia and to identify other key regulators of hypoxic metabolic reprogramming in primary cells.
+In light of our findings, additional studies are warranted to clarify the role of HIFs in mediating the metabolic response to hypoxia, to determine how MYC activity is regulated by hypoxia, and to identify other key regulators of hypoxic metabolic reprogramming in primary cells.
 Moreover, these data strongly caution investigators against drawing conclusions about metabolite flux from measures of gene transcription alone.
@@
 (**E**) Transcription factor enrichment analysis suggests mechanisms for differential regulation of gene expression following hypoxia or BAY treatment.
-(**F**) Relative change in Myc protein levels after 72 h of treatment (n = 4) and a representative immunoblot.
-Comparisons were performed using a mixed-effects linear model with replicate as a random effect.
+
+(ref:m7) **MYC regulates HIF-dependent glycolytic flux.** (**A**) Representative immunoblot of MYC protein expression in lung fibroblsts following 72 h of treatment with 0.5% oxygen or molidustat (BAY, B).
+(**B**) Quantification of immunoblots shown in (A).
+(**C**) Representative immunoblot of lung fibroblasts (LFs) treated with siRNA targeting MYC (M) demonstrating adequate protein knockdown.
+(**D**) Growth rates of MYC-knockdown cells cultured in hypoxia.
+(**E**) Lactate efflux rates of MYC-knockdown cells cultured in hypoxia.
+(**F**) Representative immunoblot of LFs treated with MYC adenovirus.
+(**G**) Growth rates of MYC overexpressing cells cultured with BAY.
+(**H**) Lactate efflux rates of MYC overexpressing cells cultured with BAY.
+Data are mean Comparisons were performed using a mixed-effects linear model with replicate as a random effect.
 Adjusted p-values for the indicated comparisons were determined using Tukey's *post hoc* test.
-Black * indicate a significant effect of treatment, colored * indicate a significant effect of oxygen within a given treatment as indicated by the color.
+Black * indicates a significant effect of treatment, colored * indicate a significant effect of oxygen within a given treatment as indicated by the color.
+Data are mean ± SEM of n = 4-5 biological replicates.
+Comparisons were made using linear mixed effects models with treatment group as a fixed effect and biological replicate as a random effect.
+The multivariate *t* distribution was used to determine significant differences as indicated by black * for difference within groups defined by the x-axis and colored * for differences between groups defined by the x-axis as indicated by the color with adjusted p-values < 0.05 considered significant.
 
@@
     "manuscript/figures/m6.pdf",
+    package = "Copeland.2021.hypoxia.flux"
+  )
+)
+```
+
+\newpage
+
+```{r m7, fig.cap = "(ref:m7)"}
+knitr::include_graphics(
+  system.file(
+    "manuscript/figures/m7.pdf",
     package = "Copeland.2021.hypoxia.flux"
@@
 Amino acids in deproteinized medium were derivatized with *o*-phthalaldehyde (OPA) and 9-fluorenylmethylchloroformate (FMOC) immediately prior to separation with a Poroshell HPH-C18 column (4.6 × 100 mm, 2.7 μm) on an Infinity II high-performance liquid chromatography system with ultraviolet and fluorescence detection of OPA- and FMOC-derivatized amino acids, respectively, according to the manufacturer's protocol (Agilent) [@RN1294].
+
+### MYC knockdown
+
+Approximately 1.25 M lung fibroblasts were reverse transfected in 6-cm dishes with 40 pmol siMYC or non-targeting siCTL pools (Dharmacon) in 20 uL RNAiMAX (Thermo).
+After 24 h, cells were collected by trypsinization and re-seeded as described in *Metabolic flux protocol* above for growth rate and lactate efflux measurements.
+
+### MYC overexpression
+
+Lung fibroblasts were seeded at 25,000 cells per 35 mm dish on Day -2.
+On Day -1, cells were transduced with adenovirus for MYC (Vector Biolabs) or YFP overexpression [@RN1603].
+After 24 h, the medium was changed and samples were collected as described in *Metabolic flux protocol* above.
 

### 2021-12-07 manuscript/supplement.Rmd
update manuscript with MYC data

@@
 title: "Supplemental Information"
-subtitle: "Hypoxia uncouples HIF gene transcription and metabolic flux in proliferating primary cells"
+subtitle: "MYC uncouples HIF target gene expression from glycolytic flux in hypoxic proliferating primary cells"
 
@@
 link-citations: yes
-
-output: 
-  bookdown::word_document2:
-    reference_docx: !expr system.file("manuscript", "template.docx", package = "Copeland.2021.hypoxia.flux")
-    df_print: kable
-    number_sections: false
-    pandoc_args:
-      - '--lua-filter=scholarly-metadata.lua'
-      - '--lua-filter=author-info-blocks.lua'
-      - '--lua-filter=pagebreak.lua'
-      - '--lua-filter=multiple-bibliographies.lua'
-
 compact-title: true

### 2021-12-07 R/1_figures.R

@@ -253,6 +253,8 @@ create_resources <- function() {
     "Antibodies", "LDHA", "Cell Signaling Technologies", "2012",
     "Antibodies", "HRP-α-Rabbit IgG", "Cell Signaling Technologies", "7074",
     "Antibodies", "HRP-α-Mouse IgG", "Cell Signaling Technologies", "7076",
+    "Bacterial and virus strains", "c-MYC adenovirus", "Vector Biolabs", "1285",
+    "Bacterial and virus strains", "YFP adenovirus", "Oldham et al., 2015", "",
     "Chemicals, peptides, and recombinant proteins", "[1,2-^13^C~1~] glucose", "Cambridge Isotope Labs", "CLM-504-PK",
     "Chemicals, peptides, and recombinant proteins", "[U-^13^C~6~] glucose", "Cambridge Isotope Labs", "CLM-1396-PK",
     "Chemicals, peptides, and recombinant proteins", "[U-^13^C~5~] glutamine", "Cambridge Isotope Labs", "CLM-1822-H-PK",
@@ -268,7 +270,9 @@ create_resources <- function() {
     "Experimental models: Cell lines", "Pulmonary artery smooth muscle cells", "Lonza", "CC-2581",
     "Oligonucleotides", "ACTB (Hs03023943_g1)", "Life Technologies", "4351370",
     "Oligonucleotides", "GLUT1 (Hs00892681_m1)", "Life Technologies", "4351370",
-    "Oligonucleotides", "LDHA (Hs00855332_g1)", "Life Technologies", "4351370"
+    "Oligonucleotides", "LDHA (Hs00855332_g1)", "Life Technologies", "4351370",
+    "Oligonucleotides", "MYC ON-TARGETplus siRNA", "Dharmacon", "L-003282-02-0005",
+    "Oligonucleotides", "ON-TARGETplus non-targeting control pool", "Dharmacon", "D-001810-10-05"
   ) %>%
     flextable::as_grouped_data(groups = c("category")) %>%
     flextable::as_flextable(hide_grouplabel = TRUE) %>%

### 2021-12-14 manuscript/manuscript.Rmd
final changes to manuscript before initial submission

@@
 - Benjamin A. Olenchock: 
+    institute: 1
+- David R. Ziehr:
     institute: 1, 2
-- David R. Ziehr:
-    institute: 1, 4
 - Sarah McGarrity: 
-    institute: 1, 5
+    institute: 1, 3
 - Kevin Leahy: 
@@
 - Jamey D. Young: 
-    institute: 3
+    institute: 4
 - Joseph Loscalzo: 
@@
 - 1: Department of Medicine, Brigham and Women's Hospital and Harvard Medical School, Boston, MA
-- 2: Regeneron Pharmaceuticals, Tarrytown, NY
-- 3: Departments of Chemical & Biomolecular Engineering and Molecular Physiology & Biophysics, Vanderbilt University, Nashville, TN
-- 4: Department of Medicine, Massachusetts General Hospital and Harvard Medical School, Boston, MA
-- 5: Center for Systems Biology, School of Health Sciences, University of Iceland, Reykjavik, Iceland
+- 2: Department of Medicine, Massachusetts General Hospital and Harvard Medical School, Boston, MA
+- 3: Center for Systems Biology, School of Health Sciences, University of Iceland, Reykjavik, Iceland
+- 4: Departments of Chemical & Biomolecular Engineering and Molecular Physiology & Biophysics, Vanderbilt University, Nashville, TN
 
@@
 MYC knockdown in hypoxia increased lactate efflux while MYC overexpression in normoxia blunted the effects of molidustat treatment.
-Together, these data suggest that other factors, including MYC, supersede the anticipated effects of HIF-dependent up-regulation of glycolytic gene expression on glycolytic flux in hypoxic proliferating primary cells.
+Together, these data suggest that other factors, notably MYC, supersede the anticipated effects of HIF-dependent up-regulation of glycolytic gene expression on glycolytic flux in hypoxic proliferating primary cells.
 
@@
 To address these questions, we have developed models of bioenergetic carbon flux in human lung fibroblasts (LFs) and pulmonary artery smooth muscle cells (PASMCs) cultured in 21% or 0.5% oxygen.
-These cells were selected as they may be exposed to a wide range of oxygen concentrations *in vivo*, continue to proliferate despite hypoxic culture conditions *in vitro*, and play important roles in the pathology of human lung diseases where tissue hypoxia is a prominent feature.
+These cells were selected as they may be exposed to a wide range of oxygen concentrations *in vivo*, continue to proliferate despite hypoxic culture conditions *in vitro*, and play important roles in the pathology of non-cancerous diseases in which tissue hypoxia is a prominent feature.
 We found that hypoxia fails to increase glycolysis in these primary cells despite robust up-regulation of the HIF-1 transcriptional program.
@@
 Transcriptomic profiling suggested a critical role for the transcription factor c-MYC (MYC) in the adaptive response to hypoxia.
-Using knockdown and overexpression approaches, we demonstrate that MYC attenuates HIF-driven glycolysis in hypoxia and following BAY treatment.
+Using knockdown and overexpression approaches, we demonstrate that MYC attenuates HIF-driven glycolysis in hypoxia and following BAY treatment in these primary cells.
 
@@
 Cells were seeded and placed into hypoxia for 24 h prior to sample collection to provide adequate time for activation of the hypoxia-dependent transcriptional program.
+We selected 0.5% oxygen for hypoxia as this level yielded the most reproducible phenotypic differences compared to 21% oxygen culture while being physiologically relevant and above the K~M~ of cytochrome *c* oxidase (electron transport chain complex IV) for oxygen [@RN1198; @wenger2015].
 From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (**Figure \@ref(fig:m1)A**), thought to be an indicator of metabolic steady state.
@@
 To validate the findings from our extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate and trace the incorporation of ^13^C into key downstream carbon utilization pathways (**Figures \@ref(fig:m2), S4-S5**).
-Consistent with our extracellular flux measurements, we found that the rate of label incorporation from [U-^13^C~6~] glucose into pyruvate was similar across 21%, 0.5%, and DMSO-treated conditions, but markedly increased with BAY treatment (**Figures \@ref(fig:m3)A-B**).
+Consistent with our extracellular flux measurements, we found that the rate of label incorporation from [U-^13^C~6~] glucose into pyruvate was similar across 21%, 0.5%, and DMSO-treated conditions, but markedly increased with BAY treatment (**Figures \@ref(fig:m2)A-B**).
 
-Several patterns of isotope labeling emerge across the treatment conditions (**Figures \@ref(fig:m3)C, S4**).
+Several patterns of isotope labeling emerge across the treatment conditions (**Figures \@ref(fig:m2)C, S4**).
 For the glycolytic intermediates fructose bisphosphate (FBP), 3-phosphoglycerate (3PG), and pyruvate (PYR), label incorporation from the glucose tracers is decreased with hypoxia and BAY treatments (*i.e.*, the unlabeled, or M0, fraction was greater).
@@
 Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (**Figure S6**).
-Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by INCA [@RN1522; @RN2501; @RN964] (**Figures \@ref(fig:m3), S7, Tables S1-S3**).
+Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by Isotopomer Network Compartment Analysis (INCA) [@RN1522; @RN2501; @RN964] (**Figures \@ref(fig:m3), S7, Tables S1-S3**).
 Overall, proliferating LFs demonstrated high rates of glucose uptake and glycolysis.
@@
 A significant increase in pentose phosphate pathway flux was also observed, although the absolute flux through this pathway is low (**Figure \@ref(fig:m3)A**).
-By contrast, HIF-1 activation by BAY in 21% oxygen increased glycolysis and lactate fermentation by ~50% (**Figure \@ref(fig:m3)B**), but had a similar effect to hypoxia in decreasing serine and glutamine incorporation.
+By contrast, HIF-1 activation by BAY in 21% oxygen increased glycolysis and lactate fermentation by ~50% (**Figure \@ref(fig:m3)B**), but had a similar effect as hypoxia in decreasing serine and glutamine incorporation.
 Metabolite fluxes in DMSO-treated cells were similar to 21% oxygen controls.
 
-In normoxia, the magnitude of intracellular metabolite fluxes were generally similar in LFs and PASMCs (**Figures S7A, S7B, Tables S1, S3**).
+In normoxia, the magnitude of intracellular metabolite fluxes was generally similar in LFs and PASMCs (**Figures S7A, S7B, Tables S1, S3**).
 LFs had greater consumption of serine while PASMCs had greater efflux of alanine.
@@
 The increased TCA flux in PASMCs was driven by increased glutamine consumption in these cells.
-This is similar to a prior report of glutamine-driven oxidative phosphorylation in hypoxic cancer cells [@RN528], although glycolysis was also up-regulated in the hypoxia-exposed cancer cells studied in this report.
+This finding is similar to a prior report of glutamine-driven oxidative phosphorylation in hypoxic cancer cells [@RN528], although glycolysis was also up-regulated in the hypoxia-exposed cancer cells studied in this report.
 
@@
 This observation is consistent with prior reports describing hypoxia-mediated increases in glycogen synthesis [@RN1678; @RN545; @RN1668].
-Together, these data suggest that lactate makes a modest (~5% carbon) contribution to glycogen precursors.
+Together, these data suggest that lactate also makes a small (~5% carbon) contribution to glycogen precursors.
 
@@
 Of the differentially regulated metabolites, 44 were affected by both 0.5% oxygen and BAY treatment (**Figure S8C**).
-Metabolite set enrichment analysis of KEGG biochemical pathways were consistent with the results of the metabolic models demonstrating significant enrichment of the pentose phosphate pathway with 0.5% oxygen or BAY treatment; decreased alanine, aspartate, and glutamate metabolism with hypoxia; and increased glycolysis/gluconeogenesis with BAY (**Figures S8D, S8E**).
-Indeed, aspartate was the most decreased metabolite with both treatments, consistent with prior reports demonstrating an important role for HIF-1 regulation of aspartate biosynthesis in cancer cells [@RN634; @RN1438].
+Metabolite set enrichment analysis of KEGG biochemical pathways was consistent with the results of the metabolic models demonstrating significant enrichment of the pentose phosphate pathway with 0.5% oxygen or BAY treatment; decreased alanine, aspartate, and glutamate metabolism with hypoxia; and increased glycolysis/gluconeogenesis with BAY (**Figures S8D, S8E**).
+Indeed, aspartate was the most significantly decreased metabolite with both treatments, consistent with prior reports demonstrating an important role for HIF-1 regulation of aspartate biosynthesis in cancer cells [@RN634; @RN1438].
 
@@
 Moreover, these hypoxia-treated cells were well-segregated from BAY-treated cells.
-These observations are again consistent with the results of the metabolic flux models demonstrating an overriding effect of hypoxia *per se* on metabolic flux and highlighting important differences between hypoxic and pharmacologic PHD inhibition.
-To identify the key differences among these treatments, we performed linear modeling focusing on those metabolites with a significant interaction term in the model described by treatment × oxygen (**Figure \@ref(fig:m5)E**) (*i.e.*, metabolites where the change with hypoxia treatment was different than the change with BAY treatment).
+These observations are, again, consistent with the results of the metabolic flux models demonstrating an overriding effect of hypoxia *per se* on metabolic flux and highlighting important differences between hypoxic and pharmacologic PHD inhibition.
+To identify the key differences among these treatments, we performed linear modeling focusing on those metabolites with a significant interaction term in the model described by treatment × oxygen (**Figure \@ref(fig:m5)E**) (*i.e.*, metabolites where the change with hypoxia treatment was different from the change with BAY treatment).
 Of 133 metabolites, 77 were significantly differentially regulated by hypoxia and BAY treatments.
@@
 Interestingly, the hypoxia-regulated metabolite, 2-hydroxyglutarate [@RN1603], increased following BAY treatment, but was unaffected by hypoxia treatment in proliferating LFs.
-The tricarboxylic acid (TCA) cycle metabolite aconitate was decreased in hypoxia, but increased by BAY treatment while the opposite effect was observed for hydroxyproline and taurine.
+The tricarboxylic acid (TCA) cycle metabolite aconitate was decreased in hypoxia, but increased by BAY treatment, while the opposite effect was observed for hydroxyproline and taurine.
 The glycolytic intermediate glyceraldehyde 3-phosphate (GAP) was decreased by both hypoxia and BAY, while hypoxia reversed the BAY-mediated decrease in γ-aminobutyric acid (GABA).
@@
 Leading edge analysis demonstrated negative enrichment scores associated with all of the TCA metabolites detected by our platform (**Figure \@ref(fig:m5)H**).
-This result suggests a more modest impact of BAY treatment on the TCA cycle than hypoxia, as suggested by our metabolic flux models where hypoxia resulted in a 1.5-2-fold reduction of TCA flux compared to a 1.1-1.5-fold reduction with BAY treatment in normoxia (**Figure \@ref(fig:m3)**).
+This result indicates a more modest impact of BAY treatment on the TCA cycle than hypoxia, as suggested by our metabolic flux models where hypoxia resulted in a 1.5-2-fold reduction of TCA flux compared to a 1.1-1.5-fold reduction with BAY treatment in normoxia (**Figure \@ref(fig:m3)**).
 
@@
 As anticipated, both hypoxia and BAY treatment induced substantial changes in gene expression (**Figure S9A, S9B**).
-Of the 7,789 differentially expressed genes across both conditions, 891 (11%) were unique to BAY treatment in normoxia, 1,649 (21%) were shared between BAY and hypoxia, while 5,249 (67%) were unique to 0.5% culture (**Figure S9C**).
+Of the 7,789 differentially expressed genes across both conditions, 891 (11%) were unique to BAY treatment in normoxia, 1,649 (21%) were shared between BAY and hypoxia, while 5,249 (67%) were unique to 0.5% hypoxia culture (**Figure S9C**).
 Gene set enrichment analysis of these differentially regulated metabolites was performed using Molecular Signatures Database "Hallmark" gene sets [@RN3112] (**Figures S9D, S9E, S9F**).
@@
 The first and second principal components correspond to 0.5% oxygen and BAY treatments, respectively.
-Consistent with our prior observations, the combination of 0.5% oxygen plus BAY was more similar to 0.5% oxygen alone with decreased distance between both hypoxia-treated groups along the axis of the second principal component.
-Again, consistent with the hypothesis that hypoxia uncouples the effects of BAY treatment.
+Consistent with our prior observations, the combination of 0.5% oxygen plus BAY was more similar to 0.5% oxygen alone with decreased distance between both hypoxia-treated groups along the axis of the second principal component, again, consistent with the hypothesis that hypoxia uncouples the effects of BAY treatment.
 To identify those transcripts that were differentially affected by hypoxia compared to BAY treatment alone, we performed linear modeling to identify transcripts with a significant interaction term in the model described by treatment × oxygen (**Figure \@ref(fig:m6)B**).
@@
 RBM3 supports the viability and proliferation of neural stem cells in hypoxia [@RN3111].
-Histone deacetylase 9 (HDAC9) and prolyl 4-hydroxylase subunit α2 (P4HA2) demonstrate opposing effects following hypoxia and BAY treatment, where BAY decreases HDAC9 expression and increases P4HA2 expression.
+Histone deacetylase 9 (HDAC9) and prolyl 4-hydroxylase subunit α2 (P4HA2) demonstrated opposing effects following hypoxia and BAY treatment, where BAY decreased HDAC9 expression and increased P4HA2 expression.
 Together, this subset of transcriptional changes illustrates important differences between hypoxia and HIF stabilization in normoxia.
@@
 To test the hypothesis that hypoxia-induced MYC expression inhibits glycolysis in hypoxic primary cells, we first combined MYC knockdown with hypoxia treatment (**Figure \@ref(fig:m7)C-E**).
-As expected, MYC deficient cells proliferated more slowly in normoxia and MYC was absolutely essential to sustain cell proliferation in hypoxia (**Figure \@ref(fig:m7)D**).
+As expected, MYC-deficient cells proliferated more slowly in normoxia and MYC was absolutely essential for sustaining cell proliferation in hypoxia (**Figure \@ref(fig:m7)D**).
 Consistent with our hypothesis, MYC knockdown cells demonstrated increased lactate efflux upon hypoxia treatment, unlike control siRNA-treated cells (**Figure \@ref(fig:m7)E**).
 We next performed the complementary experiment to determine whether MYC overexpression could attenuate the increase in glycolysis observed with BAY treatment (**Figure \@ref(fig:m7)F-H**).
-MYC increased the proliferation rate of DMSO-treated cells, although did not augment the proliferation rate of BAY-treated cells.
+MYC increased the proliferation rate of DMSO-treated cells, although it did not augment the proliferation rate of BAY-treated cells.
 As expected, MYC overexpression blocked the BAY-stimulated increase in lactate efflux.
@@
 Certainly, the LFs studied here are capable of augmenting glycolysis in response to HIF stabilization, as demonstrated by experiments with the PHD inhibitor BAY; however, these effects are completely attenuated when BAY-treated cells are cultured in hypoxia.
-Together, these findings suggest that changes in enzyme levels alone are insufficient to alter metabolic flux in hypoxia and point to the importance of regulatory mechanisms that supersede the effects of HIF-dependent gene transcription.
+Together, these findings suggest that changes in enzyme levels alone are insufficient to alter metabolic flux in hypoxia and point to the importance of regulatory mechanisms that supersede the effects of HIF-dependent gene transcription in primary cells.
 
@@
 MYC is a transcription factor that regulates the expression of numerous genes involved in many biological processes, including metabolism, proliferation, apoptosis, and differentiation [@dang2006; @li2020d; @stine2015].
-As deregulated MYC activity has been associated with the majority of human cancers [@vita2006], much of our understanding of MYC regulation comes from studies using cancer cell models [@li2020d; @stine2015; @dang2012a; @madden2021], while the role of MYC in the biology of transformed cells is less well understood.
+As deregulated MYC activity has been associated with the majority of human cancers [@vita2006], much of our understanding of MYC regulation comes from studies using cancer cell models [@li2020d; @stine2015; @dang2012a; @madden2021], while the role of MYC in the biology of untransformed cells is less well understood.
 The literature describes a complex and reciprocal relationship between HIF and MYC that depends on both environmental (*e.g.*, hypoxia) and cellular context [@li2020d].
@@
 MYC stimulates the expression of nuclear-encoded mitochondrial genes and promotes mitochondrial biogenesis, both directly and through activation of mitochondrial transcriptional factor A (TFAM) [@li2005a].
-Indeed, the oxidative phosphorylation gene set was relatively enriched with hypoxia treatment compared to BAY treatment (**Figure \@ref(fig:m6)D**).
+Indeed, we found that the oxidative phosphorylation gene set was relatively enriched with hypoxia treatment compared to BAY treatment (**Figure \@ref(fig:m6)D**).
 In this way, hypoxic MYC activation may sustain energy production by oxidative phosphorylation, thereby decreasing the energetic demands driving increased glycolytic flux.
@@
 Beyond MYC, the identification of other HIF-independent mechanisms regulating primary cell adaption to hypoxia is of critical importance.
-Cells express several oxygen-dependent enzymes in addition to PHD whose activities may be impacted in hypoxia but not by PHD inhibition.
+Cells express several oxygen-dependent enzymes in addition to PHD whose activities may be altered in hypoxia but not by PHD inhibition.
 For example, PHD is one of many α-ketoglutarate-dependent dioxygenase enzymes that rely on molecular oxygen for their catalytic activity [@RN2611].
@@
 In addition to dioxygenase enzymes, electron transport chain dysfunction resulting from impaired Complex IV activity leads to increased ROS production in hypoxia [@RN295].
-Mitochondrial ROS increase the half-lives of several mRNAs in hypoxia, including MYC, as we observe in this work, independent of HIF stabilization [@RN742].
+Mitochondrial ROS increase the half-lives of several mRNAs in hypoxia, including MYC, as we observed in this work, independent of HIF stabilization [@RN742].
 Finally, hypoxia imposes a reductive stress on cells associated with an increase in the NADH/NAD^+^ ratio secondary to impaired electron transport (**Figure S5K**) [@RN2618; @RN2619].
@@
 
-In addition to its effects on cellular metabolism, another canonical role of HIF-1 activation is slowing of cellular proliferation rate in the face of limited oxygen availability [@RN615] and this is what we observed here (**Figures \@ref(fig:m1)B, \@ref(fig:m5)A**).
-The effects of HIF-1 on cell proliferation rate are mediated, in part, by increased expression of cyclin-dependent kinase inhibitor p21 (*CDKN1A*) and inhibition of E2F targets [@RN3152] and inhibition of pro-proliferative MYC signaling [@RN3151].
+In addition to its effects on cellular metabolism, another canonical role of HIF-1 activation is slowing of cellular proliferation rate in the face of limited oxygen availability [@RN615], which is precisely what we observed here (**Figures \@ref(fig:m1)B, \@ref(fig:m5)A**).
+The effects of HIF-1 on cell proliferation rate are mediated, in part, by increased expression of cyclin-dependent kinase inhibitor p21 (*CDKN1A*), inhibition of E2F targets [@RN3152], and inhibition of pro-proliferative MYC signaling [@RN3151].
 These transcriptional effects are precisely what is observed in BAY treated LFs in normoxia.
@@
 
-Altogether, these findings raise important questions regarding the cell-autonomous role of HIFs in the hypoxia response.
+Taken together, these findings raise important questions regarding the cell-autonomous role of HIFs in the hypoxia response.
 On an organismal level, HIFs drive expression of angiogenic and erythropoietic factors to increase oxygen delivery to hypoxic tissues.
@@
 1. Protein turnover occurred at a negligible rate compared to glucose and glutamine consumption.
-1. Acetyl-CoA, aspartate, fumarate, malate, oxaloacetate, pyruvate existed in cytosolic and mitochondrial pools. Aspartate and malate were allowed to exchange freely between the compartments.
+1. Acetyl-CoA, aspartate, fumarate, malate, oxaloacetate, and pyruvate existed in cytosolic and mitochondrial pools. Aspartate and malate were allowed to exchange freely between the compartments.
 1. The per cell biomass requirements of proliferating lung fibroblasts were similar to published estimated in other cells [@RN1741; @RN2005]. 
@@
 Summary data show the mean ± SEM.
-Outliers were identified using twice the median absolute deviation.
+Outliers were identified using twice the median absolute deviation as a cutoff threshold.
 Two group comparisons (*e.g.*, 21% *v.* 0.5% oxygen) were performed using Student's *t*-test on log-transformed values with pairing among biological replicates.

### 2021-12-14 manuscript/supplement.Rmd
final changes to manuscript before initial submission

@@
 - Benjamin A. Olenchock: 
+    institute: 1
+- David R. Ziehr:
     institute: 1, 2
-- David R. Ziehr:
-    institute: 1, 4
 - Sarah McGarrity: 
-    institute: 1, 5
+    institute: 1, 3
+- Kevin Leahy: 
+    institute: 1
 - Jamey D. Young: 
-    institute: 3
+    institute: 4
 - Joseph Loscalzo: 

### 2021-12-20 manuscript/manuscript.Rmd
reformat for nature comms and fix figure legend (*) by escaping

@@
 - packages.bib
-csl: cell-metabolism.csl
+csl: nature-communications.csl
 link-citations: yes
@@
 
-# Summary
+# Abstract
 
@@
 Comparisons were made using linear mixed effects models with treatment group as a fixed effect and biological replicate as a random effect.
-Tukey's *post hoc* test was applied to determine differences between 21% and 0.5% oxygen (*), between DMSO and BAY treatment (†), or between 0.5% oxygen and BAY treatment (‡) with adjusted p-values < 0.05 considered significant.
+Tukey's *post hoc* test was applied to determine differences between 21% and 0.5% oxygen (\*), between DMSO and BAY treatment (†), or between 0.5% oxygen and BAY treatment (‡) with adjusted p-values < 0.05 considered significant.
 
@@
 Data are the mean ± SEM of 4 biological replicates.
-Significant differences in labeling patterns between 21% and 0.5% oxygen (*), DMSO and BAY treatment (†), and 0.5% oxygen and BAY treatment (‡) for each combination of metabolite and tracer are highlighted.
+Significant differences in labeling patterns between 21% and 0.5% oxygen (\*), DMSO and BAY treatment (†), and 0.5% oxygen and BAY treatment (‡) for each combination of metabolite and tracer are highlighted.
 FBP, fructose bisphosphate; 3PG, 3-phosphoglycerate; AKG, α-ketoglutarate.
@@
 
-# STAR Methods
+# Methods
 
@@
 
-### Lung fibrolbasts
+### Lung fibroblasts
 

### 2021-12-20 manuscript/supplement.Rmd
reformat for nature comms and fix figure legend (*) by escaping

@@
 bibliography_software: packages.bib
-csl: cell-metabolism.csl
+csl: nature-communications.csl
 link-citations: yes
@@
 Comparisons were made using linear mixed effects models with treatment group as a fixed effect and biological replicate as a random effect.
-Tukey's *post hoc* test was applied to determine differences between 21% and 0.2% oxygen (*) with p-values < 0.05 considered significant.
+Tukey's *post hoc* test was applied to determine differences between 21% and 0.2% oxygen (\*) with p-values < 0.05 considered significant.
 
@@
 Comparisons were made using linear mixed effects models with treatment group as a fixed effect and biological replicate as a random effect.
-Tukey's *post hoc* test was applied to determine differences between 21% and 0.5% oxygen (*) with p-values < 0.05 considered significant.
+Tukey's *post hoc* test was applied to determine differences between 21% and 0.5% oxygen (\*) with p-values < 0.05 considered significant.
 
@@
 Data are the mean ± SEM of 4 biological replicates.
-Significant differences in labeling patterns between 21% and 0.5% oxygen (*), DMSO and BAY treatment (†), and 0.5% oxygen and BAY treatment (‡) for each combination of metabolite and tracer are highlighted.
+Significant differences in labeling patterns between 21% and 0.5% oxygen (\*), DMSO and BAY treatment (†), and 0.5% oxygen and BAY treatment (‡) for each combination of metabolite and tracer are highlighted.
 
@@
 Data are the mean ± SEM of 4 biological replicates.
-Significant differences in labeling patterns between 21% and 0.5% oxygen (*) for each combination of metabolite and tracer are highlighted.
+Significant differences in labeling patterns between 21% and 0.5% oxygen (\*) for each combination of metabolite and tracer are highlighted.
 

### 2022-02-01 manuscript/manuscript.Rmd
send to Cell Reports

@@
 - packages.bib
-csl: nature-communications.csl
+csl: cell-metabolism.csl
 link-citations: yes
@@
 The literature describes a complex and reciprocal relationship between HIF and MYC that depends on both environmental (*e.g.*, hypoxia) and cellular context [@li2020d].
-Generally, HIF-1 has been observed to inhibit MYC through multiple mechanisms [koshiji2004; koshiji2005; @gordan2007; @zhang2007], and this previous work is consistent with our observations that HIF stabilization following BAY treatment decreased MYC protein and target gene expression.
+Generally, HIF-1 has been observed to inhibit MYC through multiple mechanisms [@koshiji2004; @koshiji2005; @gordan2007; @zhang2007], and this previous work is consistent with our observations that HIF stabilization following BAY treatment decreased MYC protein and target gene expression.
 Conversely, MYC has been implicated in increased HIF activity through transcriptional and post-transcriptional mechanisms [@zhang2009a; @chen2013; @doe2012], primarily in the context of malignant transformation.
@@
 
-# Methods
+# STAR Methods
 

### 2022-02-01 manuscript/supplement.Rmd
send to Cell Reports

@@
 bibliography_software: packages.bib
-csl: nature-communications.csl
+csl: cell-metabolism.csl
 link-citations: yes
